---
title: "An instrument does not have to be exogenous to be consistent"
author: "Declare Design Team"
slug: "2020-02-05-instrumental-variables"
date: 2020-02-19
output:
  html_document:
    highlight: tango
    theme: cerulean
    code_folding: show
bibliography: bib/blog.bib
---

```{r setup, include=FALSE}
library(tidyverse)
library(DeclareDesign)
library(DesignLibrary)

library(knitr)
opts_chunk$set(echo = TRUE)

library(future)
plan(multiprocess)

sims   <- 200
b_sims <- 100

do_diagnosis <- FALSE

hj_dag <- function(x,
                   y,
                   names,
                   arcs = cbind(0,0),
                   add_points = FALSE,
                   solids = rep(1, length(x)),
                   title = "",
                   contraction = .1,
                   add_functions = 0,
                   add_functions_text = NULL,
                   text_shift = .2*add_points,
                   padding = .5,
                   length = 0.2,
                   cex = 1,
                   box = TRUE) {
  if(add_points)  plot(x, y, pch=ifelse(solids == 1, 19, 1), cex = 2, axes = FALSE, xlab = "", ylab = "",
                       xlim = c(min(x)-padding, max(x)+padding),
                       ylim = c(min(y)-padding-add_functions, max(y)+padding),
                       main = title)
  if(!add_points)  plot(x, y, type = "n", cex = 2, axes = FALSE, xlab = "", ylab = "",
                        xlim = c(min(x)-padding, max(x)+padding),
                        ylim = c(min(y)-padding-add_functions, max(y)+padding),
                        main = title)

  arrows(x[arcs[,1]]*(1-contraction) + x[arcs[,2]]*contraction,
         y[arcs[,1]]*(1-contraction) + y[arcs[,2]]*contraction,
         x[arcs[,2]]*(1-contraction) + x[arcs[,1]]*contraction,
         y[arcs[,2]]*(1-contraction) + y[arcs[,1]]*contraction, length = length)
  text(x, y + text_shift, names, cex = cex)
  if(!is.null(add_functions_text)) text(((min(x)+max(x))+2), min(y)-1, add_functions_text)
  if(box) box()
}
```

We often think of an instrumental variable ($Z$) as a random shock that generates exogenous variation in a treatment of interest $X$. The randomness of $Z$ lets us identify the effect of $X$ on $Y$, at least for units for which $Z$ perturbs $X$ in a way that's not possible by just looking at the relationship between $X$ and $Y$.  But surprisingly, we think, if effects are constant the instrumental variables estimator can be consistent for the effect of $X$ on $Y$ even when the relationship between the instrument ($Z$) and the endogenous variable ($X$) is confounded (for example, @hernan2006instruments). That's the good news. Less good news is that when there is effect heterogeneity you can get good estimates for some units but it can be hard to know *which* units those are [@Swanson2017]. We use a declaration and diagnosis to illustrate these insights. 
<!-- We show first that instrumental variables approaches can do well even with confounded instruments if there are constant treatment effects. We then show that with a confounded instrument and effect heterogeneity  the interpretation of who the compliers are for "complier average effects" really depends on the nature of confounding.  -->

Consider the causal graph below (a directed acyclic graph, or DAG). Here $U_2$ confounds the relationship between $X$ and $Y$. That's the usual thing and the reason for why you need something like an instrument in the first place. But now a binary $U_1$ also confounds the relationship between $Z$ and $X$ (see for instance Fig 2d @swanson2018partial). To emphasize, in this set up, the instrumental variable and the **instrumented** variable share a common cause in $U_1$.

```{r, fig.cap="Figure: DAG for a non-randomized instrument. Z is an instrument for the effect of X on Y",  out.height = '30%', out.width = '50%', fig.align = "center", echo = FALSE}
hj_dag(x = 1:5, 
       y = c(0,1,0,1,0), 
       arcs = cbind(c(1,2,2,3,4,4), c(3, 1, 3, 5, 3, 5)),
       names = c("Z", expression(U[1]), "X", expression(U[2]), "Y"), box = FALSE, padding = 0)
```

Using `DeclareDesign`, we write down a declaration that is consistent with this DAG. Our declaration specifies some "compliance types" or "principal strata." In the standard IV setup with a binary instrument $Z$ and a binary treatment $X$, there are only four types (corresponding to all conceivable ways that the endogenous variable can respond to the exogenous variable): compliers, never-takers, always-takers, and defiers.^[Denoting $x$ the value that the endogenous variable $X$ takes, compliers have $x = f_X(Z) = Z$, never-takers have $x = f_X(Z) = 0$, always-takers have $x = f_X(Z) = 1$, and defiers have $x = f_X(Z) = (1 - Z)$.] But here, the binary $U_1$ and binary $Z$ both affect $X$, so there are actually **sixteen** conceivable ways to comply!^[The value $X$ takes could depend on all four potential combinations of $U_1$ and $Z$. So if $x = f_X(U_1,Z)$, you have to think about every possible way in which $f_X(0,0)$, $f_X(1,0)$, $f_X(0,1)$ and $f_X(0,0)$ can map into values of $x$.]

We simplify  by imagining a world with only three types: $U_1$-compliers, $Z$-compliers, and never-takers.^[The assumption that there are no types who "defy" their $U_1$ or $Z$ assignments amounts to a monotonicity assumption common in the IV framework.] $U_1$-compliers have $X = 1$ if and only if $U_1 = 1$. $Z$-compliers have $X = 1$ if and only if $Z = 1$. Never-takers have $X = 0$ regardless of the values of $U_1$ or $Z$. $U_2$ takes three values, 1-3, and plays two roles. First, it determines whether you are a $U_1$-complier, $Z$-complier, or never-taker with respect to $X$. Second, it is positively correlated with your value in $Y$. 

We'll define five estimands, which is a lot. 

1. The Average Treatment Effect (ATE) of $X$ on $Y$. 
2. The Intention-To-Treat (ITT) effect. The average effect of $Z$ on $Y$.
3. The first-stage effect. This is the average effect of $Z$ on $X$. 
4. The Local Average Treatment Effect (LATE_U1) of $X$ on $Y$, among the subgroup of units that are $U_1$-compliers
4. The Local Average Treatment Effect (LATE_Z) of $X$ on $Y$, among the subgroup of units that are $Z$-compliers

We'll use three estimators:

1. The ITT estimator, which is an OLS regression of $Y$ on $Z$. 
2. The first-stage estimator, which is an OLS regression of $X$ on $Z$. 
3. The IV estimator, which is two-stage least squares regression of $Y$ on $X$, instrumented by $Z$.

The declaration also includes two additional parameters (bear with us!) that we'll vary down below.  They are:

1. `heterogeneity`: the amount of treatment effect heterogeneity. When this is 0, the effect of $X$ on $Y$ is the same for $U_1$-compliers as it is for $Z$-compliers. Otherwise, the effect for the two subgroups is different. 
2. `path_weight`: describes whether the `U1 --> X` or the `Z --> X` path is stronger. This is equivalent to the fraction of compliers that are $U_1$-compliers.

OK, here's the design:

```{r}
heterogeneity <- 0   # how much does LATE_Z differ from LATE_U1?
path_weight   <- .5  # what fraction of compliers are U1 compliers?

design <-
  
  # U2 is the compliance "type" (U1 complier, Z complier, or Never Taker)
  declare_population(N = 2000, 
                     U2 = complete_ra(
                             N, 
                             prob_each = c(path_weight * 0.6, (1 - path_weight) * 0.6, 0.4), 
                             conditions = c("U1_c", "Z_c", "NT")),
                     noise = rnorm(N)) +
  
  # U1 is binary
  declare_assignment(prob = 0.5, assignment_variable = "U1") +

  # Z is affected by U1: among U1 = 0, Z has probability 0.3; among U1 = 1, probability 0.6
  declare_assignment(blocks = U1, block_prob = c(0.3, 0.6), assignment_variable = "Z") +
  
  # X potential outcomes in terms of U1 and Z, given U2 (there are **four**)
  declare_potential_outcomes(X ~ as.numeric((U2 == "U1_c") * U1 +
                                            (U2 == "Z_c")  * Z),
                             assignment_variables = c(U1, Z)) +

  # X potential outcomes in terms of Z given U1 and U2 (there are **two**)
  declare_potential_outcomes(X ~ as.numeric((U2 == "U1_c") * U1 +
                                            (U2 == "Z_c")  * Z),
                             assignment_variables = Z) +
  
  declare_reveal(X, assignment_variables = c(U1, Z)) +
  
  # Y potential outcomes in terms of X (there are **two**)
  declare_potential_outcomes(Y ~ X * (1 + heterogeneity * as.numeric(U2)) + noise, 
                             assignment_variables = X) +
  
  # Y potential outcomes in terms of Z (there are **two**)
  # this is same as above, but X is written in terms of Z
  declare_potential_outcomes(Y ~ as.numeric((U2 == "U1_c") * U1 + (U2 == "Z_c") * Z) * 
                              (1 + heterogeneity * as.numeric(U2)) + noise,
                              assignment_variables = Z) +
  
  # reveal Y (doesn't matter "how" this is revealed, either via the X pos or the Z pos)
  declare_reveal(Y, assignment_variables = X) +
  
  # Use X POs to learn about the two complier definitions
  declare_step(
    U1_complier = X_U1_0_Z_0 == 0 & X_U1_0_Z_1 == 0 & X_U1_1_Z_0 == 1 & X_U1_1_Z_1 == 1,
    Z_complier  = X_U1_0_Z_0 == 0 & X_U1_1_Z_0 == 0 & X_U1_0_Z_1 == 1 & X_U1_1_Z_1 == 1, 
    handler = fabricate) +
  
  # 5 (!) estimands
  declare_estimand(
    ATE =         mean(Y_X_1 - Y_X_0),
    LATE_U1 =     mean(Y_X_1[U1_complier] - Y_X_0[U1_complier]),
    LATE_Z =      mean(Y_X_1[Z_complier] - Y_X_0[Z_complier]),
    ITT =         mean(Y_Z_1 - Y_Z_0),
    first_stage = mean(X_Z_1) - mean(X_Z_0)) + 
  
  # Three estimators
  declare_estimator(Y ~ Z, estimand = "ITT", model = lm_robust, label =  "itt") +
  declare_estimator(X ~ Z, estimand = "first_stage", model = lm_robust, label =  "first") +
  declare_estimator(Y ~ X | Z, estimand = c("ATE", "LATE_U1", "LATE_Z"), 
                              model = iv_robust, label =  "iv_robust")
```

Let's now diagnose this design to show that -- pretty counterintuitively for us at least! -- the IV estimator still recovers the LATE even when $Z$ and $X$ have a common cause.

```{r, eval = FALSE, message = FALSE}
diagnosis_1 <- 
  diagnose_design(
    design, diagnosands = declare_diagnosands(select = c(bias, mean_estimate, mean_estimand)))
```

```{r, echo = FALSE, warning = FALSE}
if(do_diagnosis){
diagnosis_1 <- 
  diagnose_design(
    design, 
    diagnosands = declare_diagnosands(select = c(bias, mean_estimate, mean_estimand)),
    sims = sims, bootstrap_sims = b_sims)
  write_rds(diagnosis_1, path = "rfiles/iv1.rds")
}

diagnosis_1 <- read_rds("rfiles/iv1.rds")

kable(reshape_diagnosis(diagnosis_1))
```

Let's take the diagnosis estimand-by-estimand:

1. The ATE is well-estimated by the instrumental variables estimator. This, despite the fact that:
2. The first stage regression estimates are biased for the average effect of $Z$ on $X$ because of the unobserved confounder $U_1$. 
3. The ITT estimates are **also** biased because of the confounding between $Y$ and $Z$. 
4. The instrumental variables performs well for the LATE, as defined here. But in this case the LATE is the same as the ATE so this is not surprising.

You see in the estimator column that the bias in the ITT and first stage estimates are proportionate, such that, in the ratio, the bias cancels out. We provide some intuition for this result below, where we show how it can also break down.

The upshot is that when assessing the validity of IV you might be imposing an additional assumption that you might not need to impose. This may be good news in some specific research settings where the more stringent assumption of exogenous assignment of $X$ to $Z$ is not well justified.^[Of course, the result also holds when $Z$ and $X$ are unconfounded, which also allows for unbiased estimation of the ITT and the first stage, which may be of practical or theoretical interest in and of themselves.]

# Allowing heterogeneity

We show here that we lose the ability to estimate the effect of $X$ on $Y$ when this effect varies among compliance types  -- the $U_1$-compliers, the $Z$-compliers, and the never-takers.^[More precisely, weaker conditions than homogeneity can be sufficient for identification. As put by @hernan2006instruments, a sufficient assumption is that "the X-Y causal risk difference is the same among treated subjects with $Z=1$ as among treated subjects with $Z=0$, and similarly among untreated subjects." See other conditions discussed in @swanson2018partial.] We can quickly redesign to allow heterogeneity, and see that now, we are biased for all the estimands!

```{r, eval = FALSE, message = FALSE}
design_h <- redesign(design, heterogeneity = 1)

diagnosis_2 <- 
  diagnose_design(
    design_h, diagnosands = declare_diagnosands(select = c(bias, mean_estimate, mean_estimand)))
```

```{r, echo = FALSE, warning = FALSE}
if(do_diagnosis){
design_h <- redesign(design, heterogeneity = 1)

diagnosis_2 <- 
  diagnose_design(
    design_h, 
    diagnosands = declare_diagnosands(select = c(bias, mean_estimate, mean_estimand)),
    sims = sims, bootstrap_sims = b_sims)
  write_rds(diagnosis_2, path = "rfiles/iv2.rds")
}

diagnosis_2 <- read_rds("rfiles/iv2.rds")

kable(reshape_diagnosis(diagnosis_2))
```

Why was the bias so much lower under homogeneous effects of $X$ on $Y$ for compliers? 

Here's some intuition. In both the heterogeneous and homogeneous cases, we misattribute some of the effect of $U_1$ on $X$ to the effect of $Z$ on $X$, because those for whom $U_1 = 1$ are also more likely to have $Z = 1$. Thus, we overestimate the first stage in proportion to this misattribution. In both the homogeneous and heterogeneous cases, we estimate that $Z$ increases the probability of $X = 1$ by 39pp on average, whereas the true effect is 30pp. When we compute the effect of $Z$ on $Y$ for the ITT, those $U_1$-compliers that we mistook for $Z$-compliers also produce an effect on $Y$ that we again misattribute to $Z$, though in fact it was caused by the effect of $U_1$ running through $X$. The extent of this misattribution in the ITT is the same as the extent of the misattribution in the first stage when the effects of $X$ on $Y$ are homogeneous. Thus, even though we wanted to divide 0.3 / 0.3 to get 1, we ended up dividing 0.39 / 0.39 -- and we still got 1! When the effects are heterogeneous, however, we lose this proportionality: because the effect of $X$ on $Y$ is bigger now for $Z$-compliers than for $U_1$ compliers, *our estimate of the ITT is not biased enough!*. We would have wanted to estimate an ITT of 1.20 so that our IV estimator would have given us the true ATE via 1.20 / 0.39 = 3.10. But instead, we misattribute the $U_1$-compliers' effect of $X$ on $Y$ to the $Z$-compliers in a way that is disproportionate to the misattribution of the first stage, $Z$ on $X$. We don't have enough bias to unbias our estimates!

# With non causal instruments the "compliers" are those that comply with respect to the unobserved causal instrument

Although we have described $Z$ as being the instrument, you can see in this DAG that both $Z$ and $U_1$ can affect $X$. So perhaps $U_1$ should thought of as an instrument even if it is unobserved? And perhaps compliers should be thought of as units that comply with $U_1$ and not $Z$? 

To explore this possibility, we diagnose design modifications that shift probability mass between two special cases of the DAG.

The first special case is the standard IV setup with no arrow between $U_1$ and $X$. The second special case is one discussed in @Swanson2017 in which $Z$ is a "noncausal instrument" and the true causal instrument is the unobserved $U_1$. We get this by removing the path from $Z$ to $X$. @Swanson2017 show that **even though** $Z$ does not itself affect $X$, instrumenting $X$ with $Z$ will recover good estimates of a well-defined estimand. You guessed it, that estimand is the LATE for $U_1$ compliers. Amazing!

These cases are the two extremes of a continuum of the relative strength of the two paths. To be clear those two paths are `U1 --> X` and `Z --> X`.  We can adjust the strength of those two paths with the `path_weight` parameter in our design. For this last simulation, we'll consider 5 values for this parameter between 0 and 1.

```{r, eval = FALSE, message = FALSE}
designs <- redesign(design, heterogeneity = 1, path_weight = seq(0.01, .99, length = 5))

diagnosis_3 <- diagnose_design(
    designs, diagnosands = declare_diagnosands(select = c(bias, mean_estimate, mean_estimand)))
```

```{r, echo = FALSE, warning = FALSE}
library(ggrepel)
if(do_diagnosis){
designs <- redesign(design, heterogeneity = 1, path_weight = seq(0.01, .99, length = 5))
diagnosis_3 <- 
  diagnose_design(
    designs, 
    diagnosands = declare_diagnosands(select = c(bias, mean_estimate, mean_estimand)),
    sims = sims, bootstrap_sims = b_sims)
  write_rds(diagnosis_3, path = "rfiles/iv3.rds")
}

diagnosis_3 <- read_rds("rfiles/iv3.rds")


gg_df <- 
diagnosis_3 %>%
  get_diagnosands()

label_df <- gg_df %>% filter(path_weight == 0.5)

ggplot(gg_df, aes(path_weight, bias, color = estimand_label)) +
  geom_point() +
  geom_line() +
  geom_label_repel(data = label_df, aes(label = estimand_label)) +
  xlab("Fraction of all compliers who are U1 compliers (the rest are Z compliers!)") +
  theme_bw() +
  theme(legend.position = "none")
```

Let's consider the graph from left to right. 

All the way to the left, all the compliers are $Z$ compliers. Equivalently, all the way to the left, $U_1$ has no effect on $X$. That's the standard IV setup in which $Z$ is as-if randomized, so our estimates of the ITT, the first stage, and the LATE are all on target.

All the way to the right, all the compliers are $U_1$-compliers, because $Z$ has no effect on $X$. That's the "non causal instrument" case described in @Swanson2017 (see also @brito2002generalized). The estimates of everything else -- the ITT, the first stage, and the ATE are all quite biased, but the IV estimator does a good job of estimating the LATE among $U_1$-compliers.  

In between the leftmost and rightmost extremes, however, our estimates of all of the estimands are biased. Indeed, the middle case (at the `0.50` mark) is exactly the situation we described in the "Allowing heterogeneity" section above. It's possible (and we'd love to hear from you if you have ideas along these lines!) that an alternative estimation strategy would recover good estimates of one or more of these estimands.

# Blog update

This will be the last of our weekly blog posts on [declaredesign.org](declaredesign.org). When we started blogging, we set ourselves a goal of six months of posts. We really enjoyed putting these posts together, but we're also excited to shift attention to other parts of the project. We expect to continue to post occasionally here and to put up more informal posts on [http://discuss.declaredesign.org](http://discuss.declaredesign.org) and we'd love it if you joined us there.

Yours in declaration and diagnosis,

Graeme, Jasper, Alex, and Macartan

# References
