---
title: Statistical Adjustment and Assessment of Adjustment in Observational Studies --- Matched Stratification for One and Multiple Variables.
date: '`r format(Sys.Date(), "%B %d, %Y")`'
author: ICPSR 2018 Session 2
bibliography:
 - BIB/abbrev-long.bib
 - BIB/refs.bib
 - BIB/master.bib
 - BIB/misc.bib
fontsize: 10pt
geometry: margin=1in
graphics: yes
biblio-style: authoryear-comp
biblatexoptions:
  - natbib=true
output:
  revealjs::revealjs_presentation:
    slide_level: 2
    incremental: true
    transition: none
    fig_caption: true
    self_contained: false
    reveal_plugins: ["notes","search","zoom","chalkboard"]
    pandoc_args: [ "--csl", "chicago-author-date.csl" ]
  beamer_presentation:
    slide_level: 2
    keep_tex: true
    latex_engine: xelatex
    citation_package: biblatex
    template: icpsr.beamer
    incremental: true
    includes:
        in_header:
           - defs-all.sty
---


<!-- Make this document using library(rmarkdown); render("day12.Rmd") -->


```{r include=FALSE, cache=FALSE}
# Some customization.  You can alter or delete as desired (if you know what you are doing).
# knitr settings to control how R chunks work.
rm(list=ls())

require(knitr)

## This plus size="\\scriptsize" from https://stackoverflow.com/questions/26372138/beamer-presentation-rstudio-change-font-size-for-chunk

knitr::knit_hooks$set(mysize = function(before, options, envir) {
			      if (before){
				      return(options$size)
			      } else {
				      return("\\normalsize")}

})


knit_hooks$set(plotdefault = function(before, options, envir) {
    if (before) par(mar = c(3, 3, .1, .1),oma=rep(0,4),mgp=c(1.5,.5,0))
})

opts_chunk$set(
  tidy=FALSE,     # display code as typed
  echo=TRUE,
  results='markup',
  strip.white=TRUE,
  fig.path='figs/fig',
  cache=FALSE,
  highlight=TRUE,
  width.cutoff=132,
  size='\\scriptsize',
  out.width='.8\\textwidth',
  fig.retina=FALSE,
  message=FALSE,
  comment=NA,
  mysize=TRUE,
  plotdefault=TRUE)

if(!file.exists('figs')) dir.create('figs')

options(digits=4,
	scipen=8,
	width=132,
	show.signif.stars=FALSE)
```

```{r eval=FALSE, include=FALSE, echo=FALSE}
## Run this only once and then not again until we want a new version from github
library('devtools')
library('withr')
with_libpaths('./lib', install_github("markmfredrickson/RItools"), 'pre')
```

```{r echo=FALSE}
library(dplyr)
library(ggplot2)
library(RItools,lib.loc="./lib")
library(optmatch)
```
## Today

\begin{enumerate}
  \item Agenda:  Continue to think about an adjustment strategy to enhance counterfactual causal interpretations in non-randomized studies: \\	
Need adjust $\rightarrow$ "fair comparison" $\rightarrow$
  stratification $\rightarrow$ evaluation of the stratification;
  How to do this with one variable and/or more than one
  variable.
\item Reading for tomorrow and next week: DOS 8--9, 13 and \cite[\S~9.5]{gelman2006dau}, and \cite{hans:04} \cite{ho:etal:07}
\item Questions arising from the reading or assignments or life?
\end{enumerate}

# But first, review:

## Review of Instrumental Variables

Say you think you have an instrument: "rainfall", "economic liberalization as
a result of NAFTA", "a housing/school lottery". You then need to convince
yourself about:

 - SUTVA
 - "As if randomized"/Ignorability
 - That the instrument detectibly changes the dose / The instrument is not
   weak.
 - That the instrument influences the outcome **only** via the dose
   (Excludable)
 - (for estimation of CACE/LATE) That there are no defiers / the instrument
   changes the dose in one direction and the dose changes the outcome in one
   direction.

## Review of randomization assessment in a randomized experiment

How, in principle, might one do this?


```{r echo=FALSE, cache=TRUE}
load(url("http://jakebowers.org/Data/meddat.rda"))
```

## Review of linear model "control for" 

Did the Metrocable intervention decrease
violence in those neighborhoods? We have Homicides per 1000 people in 2008 (`HomRate08`) as a function of Metrocable.

```{r}
meddat<- mutate(meddat, HomRate03=(HomCount2003/Pop2003)*1000,
                HomRate08=(HomCount2008/Pop2008)*1000)
row.names(meddat) <- meddat$nh

lmRaw <- lm(HomRate08~nhTrt,data=meddat)
coef(lmRaw)[["nhTrt"]]
```

What do we need to believe or know in order to imagine that we have done a good job adjusting for Proportion with more than HS Education below? (concerns about extrapolation, interpolation, linearity, influential points, parallel slopes, biased estimation of the ACE)

```{r}
lmAdj1 <- lm(HomRate08 ~ nhTrt + nhAboveHS, data=meddat)
coef(lmAdj1)["nhTrt"]
```

What about when we try to adjust for more than one variable? (+the curse of
dimensionality)

```{r}
lmAdj2 <- lm(HomRate08 ~ nhTrt + nhAboveHS + nhRent, data=meddat)
coef(lmAdj2)["nhTrt"]
```


## Another problem with the linear model: How to choose?

Here are 10,000 regression models, each with a different way to adjust for
between 1 and 4 of the variables below. Restricted to additive functional
forms, no interaction effects, no smooth non-linearity (no $x^2$).

```{r manyates, cache=TRUE}
lmadjfn<-function(){
	covs <- c("nhAboveHS","nhRent","nhMale","nhEmp")
	ncovs <- sample(1:length(covs),1)
	somecovs <- sample(covs,size=ncovs)
	ncuts <- round(runif(ncovs,min=1,max=8))
	theterms <- ifelse(ncuts==1,somecovs,
			   paste("cut(",somecovs,",",ncuts,")",sep=""))
	thefmla <- reformulate(c("nhTrt",theterms),response="HomRate08")
	thelm <- lm(thefmla,data=meddat)
	theate <- coef(thelm)[["nhTrt"]]
	return(theate)
}

set.seed(12345)
res <- replicate(10000,lmadjfn())
```

```{r}
summary(res)
```

# Matching on one variable to create strata

## The optmatch workflow: The distance matrix

The goal of optmatch: collect observations into strata which minimize the
within strata differences (calculate the absolute difference between treated and
controls within strata, sum across strata, minimize this quantity).


```{r}
tmp <- meddat$nhAboveHS
names(tmp) <- rownames(meddat)
absdist <- match_on(tmp, z = meddat$nhTrt,data=meddat)
absdist[1:3,1:3]
abs(meddat$nhAboveHS[meddat$nhTrt==1][1] - meddat$nhAboveHS[meddat$nhTrt==0][1] )
```

## Do the match

We will talk in more detail later, for now, we merely use the software.

```{r}
fm1 <- fullmatch(absdist,data=meddat)
summary(fm1, min.controls=0, max.controls=Inf )
table(meddat$nhTrt,fm1)

pm1 <- pairmatch(absdist,data=meddat)
summary(pm1, min.controls=0, max.controls=Inf )
table(meddat$nhTrt,pm1,exclude=c())
```

## Evaluate the design: Within set differences

Look within sets:

```{r echo=FALSE}
meddat$fm1 <- fm1
meddat$pm1 <- pm1
```

```{r echo=FALSE, out.width=".7\\textwidth"}
library(gridExtra)
bpfm1 <- ggplot(meddat,aes(x=fm1,y=nhAboveHS)) +
	geom_boxplot() +
	 stat_summary(fun.y=mean, geom="point", shape=20, size=3, color="red", fill="red")

bporig <- ggplot(meddat,aes(y=nhAboveHS))+
	 geom_boxplot()

grid.arrange(bpfm1,bporig,ncol=2,layout_matrix=matrix(c(1,1,1,1,2),nrow=1))
```

## Evaluate the design: Within set differences

Look within sets:

```{r echo=FALSE, out.width=".7\\textwidth"}
bppm1 <- ggplot(meddat,aes(x=pm1,y=nhAboveHS)) +
	geom_boxplot() +
	 stat_summary(fun.y=mean, geom="point", shape=20, size=3, color="red", fill="red")

bporig <- ggplot(meddat,aes(y=nhAboveHS))+
	 geom_boxplot()

grid.arrange(bppm1,bporig,ncol=2,layout_matrix=matrix(c(1,1,1,1,2),nrow=1))
```

## Evaluate the design: Within set differences


```{r}
rawmndiffs <- with(meddat, mean(nhAboveHS[nhTrt==1]) - mean(nhAboveHS[nhTrt==0]))
setdiffsfm1 <- meddat %>% group_by(fm1) %>% summarize(mneddiffs =
						   mean(nhAboveHS[nhTrt==1]) -
						   mean(nhAboveHS[nhTrt==0]),
					   mnAboveHS = mean(nhAboveHS),
					   minAboveHS = min(nhAboveHS),
					   maxAboveHS = max(nhAboveHS))

setdiffsfm1
#summary(setdiffs$mneddiffs)
#quantile(setdiffs$mneddiffs, seq(0,1,.1))

```

## Evaluate the design: Within set differences


```{r}
setdiffspm1 <- meddat %>% group_by(pm1) %>% summarize(mneddiffs =
						   mean(nhAboveHS[nhTrt==1]) -
						   mean(nhAboveHS[nhTrt==0]),
					   mnAboveHS = mean(nhAboveHS),
					   minAboveHS = min(nhAboveHS),
					   maxAboveHS = max(nhAboveHS))

setdiffspm1
```

## Evaluate the design: Compare to a randomized experiment.

The within-set differences look different from those that would be expected
from a randomized experiment.

```{r}
xbHS2 <- xBalance(nhTrt~nhAboveHS,
                  strata=list(nostrat=NULL,
                              fm = ~fm1,
			      pm = ~pm1),
                  data=meddat,report="all")
xbHS2$results
xbHS2$overall
```


## What is xBalance doing?

```{r}
setmeanDiffs <- meddat %>% group_by(fm1) %>%
  summarise(diffAboveHS=mean(nhAboveHS[nhTrt==1])-mean(nhAboveHS[nhTrt==0]),
            nb=n(),
            nTb = sum(nhTrt),
            nCb = sum(1-nhTrt),
            hwt = ( 2*( nCb * nTb ) / (nTb + nCb))
            )
setmeanDiffs
```

## What is xBalance doing with multiple sets/blocks?

The test statistic is a weighted average of the set-specific differences (same
approach as we would use to test the null in a block-randomized experiment)

```{r}
## The descriptive adj.mean diff from balanceTest
with(setmeanDiffs, sum(diffAboveHS*nTb/sum(nTb)))
## The mean diff used as the observed value in the testing
with(setmeanDiffs, sum(diffAboveHS*hwt/sum(hwt)))
## Compare to xBalance output
xbHS2$results[,,"fm"]
```

Notice that `balanceTest` prints the set-size weighted difference (the updated
version differs a little from `xBalance`):

```{r}
btHS2 <- balanceTest(nhTrt~nhAboveHS+strata(fm1) +strata(pm1),
                  data=meddat,report="all")
btHS2
```

#  Matching on Many Covariates: Using Mahalnobis Distance

## Dimension reduction using the Mahalanobis Distance

The general idea: dimension reduction. When we convert many columns into one column we reduce the dimensions of the dataset (to one column).


```{r}
X <- meddat[,c("nhAboveHS","nhPopD")]
plot(meddat$nhAboveHS,meddat$nhPopD,xlim=c(-.3,.6),ylim=c(50,700))
```

## Dimension reduction using the Mahalanobis Distance

First, let's look at Euclidean distance: $\sqrt{ (x_1 - x_2)^2 + (y_1 - y_2)^2 }$

```{r echo=FALSE, out.width=".8\\textwidth"}
par(mgp=c(1.25,.5,0),oma=rep(0,4),mar=c(3,3,0,0))
plot(meddat$nhAboveHS,meddat$nhPopD,xlim=c(-.3,.6),ylim=c(50,700))
points(mean(X[,1]),mean(X[,2]),pch=19,cex=1)
arrows(mean(X[,1]),mean(X[,2]),X["407",1],X["407",2])
text(.4,200,label=round(dist(rbind(colMeans(X),X["407",])),2))
```

## Dimension reduction using the Mahalanobis Distance

First, let's look at Euclidean distance: $\sqrt{ (x_1 - x_2)^2 + (y_1 - y_2)^2 }$

```{r echo=FALSE, out.width=".5\\textwidth"}
par(mgp=c(1.25,.5,0),oma=rep(0,4),mar=c(3,3,0,0))
plot(meddat$nhAboveHS,meddat$nhPopD,xlim=c(-.3,.6),ylim=c(50,700))
points(mean(X[,1]),mean(X[,2]),pch=19,cex=1)
arrows(mean(X[,1]),mean(X[,2]),X["407",1],X["407",2])
text(.4,200,label=round(dist(rbind(colMeans(X),X["407",])),2))
```

Distance between point 0,0 and unit "407".

```{r}
tmp <- rbind(colMeans(X),X["407",])
tmp
sqrt( (tmp[1,1] - tmp[2,1])^2 + (tmp[1,2]-tmp[2,2])^2 )
```

Problem: overweights variables with bigger scales (Population Density dominates).

## Dimension reduction using the Mahalanobis Distance

Now the Euclidean distance (on a standardized scale) so neither variable is overly dominant.

```{r echo=FALSE,out.width=".6\\textwidth"}
Xsd <-scale(X)
apply(Xsd,2,sd)
apply(Xsd,2,mean)
plot(Xsd[,1],Xsd[,2],xlab="nhAboveHS/sd",ylab="nhPopD/sd")
points(mean(Xsd[,1]),mean(Xsd[,2]),pch=19,cex=1)
arrows(mean(Xsd[,1]),mean(Xsd[,2]),Xsd["407",1],Xsd["407",2])
text(2,-1.2,label=round(dist(rbind(colMeans(Xsd),Xsd["407",])),2))
```


## Dimension reduction using the Mahalanobis Distance

The mahalanobis distance avoids the scale problem in the euclidean distance.^[For more [see here](https://stats.stackexchange.com/questions/62092/bottom-to-top-explanation-of-the-mahalanobis-distance)] Here each circle are points of the same MH distance.

```{r mhfig, echo=FALSE,out.width=".6\\textwidth"}
library(chemometrics)
par(mgp=c(1.5,.5,0),oma=rep(0,4),mar=c(3,3,0,0))
mh <- mahalanobis(X,center=colMeans(X),cov=cov(X))
drawMahal(X,center=colMeans(X),covariance=cov(X),
          quantile = c(0.975, 0.75, 0.5, 0.25))
abline(v=mean(meddat$nhAboveHS),h=mean(meddat$nhPopD))
pts <-c("401","407","411","202")
arrows(rep(mean(X[,1]),4),rep(mean(X[,2]),4),X[pts,1],X[pts,2])
text(X[pts,1],X[pts,2],labels=round(mh[pts],2),pos=1)
```

```{r}
Xsd <- scale(X)
tmp<-rbind(c(0,0),Xsd["407",])
mahalanobis(tmp,center=c(0,0),cov=cov(Xsd))
edist <- sqrt( (tmp[1,1] - tmp[2,1])^2 + (tmp[1,2]-tmp[2,2])^2 )
edist
```

## Dimension reduction using the Mahalanobis Distance


```{r}
plot(Xsd[,1],Xsd[,2],xlab="nhAboveHS/sd",ylab="nhPopD/sd")
```

## Dimension reduction using the Mahalanobis Distance


```{r}
drawMahal(X,center=colMeans(X),covariance=cov(X),quantile=c(.1,.2))
```

```{r}
covX <- cov(X)
newcovX <- covX
newcovX[1,2] <- 0
newcovX[2,1] <- 0
```

## Dimension reduction using the Mahalanobis Distance


```{r}
drawMahal(X,center=colMeans(X),covariance=newcovX,quantile=c(.1,.2))
```

## Matching on the Mahalanobis Distance

Here using the rank based Mahalanobis distance following DOS Chap. 8 (but comparing to the ordinary version).

```{r}
mhdist <- match_on(nhTrt~nhPopD+nhAboveHS,data=meddat,method="rank_mahalanobis")
mhdist[1:3,1:3]
mhdist2 <- match_on(nhTrt~nhPopD+nhAboveHS,data=meddat)
mhdist2[1:3,1:3]
mhdist2[,"407"]
```


```{r}
par(mgp=c(1.5,.5,0),oma=rep(0,4),mar=c(3,3,0,0))
drawMahal(X,center=colMeans(X),covariance=cov(X),
          quantile = c(0.975, 0.75, 0.5, 0.25))
abline(v=mean(meddat$nhAboveHS),h=mean(meddat$nhPopD))
cpts <-c("401","407","411")
tpts <-c("101","102","202")
arrows(X[tpts,1],X[tpts,2],rep(X["407",1]),rep(X["407",2]))
text(X[tpts,1],X[tpts,2],labels=round(mhdist2[tpts,"407"],2),pos=1)
mhdist2[tpts,"407"]

```


## Matching on the Mahalanobis Distance

```{r}
fmMh <- fullmatch(mhdist,data=meddat)
summary(fmMh,min.controls=0,max.controls=Inf)

fmMh1 <- fullmatch(mhdist,data=meddat,min.controls=1)
summary(fmMh1,min.controls=0,max.controls=Inf)

```

```{r}

xbMh <- xBalance(nhTrt~nhAboveHS+nhPopD,strata=list(unstrat=NULL,fmMh=~fmMh,fmMh1=~fmMh1),report="all",data=meddat)
xbMh

```


#  Matching on Many Covariates: Using Propensity Scores

## Matching on the propensity score

**Make the score**^[Note that we will be using `brglm` or `bayesglm` in the
future because of logit separation problems when the number of covariates
increases.]

```{r}
theglm <- glm(nhTrt~nhPopD+nhAboveHS+HomRate03,data=meddat,family=binomial(link="logit"))
thepscore <- theglm$linear.predictor
thepscore01 <- predict(theglm,type="response")
````

We tend to match on the linear predictor rather than the version required to
range only between 0 and 1.

```{r echo=FALSE, out.width=".7\\textwidth"}
par(mfrow=c(1,2),oma=rep(0,4),mar=c(3,3,2,0),mgp=c(1.5,.5,0))
boxplot(split(thepscore,meddat$nhTrt),main="Linear Predictor (XB)")
stripchart(split(thepscore,meddat$nhTrt),add=TRUE,vertical=TRUE)

boxplot(split(thepscore01,meddat$nhTrt),main="Inverse Link Function (g^-1(XB)")
stripchart(split(thepscore01,meddat$nhTrt),add=TRUE,vertical=TRUE)
```

## Matching on the propensity score

```{r}
psdist <- match_on(theglm,data=meddat)
psdist[1:4,1:4]
fmPs <- fullmatch(psdist,data=meddat)
summary(fmPs,min.controls=0,max.controls=Inf)
```

## Can you do better?

**Challenge:** Improve the matched design by adding covariates or functions of
covariates using either or both of the propensity score or mahalanobis distance
(rank- or not-rank based). So far we have:

```{r}
thecovs <- unique(c(names(meddat)[c(5:7,9:24)],"HomRate03"))
balfmla<-reformulate(thecovs,response="nhTrt")
xb4 <- balanceTest(update(balfmla,.~.+strata(fmMh)+strata(fmPs)),
                   data=meddat,report="all",p.adjust.method="none")
xb4$overall[,]
xb5 <- xBalance(balfmla,strata=list(fmMh=~fmMh,fmPs=~fmPs),
                   data=meddat,report="all")
xb5$overall
```

```{r}

newexp <- function(z,block){

}


```


## Can you do better?

Challenge: Improve the matched design. So far we have:

```{r}
plot(xb4)
```

## Summary:

What do you think?

 - Statistical adjustment with linear regression models is hard to justify.
 - Stratification via matching is easier to justify and assess (and describe).
 - Matching solves the problem of making comparisons that are transparent
   (Question: "Did you adjust enough for X?" Ans: "Here is some evidence about
   how well I did.")
 - You can adjust for one variable or more than one (if more than one, you
   need to choose one or more methods for reducing many columns to one
   column).
 - The workflow involves the creation of a distance matrix, asking an
   algorithm to find the best configuration of sets that minimize the
   distances within set, and checking balance. (Eventually, it will also be
   concerned about the effective sample size.)
  - Next: We will get more into the
   differences between full matching, optimal matching, greedy matching,
   matching with and without replacement, etc.. next week. (Also: handling
   missing data, calipers and other methods of improving design).
  - Next: How to estimate causal effects and test causal hypotheses with a
    matched design.



## References

