
## Covariate balance in experiments: What does it look like?

\begin{columns}
\begin{column}{.4\linewidth}
\begin{itemize}
\item \cite{arceneaux:2005}
\item Kansas City, November 2003
\item Completely randomized design: 14 precincts $\rightarrow$ Tx; 14 $\rightarrow $ Control.
\item Substantively large baseline differences (red dots)
\item Differences not large compared to other possible assignments from same design; compared to other possible experiments with the same design.
\item<2-> $\PP(\chi^{2} > x) = .91$ \citep{hansenbowers2008}. (grey lines)
\end{itemize}
\end{column}
\begin{column}{.6\linewidth}
\only<1>{\igrphx{KC-baseline}}
\only<2>{\igrphx{KC-bal+SDs}}
\end{column}
\end{columns}


## How did we do this?

```{r xb1, echo=TRUE}
acorn <- read.csv("data/acorn03.csv", row.names=1)
xb1 <- balanceTest(z ~ v_p2003 + v_m2003 + v_g2002 + v_p2002 + v_m2002 + v_s2001 +
         v_g2000 + v_p2000 + v_m2000 + v_s1999 + v_m1999 + v_g1998 +
         v_m1998 + v_s1998 + v_m1997 + v_s1997 + v_g1996 + v_p1996 +
         v_m1996 + v_s1996 + size, data=acorn, p.adjust.method = "none")

xb1$results
```

## How did we do this?

```{r xb1overall, echo=TRUE}
xb1$overall
```


```{r, out.width=".7\\textwidth"}
plot(xb1)
```


## DeMystifying balanceTest

```{r d1, echo=TRUE}
d.stat<-function(zz, mm, ss){
  ## this is the d statistic (harmonic mean weighted diff of means statistic)
  ## from Hansen and Bowers 2008 almost directly from balanceTest.Engine
  h.fn<-function(n, m){(m*(n-m))/n}
  myssn<-apply(mm, 2, function(x){sum((zz-unsplit(tapply(zz, ss, mean), ss))*x)})
  hs<-tapply(zz, ss, function(z){h.fn(m=sum(z), n=length(z))})
  mywtsum<-sum(hs)
  myadjdiff<-myssn/mywtsum
  return(myadjdiff)
}
```

## DeMystifying balanceTest

Recall our discussion of estimation "holding constant" within strata?

```{r d1v2, echo=TRUE}
## This is another version that might be more clear in regards what is going on.
dstatv2 <- function(zz,mm,ss){
    ## mm is a data.frame
    dat <- cbind(mm,z=zz,s=ss)
    datb <- dat %>% group_by(s) %>% summarize(across(.cols=all_of(names(mm)),function(x){ mean(x[z==1]) - mean(x[z==0])}),
        nb=n(),
        pib=mean(z),
        nbwt=nb/nrow(dat),
        hbwt0= pib * (1-pib) * nbwt)
    datb$hbwt <- datb$hbwt0/sum(datb$hbwt0)
    datb[,15:27]
    adjmns <- datb %>% summarize(across(.cols=all_of(names(mm)),function(x){ sum(x*hbwt) }))
    adjmnsmat <- as.matrix(adjmns)
    return(adjmnsmat)
}

```

## DeMystifying balanceTest

```{r nullddistsetup, echo=TRUE}
acorncovs<-c("v_p2003","v_m2003","v_g2002","v_p2002","v_m2002","v_s2001","v_g2000","v_p2000","v_m2000","v_s1999","v_m1999","v_g1998","v_m1998","v_s1998","v_m1997","v_s1997","v_g1996","v_p1996","v_m1996","v_s1996","size")

dstats1  <-d.stat(zz=acorn$z,mm=acorn[,acorncovs],ss=rep(1,nrow(acorn)))
dstats2  <-dstatv2(zz=acorn$z,mm=acorn[,acorncovs],ss=rep(1,nrow(acorn)))

dstats1[1:5]
dstats2[1:5]
```

## Calculate the reference distribution of the d-stat and the $d^2$ stat

For all vectors $z \in \Omega$ get `adj.diffs`. This is the distribution of the $d$ statistic for one-by-one balance assessment. Next question is about the distribution of the $d^2$ statistic: does it follow a $\chi^2$ distribution in this case?

```{r nulldist, cache=TRUE}
d.dist<-replicate(10000, d.stat(sample(acorn$z), acorn[,acorncovs], ss=rep(1,nrow(acorn))))
```

Get the randomization-based $p$-values:

```{r echo=TRUE}
xb1ds <- xb1$results[,"adj.diff",]
xb1ps <- xb1$results[,"p",]
obs.d<-d.stat(acorn$z, acorn[, acorncovs], rep(1,nrow(acorn)))
dps <- matrix(NA,nrow=length(obs.d),ncol=1)
for(i in 1:length(obs.d)){
  dps[i,] <- 2*min( mean(d.dist[i,] >= obs.d[i]),mean(d.dist[i,] <= obs.d[i]))
}
## You can compare this to the results from balanceTest
round(cbind(randinfps=dps[,1],xbps=xb1ps,obsdstats=obs.d,xbdstats=xb1ds),3)
```

## Calculate the reference distribution of the d-stat and the $d^2$ stat

The $d^2$ statistic is a linear function of the $d$-statistics that accounts
for the covariance between those statistics (across the possible assignments
under the null hypothesis of no effects).

```{r d2stat, echo=TRUE}
d2.stat <- function(dstats,ddist=NULL,theinvcov=NULL){
  ## d is the vector of d statistics
  ## ddist is the matrix of the null reference distributions of the d statistics
  if(is.null(theinvcov) & !is.null(ddist)){
    as.numeric( t(dstats) %*% solve(cov(t(ddist))) %*% dstats)
  } else {
    as.numeric( t(dstats) %*% theinvcov %*% dstats)
  }
}
```

## Calculate the reference distribution of the d-stat and the $d^2$ stat

The distribution of the $d^2$ statistic arises from the distribution of the d statistics --- for each draw from the set of treatment assignments we can collapse the $d$-statistics into one $d^2$. And so we can calculate the $p$-value for the $d^2$.

```{r d22, echo=TRUE}
## Here we have the inverse of the covariance/variance matrix of the d statistics
invCovDDist <- solve(cov(t(d.dist)))
obs.d2<- d2.stat(obs.d,d.dist,invCovDDist)

d2.dist<-apply(d.dist, 2, function(thed){
                 d2.stat(thed,theinvcov=invCovDDist)
         })
## The chi-squared reference distribution only uses a one-sided p-value going in the positive direction
d2p<-mean(d2.dist>=obs.d2)
cbind(obs.d2,d2p)
xb1$overall
```



## Why differences between balanceTest and d2?

I suspect that $N=28$ is too small. `balanceTest` uses an asymptotic
approximation to the randomization distribution.

```{r echo=FALSE, out.width=".8\\textwidth"}
## Notice that the distribution of d2.dist is not that close to the
## chi-squared distribution in this case with N=28
par(mfrow=c(1,2))
qqplot(rchisq(10000,df=21),d2.dist)
abline(0,1)

plot(density(d2.dist))
rug(d2.dist)
curve(dchisq(x,df=21),from=0,to=40,add=TRUE,col="grey")
```

## Does balanceTest have a controlled false positive rate here?

```{r xberror, echo=TRUE, cache=TRUE}
xbfn <- function(){
	acorn$newz <- sample(acorn$z)
	xb1 <- balanceTest(newz ~ v_p2003 + v_m2003 + v_g2002 + v_p2002 + v_m2002 + v_s2001 +
			v_g2000 + v_p2000 + v_m2000 + v_s1999 + v_m1999 + v_g1998 +
			v_m1998 + v_s1998 + v_m1997 + v_s1997 + v_g1996 + v_p1996 +
			v_m1996 + v_s1996 + size, data=acorn)
	return(xb1$overall[["p.value"]])
}

res <- replicate(1000,xbfn())
```

```{r resout, echo=TRUE}
summary(res)
mean(res <= .05)
mean(res <= .2)
```

## Does balanceTest have a controlled false positive rate here?

Ex. are fewer than 5% of the p-values less than .05?

```{r}
plot(ecdf(res))
abline(0,1)
abline(v=c(.01,.05,.1))
```


## Does the simulation based approach have a controlled false positive rate here?


```{r resdirecterror, cache=TRUE}
d2pfn <- function(z,X){
	newz <- sample(z)

	d.dist<-replicate(1000, d.stat(sample(newz), X, ss=rep(1,nrow(X))))

	obs.d<-d.stat(newz, X, rep(1,nrow(X)))

	dps <- matrix(NA,nrow=length(obs.d),ncol=1)
	for(i in 1:length(obs.d)){
		dps[i,] <- 2*min( mean(d.dist[i,] >= obs.d[i]),mean(d.dist[i,] <= obs.d[i]))
	}

	invCovDDist <- solve(cov(t(d.dist)))
	obs.d2<- d2.stat(obs.d,d.dist,invCovDDist)

	d2.dist<-apply(d.dist, 2, function(thed){
			       d2.stat(thed,theinvcov=invCovDDist)
		})

	d2p<-mean(d2.dist>=obs.d2)

	return(d2p)
}
```

```{r doresdirect, eval=FALSE, cache=TRUE}
resdirect <- replicate(1000,d2pfn(z=acorn$z,X=acorn[,acorncovs]))
```

```{r doresdirectparallel, eval=TRUE, cache=TRUE}
library(parallel)
resdirectlst <- mclapply(1:1000,function(i){ d2pfn(z=acorn$z,X=acorn[,acorncovs]) },mc.cores=detectCores())
resdirect <- unlist(resdirectlst)
save(resdirect,file="day9-resdirect.rda")
```

## Does the simulation based approach have a controlled false positive rate here?

It looks like it is a bit too high. Hmm... Maybe the simulation needs to be fixed.

```{r lazyload, echo=TRUE}
##lazyLoad("day9-AdjustmentBalance_cache/beamer/doresdirectparallel_5ec4fa8cdcbcf586138e928bc0f9fc0b")
##load("day9-resdirect.rda")
summary(resdirect)
mean(resdirect <= .05)
mean(resdirect <= .2)
```

## Does balanceTest have a controlled false positive rate here?


```{r}
plot(ecdf(resdirect))
abline(0,1)
abline(v=c(.01,.05,.1))
```

