%  For slides only
%\input{slidesonly}

% % For handout
%\input{handout}

%For handout + mynotes
\input{handout+mynotes}

\input{beamer-preamble-bbh-all}
\input{defs-all}
\input{courseedition}

\title[Day 18]{Day 18: $H^{3}$ sensitivity analysis; various}
% \author, date moved to beamer-preamble-*-all.tex


% copied from CSCAR svn repo, `dissdefs.tex`
\usepackage{xspace}
\newcommand{\satm}{\mbox{\textsc{sat-m}}\xspace}
\newcommand{\satv}{\mbox{\textsc{sat-v}}\xspace}
\newcommand{\upm}{\mbox{\textsc{urm}}\xspace}
\newcommand{\asian}{\mbox{\textsc{asian}}\xspace}
\newcommand{\presatm}{\mbox{\textsc{pre-m}}\xspace}
\newcommand{\premath}{\mbox{\textsc{pre-m}}\xspace}
\newcommand{\presatv}{\mbox{\textsc{pre-v}}\xspace}
\newcommand{\preverb}{\mbox{\textsc{pre-v}}\xspace}
\newcommand{\parentsinc}{\mbox{\textsc{incm}}\xspace}
\newcommand{\gpa}{\mbox{\textsc{gpa}}\xspace}
\newcommand{\dadsed}{\mbox{\textsc{dadsed}}\xspace}
\newcommand{\momsed}{\mbox{\textsc{momsed}}\xspace}
\newcommand{\avgeng}{\mbox{\textsc{e-gpa}}\xspace}
\newcommand{\avgmath}{\mbox{\textsc{m-gpa}}\xspace}
\newcommand{\avgnatsci}{\mbox{\textsc{ns-gpa}}\xspace}
\newcommand{\avgssci}{\mbox{\textsc{ss-gpa}}\xspace}
\newcommand{\coach}{\mbox{\textsc{coach}}\xspace}
\newcommand{\dadcoll}{\mbox{\textsc{dadcoll}}\xspace}
\newcommand{\aavg}{\mbox{\textsc{a-avg}}\xspace}

\begin{document}

\input{announcement-of-the-day}

\section{$H^{3}$ sensitivity analysis; PS $\leftrightarrow$ OV ``interaction''}


\begin{frame}
  \frametitle{Background: omitted variables}
  
  Almost all findings from observational studies could in principle be
  explained by an unmeasured variable. But how strong a confounder
  would be needed?  Rosenbaum (1987,\ldots) quantifies confounding in
  terms of propensity scores.
  
  \begin{center}
  \igrphx[height=.6\textheight]{rb2010p77}
\end{center}

  An analytically easier course uses regression to quantify confounding. 
\end{frame}

\begin{frame}
  \frametitle{Causal inference via covariance adjustment}


When we fit a model
$$
Y = a + Zb +  \mathbf{X} c + e,\,\,\, e \sim \mathcal{N}(0, \sigma^{2}),
$$
the estimated $b$ merits interpretation as the causal effect of treatment $Z$ if\\
\begin{enumerate}
\item $(Y_{t}, Y_{c}) \perp Z | X$
\item The linear relationship well-enough describes the finite population of interest.
\end{enumerate}

The first of these assumptions is usually untestable, and more central. 

\end{frame}

\begin{frame}
  \frametitle{Sensitivity analysis for the linear model}
\framesubtitle{An omitted variable}

We fit
$$
Y = a + Zb +  \mathbf{X} c + e,\,\,\, e \sim \mathcal{N}(0, \sigma^{2}),
$$
but would have liked to have fit
$$
Y = \alpha + Z\beta +  \mathbf{X} \gamma + W\zeta + e,\,\,\, e \sim \mathcal{N}(0, \sigma^{2}).
$$

\begin{enumerate}
\item<2-> How different are $b$ and $\beta$?
\item<3-> How different are the confidence intervals for $b$ and $\beta$?
\end{enumerate}

\end{frame}

 
\begin{frame}
  \frametitle{Sensitivity analysis for the linear model}
\framesubtitle{``speculation parameters''}

We can quantify the relationship of $b$ to $\beta$, and of $\mathrm{se}(b)$ to
$\mathrm{se}(\beta)$, in terms of 2 ``speculation parameters'':
\begin{enumerate}
\item $t_{w}$, the $t$-statistic associated with $W$'s coefficient in the (OLS) regression of $Z$ on $W$ and $\mathbf{X}$;
\item $R^2_{y\cdot z\mathbf{x} w}$, the
coefficient of multiple determination of the regression of $Y$ on $Z$,
$\mathbf{X}$, and $W$.
\end{enumerate}

More specifically, we can bound the upper and lower limits of conventional confidence intervals in terms of $t_{w}$ alone, or (often more sharply) in terms of $t_{w}$ and $R^2_{y\cdot z\mathbf{x} w}$.
\end{frame}
\begin{frame}
  \frametitle{$W$-insensitive confidence bounds as a function of $t_{w}$}

  \begin{prop}[Hosman, Hansen \& Holland, 2010]
Let $T>0$.  If $|t_{w}| \leq T$ then 
$$
\hat{\beta}\pm q\widehat{\mathrm{se}} (\hat{\beta}) \subseteq 
\hat{b} \pm \left( \sqrt{T^2 + q^{2} \cdot \frac{T^2 + 
n-r(\mathbf{X})-2}{n-r(\mathbf{X})-3} } \right) \widehat{\mathrm{se}}(\hat{b}).
$$
\end{prop}
\end{frame}

\frame<1-4>[label=probsstudy]
{
  \frametitle{Application: An observational study of coaching for the SAT}

Powers \& Rock (1999) sampled one in 200 SAT-I registrants in 1995-96.
\begin{itemize}
\item<2-| alert@+> The ``treatment'' is being coached for the SAT.  This
  information comes from survey responses.
\item<3-| alert@+> Outcomes, \textit{i.e.} SAT scores, come from the College Board's
administrative records.
\item<4-| alert@+> Many students took the SAT or PSAT before being coached; so
  there are pretest scores too. \pause
\item<5-| alert@+> Covariate is high-dimensional (and a little bit messy).
\end{itemize}
}

\begin{frame}  \frametitle{Sensitivity of estimates of \satm benefit of coaching }
\framesubtitle{\satm \textasciitilde \coach + \presatm + \presatv + \asian + \upm + \dadcoll + \aavg + \avgmath}

\newlength{\pholderlngth}
\settowidth{\pholderlngth}{[aims for postgrad. deg.]}
\begin{center}
\begin{tabular}{|l|l|r|} \hline
\parbox{\pholderlngth}{Excluding from predictors in regression of \coach on 
\asian,\ldots, \avgmath the
 variable}  & 
\parbox{2cm}{dullens prediction of \coach by $t=$}
    & \parbox{2.5cm}{Then if $W$ is s.t. $t_w \leq t$,
$\beta \pm 2\widehat{\mathrm{se}}(\hat{\beta})  \subseteq$} \\  
\hline
\asian &       8.1 & [-2,44] \\
\upm  &       3.2 &  [11,32] \\
\dadcoll &       7.9 & [-1,44]\\
\presatm &       2.4 &  [13,30 ]\\
\presatv  &       -2.4&  [13,30 ]  \\ 
\aavg    &      0.5 &  [16,27] \\ 
\avgmath &      -2.0& [13,29 ] \\\hline 
 \end{tabular}
\end{center}

\end{frame}

\begin{frame}
  \frametitle{Confidence limits for math effect under several omitted variable scenarios }


\begin{columns}
\column{.5\linewidth}%
{
\igrphx[width=\linewidth]{SensAnal-mathSA}

\bigskip
{\footnotesize Propensity score stratification + fixed effects regression}
}

\column{.5\linewidth}%
{
\only<2->{
\igrphx[width=\linewidth]{climits_ovsa1}

\bigskip
{\footnotesize Covariate adjustment via OLS}
}
}
\end{columns}

$H^{3}$ (2010) show this phenomenon in another example.
\end{frame}

\section[R.I. w/ modeled outcomes]{Randomization inference w/ modeled outcomes}
\subsection[U3 review]{Unit 3 review: Confidence intervals by inversion of tests}
\begin{frame}{Confidence intervals and models of effects}

  \begin{itemize}[<+->]
  \item The 3 models just considered fall under the broader model that the GOTV
generated $v$ votes per contact, some $v \in [-1, 1]$.
\item Assuming that model, the 95\% CI for $v$ is the collection of $v$s corresponding to incomplete response schedules that would not be rejected at level .05 - a confidence interval by inversion of a family of hypothesis tests.
\item This enables us to back out a confidence interval for the
  ``Complier average treatment effect,'' w/o violating the
  intention-to-treat principle! (Rosenbaum, 1997).
\item In addition, this method of confidence interval construction is
  immune to the problem of weak instruments (cf. \textit{DOS} \S~5.3 \&
  contained references).
  \end{itemize}


\end{frame}

\begin{frame}{Estimates and models of effects}
  
  \begin{itemize}
\item If you want an estimate to go with such a confidence interval,
  the convention is to report a \textit{Hodges-Lehmann} estimate ---
  the limit of 100*$(1-\alpha)$\% CIs, as $\alpha \uparrow
  1$\footnote{Assuming the limit exists and contains a single
    point. If you want to compute (rather than interpret) an HL
    estimate, get the precise definition, as e.g. in \textit{DOS} \S~2.4.3. }. 
\item There's no precise analogue of the ``standard error''\ldots
\item So you can't use $\hat{\theta} \pm z_{*}\mathrm{s.e.}(\hat
  \theta)$ as a CI. 
\item However, for the spirit of a ``standard error'', some report a 2/3 CI alongside a 95\% CI  (Mosteller \& Tukey, 1977, \textit{Data analysis and regression}).
  \end{itemize}
\end{frame}
\subsection{Example: Violence in Medell{\'i}n} 
\begin{frame}
\frametitle{Example: Violence and public infrastructure in Medell{\'i}n, Colombia}
\begin{columns}
  \begin{column}{.5\linewidth}
    \begin{itemize}
    \item 2 million residents; 16 districts
    \item Pre-intervention: 60\% poverty rate, 20\% unemployment, homicide 185 p
er 100K
    \item High residential segregation
    \item<2-> 2004-2006: infrastructure intervention for certain poor neighborho
ods.
    \end{itemize}
  \end{column}
  \begin{column}{.5\linewidth}
    \only<1>{\igrphx{medellin-conc-pov}}
\only<2-\mynoteonly>{\igrphx{medellin-gondola}}
  \end{column}
\end{columns}

\end{frame}
\itnote{
\item 2017+: Cycled back to this later in course, not during unit 3.
  Ordinarily it's too much at first pass.
\item (strata will enter the story because the end analysis involved
  matching)
}



\begin{frame}[fragile]{A model of effects for homicide rates in Medellin}

Each Medellin neighborhood provides several years of non-independent
data.  If we're willing to model Metrocable as a natural experiment,
we can borrow longitudinal data methods, without having to adopt their
dependence assumptions.   For instance, using random
 effects: 

 \begin{enumerate}[<+->]
 \item Let
   $y(T)=$ neighborhood homicide rate in year $T=2002, \ldots, 2008$.  We'll test models of
   effects of form, for rates $r \leq 0$:
$$H_{0}: y_{t}(T) = \left\{ \begin{array}{lr}\exp((T-2004) r/4)
                              y_{c}(T),& T> 2004; \\
y_{c}(T),& T\leq 2004.\end{array} \right.$$ 
 \item Given $H_{0}$, fit a random effects model of this form:
\begin{semiverbatim}
 lmer(Count \textasciitilde\  year + (year+1|nh) +\ \pause
            \alert<@+| handout:0>{offset}( nhTrt*(yr>2004)*(r/4)*(yr-2004) ),
          family=poisson, data=homd)
\end{semiverbatim}
\item Fitted params include  ``fixed'' intercept and slope,
  $\hat{\beta}_{0(r)}$ and $ \hat{\beta}_{1(r)}$,\pause  plus a
  ``random'' slope and intercept (also specific to $r$) for
  each n-hood.  
\item To test $H_{0}$, I used a test statistic comparing $t$ vs $c$
  n-hoods' fitted intercepts. For reference distribution, 
  intercepts shuffled within matched sets.
 \end{enumerate}

  
\end{frame}

\begin{frame}<\nottheirhandout>{Outcome analysis for the Medellin study}
\framesubtitle{Official statistics on n-hood violence}
  
  \begin{center}
    \igrphx[width=\linewidth]{medellin-TEs-homicide}
  \end{center}

\end{frame}



\begin{frame}<\nottheirhandout>{Outcome analysis for the Medellin
    study}
\framesubtitle{Outcomes constructed from survey responses}

\begin{center}
      \igrphx[width=\linewidth]{medellin-TEs-amenities}
\end{center}
\end{frame}

\section{matching prior to probability sampling}


\begin{frame}{Example: propensity matching prior to cluster sampling}
  \framesubtitle{to balance internal and external validity}
  \begin{itemize}[<+->]
  \item \only<2->{\sout}{Question: how did Grutter v. Bollinger affect applicants'
    responses to the diversity essay?}
\item Question: how applicants responses to diversity essay vary by
  race of applicant?
\item B/c race was so confounded w/ class, Kirkland \& Hansen (2011)
  propensity-matched, then studied a sample of propensity matched
  sets (clusters) and unmatched applicants.
  \end{itemize}
  
  \addtolength{\tabcolsep}{-\tabcolsepadj}
  {\footnotesize
  \begin{center}
\begin{tabular}{rrrrrrrrr}
  \hline & \multicolumn{2}{c}{Matched:} & \multicolumn{2}{c}{Matches:} & \multicolumn{1}{c}{Sampling} & \multicolumn{2}{c}{Unmatched:} & \multicolumn{1}{c}{Smpling} \\ & Tot. \# & Spl. \# & Tot. \# & Spl. \# & weight & Tot. \# & Spl. \# & weight. \\ 
  \hline
[0,30] & 1400 & 54 & 540 & 20 & 27 & 1600 & 10 & 160 \\ 
  (30,60] & 2600 & 45 & 880 & 15 & 59 & 9300 & 10 & 930 \\ 
  (60,100] & 610 & 47 & 200 & 15 & 14 & 2200 & 10 & 220 \\ 
   \hline
\end{tabular}
\end{center}
}
  \addtolength{\tabcolsep}{\tabcolsepadj}
\end{frame}

\end{document}
