---
title: |
 | Open Discussion of Various Topics
date: '`r format(Sys.Date(), "%B %d, %Y")`'
author: |
  | ICPSR 2021 Session 2
  | Jake Bowers, Ben Hansen, Tom Leavitt
bibliography:
 - 'BIB/MasterBibliography.bib'
 - 'BIB/master.bib'
 - 'BIB/misc.bib'
 - 'BIB/refs.bib'
fontsize: 10pt
geometry: margin=1in
graphics: yes
biblio-style: authoryear-comp
colorlinks: true
biblatexoptions:
  - natbib=true
output:
  beamer_presentation:
    slide_level: 2
    keep_tex: true
    latex_engine: xelatex
    citation_package: biblatex
    template: icpsr.beamer
    includes:
        in_header:
           - defs-all.sty
    md_extensions: +raw_attribute-tex_math_single_backslash+autolink_bare_uris+ascii_identifiers+tex_math_dollars
    pandoc_args: [ "--csl", "chicago-author-date.csl" ]
---


```{r setup1_env, echo=FALSE, include=FALSE}
library(here)
source(here::here("rmd_setup.R"))

## Print code by default
opts_chunk$set(echo = TRUE)
```

```{r setup2_loadlibs, echo=FALSE, include=FALSE}
## Load all of the libraries that we will use when we compile this file
## We are using the renv system. So these will all be loaded from a local library directory
library(dplyr)
library(ggplot2)
library(coin)
library(RItools)
library(optmatch)
```


## Today


  1. Agenda: Talk about a few topics you mentioned last time and/or any topics
     that you came to class wanting to discuss: TSCS/Longtudinal, Interference,
     Pointer to Regression-based sensitivity analysis.
  2. Questions arising from the reading or assignments or life?

# But first, review

## What have we done so far recently? {.fragile}

 1. Assessed the sensitivity of a design which adjusted well for $\mathbf{x}$ but
    which could not adjust directly for unmeasured confounders, $\mathbf{u}$
    1. Rosenbaum's approach focusing on $\Gamma$ (assuming strong relationship
       with $Y$). (And briefly mentioning how $\Gamma$ can be decomposed into
       $\delta$ (increase in odds of a positive pair difference --- i.e. effect
       on outcomes), and $\lambda$ (increase in odds of a treatment difference
       --- i.e. effect on treatment/causal driver) to aid interpretation.
    2. Fogarty's approach also using $\Gamma$ but in the context of testing
       hypotheses about no average effect.

\begin{center}
\begin{tikzcd}[column sep=large, row sep=large,every arrow/.append style=-latex]
  \mathbf{x}=\{x_1,x_2,\ldots\} \arrow[from=1-1,to=2-1, "\text{0 by adjustment}" description] \arrow[from=1-1,to=2-4, "?" description]  & & &\\
  Z  \arrow[from=2-1,to=2-4, "\tau" description] &    &                           & Y \\
  \mathbf{u} \arrow[from=3-1,to=2-1, "\Gamma \text{ or } \lambda" description] \arrow[from=3-1,to=2-4, "\delta" description]
\end{tikzcd}
\end{center}

## What have we done so far recently?

 2. Introduced the ideas behind Differences in Differences Designs (like
    interrupted time-series but with a control group) and their focus on the
    estimand of the $ATT_{\text{Population}}$ (pointers to other estimators.
    Difference-in-Differences Estimators refer to this kind of design and focus
    on this estimand.)
    1. Addresses problems with the simple interrupted time series approach
    2. Using own unit as counterfactual makes sense (but raises question about
       history/trend other $\mathbf{u}$ that could be intervening in between
       measurements)
    3. Adding the non-intervention or control unit(s) helps under some
       assumptions (or pre-sumptions) about using this group to impute
       potential outcomes for the focal/treatment group.
    4. Teaser about estimating the trends rather than assuming them (see
       Leavitt's work and citations therein).

# Longitudinal Data

## Two simple questions

 1. What is the impact of an event on a trajectory?
 2. How do trajectories differ between types? ("treated types" and "control
   types" or any types).

## Trajectories as Types

We can learn about #2 using existing tools: multiple observations can be
collapsed into a trajectory and we can work to clarify comparisons in a
categorical outcome between units. Here we create "types" directly. One could
imagine doing some latent class or cluster analysis etc.. to categorize types
of trajectories.

```{r}
## Trajectories Y_i_t for three units

time_length <- 7
times <- seq_len(time_length)
Y_1_t <- c(1, 0, 1, 1, 0, 0, 1)
Y_2_t <- c(0, 0, 0, 0, 0, 0, 1)
Y_3_t <- c(0, 1, 1, 1, 1, 1, 1)

longdat1 <- data.frame(
  id = c(rep(1, time_length), rep(2, time_length), rep(3, time_length)),
  Y = c(Y_1_t, Y_2_t, Y_3_t),
  times = rep(times, 3),
  type = c(rep(1, time_length), rep(0, time_length), rep(1, time_length))
)

longdat1

widedat1 <- longdat1 %>%
  group_by(id) %>%
  summarize(
    Trajectory = paste(Y, collapse = ""),
    Type = unique(type)
  )

widedat1
```

 - One could further classify trajectories as "Rising" versus "Falling" versus "Stable" and simplify the analysis.
 - One could add "time of switching to type=1" as another variable if you observe this and it matters

## What is the impact of an event on a trajectory?

```{r echo=FALSE}
load("/Users/jwbowers/Documents/PROJECTS/jakebowers.org/ICPSR/ps4wl.rda")
```

What if we have an event occurring in the middle of a trajectory? What is the
causal effect of this event? Here the question is about the effect of having a
child on political participation (see larger literature on gender and political
participation). This structure could be the same as effect of "treaty signing" etc..

```{r}
## One person:
ps4wl %>%
  filter(id65 == 8) %>%
  select(one_of(c(
    "id65", "year", "female", "allacts", "cumallacts",
    "futureacts", "kidbornN", "kage1", "kage2", "maxdegP"
  )))
```

What would be a good comparison person (or persons) for person 8 if we could find that person?

## Basics of matching for event data:

Here, I select only rows where people have a child or rows before people have a child (or rows where no child was born):

```{r}

wrk.df <- ps4wl %>%
  filter((ps4wl$k1born == 1) | (ps4wl$k1born == 0 & (ps4wl$k1year < 0 | is.na(ps4wl$k1year)))) %>%
  select(one_of(c(
    "id65", "year", "maxdegP", "pmdeg65", "female", "allacts", "cumallacts",
    "futureacts", "k1year", "kage1",
    "statehs", "havekids", "hsact65", "civics65", "churP", "pintP", "ybyear", "k1born", "kid1"
  )))

## For the purposes of design, we do not want years after birth of first child.
wrk.df %>% filter(id65 == 8)
```

Let's look for people who have not yet had a child but who have had a similar previous trajectory of political participation:

```{r}
## But we want allacts and cumallacts for the year before the kid was born in the treatment group.
wrk.df$baselineacts <- wrk.df$allacts
wrk.df$baselineacts[wrk.df$k1born == 1] <- wrk.df$allacts[which(wrk.df$k1born == 1) - 1]
wrk.df$baselinecumacts <- wrk.df$cumallacts
wrk.df$baselinecumacts[wrk.df$k1born == 1] <- wrk.df$cumallacts[which(wrk.df$k1born == 1) - 1]

## fix this for thos people who had a kid in 1965 (i.e. don't use someone else's 1997 data for them).
wrk.df$baselineacts[wrk.df$k1born == 1 & wrk.df$year == 65] <- wrk.df$allacts[wrk.df$k1born == 1 & wrk.df$year == 65]
wrk.df$baselinecumacts[wrk.df$k1born == 1 & wrk.df$year == 65] <- wrk.df$cumallacts[wrk.df$k1born == 1 & wrk.df$year == 65]


actsDist <- match_on(k1born ~ baselineacts + baselinecumacts,
  data = wrk.df,
  within = exactMatch(k1born ~ female, wrk.df)
)
```

Want to match people closer in year:

```{r}
## Make a distance matrix by hand
yearDist <- with(wrk.df, outer(year[k1born == 1], year[k1born == 0], "-"))
dimnames(yearDist) <- list(
  rownames(wrk.df[wrk.df$k1born == 1, ]),
  rownames(wrk.df[wrk.df$k1born == 0, ])
)
```

Don't match same person to self for purposes of design (maybe use own pre-baby
participation as a change score after matching) (Why or why not woudl we try
Difference-In-Differences here with or without matching?)

```{r}
id65Dist <- with(wrk.df, outer(id65[k1born == 1], id65[k1born == 0], function(x, y) {
  as.numeric(x == y)
}))
dimnames(id65Dist) <- list(
  rownames(wrk.df[wrk.df$k1born == 1, ]),
  rownames(wrk.df[wrk.df$k1born == 0, ])
)
```

A mahalanobis distance:


```{r}
## Make a version of the dataset with no missing data
wrk.df.noNAs <- fill.NAs(wrk.df)

## Make a mh distance matrix
mhDist <- match_on(k1born ~ baselineacts + baselinecumacts + maxdegP + pmdeg65 +
  pmdeg65.NA + hsact65 + civics65 + churP + churP.NA,
data = wrk.df.noNAs,
within = exactMatch(k1born ~ female, data = wrk.df.noNAs)
)
summary(mhDist)
```

Also a propensity score:

```{r}
## A propensity to have kids year score
## Look at cov scaling
## afmla <- k1born~baselineacts+baselinecumacts+maxdegP+pmdeg65+
## 	       pmdeg65.NA+hsact65+civics65+churP+churP.NA
## summary(wrk.df.noNAs[,all.vars(afmla)])

library(lme4)

pscore_glm <- glmer(k1born ~ female + I(scale(baselineacts)) + I(scale(baselinecumacts)) + maxdegP + pmdeg65 +
  hsact65 + civics65 + churP + churP.NA + (1 | id65),
family = binomial(link = "logit"), data = wrk.df.noNAs
)

## res <- allFit(pscore_glm)

wrk.df.noNAs$psscore <- predict(pscore_glm)
psDist <- match_on(k1born ~ psscore,
  data = wrk.df.noNAs,
  within = exactMatch(k1born ~ female, data = wrk.df.noNAs)
)
summary(psDist)
```

Now create the overall distance matrix reflecting these decision:

```{r}
## make self-matches infinite
baselineactsDist.pen1 <- actsDist + caliper(id65Dist, width = 0)
sometrtnms <- rownames(actsDist)[1:10]
as.matrix(id65Dist)[sometrtnms, grep("^8\\.", colnames(id65Dist), value = TRUE)]
as.matrix(baselineactsDist.pen1)[sometrtnms, grep("^8\\.", colnames(baselineactsDist.pen1), value = TRUE)]

## Don't match to controls in the future:
baselineactsDist.pen2 <- baselineactsDist.pen1 + caliper(yearDist, width = 0, compare = `>`)

## Require matches close in year
baselineactsDist.pen3 <- baselineactsDist.pen2 + caliper(yearDist, width = 3)

## And not too distant in mhDist or psDist
baselineactsDist.pen4 <- baselineactsDist.pen3 + caliper(mhDist, quantile(as.vector(mhDist), .9)) + caliper(psDist, quantile(as.vector(psDist), .9))
```

We have 85 moments where a child is born and we want to find moments where
children are not born for people like those who decided to have a child, near
the same moments in their lives and also the same moments in history.

```{r}
## How many "treated" observations?
table(wrk.df$k1born)

pm1 <- pairmatch(baselineactsDist.pen4,
  data = wrk.df.noNAs,
  remove.unmatchables = TRUE
)
summary(pm1)

wrk.df.noNAs$pm1 <- pm1

wrk.df.noNAs %>% filter(pm1 == 1.1)
```

## Where to go from here

See Chapter 13, "Risk Set Matching", of @rosenbaum2010 and associated references.

Promising work. I haven't read it.

 - Maybe promising review: @thomas2020matching
 - @blackwell2018telescopematching
 - @imai2019matching (Also Kim and Imai and colleagues have a series of papers
   on "Fixed effects" models that addresses the same design topic ---
   obervational studies with multiple observations per unit, and a desire to
   estimate average causal effects using stratification but **not** using
   future observations (say) to estimate effects of past events.)
 - Consider Mahalanobis distance on all past outcomes? (See Nielson on
   "Matching with Time-Series Cross-Sectional Data")


# Interference

## Overview

Just to demonstrate what is behind the presumption of "no interference" and to
show just one very simple approach (that you already know).


## Statistical inference with interference?

\includegraphics[width=.3\textwidth]{images/complete-graph.pdf}

\includegraphics[width=.95\textwidth]{images/interference-example.pdf}

On estimation see (Sobel, Aronow, Basse, Eckles, Feller, Samii, Hudgens,
Ogburn, VanderWeele, Toulis, Kao, Coppock, Sicar, Raudenbush, Hong, Volfovsky).
The key question for that work: What is the function of potential outcomes that
we can estimate using observed data?

## Statistical inference with interference?

\includegraphics[width=.3\textwidth]{images/complete-graph.pdf}

\includegraphics[width=.95\textwidth]{images/interference-example-2.pdf}

Introducing the potential outcome when no treatment has been applied to any
node aka \textbf{the uniformity trial} $\equiv \by_{i,0000}$ (Rosenbaum, 2007).


## Imagine an experiment on a network:

\includegraphics[width=.95\textwidth]{images/fishersutvaNetwork.pdf}

## A model of propagation

 - The direct effect of treatment is $\beta$ (it is a multiplicative effect).
 - Treatment effects flows from treated to control units, increasing with
   number of treated neighbors, with rate of growth of effect $\tau$.

\includegraphics[width=.7\textwidth]{images/fishersutvaModel.pdf}

## Evidence about the parameters

\includegraphics[width=.5\textwidth]{images/ssr2d.pdf}

Plot shows the proportion of $p$-values less than .05 for randomization tests of
joint hypotheses about $\tau$ and $\beta$. Darker values mean less rejection.
Truth is at $\tau = .5$, $\beta= 2$. All tests reject the truth no more than 5% of the
time at $\alpha = .05$. All simulations using permutation-based reference distributions.

## Summary of Hypotheses as Causal Models

A causal model relates potential outcomes to each other and a research design
relates potential outcomes to observed data (and to sources of uncertainty). For examples:

 -  For no effects at all: $y_{i,0} = \HH(Y_i,Z_i,\tau_i) = Y_{i}$.
 -  For constant, additive effects: $y_{i,0} = \HH(Y_i,Z_i,\tau_i) = Y_{i} - Z_i \tau$.
 - For vector valued outcomes in a network with nonlinear propagation of causal effects:
\begin{equation}
\yu = \HH(\by_{\bz}, \bzero, \beta, \tau) = ( \beta + (1 - z_i) (1 - \beta) \exp( - \tau^2 \bz^{T} \bS ) )^{-1} \by \bz
\end{equation}


In fact *any function* that produces vectors of $\yu$ could be used to
represent these kinds of causal models (including agent based models ---
imagining agents revealing potential outcomes depending on network, treatment,
and covariates).


## A General Testing-Based Causal Inference Algorithm

 1. Write a model converting uniformity trial potential outcomes (like $\yu$ or
    simply $y_{i,0}$) into observed data (like $\by_{\bz}$ or simply $Y_i$):
    What is the mechanism by which the treatment changes the outcomes of the
    units? (This is a structural model of potential outcomes. It could be an
    agent-based model.)
 2. Solve for $\yu$. What adjustment of the observed data does this model
    imply? $\HH(\by_{\bz}, \bzero, \theta_0) = \yu$ like $y_{i,0} = Y_i + Z
    \tau$ for the simple constant, additive effects model.
 3. Select a test statistic that is effect-increasing in all relevant
    dimensions (like the sum of squared residuals test statistic or the KS-test
    statistic for certain models, or, I conjecture, an energy-statistic).
 4. Compute $p$-values for substantively meaningful range of $\theta$.  Or
    calculate boundaries of regions. (Perhaps collapse or aggregate the
    rejection regions to aid interpretation.)

## Where to to go from here

**Neyman:** Estimating some average causal effects @aronowsamii2017 weighting by
   probability of exposure to the treatment via the network (see associated
   @toulis2013estimation).

   \medskip
Lots of work here now. Estimating average effects when agnostic to
     networks, for example. Etc. Check out work by Hudgens
     <http://www.bios.unc.edu/~mhudgens/>, Ogburn <https://www.eogburn.com>,
     Basse <https://gwbasse.com>, Eckles <https://www.deaneckles.com>, Toulis
     <https://www.ptoulis.com/working-papers>, Volfovsky
     <https://volfovsky.github.io>.

Check out: 
 - <https://cran.r-project.org/web/packages/inferference/index.html>
 - <https://github.com/szonszein/interference> (See for example <https://rdrr.io/github/szonszein/interference/man/estimates.html>)

\medskip

**Fisher:** Testing causal theories that generate hypotheses about effects (as
   it propagates across a network) @bowers2013sutva, @bowers2016research,
   @bowers2018models.


# Sensitivity Analysis for Regression Models

## Sensitivity Analysis for Regression Models: A pointer

Read  @cinellihazlett2020 and @hosman2010 and @chaudoinetal2018.
@cinellihazlett2020 is current state of the art (see their easy to use R
package, youtube videos, etc).

# Futher Topics and Discussion?

## Questions, Thoughts?




## References
