---
title: Matching Tools and Information in a Block-Randomized Experiment
date: '`r format(Sys.Date(), "%B %d, %Y")`'
author: ICPSR 2018 Session 2
bibliography:
 - refs.bib
 - BIB/master.bib
 - BIB/misc.bib
fontsize: 10pt
geometry: margin=1in
graphics: yes
biblio-style: authoryear-comp
biblatexoptions:
  - natbib=true
output:
  beamer_presentation:
    slide_level: 2
    keep_tex: true
    latex_engine: xelatex
    citation_package: biblatex
    template: icpsr.beamer
    includes:
        in_header:
           - defs-all.sty
---

<!-- Make this document using library(rmarkdown); render("day12.Rmd") -->


```{r include=FALSE, cache=FALSE}
# Some customization.  You can alter or delete as desired (if you know what you are doing).
# knitr settings to control how R chunks work.
rm(list=ls())

require(knitr)

## This plus size="\\scriptsize" from https://stackoverflow.com/questions/26372138/beamer-presentation-rstudio-change-font-size-for-chunk

knitr::knit_hooks$set(mysize = function(before, options, envir) {
			      if (before)
				      return(options$size)
})

knit_hooks$set(plotdefault = function(before, options, envir) {
		       if (before) par(mar = c(3, 3, .1, .1),oma=rep(0,4),mgp=c(1.5,.5,0))
})

opts_chunk$set(
  tidy=FALSE,     # display code as typed
  echo=TRUE,
  results='markup',
  strip.white=TRUE,
  fig.path='figs/fig',
  cache=FALSE,
  highlight=TRUE,
  width.cutoff=132,
  size='\\scriptsize',
  out.width='.8\\textwidth',
  fig.retina=FALSE,
  message=FALSE,
  comment=NA,
  mysize=TRUE,
  plotdefault=TRUE)

if(!file.exists('figs')) dir.create('figs')

options(digits=4,
	scipen=8,
	width=132,
	show.signif.stars=FALSE)
```

```{r eval=FALSE, include=FALSE, echo=FALSE}
## Run this only once and then not again until we want a new version from github
library('devtools')
library('withr')
with_libpaths('./lib', install_github("markmfredrickson/RItools"), 'pre')
```

```{r echo=FALSE}
library(dplyr)
library(ggplot2)
library(RItools,lib.loc="./lib")
library(optmatch)
library(chemometrics) ## for drawMahal
library(mvtnorm)
```

## Today

\begin{enumerate}
  \item Agenda: How to avoid overfit propensity score functions; How to characterize the information in a matched design (use
  what we know about the variance of estimators from block-randomized experiments).
  \item Reading for tomorrow and next week: DOS 8--9, 13 and \cite[\S~9.5]{gelman2006dau}, and \cite{hans:04} (on full matching versus fixed ratio matching versus pair matching), \cite{gelman2008weakly} on "separation" problems in logistic regression other references on the problem of "overfitting" in general.
  \item Questions arising from the reading or assignments or life?
\end{enumerate}

# But first, review:


\begin{frame}[fragile]
\frametitle{What have we done so far?}

The problem of attrition in randomized studies (simple experiments,
encouragement designs, etc.)

\medskip

Making the case for adequate adjustment in observational studies

\begin{itemize}
\item The potential for making this case for one or more variables using
stratification and the problems with making this case using the linear model
to "control for" covariates.
\item Using optimal, full matching technology to make and evaluate
stratifications with a distance-matrix as input.
\begin{itemize}
\item Mahalanobis distance matrices compare units on the covariates in the $ n \times p$ matrix $\bm{x} = \{ \bm{x}_1, \bm{x}_2,
\ldots \}$.
\item The propensity score distance matrices compare units on $\hat{Z} \leftarrow \bm{x}$ or $\hat{Z}=\hat{\bm{\beta}}\bm{x}$ (the elements of $\bm{x}$ as it relates to treatment or intervention).
\item Calipers to restrict set membership based on substantively important
variables and/or extreme differences on the score matrices.
\item Exact matching to restrict set membership based on groups or categories
that are substantively important (as a side effect this speeds up the matching
process).
\end{itemize}
\end{itemize}

\end{frame}


```{r echo=FALSE, cache=TRUE}
load(url("http://jakebowers.org/Data/meddat.rda"))
meddat<- mutate(meddat,
		HomRate03=(HomCount2003/Pop2003)*1000,
		HomRate08=(HomCount2008/Pop2008)*1000)
row.names(meddat) <- meddat$nh
```


## The separation problem in logistic regression and the benefits of shrinkage/penalization

```{r echo=FALSE}
thecovs <- unique(c(names(meddat)[c(5:7,9:24)],"HomRate03"))
balfmla<-reformulate(thecovs,response="nhTrt")
library(splines)
psfmla <- update(balfmla,.~.+ns(HomRate03,2)+ns(nhPopD,2)+ns(nhHS,2))
glm0 <- glm(balfmla,data=meddat,family=binomial(link="logit"))
glm1 <- glm(psfmla,data=meddat,family=binomial(link="logit"))

# install.packages("arm")
library(arm) 
## See also https://github.com/stan-dev/rstanarm

bayesglm0 <- bayesglm(balfmla,data=meddat,family=binomial(link="logit"))
bayesglm1 <- bayesglm(psfmla,data=meddat,family=binomial(link="logit"))

psg1 <- predict(glm1,type="response")
psg0 <- predict(glm0,type="response")
psb1 <- predict(bayesglm1,type="response")
psb0 <- predict(bayesglm0,type="response")
```

Logistic regression is excellent at discriminating between groups \ldots often **too excellent** for us \autocite{gelman2008weakly}. First evidence of this is big and/or missing coefficients in the propensity score model. See the coefficients below:

```{r echo=FALSE}
thecoefs <- rbind(glm0=coef(glm0),
      glm1=coef(glm1),
      bayesglm0=coef(bayesglm0),
      bayesglm1=coef(bayesglm1)
      )
thecoefs[,1:5]
```


## The separation problem in logistic regression and the benefits of shrinkage/penalization


```{r}
matplot(t(thecoefs),axes=FALSE)
axis(2)
axis(1,at=0:26,labels=colnames(thecoefs),las=2)
```


## The separation problem in logistic regression

```{r out.width=".8\\textwidth", echo=FALSE}
par(mfrow=c(2,2),mar=c(3,3,2,.1))
boxplot(psg0~meddat$nhTrt,main=paste("Logit",length(coef(glm0))," parms",sep=" "))
stripchart(psg0~meddat$nhTrt,vertical=TRUE,add=TRUE)
boxplot(psg1~meddat$nhTrt,main=paste("Logit",length(coef(glm1))," parms",sep=" "))
stripchart(psg1~meddat$nhTrt,vertical=TRUE,add=TRUE)
boxplot(psb0~meddat$nhTrt,main=paste("Shrinkage Logit",length(coef(bayesglm0))," parms",sep=" "))
stripchart(psb0~meddat$nhTrt,vertical=TRUE,add=TRUE)
boxplot(psb1~meddat$nhTrt,main=paste("Shrinkage Logit",length(coef(bayesglm1))," parms",sep=" "))
stripchart(psb1~meddat$nhTrt,vertical=TRUE,add=TRUE)
```

## Matching on the propensity score

```{r}
psdist <- match_on(bayesglm1,data=meddat)
psdist[1:4,1:4]
fmPS <- fullmatch(psdist,data=meddat)
summary(fmPS,min.controls=0,max.controls=Inf,propensity.model=bayesglm1)
stopifnot(all.equal(names(fmPS),row.names(meddat)))
meddat$fmPS <- fmPS ## because same order
meddat$pscore <- predict(bayesglm1) ## because no missing data
setdiffs <- meddat %>% group_by(fmPS) %>% summarize(minHR03=min(HomRate03),
						    maxHR03=max(HomRate03))


```

# Major Matching modes

##  Optimal (communitarian) vs greedy (individualistic) matching

Compare the greedy to optimal matched designs:

\begin{center}
    \begin{tabular}{l|cccc}
      & \multicolumn{4}{c}{Illustrator} \\
Writer & Mo& John & Terry  & Pat \\ \hline
Ben    & 0 & 1 & 1 & 10 \\
Jake   & 10& 0 & 10 & 10 \\
Tom    &  1& 1 & 20 & $\infty$ \\ \hline
    \end{tabular}
  \end{center}

```{r results="hide", warnings=FALSE, messages=FALSE}
bookmat <- matrix(c(0,1,1,10,10,0,10,10,1,1,20,Inf),nrow=3,byrow=TRUE)
dimnames(bookmat) <- list(c("Ben","Jake","Tom"),c("Mo","John","Terry","Pat"))
fullmatch(bookmat)
```

\only<2->{ Greedy match without replacement has mean distance (0+0+20)/3=6.67. The optimal
match keeps all the obs, and has mean distance (1+0+1)/3=.67. }


\note{
*Greedy:* Ben-Mo (0) , Jake-John (0), Tom-Terry (20)
*Optimal:* Ben-Terry (1), Jake-John (0), Tom-Mo (1)
}

## Matching with and without replacement
*Example*: Costs of nuclear plants. When expanding, is it cheaper to build on the site of an existing plant?

\begin{columns}
\column{.5\textwidth}
\begin{center}
  \igrphx[width=.5\textwidth]{coxSnell}
\end{center}

\column{.5\textwidth}
  \igrphx[width=\textwidth]{Nuclear_plant_at_Grafenrheinfeld}
\end{columns}

\note{This is a \textit{prospective} design.  Explain.

(Presentation version has Nuclear plant image on 2nd overlay)
}


## New vs refurbished nuclear plants
\note{
  \begin{itemize}
  \item This is what many understand by matching.  [Greedy pair match follows.]
  \item Sometimes matching means  ``with-replacement'' matching. We'll permit that too, but in a structured way, and keeping track of how many replacements. 
  \item Why keep track of the replacements?  --B/c they affect standard errors of the matched comparison.  (``Effective sample size'' to be introduced a few slides later.)
  \end{itemize}
}
\begin{minipage}[t]{2in}
\begin{center}
Existing site\\
{\small
% latex table generated in R 3.0.1 by xtable 1.7-1 package
% Wed Aug 14 15:26:53 2013
\begin{tabular}{lr}
  \hline
 & capacity \\ 
  \hline
A & {660} {\mlpnode{NA}} \\ 
  B & {660} {\mlpnode{NB}} \\ 
  C & {420} {\mlpnode{NC}} \\ 
  D & {130} {\mlpnode{ND}} \\ 
  E & {650} {\mlpnode{NE}} \\ 
  F & {430} {\mlpnode{NF}} \\ 
  G & {420} {\mlpnode{NG}} \\ 
   \hline
\end{tabular}}
\end{center}
\bigskip
\bigskip
{\footnotesize  "capacity" is net capacity of the power plant, in MWe above
400.

\only<9->{\alert{This is matching "without replacement."}}
\only<10->{\alert{Contrast with matching "with replacement."}}
}
\end{minipage}
\begin{minipage}[t]{2in}
\begin{center}
New site\\
{\scriptsize
% latex table generated in R 3.0.1 by xtable 1.7-1 package
% Wed Aug 14 15:26:53 2013
\begin{tabular}{lr}
  \hline
 & capacity \\ 
  \hline
{\mlpnode{NH}\mbox{}} {H} & 290 \\ 
  {\mlpnode{NI}\mbox{}} {I} & 660 \\ 
  {\mlpnode{NJ}\mbox{}} {J} & 660 \\ 
  {\mlpnode{NK}\mbox{}} {K} & 110 \\ 
  {\mlpnode{NL}\mbox{}} {L} & 420 \\ 
  {\mlpnode{NM}\mbox{}} {M} &  60 \\ 
  {\mlpnode{NN}\mbox{}} {N} & 390 \\ 
  {\mlpnode{NO}\mbox{}} {O} & 160 \\ 
  {\mlpnode{NP}\mbox{}} {P} & 390 \\ 
  {\mlpnode{NQ}\mbox{}} {Q} & 130 \\ 
  {\mlpnode{NR}\mbox{}} {R} & 650 \\ 
  {\mlpnode{NS}\mbox{}} {S} & 450 \\ 
  {\mlpnode{NT}\mbox{}} {T} & 380 \\ 
  {\mlpnode{NU}\mbox{}} {U} & 440 \\ 
  {\mlpnode{NV}\mbox{}} {V} & 690 \\ 
  {\mlpnode{NW}\mbox{}} {W} & 510 \\ 
  {\mlpnode{NX}\mbox{}} {X} & 390 \\ 
  {\mlpnode{NY}\mbox{}} {Y} & 140 \\ 
  {\mlpnode{NZ}\mbox{}} {Z} & 730 \\ 
   \hline
\end{tabular}}
\end{center}
\end{minipage}


\begin{tikzpicture}[overlay]
  \path[draw,gray]<2-> (NA) edge (NI);
 \path[draw,gray]<3-> (NB) edge (NJ);
 \path[draw,gray]<4-> (NC) edge (NL);
 \path[draw,gray]<5-> (ND) edge (NQ);
 \path[draw,gray]<6-> (NE) edge (NR);
 \path[draw,gray]<7-> (NF) edge (NU);
 \path[draw,gray]<8-> (NG) edge (NN);
 \end{tikzpicture}\only<10>{
\begin{tikzpicture}[overlay]
  \path[draw,red] (NA) edge [out=0, in=-180] (NI);
 \path[draw,red] (NB) edge [out=0, in=-180] (NJ);
 \path[draw,red] (NC) edge [out=0, in=-180] (NL);
 \path[draw,red] (NF) edge [out=0, in=-180] (NL);
 \path[draw,red] (NG) edge [out=0, in=-180] (NL);
 \path[draw,red] (ND) edge [out=0, in=-180] (NQ);
 \path[draw,red] (NE) edge [out=0, in=-180] (NR);
 \end{tikzpicture}}



## Role of the matching algorithm: greedy matching

\begin{minipage}[t]{2in}
\begin{center}
Existing site\\
{\small
% latex table generated in R 3.0.1 by xtable 1.7-1 package
% Wed Aug 14 15:26:57 2013
\begin{tabular}{lrr}
  \hline
 & z.date & z.cap \\ 
  \hline
A & -1.6 & {1.2} {\mlpnode{NA}} \\ 
  B & -0.9 & {1.2} {\mlpnode{NB}} \\ 
  C & -0.4 & {0} {\mlpnode{NC}} \\ 
  D & -0.4 & {-1.4} {\mlpnode{ND}} \\ 
  E & 0.1 & {1.1} {\mlpnode{NE}} \\ 
  F & 2.2 & {0} {\mlpnode{NF}} \\ 
  G & 1.3 & {0} {\mlpnode{NG}} \\ 
   \hline
\end{tabular}}
\end{center}
\bigskip
\bigskip
\bigskip
{\footnotesize Here, ``date'' is \emph{rank of} date of
construction, in years after 1965, and  ``capacity'' is
\emph{rank of} net capacity of the power plant, in MWe
above 400.}
\end{minipage}
\begin{minipage}[t]{2in}
\begin{center}
New site\\
{\scriptsize
% latex table generated in R 3.0.1 by xtable 1.7-1 package
% Wed Aug 14 15:26:57 2013
\begin{tabular}{lrr}
  \hline
 & z.date & z.cap \\ 
  \hline
{\mlpnode{NH}\mbox{}} {H} & -0.3 & -0.7 \\ 
  {\mlpnode{NI}\mbox{}} {I} & -1.6 & 1.2 \\ 
  {\mlpnode{NJ}\mbox{}} {J} & -0.9 & 1.2 \\ 
  {\mlpnode{NK}\mbox{}} {K} & -0.9 & -1.5 \\ 
  {\mlpnode{NL}\mbox{}} {L} & -0.7 & -0.0 \\ 
  {\mlpnode{NM}\mbox{}} {M} & -0.4 & -1.8 \\ 
  {\mlpnode{NN}\mbox{}} {N} & -0.5 & -0.2 \\ 
  {\mlpnode{NO}\mbox{}} {O} & -0.3 & -1.3 \\ 
  {\mlpnode{NP}\mbox{}} {P} & -0.1 & -0.2 \\ 
  {\mlpnode{NQ}\mbox{}} {Q} & -0.4 & -1.4 \\ 
  {\mlpnode{NR}\mbox{}} {R} & 0.1 & 1.1 \\ 
  {\mlpnode{NS}\mbox{}} {S} & 0.1 & 0.1 \\ 
  {\mlpnode{NT}\mbox{}} {T} & -0.4 & -0.2 \\ 
  {\mlpnode{NU}\mbox{}} {U} & 0.7 & 0.1 \\ 
  {\mlpnode{NV}\mbox{}} {V} & 0.4 & 1.3 \\ 
  {\mlpnode{NW}\mbox{}} {W} & -0.1 & 0.4 \\ 
  {\mlpnode{NX}\mbox{}} {X} & 0.9 & -0.2 \\ 
  {\mlpnode{NY}\mbox{}} {Y} & 1.7 & -1.4 \\ 
  {\mlpnode{NZ}\mbox{}} {Z} & 2.3 & 1.5 \\ 
   \hline
\end{tabular}}
\end{center}
\end{minipage}

\begin{tikzpicture}[overlay]
  \path[draw,red]<2-> (NA) edge (NI);
 \path[draw,red]<3-> (NA) edge (NJ);
 \path[draw,red]<4-> (NB) edge (NR);
 \path[draw,red]<5-> (NB) edge (NW);
 \path[draw,red]<6-> (NC) edge (NN);
 \path[draw,red]<7-> (NC) edge (NT);
 \path[draw,red]<8-> (ND) edge (NO);
 \path[draw,red]<9-> (ND) edge (NQ);
 \path[draw,red]<10-> (NE) edge (NS);
 \path[draw,red]<11-> (NE) edge (NV);
 \path[draw,red]<12-> (NF) edge (NU);
 \path[draw,red]<13-> (NF) edge (NX);
 \path[draw,red]<14-> (NG) edge (NM);
 \path[draw,red]<15-> (NG) edge (NZ);
 \end{tikzpicture}

## Optimal Matching

\begin{minipage}[t]{2in}
\begin{center}
Existing site\\
{\small
% latex table generated in R 3.0.1 by xtable 1.7-1 package
% Wed Aug 14 15:26:57 2013
\begin{tabular}{lrr}
  \hline
 & z.date & z.cap \\ 
  \hline
A & -1.6 & {1.2} {\mlpnode{NA}} \\ 
  B & -0.9 & {1.2} {\mlpnode{NB}} \\ 
  C & -0.4 & {0} {\mlpnode{NC}} \\ 
  D & -0.4 & {-1.4} {\mlpnode{ND}} \\ 
  E & 0.1 & {1.1} {\mlpnode{NE}} \\ 
  F & 2.2 & {0} {\mlpnode{NF}} \\ 
  G & 1.3 & {0} {\mlpnode{NG}} \\ 
   \hline
\end{tabular}}
\end{center}
\bigskip

\textcolor{blue}{Optimal vs. Greedy matching}
\bigskip

{\footnotesize By evaluating potential matches all together rather than
  sequentially, optimal matching (\textcolor{blue}{blue lines}) reduces
  the mean matched distance from 1.17 to
 0.56.}

\end{minipage}
\begin{minipage}[t]{2in}
\begin{center}
New site\\
{\scriptsize
% latex table generated in R 3.0.1 by xtable 1.7-1 package
% Wed Aug 14 15:26:57 2013
\begin{tabular}{lrr}
  \hline
 & z.date & z.cap \\ 
  \hline
{\mlpnode{NH}\mbox{}} {H} & -0.3 & -0.7 \\ 
  {\mlpnode{NI}\mbox{}} {I} & -1.6 & 1.2 \\ 
  {\mlpnode{NJ}\mbox{}} {J} & -0.9 & 1.2 \\ 
  {\mlpnode{NK}\mbox{}} {K} & -0.9 & -1.5 \\ 
  {\mlpnode{NL}\mbox{}} {L} & -0.7 & -0.0 \\ 
  {\mlpnode{NM}\mbox{}} {M} & -0.4 & -1.8 \\ 
  {\mlpnode{NN}\mbox{}} {N} & -0.5 & -0.2 \\ 
  {\mlpnode{NO}\mbox{}} {O} & -0.3 & -1.3 \\ 
  {\mlpnode{NP}\mbox{}} {P} & -0.1 & -0.2 \\ 
  {\mlpnode{NQ}\mbox{}} {Q} & -0.4 & -1.4 \\ 
  {\mlpnode{NR}\mbox{}} {R} & 0.1 & 1.1 \\ 
  {\mlpnode{NS}\mbox{}} {S} & 0.1 & 0.1 \\ 
  {\mlpnode{NT}\mbox{}} {T} & -0.4 & -0.2 \\ 
  {\mlpnode{NU}\mbox{}} {U} & 0.7 & 0.1 \\ 
  {\mlpnode{NV}\mbox{}} {V} & 0.4 & 1.3 \\ 
  {\mlpnode{NW}\mbox{}} {W} & -0.1 & 0.4 \\ 
  {\mlpnode{NX}\mbox{}} {X} & 0.9 & -0.2 \\ 
  {\mlpnode{NY}\mbox{}} {Y} & 1.7 & -1.4 \\ 
  {\mlpnode{NZ}\mbox{}} {Z} & 2.3 & 1.5 \\ 
   \hline
\end{tabular}}
\end{center}
\end{minipage}

\begin{tikzpicture}[overlay]
  \path[draw,red] (NA) edge (NI);
 \path[draw,red] (NA) edge (NJ);
 \path[draw,red] (NB) edge (NR);
 \path[draw,red] (NB) edge (NW);
 \path[draw,red] (NC) edge (NN);
 \path[draw,red] (NC) edge (NT);
 \path[draw,red] (ND) edge (NO);
 \path[draw,red] (ND) edge (NQ);
 \path[draw,red] (NE) edge (NS);
 \path[draw,red] (NE) edge (NV);
 \path[draw,red] (NF) edge (NU);
 \path[draw,red] (NF) edge (NX);
 \path[draw,red] (NG) edge (NM);
 \path[draw,red] (NG) edge (NZ);
 \end{tikzpicture}\begin{tikzpicture}[overlay]
  \path[draw,blue] (NA) edge [out=0, in=-180] (NI);
 \path[draw,blue] (NA) edge [out=0, in=-180] (NJ);
 \path[draw,blue] (NB) edge [out=0, in=-180] (NL);
 \path[draw,blue] (NB) edge [out=0, in=-180] (NW);
 \path[draw,blue] (NC) edge [out=0, in=-180] (NN);
 \path[draw,blue] (NC) edge [out=0, in=-180] (NT);
 \path[draw,blue] (ND) edge [out=0, in=-180] (NO);
 \path[draw,blue] (ND) edge [out=0, in=-180] (NQ);
 \path[draw,blue] (NE) edge [out=0, in=-180] (NR);
 \path[draw,blue] (NE) edge [out=0, in=-180] (NV);
 \path[draw,blue] (NF) edge [out=0, in=-180] (NY);
 \path[draw,blue] (NF) edge [out=0, in=-180] (NZ);
 \path[draw,blue] (NG) edge [out=0, in=-180] (NU);
 \path[draw,blue] (NG) edge [out=0, in=-180] (NX);
 \end{tikzpicture}



## The World of Matching Today

This is an active and productive research area (i.e. lots of new ideas, not everyone agrees
about everything). Here are just a few of the different approaches to creating
strata or sets or selecting subsets.

 - The work on cardinality matching and fine balance
   <http://jrzubizarreta.com/>
   <https://cran.rstudio.com/web/packages/designmatch/> )
 - The work on speedier approximate full matching <http://fredriksavje.com/>
   <https://github.com/fsavje/quickmatch>
 - The work on using genetic algorithms to (1) find approximate strata
   with-replacement <http://sekhon.berkeley.edu/matching/> and (2) to find
   an approximation to a completely randomized study (i.e. best subset
   selection) <http://cho.pol.illinois.edu/wendy/papers/bossor.pdf>
 - The work on coarsened exact matching <https://gking.harvard.edu/cem>
 - The work on entropy-balancing approach to causal effects <https://web.stanford.edu/~jhain/software.htm#ebal>.
 - The covariate-balancing propensity score (CBPS) <https://imai.princeton.edu/research/CBPS.html>.

# Information and Balance: Matching structure and effective sample size

## Matching with a varying number of controls and full matching

### Fixed vs flexible ratio matching:

\begin{itemize}[<+->]
\item Pair matching \& sample size
\item Effective vs real sample size
\item If we limit ourselves to fixed matching ratios, we gain in
  simplicity but pay a price in sample size (effective \& real).
\item How big a price?  Trying is the best way to find out.
\end{itemize}

```{r echo=FALSE}
nuke.nopt <- subset(nuclearplants, pt == 0)
```

## Matching with a varying number of controls

\begin{minipage}[t]{2in}
\begin{center}
Existing site\\
{\small
% latex table generated in R 3.0.2 by xtable 1.7-3 package
% Thu Jul 31 13:51:34 2014
\begin{tabular}{lrr}
  \hline
 & z.date & z.cap \\
  \hline
A & -1.6 & {1.2} {\mlpnode{NA}} \\
  B & -0.9 & {1.2} {\mlpnode{NB}} \\
  C & -0.4 & {0} {\mlpnode{NC}} \\
  D & -0.4 & {-1.4} {\mlpnode{ND}} \\
  E & 0.1 & {1.1} {\mlpnode{NE}} \\
  F & 2.2 & {0} {\mlpnode{NF}} \\
  G & 1.3 & {0} {\mlpnode{NG}} \\
   \hline
\end{tabular}}
\end{center}
\end{minipage}
\begin{minipage}[t]{2in}
\begin{center}
New site\\
{\scriptsize
% latex table generated in R 3.0.2 by xtable 1.7-3 package
% Thu Jul 31 13:51:34 2014
\begin{tabular}{lrr}
  \hline
 & z.date & z.cap \\
  \hline
{\mlpnode{NH}\mbox{}} {H} & -0.3 & -0.7 \\
  {\mlpnode{NI}\mbox{}} {I} & -1.6 & 1.2 \\
  {\mlpnode{NJ}\mbox{}} {J} & -0.9 & 1.2 \\
  {\mlpnode{NK}\mbox{}} {K} & -0.9 & -1.5 \\
  {\mlpnode{NL}\mbox{}} {L} & -0.7 & -0.0 \\
  {\mlpnode{NM}\mbox{}} {M} & -0.4 & -1.8 \\
  {\mlpnode{NN}\mbox{}} {N} & -0.5 & -0.2 \\
  {\mlpnode{NO}\mbox{}} {O} & -0.3 & -1.3 \\
  {\mlpnode{NP}\mbox{}} {P} & -0.1 & -0.2 \\
  {\mlpnode{NQ}\mbox{}} {Q} & -0.4 & -1.4 \\
  {\mlpnode{NR}\mbox{}} {R} & 0.1 & 1.1 \\
  {\mlpnode{NS}\mbox{}} {S} & 0.1 & 0.1 \\
  {\mlpnode{NT}\mbox{}} {T} & -0.4 & -0.2 \\
  {\mlpnode{NU}\mbox{}} {U} & 0.7 & 0.1 \\
  {\mlpnode{NV}\mbox{}} {V} & 0.4 & 1.3 \\
  {\mlpnode{NW}\mbox{}} {W} & -0.1 & 0.4 \\
  {\mlpnode{NX}\mbox{}} {X} & 0.9 & -0.2 \\
  {\mlpnode{NY}\mbox{}} {Y} & 1.7 & -1.4 \\
  {\mlpnode{NZ}\mbox{}} {Z} & 2.3 & 1.5 \\
   \hline
\end{tabular}}
\end{center}
\end{minipage}
\begin{tikzpicture}[overlay]
  \path[draw,gray] (NA) edge (NI);
 \path[draw,gray] (NB) edge (NJ);
 \path[draw,gray] (NC) edge (NH);
 \path[draw,gray] (NC) edge (NL);
 \path[draw,gray] (NC) edge (NN);
 \path[draw,gray] (NC) edge (NP);
 \path[draw,gray] (NC) edge (NS);
 \path[draw,gray] (NC) edge (NT);
 \path[draw,gray] (NC) edge (NW);
 \path[draw,gray] (ND) edge (NK);
 \path[draw,gray] (ND) edge (NM);
 \path[draw,gray] (ND) edge (NO);
 \path[draw,gray] (ND) edge (NQ);
 \path[draw,gray] (NE) edge (NR);
 \path[draw,gray] (NE) edge (NV);
 \path[draw,gray] (NF) edge (NZ);
 \path[draw,gray] (NG) edge (NU);
 \path[draw,gray] (NG) edge (NX);
 \path[draw,gray] (NG) edge (NY);
 \end{tikzpicture}

```{r}
fmnukeMC1<-fullmatch(pr~date+cap,data=nuke.nopt,min.controls=1)
summary(fmnukeMC1, min.controls=0, max.controls=Inf)
```


\note{
  Observe that now no control plants are left out.  (This is something you can change if you want.)

Discuss effective sample size.
}



## Matching so as to maximize effective sample size

\begin{minipage}[t]{2in}
\begin{center}
Existing site\\
{\small
% latex table generated in R 3.0.2 by xtable 1.7-3 package
% Thu Jul 31 13:51:34 2014
\begin{tabular}{lrr}
  \hline
 & z.date & z.cap \\
  \hline
A & -1.6 & {1.2} {\mlpnode{NA}} \\
  B & -0.9 & {1.2} {\mlpnode{NB}} \\
  C & -0.4 & {0} {\mlpnode{NC}} \\
  D & -0.4 & {-1.4} {\mlpnode{ND}} \\
  E & 0.1 & {1.1} {\mlpnode{NE}} \\
  F & 2.2 & {0} {\mlpnode{NF}} \\
  G & 1.3 & {0} {\mlpnode{NG}} \\
   \hline
\end{tabular}}
\end{center}
\bigskip
\bigskip
\bigskip
\bigskip
\end{minipage}
\begin{minipage}[t]{2in}
\begin{center}
New site\\
{\scriptsize
% latex table generated in R 3.0.2 by xtable 1.7-3 package
% Thu Jul 31 13:51:34 2014
\begin{tabular}{lrr}
  \hline
 & z.date & z.cap \\
  \hline
{\mlpnode{NH}\mbox{}} {H} & -0.3 & -0.7 \\
  {\mlpnode{NI}\mbox{}} {I} & -1.6 & 1.2 \\
  {\mlpnode{NJ}\mbox{}} {J} & -0.9 & 1.2 \\
  {\mlpnode{NK}\mbox{}} {K} & -0.9 & -1.5 \\
  {\mlpnode{NL}\mbox{}} {L} & -0.7 & -0.0 \\
  {\mlpnode{NM}\mbox{}} {M} & -0.4 & -1.8 \\
  {\mlpnode{NN}\mbox{}} {N} & -0.5 & -0.2 \\
  {\mlpnode{NO}\mbox{}} {O} & -0.3 & -1.3 \\
  {\mlpnode{NP}\mbox{}} {P} & -0.1 & -0.2 \\
  {\mlpnode{NQ}\mbox{}} {Q} & -0.4 & -1.4 \\
  {\mlpnode{NR}\mbox{}} {R} & 0.1 & 1.1 \\
  {\mlpnode{NS}\mbox{}} {S} & 0.1 & 0.1 \\
  {\mlpnode{NT}\mbox{}} {T} & -0.4 & -0.2 \\
  {\mlpnode{NU}\mbox{}} {U} & 0.7 & 0.1 \\
  {\mlpnode{NV}\mbox{}} {V} & 0.4 & 1.3 \\
  {\mlpnode{NW}\mbox{}} {W} & -0.1 & 0.4 \\
  {\mlpnode{NX}\mbox{}} {X} & 0.9 & -0.2 \\
  {\mlpnode{NY}\mbox{}} {Y} & 1.7 & -1.4 \\
  {\mlpnode{NZ}\mbox{}} {Z} & 2.3 & 1.5 \\
   \hline
\end{tabular}}
\end{center}
\end{minipage}
\begin{tikzpicture}[overlay]
  \path[draw,gray] (NA) edge (NI);
 \path[draw,gray] (NA) edge (NJ);
 \path[draw,gray] (NB) edge (NL);
 \path[draw,gray] (NB) edge (NN);
 \path[draw,gray] (NB) edge (NW);
 \path[draw,gray] (NC) edge (NH);
 \path[draw,gray] (NC) edge (NO);
 \path[draw,gray] (NC) edge (NT);
 \path[draw,gray] (ND) edge (NK);
 \path[draw,gray] (ND) edge (NM);
 \path[draw,gray] (ND) edge (NQ);
 \path[draw,gray] (NE) edge (NR);
 \path[draw,gray] (NE) edge (NS);
 \path[draw,gray] (NE) edge (NV);
 \path[draw,gray] (NF) edge (NY);
 \path[draw,gray] (NF) edge (NZ);
 \path[draw,gray] (NG) edge (NP);
 \path[draw,gray] (NG) edge (NU);
 \path[draw,gray] (NG) edge (NX);
 \end{tikzpicture}
```{r}
fmnuke<-fullmatch(pr~date+cap, min=2, max=3, data=nuke.nopt)
summary(fmnuke)
```

## Showing matches

\centering
```{r out.width=".8\\textwidth", echo=FALSE}
## perhaps try this https://briatte.github.io/ggnet/#example-2-bipartite-network next time
library(igraph)
blah1 <- outer(fmnuke,fmnuke,FUN=function(x,y){ as.numeric(x==y) })
blah2 <- outer(fmnukeMC1,fmnukeMC1,FUN=function(x,y){ as.numeric(x==y) })
par(mfrow=c(1,2),mar=c(3,3,3,1))
plot(graph_from_adjacency_matrix(blah1,mode="undirected",diag=FALSE),
     vertex.color=c("white","green")[nuke.nopt$pr+1],main="Min Ctrls=2, Max Ctrls=3")
plot(graph_from_adjacency_matrix(blah2,mode="undirected",diag=FALSE),
     vertex.color=c("white","green")[nuke.nopt$pr+1],main="Min Ctrls=1, Max Ctrls=Inf")
```

## Matching so as to maximize effective sample size

\tiny
```{r}
effectiveSampleSize(fmnuke)
nukewts <- cbind(nuke.nopt,fmnuke) %>% 
  group_by(fmnuke) %>% summarise(nb = n(),
                                 nTb = sum(pr),
                                 nCb = nb - nTb,
                                 hwt = ( 2*( nCb * nTb ) / (nTb + nCb)))
nukewts
dim(nukewts)
sum(nukewts$hwt)
```

## Matching so as to maximize effective sample size

```{r}
stratumStructure(fmnuke)
```

So effective sample size for this match = $2 * 4/3 + 5* 3/2 = 10.17$ --- compare to 7 for pairs, 9.33 for triples.

```{r}
nukemh <- match_on(pr~date+cap,data=nuke.nopt)
pmnuke <- pairmatch(nukemh,data=nuke.nopt)
levels(pmnuke)
effectiveSampleSize(pmnuke)
```

##  Tracking effective sample size

In 2-sample comparisons, total sample size can be a misleading as a measure of information content.  Example:
\begin{itemize}
\item say $Y$ has same variance, $\sigma^{2}$,in the Tx and the Ctl population.
\item Ben H. samples 10 Tx and 40 Ctls, and
\item Jake B. samples 25 Tx and 25 Ctls
\end{itemize}
--- so that total sample sizes are the same.  However,

\begin{eqnarray*}
  V_{BH}(\bar{y}_{t} - \bar{y}_{c}) &=& \frac{\sigma^{2}}{10} + \frac{\sigma^{2}}{40}=.125\sigma^{2}\mbox{;}\\
  V_{JB}(\bar{y}_{t} - \bar{y}_{c}) &=& \frac{\sigma^{2}}{25} + \frac{\sigma^{2}}{25}=.08\sigma^{2}.\\
\end{eqnarray*}

Similarly, a matched triple is roughly $[(\sigma^{2}/1 + \sigma^{2}/2)/(\sigma^{2}/1 + \sigma^{2}/1)]^{-1}= 1.33$ times as informative as a matched pair.

## Details

Use pooled 2-sample t statistic SE formula to compare 1-1 vs 1-2 matched sets' contribution to variance:

$$
\begin{array}{c|c}
  \atob{1}{1} & \atob{1}{2} \\
M^{-2}\sum_{m=1}^{M} (\sigma^{2}/1 + \sigma^{2}/1) & M^{-2}\sum_{m=1}^{M} (\sigma^{2}/1 + \sigma^{2}/2) \\
\frac{2\sigma^{2}}{M} & \frac{1.5\sigma^{2}}{M} \\
\end{array}
$$

So 20 matched pairs is comparable to 15 matched triples.

(Correspondingly, h-mean of $n_{t},n_{c}$ for a pair is 1, while for a triple it's $[(1/1 + 1/2)/2]^{-1}=4/3$.)

The variance of the `pr`-coeff in `v~pr + match` is
$$
 \frac{2 \sigma^{2}}{\sum_{s} h_{s}}, \hspace{3em} h_{s} = \left( \frac{n_{ts}^{-1} + n_{cs}^{-1} }{2}  \right)^{-1} ,
$$

assuming the OLS model and homoskedastic errors.  (This is b/c the anova formulation is equivalent to harmonic-mean weighting, under which $V(\sum_{s}w_{s}(\bar{v}_{ts} - \bar v_{cs})) = \sum_{s} w_{s}^{2}(n_{ts}^{-1} + n_{cs}^{-1}) \sigma^{2} = \sigma^{2} \sum_{s} w_{s}^{2} 2 h_{s}^{-1} = 2\sigma^{2} \sum_{s}w_{s}/\sum_{s}h_{s} = 2\sigma^{2}/\sum_{s} h_{s}$.)

For matched pairs, of course, $h_{s}=1$.  Harmonic mean of 1, 2 is $4/3$. Etc.


## Matching so as to maximize effective sample size

```{r}
mds <- matched.distances(matchobj=fmnuke,distance=nukemh)
mds[1:2]
summary(unlist(mds))
```

Mean of matched distances is `r mean(unlist(mds))` --- compare to 0.29 for pairs, 0.57 for triples.

Note variance/bias tradeoff. Covariate balance has to do with bias and extrapolation (i.e. confounding). We want balance **and** precision.

## Inspect distances after matching

What kinds of distances remain after matching?

```{r}
str(mds)
quantile(unlist(mds))
```


## Playing with Caliper Matching

```{r eval=FALSE}

install.packages("/Library/gurobi702/mac64/R/gurobi_7.0-2.tgz",repos=NULL)
library(gurobi)
library(designmatch)
distmat <- as.matrix(dist3)
distmat[is.infinite(distmat)]<-1000
res <- bmatch(t_ind=meddat$nhTrt,dist_mat=distmat,
	      n_controls=1,
	      solver=list(name='gurobi',approximate=0))


```



## Summary:

What do you think?



## References
