---
title: Non-bipartite Matching
date: '`r format(Sys.Date(), "%B %d, %Y")`'
author: ICPSR 2019 Session 2
bibliography:
  - BIB/abbrev-long.bib
  - BIB/refs.bib
  - BIB/master.bib
  - BIB/misc.bib
fontsize: 10pt
geometry: margin=1in
graphics: yes
biblio-style: authoryear-comp
output:
  beamer_presentation:
    slide_level: 2
    keep_tex: true
    latex_engine: xelatex
    citation_package: biblatex
    template: icpsr.beamer
    includes:
      in_header:
         - defs-all.sty
    md_extensions: +raw_attribute
---

<!-- Make this document using library(rmarkdown); render("day12.Rmd") -->

```{r include=FALSE, cache=FALSE}
# Some customization.  You can alter or delete as desired (if you know what you are doing).
# knitr settings to control how R chunks work.
rm(list=ls())

require(knitr)

## This plus size="\\scriptsize" from https://stackoverflow.com/questions/26372138/beamer-presentation-rstudio-change-font-size-for-chunk

knitr::knit_hooks$set(mysize = function(before, options, envir) {
			      if (before){
				      return(options$size)
			      } else {
				      return("\\normalsize")}
})

knit_hooks$set(plotdefault = function(before, options, envir) {
		       if (before) par(mar = c(3, 3, .1, .1),oma=rep(0,4),mgp=c(1.5,.5,0))
})

opts_chunk$set(
	       tidy='styler',
	       echo=TRUE,
	       results='markup',
	       strip.white=TRUE,
	       fig.path='figs/fig',
	       cache=FALSE,
	       highlight=TRUE,
	       width.cutoff=132,
	       size='\\scriptsize',
	       out.width='.8\\textwidth',
	       fig.retina=FALSE,
	       message=FALSE,
	       comment=NA,
	       mysize=TRUE,
	       plotdefault=TRUE)

if(!file.exists('figs')) dir.create('figs')

options(digits=4,
	scipen=8,
	width=132
)
```

```{r eval=FALSE, include=FALSE, echo=FALSE}
## Run this only once and then not again until we want a new version from github
library('devtools')
library('withr')
with_libpaths('./lib', install_github("markmfredrickson/RItools"), 'pre')
```

```{r echo=FALSE}
library(dplyr)
library(RItools,lib.loc="./lib")
library(optmatch)
library(nbpMatching)
library(lmtest)
library(sandwich)
library(estimatr)
library(coin)
```

## Today

  1. Agenda: Non-bipartite matching.
  2. Reading for today and next week: (1) Non-bipartite matching DOS Chap 11
     and \autocite{lu2011optimal}  and DOS 12 (longitudinal applications of
     non-bipartite matching) and (2) Sensitivity analysis DOS Chap 3
     \autocite{hhh2010}
  3. Questions arising from the reading or assignments or life?
 
## But first review

 - Statistical inference for Causal Effects and Causal Hypotheses in Randomized
   Experiments
 - Adjustment by stratification;
   - Matching to generate optimal stratifications  (decisions and strategies that are part of research design;
     matching on missingness and `fill.NAs`; `exactMatch`; `caliper`;
     `min.controls`; `effectiveSampleSize`);
   - Assessing success of stratified research designs in adjustment;
   - The As-If-Randomized mode of statistical inference for stratified research
   designs (treat a matched design as a block-randomized experiment).

## Review 2: An Adjustment Strategy to Address Alternative Explanations Effectively

How to strengthen evidence about the claim that Metrocable caused a decrease in crime?

 1. **List main alternative explanations** (could crime have caused Metrocable
    stations; socio-economic status differences; \ldots). Can we operationalize
    these explanations?
 2. **Stratify data to minimize heterogeneity within set.** If education does
    not vary within set, then we have "adjusted for" education by conditioning
    on the set. The `optmatch` package for R finds sets that minimize the
    weighted sum of distances across the sets. (See also `rcbalance`, `DiPs`,
    `bigmatch`, `designmatch`, `quickmatch`).
    1. Create distance matrices using `match_on` (and `caliper` and
       `exactMatch`) (Scalar distances on  especially important variables like
       baseline outcomes; Multivariate distances in terms of other covariates
       via Mahalanobis or Propensity scores distances.)
    2. Find stratifications using `fullmatch`.



```{r echo=FALSE, cache=TRUE}
load(url("http://jakebowers.org/Data/meddat.rda"))
meddat<- mutate(meddat,
		HomRate03=(HomCount2003/Pop2003)*1000,
		HomRate08=(HomCount2008/Pop2008)*1000)
row.names(meddat) <- meddat$nh
```


```{r echo=FALSE}
covs <- unique(c(names(meddat)[c(5:7,9:24)],"HomRate03"))
balfmla <- reformulate(covs,response="nhTrt")
mhdist <- match_on(balfmla,data=meddat, method="rank_mahalanobis")
psmod <- arm::bayesglm(balfmla,data=meddat,family=binomial(link="logit"))
stopifnot(any(abs(coef(psmod))<10))
psdist <- match_on(psmod,data=meddat)
tmp <- meddat$HomRate03
names(tmp) <- rownames(meddat)
absdist <- match_on(tmp, z = meddat$nhTrt,data=meddat)
```

## Example design for the day

```{r}
## Inspect the distance matrices
quantile(as.vector(psdist),seq(.9,1,.01))
quantile(as.vector(mhdist),seq(.9,1,.01))
quantile(as.vector(absdist),seq(.9,1,.01))
matchdist <- psdist + caliper(psdist,9) + caliper(absdist,2) + caliper(mhdist,60)
summary(matchdist)
fm1 <- fullmatch(matchdist, min.controls= 1, max.controls=Inf, data=meddat,tol=.00001)
summary(fm1,min.controls=0,max.controls=Inf,propensity.model=psmod)
meddat$fm1 <- factor(fm1)
meddat$nhTrtF <- factor(meddat$nhTrt)
```


## An Adjustment Strategy to Address Alternative Explanations

 3. **Assess the stratification in substantive terms** If we look within the
    sets, are the differences we see substantively concerning or trivial?



```{r echo=FALSE}
meddat[names(fm1),"fm1"] <- fm1
setmeanDiffs <- meddat %>% filter(!is.na(fm1)) %>% group_by(fm1) %>%
  summarise(ateb=mean(HomRate08[nhTrt==1])-mean(HomRate08[nhTrt==0]),
            nb=n(),
            nTb = sum(nhTrt),
            nCb = sum(1-nhTrt),
	    baselinediffs = mean(HomRate03[nhTrt==1])-mean(HomRate03[nhTrt==0]),
	    minbaselines = min(HomRate03),
	    maxbaseline = max(HomRate03)
            ) %>% arrange(abs(baselinediffs))
setmeanDiffs
```

## An Adjustment Strategy to Address Alternative Explanations

 4. **Assess the stratification by comparison to a model of a block-randomized
    experiment** Does our research design look like a block-randomized
    experiment in terms of covariate balance? If so, move onto step 4.
    Otherwise, work to improve the research design by (a) changing scores; (b)
    combining scores (for example, using calipers); (c) excluding units (using
    calipers); (d) exact matching on subgroups; (e) reducing variation in
    set-size.

## An Adjustment Strategy to Address Alternative Explanations

 5. **Estimate effects and test hypothesis as-if-block-randomized** Estimators
    and tests refer to the finite "population" of the study pool and the fixed
    stratification in the same way common in the analysis of block-randomized
    experiments.

```{r  estandtest,  cache=TRUE, warning=FALSE}
meddat <-  meddat %>%  filter(!is.na(fm1)) %>% group_by(fm1) %>%
	mutate(trtprob=mean(nhTrt), nbwt=nhTrt/trtprob + (1-nhTrt)/(1-trtprob))
estate <-  lm_robust(HomRate08~nhTrt,data=meddat,weights=nbwt,subset=!is.na(fm1))
estate
with(setmeanDiffs,sum(ateb*nb/sum(nb)))
ranktest1  <-  wilcox_test(HomRate08~nhTrtF|fm1,data=meddat,
		       distribution=approximate(nresample=10000))
pvalue(ranktest1)
```

## An Adjustment Strategy to Address Alternative Explanations

 6. **Assess the sensitivity of the analysis to the assumptions of
    as-if-randomized** The design is not a randomized design. Is this likely to
    cause small or large changes in the substantive interpretation of our
    results? (R packages `sensitivityfull`, `sensitivitymv`, `sensitivitymw`, `sensitivitymult`,
    `rbounds`)

# Non-bipartite Matching: An Application with the Study of Race and Place

## How do perceptions of place influence attitudes?

\autocite{wong2012jop} set out to measure perceptions of environments using an
internet survey of Canadians during 2012 which each respondent drew a free hand
map of their "local community" and then reported their understanding of the
demographic breakdown of this place.

```{r echo=FALSE, results='hide'}
## White English Speaking Canadians only
load(url("http://jakebowers.org/ICPSR/canadamapdat.rda"))
## summary(canadamapdat)
```
\centering
\igrphx{TwoMapsToronto.png}


## Capturing perceptions

Here are 50 maps drawn by people based in Toronto.

\centering
\igrphx{TorontoAllCommunities1.png}

## Capturing perceptions

And here is the question people were asked (groups in random order).

\centering
\igrphx{MLCCPerceptionsQuestion.pdf}

## Capturing perceptions

White Canadian respondents' reports about "visible minorities" in their hand drawn "local communities".

\centering
```{r echo=FALSE}
par(mfrow=c(1,2))
with(canadamapdat,scatter.smooth(vm.da,vm.community.norm2,col="gray",
				 ylab="Perceptions",xlab="Census Neighborhood (DA)",
				 xlim=c(0,1),ylim=c(0,1),lpars=list(lwd=2)))
with(canadamapdat,scatter.smooth(vm.csd,vm.community.norm2,col="gray",
				 ylab="Perceptions",xlab="Census Municipality (CSD)",
				 xlim=c(0,1),ylim=c(0,1),lpars=list(lwd=2)))
##summary(canadamapdat$vm.community.norm2)
```

## Codebook: Mainly for Rmd file

The variables are: age in years, income as a scale, sex in categories, a
social.capital scale coded to run 0 to 1, country of ancestry in categories,
csd.pop is population of the Census Subdivision (like a municipality), vm.csd
is 2006 proportion visible minority in the CSD, vm.da is proportion visible
minority in the Census Dissemination Area (a small area containing 400--700
persons), and vm.community.norm2 is the proportion of visible minorities
reported by respondents in their map of their local community,
community_area_km is the area within their drawing in square km.

## How to make the case for perceptions?

If we could randomly assign different perceptions to people, we could claim
that differences of perceptions matter (above and beyond and independent of
objective characteristics of the context).

\medskip

What is an observational design that would do this? Match people on objective
context (and maybe covariates) who differ in perceptions.

\medskip

But objective context is continuous not binary: rather than matching $m$ "treated"
to $n-m$ "controls", we want to compare all $n$ with all $n$ respondents.

```{r echo=FALSE}
## Exclude people who did not offer a perception or an outcome
wrkdat<-canadamapdat[!is.na(canadamapdat$vm.community.norm2) &
		     !is.na(canadamapdat$social.capital01),]
	     wrkdat$vmdaPct <- wrkdat$vm.da * 100 ## express in pct
```

## Create $n \times n$ distance matrices

Our main design compares white, English-speaking, Canadians people with similar
neighborhood proportions of visible minorities.

```{r }
scalar.dist<-function(v){
	## Utility function to make n x n abs dist matrices
	outer(v,v,FUN=function(x,y){ abs(x-y) })
}

vmdaDist<-scalar.dist(wrkdat$vmdaPct)
dimnames(vmdaDist)<-list(row.names(wrkdat), row.names(wrkdat))
## The nbpmatching way (Mahalanobis \equiv standardized in one dimension) takes a while:
##obj.com.dist.mat2<-distancematrix(gendistance(wrkdat[,"vmdaPct",drop=FALSE]))
## compare to tmp<-scalar.dist(wrkdat$vmdaPct/sd(wrkdat$vmdaPct))
wrkdat$vmdaPct[1:4]
diff(wrkdat$vmdaPct[1:4])
vmdaDist[1:4,1:4]
```

## Non-bipartite match

```{r nbp1, cache=TRUE}
vmdaDistMat <- distancematrix(vmdaDist)
nbp1match<-nonbimatch(vmdaDistMat)
nbp1<-get.sets(nbp1match$matches,remove.unpaired=TRUE)
wrkdat[names(nbp1),"nbp1"]<-nbp1
nbp1[1:5]
table(is.na(wrkdat$nbp1)) ## recall the "ghost message"
```

## Inspect the solution

```{r }
wrkdat[order(wrkdat$nbp1),c("nbp1","vmdaPct","vm.community.norm2")][1:6,]
## table(wrkdat$nbp1)
nbp1vmdiffs<-tapply(wrkdat$vmdaPct,wrkdat$nbp1,function(x){ abs(diff(x)) })
nbp1percdiffs<-tapply(wrkdat$vm.community.norm2,wrkdat$nbp1,function(x){ abs(diff(x)) })
summary(nbp1vmdiffs)
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
source(url("http://jakebowers.org/Matching/nonbimatchingfunctions.R"))
```

## Inspect the solution

\centering
```{r out.width=".8\\textwidth"}
nbmplot(wrkdat,yvar="vmdaPct",xvar="vm.community.norm2",strata="nbp1",points=FALSE,
	ylim=range(wrkdat$vmdaPct))
```

## Assess balance

No treatment and control groups to compare. But we can still compare the **relationships** between the adjusted variable (`vmdaPct`) and other covariates conditional on pair.

```{r balnbp1, cache=TRUE }
thecovs<-c("age","income.coded","education","x.years","sex",
	   "csd.pop","vm.csd","community_area_km")
balfmla<-reformulate(thecovs,response="vmdaPct")
xb1<-xBalance(balfmla,strata=list(unstrat=NULL,nbp1=~nbp1), report="all",data=wrkdat)
xb1$overall
xb1$results[,c("z","p"),"nbp1"]
```

## Improve balance using penalties and dropping observations

  For example, (a) require matches within Province, (b) do not compare people
  in small towns to people in large towns, (c) do not compare people who drew
  big maps to people who drew small maps, (d) drop the 8 least well matched
  observations. (Choosing 8 arbitrarily to demonstrate.)

```{r echo=FALSE}
rescale01<-function(x){
	(x-min(x,na.rm=TRUE))/(max(x,na.rm=TRUE)-min(x,na.rm=TRUE))
}
```

```{r results="hide"}
csdpopDist<-scalar.dist(wrkdat$csd.pop)
dimnames(csdpopDist)<-list(row.names(wrkdat),row.names(wrkdat))

## Since we have some missing values on community area, and we would like to
## match people who are both missing, we will give it a very large value.
wrkdat$commarea<-ifelse(is.na(wrkdat$community_area_km),
			max(wrkdat$community_area_km,na.rm=TRUE)*10,
			wrkdat$community_area_km)

areaDist<-scalar.dist(log(wrkdat$commarea))
dimnames(areaDist)<-list(row.names(wrkdat),row.names(wrkdat))

csdpopDist01<-rescale01(csdpopDist)
areaDist01<-rescale01(areaDist)

summary(as.vector(csdpopDist01))
summary(as.vector(areaDist01))
summary(as.vector(vmdaDist))

maxvmdaDist<-max(as.vector(vmdaDist))
```

## Improve balance using penalties and dropping observations

```{r cache=TRUE}
vmdaPen1<-vmdaDist+(maxvmdaDist*csdpopDist01)+(maxvmdaDist*areaDist01)
vmdaDist[1:5,1:5]
csdpopDist01[1:5,1:5]
areaDist01[1:5,1:5]
vmdaPen1[1:5,1:5]
```

## Dropping some observations

```{r cache=TRUE}
## now decide how many to drop
vmdaPhPen<-make.phantoms(vmdaPen1,8,maxval = max(as.vector(vmdaPen1))*10)
```

## Improve balance using penalties and dropping observations

```{r cache=TRUE}
vmdaPhPenMat <- distancematrix(vmdaPhPen)
nbp2match<-nonbimatch(vmdaPhPenMat)
nbp2<-get.sets(nbp2match$matches,remove.unpaired=TRUE)
wrkdat[names(nbp2),"nbp2"]<-nbp2
```

## Assess this new match

Is this match better or worse (in terms of balance? in terms of within-set distances?)

```{r echo=FALSE,results="hide"}
xb2<-xBalance(balfmla,strata=list(unstrat=NULL,nbp1=~nbp1,nbp2=~nbp2),
	      report="all",data=wrkdat)
xb2$overall[2:3,]
xb2$results[,"p",c("nbp1","nbp2")]
```

## Assess this new match

```{r echo=FALSE,out.width=".8\\textwidth"}
nbmplot(wrkdat,yvar="vmdaPct",xvar="vm.community.norm2",strata="nbp2",points=FALSE,ylim=range(wrkdat$vmdaPct))
```

## Strength of the treatment

The difference in "treatment" within sets varies --- and so we expect the size
of the effect to vary. For example, consider the ratio of  objective context
differences to  perceived context differences:

```{r}
summary(nbp1vmdiffs)
summary(nbp1percdiffs)
percDist <- scalar.dist(wrkdat$vm.community.norm2*100)
da <- vmdaDist[1:5,1:5]
perc <- percDist[1:5,1:5]

da/perc
```

## Strength of the treatment

To prevent many sets with no variation on "treatment" we could add a penalty
for pairs that are too close on treatment: to maximize size of treatment (i.e.
differences in perceptions) while minimizing differences on covariates. That
is, we should be able to get more precision / power about the perceptions
related differences if they are larger.

For example: if difference in  perceptions is  small,  make the distance very large.


```{r}
da + 1000*(perc < 2)

```

## Assess hypotheses about effects

Test the hypothesis of no relationship between perceptions
`vm.community.norm2` and `social capital`.

```{r eval=TRUE,echo=TRUE}
## These are the same test in this case
test1<-independence_test(social.capital01~vm.community.norm2|nbp1,data=wrkdat[!is.na(wrkdat$nbp1),])
test2<-xBalance(vm.community.norm2~social.capital01,
		strata=list(nbp1=~nbp1), data=wrkdat, report=c("adj.mean.diffs","chisquare.test","z.scores","p.values"))
test1
test2$overall
```


## Describe the differences within pairs

Does the person who perceives more visible minorities in their community tend
to be higher (or lower) in `social.capital` than the other person in the pair?

```{r echo=FALSE}
rank.pairs<-function (x, block) { ## Identify the low and high subj in each pair
	unsplit(lapply(split(x, block), function(x) {
			       rank(x)
				 }), block)
}
```

```{r}
wrkdat$scRank<-with(wrkdat,rank.pairs(social.capital01,nbp1))
wrkdat$vmCRank<-with(wrkdat,rank.pairs(vm.community.norm2,nbp1))
wrkdat[order(wrkdat$nbp1),c("nbp1","social.capital01","scRank","vm.community.norm2","vmCRank")][1:6,]
with(wrkdat,tapply(scRank,vmCRank,mean))
```

## Summarize mean differences within pairs

If perceptions matters for social capital then we would expect pairs differing
greatly in subjective context to display greater differences in social capital
than pairs that differ a little.


```{r echo=FALSE,results="hide"}
align.by.block<-function (x, block, fn = mean, thenames=NULL) { ## By default, this rescales each observation to be the distance from the group mean.
	newx<-unsplit(lapply(split(x, block), function(x) {
				     x - fn(x)
				 }), block)
	if(!is.null(names)){ names(newx)<-thenames }
	return(newx)
}
```

```{r}
wrkdat$scMD<-with(wrkdat,align.by.block(social.capital01,nbp1))
wrkdat$vmcn2MD<-with(wrkdat,align.by.block(vm.community.norm2,nbp1))
wrkdat[order(wrkdat$nbp1),c("social.capital01","scMD","vm.community.norm2","vmcn2MD","nbp1")][1:4,]
## notice that aligning or pair-mean-centering the data preserves the within
## set relationships
## summary(tapply(wrkdat$scMD,wrkdat$nbp1,function(x){ abs(diff(x)) }))
## summary(tapply(wrkdat$social.capital01,wrkdat$nbp1,function(x){ abs(diff(x)) }))
lm1<-lm(scMD~vmcn2MD,data=wrkdat[!is.na(wrkdat$nbp2),])
coefci(lm1,vcov=vcovHC(lm1,type="HC2"))
coeftest(lm1,vcov.=vcovHC(lm1,type="HC2"))
```
## Summarize mean differences within pairs

```{r warning=FALSE}
lm2 <- lm_robust(scMD~vmcn2MD,data=wrkdat[!is.na(wrkdat$nbp2),])
lm2
lm3 <- lm_robust(social.capital01~vm.community.norm2,fixed_effects=~nbp2,data=wrkdat)
lm3
table(wrkdat$vmCRank,exclude=c())
lm4 <- lm_robust(social.capital01~I(vmCRank-1),fixed_effects=~nbp2,data=wrkdat)
lm4
```

## Summarize mean differences within pairs

If perceptions matter for social capital above and beyond objective context
then we would expect pairs differing greatly in subjective context to display
greater differences in social capital than pairs that differ a little.

```{r}
coefci(lm1,vcov=vcovHC(lm1,type="HC2"))
coeftest(lm1,vcov.=vcovHC(lm1,type="HC2"))
lm2
lm3
pairdiffs <- wrkdat %>% filter(!is.na(vmCRank)&!is.na(social.capital01)&!is.na(nbp2)) %>%
       group_by(vmCRank) %>% summarize(mnsc=mean(social.capital01))
wrkdat[order(wrkdat$nbp2),c("social.capital01","scRank","scMD","vm.community.norm2","vmcn2MD","vmCRank","nbp2")][1:4,]
lm4
```

## Summarize mean differences within pairs

```{r}
summary(wrkdat$vmcn2MD)
summary(wrkdat$scMD)
```

Within matched pair, the person who perceives more visible minorities within set tends to report
lower social capital than the person who perceives fewer visible minorities
within set.

\medskip

The largest difference is about `r round(max(wrkdat$vmcn2MD,na.rm=TRUE),2)`. The model
predicts that social capital would differ by about `r coef(lm1)[[2]]*.48` for such a difference. This is about
`r coef(lm1)[[2]]*.48/sd(wrkdat$scMD,na.rm=TRUE)` of a standard deviation
of the social capital scale. Or about
`r coef(lm1)[[2]]*.48/abs(diff(range(wrkdat$scMD,na.rm=TRUE)))` of the range.


## Summarize mean differences within pairs

Here is a look at the within-pair differences in perceptions of visible minorities as well as social capital.

```{r smoothplot, out.width=".7\\textwidth", echo=FALSE}
with(wrkdat,scatter.smooth(vmcn2MD,scMD,span=.3,cex=.7,col="gray",pch=19,lpars=list(lwd=2)))
abline(h=0,lwd=.5)
```


## Summary of matching without groups

 - Workflow in general is the same as matching with groups (covariates,
   distance matrices, optimization to select a stratification, assessment of
   the stratification by comparison to an experiment)
 - Estimation is more flexible --- could look simply at "higher versus lower"
   within  pair, or could average over scores.



## Another estimation approach

\autocite{smith:1997} presents a multi-level modelling approach to taking
matched sets into account. The weights implied here are a bit different from
the weights that we've discussed before (although with pairs they might be more
or less the same). What is the data model? What additional assumptions are involved
here?

```{r lmer, cache=FALSE, warning=FALSE}
library(lme4)
wrkdat$vmCbi<-ave(wrkdat$vm.community.norm2,wrkdat$nbp1)
lmer1<-lmer(social.capital01~vm.community.norm2+vmCbi+(1|nbp1),data=wrkdat)
confint(lmer1)["vm.community.norm2",]
## Notice that we may have less information than we thought (some clustered of observation by DA).
table(table(wrkdat$dauid))
## So maybe:
lmer2<-lmer(social.capital01~vm.community.norm2+vmCbi+(1|nbp1)+(1|dauid),data=wrkdat)
confint(lmer2)["vm.community.norm2",]
```

## Other applications of non-bipartite matching?

See: DOS Chapter 11.

Also: this has a lot of applications in experimental design (see `blockTools` and \autocite{moore2012blocktools,moore2012multivariate}).


## Anything Else?

## References
