---
title: Information, Balance, Estimation, Testing
date: '`r format(Sys.Date(), "%B %d, %Y")`'
author: ICPSR 2017 Session 2
bibliography:
 - refs.bib
 - BIB/master.bib
 - BIB/misc.bib
fontsize: 10pt
geometry: margin=1in
graphics: yes
biblio-style: authoryear-comp
output:
  beamer_presentation:
    slide_level: 2
    keep_tex: true
    latex_engine: xelatex
    citation_package: biblatex
    template: icpsr.beamer
    includes:
        in_header:
           - defs-all.sty
---


<!-- Make this document using library(rmarkdown); render("day12.Rmd") -->


```{r include=FALSE, cache=FALSE}
# Some customization.  You can alter or delete as desired (if you know what you are doing).
# knitr settings to control how R chunks work.
rm(list=ls())

require(knitr)

## This plus size="\\scriptsize" from https://stackoverflow.com/questions/26372138/beamer-presentation-rstudio-change-font-size-for-chunk

knitr::knit_hooks$set(mysize = function(before, options, envir) {
  if (before)
    return(options$size)
})

knit_hooks$set(plotdefault = function(before, options, envir) {
    if (before) par(mar = c(3, 3, .1, .1),oma=rep(0,4),mgp=c(1.5,.5,0))
})

opts_chunk$set(
  tidy=FALSE,     # display code as typed
  echo=TRUE,
  results='markup',
  strip.white=TRUE,
  fig.path='figs/fig',
  cache=FALSE,
  highlight=TRUE,
  width.cutoff=132,
  size='\\scriptsize',
  out.width='.8\\textwidth',
  fig.retina=FALSE,
  message=FALSE,
  comment=NA,
  mysize=TRUE,
  plotdefault=TRUE)

if(!file.exists('figs')) dir.create('figs')

options(digits=4,
	scipen=8,
	width=132
	)
```

```{r eval=FALSE, include=FALSE, echo=FALSE}
## Run this only once and then not again until we want a new version from github
library('devtools')
library('withr')
with_libpaths('./lib', install_github("markmfredrickson/RItools"), 'pre')
```


```{r eval=FALSE, echo=FALSE,include=FALSE}
## Having downloaded optmatch from Box OR http://jakebowers.org/ICPSR for mac (tgz) or windows (zip)
## For Mac
## with_libpaths('./lib',install.packages('optmatch_0.9-8.9003.tgz', repos=NULL),'pre')
## For Windows
## with_libpaths('./lib',install.packages('optmatch_0.9-8.9003.zip', repos=NULL),'pre')
```

```{r eval=FALSE, echo=FALSE}
## Or if you have all of the required libraries (like fortran and c++) for compilation use
with_libpaths('./lib', install_github("markmfredrickson/optmatch"), 'pre')
```

```{r echo=FALSE}
library(dplyr)
library(chemometrics) ## for the nice Mahalanobis ellipse plot
library(RItools,lib.loc="./lib")
library(optmatch,lib.loc="./lib")
library(lmtest)
library(sandwich)
```

## Today

\begin{enumerate}
\item Agenda: Estimation and Testing after matching; Plus more on  Navigating implementation tradeoffs in matching: Information and blocked designs (some blockings have more information about treatment effects than others); Balance tests and the Sequential Intersection Union Principle
\item Reading for this week: DOS 8--9, 13 and \cite[\S~9.5]{gelman2006dau}, and \cite{ho:etal:07} and \cite{hansen2011}.
\item Questions arising from the reading or assignments or life?
\item Next week: Sensitivity Analysis; Non-bipartite matching (matching where the treatment variable not binary).
\end{enumerate}


```{r loaddat, echo=FALSE}
load(url("http://jakebowers.org/Data/meddat.rda"))
meddat$id <- row.names(meddat)
meddat<- mutate(meddat, HomRate03=(HomCount2003/Pop2003)*1000,
                HomRate08=(HomCount2008/Pop2008)*1000)
## mutate strips off row names
row.names(meddat) <- meddat$id
options(show.signif.stars=FALSE)
```

# Matching structure and effective sample size

## Matching with a varying number of controls and full matching

### Fixed vs flexible ratio matching:

\begin{itemize}[<+->]
\item Pair matching \& sample size
\item Effective vs real sample size
\item If we limit ourselves to fixed matching ratios, we gain in
  simplicity but pay a price in sample size (effective \& real).
\item How big a price?  Trying is the best way to find out.
\end{itemize}

```{r echo=FALSE}
nuke.nopt <- subset(nuclearplants, pt == 0)
```

## Matching with a varying number of controls

\begin{minipage}[t]{2in}
\begin{center}
Existing site\\
{\small
% latex table generated in R 3.0.2 by xtable 1.7-3 package
% Thu Jul 31 13:51:34 2014
\begin{tabular}{lrr}
  \hline
 & z.date & z.cap \\
  \hline
A & -1.6 & {1.2} {\mlpnode{NA}} \\
  B & -0.9 & {1.2} {\mlpnode{NB}} \\
  C & -0.4 & {0} {\mlpnode{NC}} \\
  D & -0.4 & {-1.4} {\mlpnode{ND}} \\
  E & 0.1 & {1.1} {\mlpnode{NE}} \\
  F & 2.2 & {0} {\mlpnode{NF}} \\
  G & 1.3 & {0} {\mlpnode{NG}} \\
   \hline
\end{tabular}}
\end{center}
\end{minipage}
\begin{minipage}[t]{2in}
\begin{center}
New site\\
{\scriptsize
% latex table generated in R 3.0.2 by xtable 1.7-3 package
% Thu Jul 31 13:51:34 2014
\begin{tabular}{lrr}
  \hline
 & z.date & z.cap \\
  \hline
{\mlpnode{NH}\mbox{}} {H} & -0.3 & -0.7 \\
  {\mlpnode{NI}\mbox{}} {I} & -1.6 & 1.2 \\
  {\mlpnode{NJ}\mbox{}} {J} & -0.9 & 1.2 \\
  {\mlpnode{NK}\mbox{}} {K} & -0.9 & -1.5 \\
  {\mlpnode{NL}\mbox{}} {L} & -0.7 & -0.0 \\
  {\mlpnode{NM}\mbox{}} {M} & -0.4 & -1.8 \\
  {\mlpnode{NN}\mbox{}} {N} & -0.5 & -0.2 \\
  {\mlpnode{NO}\mbox{}} {O} & -0.3 & -1.3 \\
  {\mlpnode{NP}\mbox{}} {P} & -0.1 & -0.2 \\
  {\mlpnode{NQ}\mbox{}} {Q} & -0.4 & -1.4 \\
  {\mlpnode{NR}\mbox{}} {R} & 0.1 & 1.1 \\
  {\mlpnode{NS}\mbox{}} {S} & 0.1 & 0.1 \\
  {\mlpnode{NT}\mbox{}} {T} & -0.4 & -0.2 \\
  {\mlpnode{NU}\mbox{}} {U} & 0.7 & 0.1 \\
  {\mlpnode{NV}\mbox{}} {V} & 0.4 & 1.3 \\
  {\mlpnode{NW}\mbox{}} {W} & -0.1 & 0.4 \\
  {\mlpnode{NX}\mbox{}} {X} & 0.9 & -0.2 \\
  {\mlpnode{NY}\mbox{}} {Y} & 1.7 & -1.4 \\
  {\mlpnode{NZ}\mbox{}} {Z} & 2.3 & 1.5 \\
   \hline
\end{tabular}}
\end{center}
\end{minipage}
\begin{tikzpicture}[overlay]
  \path[draw,gray] (NA) edge (NI);
 \path[draw,gray] (NB) edge (NJ);
 \path[draw,gray] (NC) edge (NH);
 \path[draw,gray] (NC) edge (NL);
 \path[draw,gray] (NC) edge (NN);
 \path[draw,gray] (NC) edge (NP);
 \path[draw,gray] (NC) edge (NS);
 \path[draw,gray] (NC) edge (NT);
 \path[draw,gray] (NC) edge (NW);
 \path[draw,gray] (ND) edge (NK);
 \path[draw,gray] (ND) edge (NM);
 \path[draw,gray] (ND) edge (NO);
 \path[draw,gray] (ND) edge (NQ);
 \path[draw,gray] (NE) edge (NR);
 \path[draw,gray] (NE) edge (NV);
 \path[draw,gray] (NF) edge (NZ);
 \path[draw,gray] (NG) edge (NU);
 \path[draw,gray] (NG) edge (NX);
 \path[draw,gray] (NG) edge (NY);
 \end{tikzpicture}

```{r}
fmnukeMC1<-fullmatch(pr~date+cap,data=nuke.nopt,min.controls=1)
summary(fmnukeMC1)
```


\note{
  Observe that now no control plants are left out.  (This is something you can change if you want.)

Discuss effective sample size.
}



## Matching so as to maximize effective sample size

\begin{minipage}[t]{2in}
\begin{center}
Existing site\\
{\small
% latex table generated in R 3.0.2 by xtable 1.7-3 package
% Thu Jul 31 13:51:34 2014
\begin{tabular}{lrr}
  \hline
 & z.date & z.cap \\
  \hline
A & -1.6 & {1.2} {\mlpnode{NA}} \\
  B & -0.9 & {1.2} {\mlpnode{NB}} \\
  C & -0.4 & {0} {\mlpnode{NC}} \\
  D & -0.4 & {-1.4} {\mlpnode{ND}} \\
  E & 0.1 & {1.1} {\mlpnode{NE}} \\
  F & 2.2 & {0} {\mlpnode{NF}} \\
  G & 1.3 & {0} {\mlpnode{NG}} \\
   \hline
\end{tabular}}
\end{center}
\bigskip
\bigskip
\bigskip
\bigskip
\end{minipage}
\begin{minipage}[t]{2in}
\begin{center}
New site\\
{\scriptsize
% latex table generated in R 3.0.2 by xtable 1.7-3 package
% Thu Jul 31 13:51:34 2014
\begin{tabular}{lrr}
  \hline
 & z.date & z.cap \\
  \hline
{\mlpnode{NH}\mbox{}} {H} & -0.3 & -0.7 \\
  {\mlpnode{NI}\mbox{}} {I} & -1.6 & 1.2 \\
  {\mlpnode{NJ}\mbox{}} {J} & -0.9 & 1.2 \\
  {\mlpnode{NK}\mbox{}} {K} & -0.9 & -1.5 \\
  {\mlpnode{NL}\mbox{}} {L} & -0.7 & -0.0 \\
  {\mlpnode{NM}\mbox{}} {M} & -0.4 & -1.8 \\
  {\mlpnode{NN}\mbox{}} {N} & -0.5 & -0.2 \\
  {\mlpnode{NO}\mbox{}} {O} & -0.3 & -1.3 \\
  {\mlpnode{NP}\mbox{}} {P} & -0.1 & -0.2 \\
  {\mlpnode{NQ}\mbox{}} {Q} & -0.4 & -1.4 \\
  {\mlpnode{NR}\mbox{}} {R} & 0.1 & 1.1 \\
  {\mlpnode{NS}\mbox{}} {S} & 0.1 & 0.1 \\
  {\mlpnode{NT}\mbox{}} {T} & -0.4 & -0.2 \\
  {\mlpnode{NU}\mbox{}} {U} & 0.7 & 0.1 \\
  {\mlpnode{NV}\mbox{}} {V} & 0.4 & 1.3 \\
  {\mlpnode{NW}\mbox{}} {W} & -0.1 & 0.4 \\
  {\mlpnode{NX}\mbox{}} {X} & 0.9 & -0.2 \\
  {\mlpnode{NY}\mbox{}} {Y} & 1.7 & -1.4 \\
  {\mlpnode{NZ}\mbox{}} {Z} & 2.3 & 1.5 \\
   \hline
\end{tabular}}
\end{center}
\end{minipage}
\begin{tikzpicture}[overlay]
  \path[draw,gray] (NA) edge (NI);
 \path[draw,gray] (NA) edge (NJ);
 \path[draw,gray] (NB) edge (NL);
 \path[draw,gray] (NB) edge (NN);
 \path[draw,gray] (NB) edge (NW);
 \path[draw,gray] (NC) edge (NH);
 \path[draw,gray] (NC) edge (NO);
 \path[draw,gray] (NC) edge (NT);
 \path[draw,gray] (ND) edge (NK);
 \path[draw,gray] (ND) edge (NM);
 \path[draw,gray] (ND) edge (NQ);
 \path[draw,gray] (NE) edge (NR);
 \path[draw,gray] (NE) edge (NS);
 \path[draw,gray] (NE) edge (NV);
 \path[draw,gray] (NF) edge (NY);
 \path[draw,gray] (NF) edge (NZ);
 \path[draw,gray] (NG) edge (NP);
 \path[draw,gray] (NG) edge (NU);
 \path[draw,gray] (NG) edge (NX);
 \end{tikzpicture}
```{r}
fmnuke<-fullmatch(pr~date+cap, min=2, max=3, data=nuke.nopt)
summary(fmnuke)
```

## Showing matches

\centering
```{r out.width=".8\\textwidth", echo=FALSE}
library(igraph)
blah1 <- outer(fmnuke,fmnuke,FUN=function(x,y){ as.numeric(x==y) })
blah2 <- outer(fmnukeMC1,fmnukeMC1,FUN=function(x,y){ as.numeric(x==y) })
par(mfrow=c(1,2),mar=c(3,3,3,1))
plot(graph_from_adjacency_matrix(blah1,mode="undirected",diag=FALSE),
     vertex.color=c("white","green")[nuke.nopt$pr+1],main="Min C=2, Max C=3")
plot(graph_from_adjacency_matrix(blah2,mode="undirected",diag=FALSE),
     vertex.color=c("white","green")[nuke.nopt$pr+1],main="Min C=1, Max C=Inf")
```

## Matching so as to maximize effective sample size

\tiny
```{r}
effectiveSampleSize(fmnuke)
nukewts <- cbind(nuke.nopt,fmnuke) %>% 
  group_by(fmnuke) %>% summarise(nb = n(),
                                 nTb = sum(pr),
                                 nCb = nb - nTb,
                                 hwt = ( 2*( nCb * nTb ) / (nTb + nCb)))
nukewts
dim(nukewts)
sum(nukewts$hwt)
```

## Matching so as to maximize effective sample size

```{r}
stratumStructure(fmnuke)
```

So effective sample size for this match = $2 * 4/3 + 5* 3/2 = 10.17$ --- compare to 7 for pairs, 9.33 for triples.

```{r}
nukemh <- match_on(pr~date+cap,data=nuke.nopt)
pmnuke <- pairmatch(nukemh,data=nuke.nopt)
levels(pmnuke)
effectiveSampleSize(pmnuke)
```

## Matching so as to maximize effective sample size

```{r}
mds <- matched.distances(fmnuke,nukemh)
mean(unlist(mds))
```

Mean of matched distances is `r mean(unlist(mds))` --- compare to 0.29 for pairs, 0.57 for triples.

Note variance/bias tradeoff. Covariate balance has to do with bias and extrapolation (i.e. confounding). We want balance **and** precision.


## Inspect distances after matching

What kinds of distances remain after matching?

```{r}
str(mds)
quantile(unlist(mds))
```


##  Tracking effective sample size

In 2-sample comparisons, total sample size can be a misleading as a measure of information content.  Example:
\begin{itemize}
\item say $Y$ has same variance, $\sigma^{2}$,in the Tx and the Ctl population.
\item Ben H. samples 10 Tx and 40 Ctls, and
\item Justin M. samples 25 Tx and 25 Ctls
\end{itemize}
--- so that total sample sizes are the same.  However,

\begin{eqnarray*}
  V_{BH}(\bar{y}_{t} - \bar{y}_{c}) &=& \frac{\sigma^{2}}{10} + \frac{\sigma^{2}}{40}=.125\sigma^{2}\mbox{;}\\
  V_{JM}(\bar{y}_{t} - \bar{y}_{c}) &=& \frac{\sigma^{2}}{25} + \frac{\sigma^{2}}{25}=.08\sigma^{2}.\\
\end{eqnarray*}

Similarly, a matched triple is roughly $[(\sigma^{2}/1 + \sigma^{2}/2)/(\sigma^{2}/1 + \sigma^{2}/1)]^{-1}= 1.33$ times as informative as a matched pair.

## Details

Use pooled 2-sample t statistic SE formula to compare 1-1 vs 1-2 matched sets' contribution to variance:

$$
\begin{array}{c|c}
  \atob{1}{1} & \atob{1}{2} \\
M^{-2}\sum_{m=1}^{M} (\sigma^{2}/1 + \sigma^{2}/1) & M^{-2}\sum_{m=1}^{M} (\sigma^{2}/1 + \sigma^{2}/2) \\
\frac{2\sigma^{2}}{M} & \frac{1.5\sigma^{2}}{M} \\
\end{array}
$$

So 20 matched pairs is comparable to 15 matched triples.

(Correspondingly, h-mean of $n_{t},n_{c}$ for a pair is 1, while for a triple it's $[(1/1 + 1/2)/2]^{-1}=4/3$.)

The variance of the `pr`-coeff in `v~pr + match` is
$$
 \frac{2 \sigma^{2}}{\sum_{s} h_{s}}, \hspace{3em} h_{s} = \left( \frac{n_{ts}^{-1} + n_{cs}^{-1} }{2}  \right)^{-1} ,
$$

assuming the OLS model and homoskedastic errors.  (This is b/c the anova formulation is equivalent to harmonic-mean weighting, under which $V(\sum_{s}w_{s}(\bar{v}_{ts} - \bar v_{cs})) = \sum_{s} w_{s}^{2}(n_{ts}^{-1} + n_{cs}^{-1}) \sigma^{2} = \sigma^{2} \sum_{s} w_{s}^{2} 2 h_{s}^{-1} = 2\sigma^{2} \sum_{s}w_{s}/\sum_{s}h_{s} = 2\sigma^{2}/\sum_{s} h_{s}$.)

For matched pairs, of course, $h_{s}=1$.  Harmonic mean of 1, 2 is $4/3$. Etc.

# Balance


 Hansen and Sales (2015) suggest one way to stop iterating between
	\texttt{fullmatch} and \texttt{xBalance} when you have one caliper. The idea
	is that if you would reject the null of balance with one caliper, you would
	also certainly reject it with a wider caliper. That is, the idea is that
	hypothesis tests about balance using calipers can be understood as nested,
	or ordered. Rosenbaum (2008) talks about this in his paper ``Testing
	Hypotheses in Order'' and Hansen and Sales (2008) how these ideas can
	help us choose a matched design:

  ``The SIUP[sequential intersection union principle] states that if a
  researcher pre-specifies a sequence of hypotheses and corresponding
  level-$\alpha$ tests, tests those hypotheses in order, and stops testing
  after the first non-rejected hypothesis, then the probability of incorrectly
  rejecting at least one correct hypothesis is at most $\alpha$.'' (page 2)

	Let us try this out and also try to assess it. Say, we start by saying that
	we will reject the null of balance at $\alpha=.50$.


Imagine, for example we had this matched design:

```{r}
balfmla <- nhTrt ~ nhPopD + nhAboveHS + HomRate03
mhdist <- match_on(balfmla,data=meddat)
psmod <- arm::bayesglm(balfmla,data=meddat,family=binomial(link="logit"))
psdist <- match_on(psmod,data=meddat)
tmp <- meddat$HomRate03
names(tmp) <- rownames(meddat)
absdist <- match_on(tmp, z = meddat$nhTrt,data=meddat)

summary(psdist)
summary(mhdist)
summary(absdist)

fmMh <- fullmatch(mhdist,data=meddat,tol=.00001)
summary(fmMh,min.controls=0,max.controls=Inf)
```

## Principled Balance Search using SIUP

If we just want to find the set of calipers and optmatch options which maximize
balance, why not do a search?

> The SIUP[sequential intersection union principle] states that if a researcher pre-specifies a sequence of hypotheses and corresponding level-$\alpha$ tests, tests those hypotheses in order, and stops testing after the first non-rejected hypothesis, then the probability of incorrectly rejecting at least one correct hypothesis is at most $\alpha$.'' \autocite{saleshansen2014}(page 2)


```{r}
matchAndBalance<-function(x,balfmla,distmat,thedata){
	#x is a caliper width
	thefm<-fullmatch(distmat+caliper(distmat,x),data=thedata,tol=.00001)
  ## This next is very annoying but there are scope problems with balanceTest and balfmla
  ## And I don't want to add thefm to meddat at each iteration, which would really slow things down.
	thexb<-balanceTest(update(balfmla,.~.+strata(thefm)),
			data=cbind(thedata,thefm),
			report=c("chisquare.test"))
	return(c(x=x,d2p=thexb$overall["thefm","p.value"]))
}
```


## Balance Search using SIUP


```{r cache=FALSE}
## Start with the the largest distance between a treated and control unit.
maxpsdist<-max(as.vector(psdist))
minpsdist<-min(as.vector(psdist))
psdistsum <- summary(psdist)
quantile(as.vector(psdist),seq(0,1,.1))
```

```{r cache=TRUE}
results1<-sapply(seq(3,minpsdist,length=100),function(thecal){
                   matchAndBalance(thecal,balfmla,distmat=psdist,thedata=meddat)})

apply(results1,1,summary)
apply(results1[,results1["d2p",]>.8],1,summary)
```


```{r echo=FALSE}
## Reorder the data from low to high to make cummax work better
results1 <- data.frame(t(results1))
results1o <- results1[order(results1$x,decreasing=TRUE),]
results1o$maxp <- cummax(results1o$d2p)
```

## Balance Search using SIUP

So, you can see that keeping the maximum produces a set of nested tests so
that if I reject some caliper at some $p$, I know that any caliper tighter
than the chosen one would have less balance (a smaller $p$, more information
against the null that our design is like a well randomized block randomized
study).


```{r echo=FALSE, out.width=".8\\textwidth"}
with(results1o,{
       plot(x,d2p,xlab="PS Caliper",ylab="d2 p",cex=.6)
       points(x,maxp,col="blue")
})
```

\note{
  Sometimes we want our matched designs to relate well not only to an
	equivalent block-randomized experiment, but also to help us make the
	argument that our comparisons are comparing specific kinds of like
	with like and/or that our comparisons are statistically powerful.
	That is, among matched designs that we might call "balanced", we might
	one which drops the fewest observations, and perhaps one that has
	specially good balance on certain special covariates (like baseline
	outcomes). So, here is one example, of doing such a search.

	In this case, we are not doing strictly nested hypothesis testing, but are
	using the $p$ values to tell us about information against the null of
	balance rather than using them strictly speaking to reject this null, or
	not-reject it.
  }

## Design Search for both precision and balance

Here I demonstrate searching for two calipers.

```{r gridsearch, cache=FALSE}

findbalance<-function(x){
	##message(paste(x,collapse=" "))
	thefm<-try(fullmatch(psdist+caliper(mhdist,x[2])+caliper(psdist,x[1]),data=meddat,tol=.00001))

	if(inherits(thefm,"try-error")){
		return(c(x=x,d2p=NA,maxHR03diff=NA,n=NA,effn=NA))
	}

	thexb<-try(balanceTest(update(balfmla,.~.+strata(thefm)),
			    data=cbind(meddat,thefm),
			    report=c("chisquare.test","p.values")),silent=TRUE)

	if(inherits(thexb,"try-error")){
		return(c(x=x,d2p=NA,maxHR03diff=NA,n=NA,effn=NA))
	}

	maxHomRate03diff<-max(unlist(matched.distances(thefm,distance=absdist)))

	return(c(x=x,d2p=thexb$overall["thefm","p.value"],
           maxHR03diff=maxHomRate03diff,
           n=sum(!is.na(thefm)),
           effn=summary(thefm)$effective.sample.size))

}

```

```{r echo=FALSE, cache=TRUE}
## Test the function
## findbalance(c(3,3))
## Don't worry about errors for certain combinations of parameters
maxmhdist<-max(as.vector(mhdist))
minmhdist<-min(as.vector(mhdist))
set.seed(123455)
system.time({
	results<-replicate(1000,findbalance(c(runif(1,minpsdist,maxpsdist),
					      runif(1,minmhdist,maxmhdist))))
}
)

```

```{r eval=FALSE, echo=FALSE}
## If you have a mac or linux machine you can speed this up:
library(parallel)
system.time({
	resultsList<-mclapply(1:5000,function(i){
				      findbalance(c(runif(1,minpsdist,maxpsdist),
						    runif(1,minmhdist,maxmhdist)))},
			      mc.cores=detectCores())
	resultsListNA<-sapply(resultsList,function(x){ any(is.na(x)) })
	resultsArr<-simplify2array(resultsList[!resultsListNA])
}
)

```


## Which matched design might we prefer?

Now, how might we interpret the results of this search for matched designs?
Here are a few ideas.

```{r }
if(class(results)=="list"){
resAnyNA<-sapply(results,function(x){ any(is.na(x)) })
resNoNA<-simplify2array(results[!resAnyNA])
} else {
resAnyNA<-apply(results,2,function(x){ any(is.na(x)) })
resNoNA<-simplify2array(results[,!resAnyNA])
}
apply(resNoNA,1,summary)
highbalres<-resNoNA[,resNoNA["d2p",]>.5]
apply(highbalres,1,summary)
```

## Which matched design might we prefer?

```{r eval=TRUE, echo=FALSE}
# color points more dark for smaller differences
plot(resNoNA["d2p",],resNoNA["n",],
     xlab='d2p',ylab='n',
     col=gray(1- ( resNoNA["maxHR03diff",]/max(resNoNA["maxHR03diff",]))),
     pch=19)

## identify(resNoNA["d2p",],resNoNA["n",],labels=round(resNoNA["maxHR03diff",],3),cex=.7)
```

```{r eval=TRUE,echo=TRUE}
interestingDesigns<- (resNoNA["d2p",]>.5 & resNoNA["n",]>=10 &
		      resNoNA["maxHR03diff",]<=1 & resNoNA["effn",] > 6)
candDesigns <- resNoNA[,interestingDesigns]
str(candDesigns)
apply(candDesigns,1,summary)
candDesigns<-candDesigns[,order(candDesigns["d2p",],decreasing=TRUE)]
```

## How would we use this information in \texttt{fullmatch}?

```{r }
fm4<-fullmatch(psdist+caliper(psdist,candDesigns["x1",2])
	       +caliper(mhdist,candDesigns["x2",2]),data=meddat,tol=.00001)

summary(fm4,min.controls=0,max.controls=Inf)

meddat$fm4<-NULL ## this line exists to prevent confusion with new fm4 objects
meddat[names(fm4),"fm4"]<-fm4

xb3<-balanceTest(update(balfmla,.~.+strata(fm4)),
	      data=meddat, report=c("all"))
xb3$overall[,1:3]
zapsmall(xb3$results["HomRate03",,])
```

## Another approach: more fine tuned optimization

```{r eval=TRUE,cache=FALSE}
matchAndBalance2<-function(x,distmat,alpha){
	#x is a caliper widths
	if(x>max(as.vector(distmat)) | x<min(as.vector(distmat))){ return(99999) }
	thefm<-fullmatch(distmat+caliper(distmat,x),data=meddat,tol=.00001)
	thexb<-xBalance(balfmla,
			strata=data.frame(thefm=thefm),
			data=meddat,
			report=c("chisquare.test"))
	return(thexb$overall[,"p.value"])
}

maxpfn<-function(x,distmat,alpha){
	## here x is the targeted caliper width and x2 is the next wider
	## caliper width
	p1<-matchAndBalance2(x=x[1],distmat,alpha)
	p2<-matchAndBalance2(x=x[2],distmat,alpha)
	return(abs( max(p1,p2) - alpha) )
}

maxpfn(c(minpsdist,minpsdist+1),distmat=psdist,alpha=.25)
quantile(as.vector(psdist),seq(0,1,.1))
sort(as.vector(psdist))[1:10]
```

## Another approach: more fine tuned optimization

```{r solnp, cache=TRUE}
library(Rsolnp)
### This takes a long time
results3<-gosolnp(fun=maxpfn,
		ineqfun=function(x,distmat,alpha){ x[2] - x[1] },
		ineqLB = 0,
		ineqUB = maxpsdist,
		LB=c(minpsdist,minpsdist+.01),
		UB=c(maxpsdist-.01,maxpsdist),
		n.restarts=2,
		alpha=.25,distmat=psdist,
		n.sim=500,
		rseed=12345,
		control=list(trace=1)
		)
```

## Another approach: more fine tuned optimization

```{r}
maxpfn(results3$pars,distmat=psdist,alpha=.25)
matchAndBalance2(results3$pars[1],distmat=psdist,alpha=.25)
matchAndBalance(results3$par[1],balfmla=balfmla,distmat=psdist,thedata=meddat)
```

# Estimation and Testing

## Overview: Estimate and Test "as if block-randomized"

This means we have to *define* our estimands in weighted terms (because different blocks provide different amounts of information).  Here are some matched sets and their associated weights:

```{r echo=TRUE}
meddat[names(fmMh),"fmMh"] <- fmMh
setmeanDiffs <- meddat %>% group_by(fmMh) %>%
  summarise(Y=mean(HomRate08[nhTrt==1])-mean(HomRate08[nhTrt==0]),
            nb=n(),
            nTb = sum(nhTrt),
            nCb = sum(1-nhTrt),
            hwt = ( 2*( nCb * nTb ) / (nTb + nCb))
            )
setmeanDiffs
```

## Using the weights: Set size weights

First, we could calculate the set-size weighted ATE:

```{r}
## The set-size weighted version
with(setmeanDiffs, sum(Y*nb/sum(nb)))
```

## Using the weights: Set size weights

Sometimes it is convenient to use `lm` because there are R functions for standard errors and confidence intervals. This requires a bit more work:

```{r}
source("http://jakebowers.org/ICPSR/confintHC.R")
## See Gerber and Green section 4.5 and also Chapter 3 on block randomized experiments. Also Hansen and Bowers 2008.
meddat <- meddat %>% group_by(fmMh) %>% mutate(trtprob=mean(nhTrt),
                                               nbwt=nhTrt/trtprob + (1-nhTrt)/(1-trtprob),
                                               gghbwt= 2*( n()/nrow(meddat) )*(trtprob*(1-trtprob)), ## GG version,
                                               nb = n(),
                                               nTb = sum(nhTrt),
                                               nCb = nb - nTb,
                                               hbwt = ( 2*( nCb * nTb ) / (nTb + nCb))
                                               )
row.names(meddat) <- meddat$id ## dplyr strips row.names
lm0b<-lm(HomRate08~nhTrt,data=meddat,weight=nbwt)
coef(lm0b)["nhTrt"]
coeftest(lm0b,vcov=vcovHC(lm0b,type="HC2"))[1:2,]
theci0 <- confint.HC(lm0b,parm="nhTrt",thevcov=vcovHC(lm0b,type="HC2"))
```


## Using the weights: Set size weights

There is an even more elaborate version of this that we illustrate below (from Winston Lin via the Green Lab SOP).

```{r}
meddat <- meddat %>% group_by(fmMh) %>% mutate(fmwt=n()/nrow(meddat),nb=n())
X <- model.matrix(~fmMh-1,data=meddat)
XminusXbar <- apply(X,2,function(x){ x - mean(x) })
wrkdat <- cbind.data.frame(meddat,data.frame(XminusXbar))
tmpfmla <- reformulate(grep("fmMh1",names(wrkdat),value=TRUE)[-1],response="HomRate08")
lmfmla <- update(tmpfmla,.~nhTrt*(.))
lm0 <- lm(lmfmla,data=wrkdat)
coef(lm0)["nhTrt"]
## But, in this case, the HC2 Standard Error is undefined because some of our blocks have too few observations.
coeftest(lm0,vcov=vcovHC(lm0,type="HC2"))[1:2,]

## See Gerber and Green 4.5
meddat$Zf <- factor(meddat$trtprob)
Z <- model.matrix(~Zf-1,data=meddat)
ZminusZbar <- apply(Z,2,function(z){ z - mean(z) })
wrkdat <- cbind.data.frame(wrkdat,ZminusZbar)
tmpfmla <- reformulate(grep("Zf0",names(wrkdat),value=TRUE)[-1],response="HomRate08")
lmfmlaZ <- update(tmpfmla,.~nhTrt*(.))
lm0a <- lm(lmfmlaZ,data=wrkdat)
coef(lm0a)["nhTrt"]
coeftest(lm0a,vcov=vcovHC(lm0a,type="HC2"))[1:2,]
```

## Using the weights: precision weights

Set size weighting is easy to explain. But set size weighting leaves information on the table. (For example, if we used a set-size weighted mean difference as a test statistic, the null reference distribution would be wider than if we used a harmonic weighted mean difference test statistic.) See this for example:

```{r}
## The mean diff used as the observed value in the testing
## or from lm
with(setmeanDiffs, sum(Y*hwt/sum(hwt)))
lm1 <- lm(HomRate08~nhTrt+fmMh,data=meddat)
xbOutcome <- balanceTest(nhTrt~HomRate08+strata(fmMh),data=meddat,report="all")
## lm1a <- lm(HomRate08~nhTrt,data=meddat,weights=1/hbwt)
coeftest(lm1,parm="nhTrt",vcov=vcovHC(lm1,type="HC2"))["nhTrt",]
coeftest(lm0b,parm="nhTrt",vcov=vcovHC(lm0b,type="HC2"))["nhTrt",]
theci1 <- confint.HC(lm1,parm="nhTrt",thevcov=vcovHC(lm1,type="HC2"))
theci1
diff(theci1[1,])
diff(theci0[1,]) ## set size weighting
```

## `balanceTest`

`balanceTest` reports the ATT (the set mean differences weighted by number of treated) but uses the precision-weighted mean for the $p$-value.

```{r echo=FALSE,eval=FALSE}
## The descriptive adj.mean diff from balanceTest
with(setmeanDiffs, sum(Y*nTb/sum(nTb)))
xbOutcome$results[,,"fmMh"]
```

## The direct permutation approach

```{r}
library(permute)

newExp <- function(z,b){
  n <- length(z)
  h1 <- how(blocks=b)
  z[shuffle(n,control=h1)]
}

hwtfn <- function(data){
  tapply(data$Tx.grp,data$stratum.code,function(x){ 2* sum((x-mean(x))^2) })
}

setsizewtfn<- function(data){
  tapply(data$Tx.grp,data$stratum.code,function(x){ length(x) })
}

trtsizewtfn<- function(data){
  ## Assumes Tx.grp \in \{0,1\} and 1=assigned to treatment
  tapply(data$Tx.grp,data$stratum.code,function(x){ sum(x) })
}

wtMeanDiffTZ<-function(y,z,b,wtfn){
  tzb <- mapply(function(yb,zb){ mean(yb[zb==1]) - mean(yb[zb==0]) },
                yb=split(y,b),
                zb=split(z,b))
  wts <- wtfn(data.frame(Tx.grp=z,stratum.code=b))
  sum(tzb*wts/sum(wts))
}

## Testing (compare to above)
obsHTZ<-wtMeanDiffTZ(meddat$HomRate08,meddat$nhTrt,fmMh,hwtfn)
obsNTZ<-wtMeanDiffTZ(meddat$HomRate08,meddat$nhTrt,fmMh,setsizewtfn)
obsTTZ<-wtMeanDiffTZ(meddat$HomRate08,meddat$nhTrt,fmMh,trtsizewtfn)
```

Test the null hypothesis of no effects:

```{r cache=TRUE}

set.seed(12345)
nulldistHwt <- replicate(10000,wtMeanDiffTZ(meddat$HomRate08,newExp(meddat$nhTrt,fmMh),fmMh,hwtfn))
set.seed(12345)
nulldistNwt <- replicate(10000,wtMeanDiffTZ(meddat$HomRate08,newExp(meddat$nhTrt,fmMh),fmMh,setsizewtfn))

## Notice more precision with the Harmonic weight in thes p-values.
2*min(mean(nulldistHwt>=obsHTZ),mean(nulldistHwt<=obsHTZ))
2*min(mean(nulldistNwt>=obsNTZ),mean(nulldistNwt<=obsNTZ))

```

Comparing the reference distributions to each other and to their Normal approximations.

```{r}
plot(density(nulldistHwt),ylim=c(0,3))
lines(density(nulldistNwt),lty=2)
curve(dnorm(x,sd=sd(nulldistHwt)),from=-1,to=1,col="gray",add=TRUE)
curve(dnorm(x,sd=sd(nulldistNwt)),from=-1,to=1,col="gray",lty=2,add=TRUE)
```


```{r echo=FALSE,eval=FALSE}
## balanceTest does not yet take functions.
hwtfn <- function(data){
  tapply(data$Tx.grp,data$stratum.code,function(x){ 2* sum((x-mean(x))^2) })
}

setsizewtfn<- function(data){
  tapply(data$Tx.grp,data$stratum.code,function(x){ length(x) })
}

wtMeanDiffTZ<-function(y,z,b,wtfn){
  temp <- balanceTest(z~y+strata(b),data=data.frame(y,z,b),
                      report="adj.mean.diffs",
                      stratum.weights=wtfn)
  return(temp$results[1,"adj.diff",1])
}

## Test the function
## table(newExp(meddat$nhTrt,meddat$fmMh),meddat$fmMh)

```

# Your reports

## Your reports?

What did you do? What kind of overall balance did you achieve so far? What did you do? What kinds of confidence intervals and/or hypothesis tests did you produce?

## Anything Else?

 - Recall the utility of `fill.NAs()`: You can and should match on missingness. No reason to throw away observations only because of covariate missingness.

 - EXTRA: How would we assess the claim that the sequential intersection union principle controls the family-wise error rate for balance tests?



## References
