---
title: |
 | "Challenges to Randomization:
 |  Noncompliance and Missing Data"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
author: |
  | ICPSR 2025 Session 1
  | Jake Bowers, Ben Hansen \& Tom Leavitt
bibliography:
 - "BIB/MasterBibliography.bib"
fontsize: 10pt
geometry: margin=1in
graphics: yes
biblio-style: authoryear-comp
biblatexoptions:
  - natbib=true
colorlinks: true
---

<!-- Make this document using library(rmarkdown); render("day12.Rmd") -->


```{r setup1_env, echo=FALSE, include=FALSE}
library(here)
source(here::here("rmd_setup.R"))
```


```{r setup2_loadlibs, echo=FALSE, include=FALSE}
## Load all of the libraries that we will use when we compile this file
## We are using the renv system. So these will all be loaded from a local library directory
library(dplyr)
library(ggplot2)
library(coin)
library(RItools)
library(DeclareDesign)
```

## Today

1. Agenda: One step away from easy to interpret experiments: non-random
   doses/compliance [@gerbergreen2012] Chapter 5, non-random missing data
   [@gerbergreen2012] Chapter 7 and the Threats module of [The Theory and Practice of Field Experiments](https://egap.github.io/theory_and_practice_of_field_experiments/).
2. Recap: We use statistics to **infer** about unobservable counterfactual
   quantities (functions of potential outcomes); we can estimate unobservable
   averages; we can test unobservable hypotheses; we can test unobservable
   hypotheses about averages.
3. Questions arising from the reading or assignments or life?

# Causal effects when we do not control the dose

## Defining causal effects I

Imagine a door-to-door communication experiment where some houses are randomly assigned to receive a visit.  Note that we now use $Z$ and $d$ instead of $T$.

 - $Z_i$ is random assignment to a visit ($Z_i=1$) or not ($Z_i=0$).
 - $d_{i,Z_i=1}=1$ means that person $i$ would open the door to have a conversation when assigned a visit.
 - $d_{i,Z_i=1}=0$ means that person $i$ would not open the door to have a conversation when assigned a visit.
 - Opening the door is an outcome of the treatment.

\begin{center}
\begin{tikzcd}[ampersand replacement=\&]
Z  \arrow[from=1-1,to=1-2, "\ne 0"] \arrow[from=1-1, to=1-4, bend left, "\text{0 (exclusion)}"] \& d  \arrow[from=1-2,to=1-4] \& \& y \\
(x_1 \ldots x_p) \arrow[from=2-1,to=1-1, "\text{0 (as if randomized)}"]  \arrow[from=2-1,to=1-2] \arrow[from=2-1,to=1-4]
\end{tikzcd}
\end{center}

<!--
## Defining causal effects II
 - $y_{i,Z_i = 1, d_{i,Z_i=1}=1}$ is the potential outcome for people who were assigned a visit and who opened the door. ("Compliers" or "Always-takers")

 - $y_{i,1, d_{i,Z_i=1}=0}$ is the potential outcome for people who were assigned a visit and who did not open the door. ("Never-takers" or "Defiers")

 - $y_{i,0, d_{i,0}=1}$ is the potential outcome for people who were not assigned a visit and who opened the door. ("Defiers" or "Always-takers")

 - $y_{i,0, d_{i,0}=0}$ is the potential outcome for people who were not assigned a visit and who would not have opened the door. ("Compliers" or "Never-takers")

## Defining causal effects III
 We could also write $y_{i,Z_i = 0, d_{i,Z_i=1}=1}$ for people who were not assigned a visit but who would have opened the door had they been assigned a visit etc.

In this case we can simplify our potential outcomes:

  - $y_{i,0, d_{i,1}=1} = y_{i,0, d_{i,1}=0} = y_{i,0, d_{i,0}=0}$ because your outcome is the same regardless of how you don't open the door.

-->

## Defining causal effects II <!--IV--> 

We can simplify the ways in which people get a dose of the treatment like so
(where $d$ is lower case reflecting the idea that whether you open the door
when visited or not is a fixed attribute like a potential outcome).

 - $Y$ : outcome ($y_{i,Z}$ or $y_{i,Z_i=1}$ for potential outcome to
   treatment for person $i$, fixed)
 - $X$ : covariate/baseline variable
 - $Z$ : treatment assignment ($Z_i=1$ if assigned to a visit, $Z_i=0$ if not
   assigned to a visit)
 -  $D$ : treatment received ($D_i=1$ if answered door, $D_i=0$ if person $i$
   did not answer the door) (using $D$ here because $D_i = d_{i,1} Z_{i} + d_{i,0} (1-Z_i)$)

## Defining causal effects III <!--V--> 

We have two causal effects of $Z$: $Z \rightarrow Y$ (known as $\delta$, ITT, ITT$_Y$), and $Z \rightarrow D$ (known as ITT$_D$, $p_c$).

And different types of people can react differently to the attempt to move the
dose with the instrument.

\centering
\begin{tabular}{llcc}
                       &        & \multicolumn{2}{c}{$Z=1$} \\
		       &       & $D=0$ & $D=1$ \\
		       \midrule
\multirow{2}{*}{$Z=0$} & $D=0$ & Never taker & Complier \\
                       & $D=1$ & Defier     & Always taker \\
		       \bottomrule
\end{tabular}


##  Defining causal effects IV <!--VI--> 


The $ITT=ITT_Y=\delta= \bar{y}_{Z=1} - \bar{y}_{Z=0}$.

\medskip

But, in this design, $\bar{y}_{Z=1}=\bar{y}_{1}$ is split into pieces: the
outcome of those who answered the door (Compliers and Always-takers and
Defiers). Write $p_C$ for the proportion of compliers in the study, $p_A$ for
proportion always-takers, etc... The proportions have to sum to 1. So, we have weighted averages:

\begin{equation*}
\bar{y}_{1}=(\bar{y}_{1}|C)p_C + (\bar{y}_{1}|A)p_A + (\bar{y}_1|N)p_N + (\bar{y}_1|D)p_D.
\end{equation*}

And $\bar{y}_{0}$ is also split into pieces:

\begin{equation*}
\bar{y}_{0}=(\bar{y}_{0}|C)p_C + (\bar{y}_{1}|A)p_A + (\bar{y}_{0}|N)p_N + (\bar{y}_0|D)p_D.
\end{equation*}

##  Defining causal effects V <!--VII--> 

So, the ITT itself is a combination of the effects of $Z$ on $Y$ within these
different groups.

\medskip

People who are compliers tend to be different types of people
than people who are always takers: comparisons across types would raise
questions about how to interpret the results --- interpretations that would
focus more on differences in types than in differences caused by $Z$.

\medskip

But, we
can still estimate it because we have unbiased estimators of $\bar{y}_1$ and
$\bar{y}_0$ within each type.

## Learning about the ITT I

First, let's learn about the effect of the policy itself.

\medskip

Let's assume we have no defiers ($p_D=0$). The we can write the ITT more simply.

\begin{equation*}
\bar{y}_{1}=(\bar{y}_{1}|C)p_C + (\bar{y}_{1}|A)p_A + (\bar{y}_1|N)p_N
\end{equation*}

\begin{equation*}
\bar{y}_{0}=(\bar{y}_{0}|C)p_C + (\bar{y}_{0}|A)p_A + (\bar{y}_{0}|N)p_N
\end{equation*}


## Learning about the ITT II

First, let's learn about the effect of the policy itself. We assume no defiers
($p_D=0$), which allows us to write the ITT more simply. 


\begin{align*}
ITT    = & \bar{y}_{1} - \bar{y}_{0} \\
        = & ( (\bar{y}_{1}|C)p_C + (\bar{y}_{1}|A)p_A + (\bar{y}_1|N)p_N ) - \\
       & ( (\bar{y}_{0}|C)p_C + (\bar{y}_{0}|A)p_A + (\bar{y}_{0}|N)p_N )  \\
       \intertext{collecting each type together --- to have an ITT for each type}
       = & ( (\bar{y}_{1}|C)p_C -  (\bar{y}_{0}|C)p_C )  +   ( (\bar{y}_{1}|A)p_A - (\bar{y}_{0}|A)p_A ) + \\
       & ( (\bar{y}_1|N)p_N  - (\bar{y}_{0}|N)p_N ) \\
       = & \underbrace{\left( (\bar{y}_{1}|C) -  (\bar{y}_{0}|C) \right)}_{\color{blue}\text{ITT among Compliers}}p_C   +  \\
       & \underbrace{\left( (\bar{y}_{1}|A)- (\bar{y}_{0}|A) \right)}_{\color{blue}\text{ITT among Always-Takers}}p_A  +  \underbrace{\left( (\bar{y}_1|N) - (\bar{y}_{0}|N) \right)}_{\color{blue}\text{ITT among Never-Takers}}p_N
\end{align*}

## Learning about the ITT III

\begin{align*}
ITT     = &   \bar{y}_{1} - \bar{y}_{0} \\
        = &  ( (\bar{y}_{1}|C)p_C + (\bar{y}_{1}|A)p_A + (\bar{y}_1|N)p_N ) - \\
       & ( (\bar{y}_{0}|C)p_C + (\bar{y}_{0}|A)p_A + (\bar{y}_{0}|N)p_N )  \\
        = &   ( (\bar{y}_{1}|C)p_C -  (\bar{y}_{0}|C)p_C )  +   ( (\bar{y}_{1}|A)p_A - (\bar{y}_{0}|A)p_A ) + \\
       & ( (\bar{y}_1|N)p_N  - (\bar{y}_{0}|N)p_N ) \\
        = &   ( (\bar{y}_{1}|C) -  (\bar{y}_{0}|C))p_C   +   ( (\bar{y}_{1}|A)- (\bar{y}_{0}|A))p_A  + \\
       & ( (\bar{y}_1|N) - (\bar{y}_{0}|N) )p_N \\
       = & (\text{ITT among compliers})(\text{proportion of compliers}) + \\
       & (\text{ITT among always takers})(\text{proportion of always takers}) + \ldots
\end{align*}


## Learning about the ITT IV

And, if the effect of the dose can only occur for those who open the door, and you can only open the door when assigned to do so then:

\begin{equation*}
( (\bar{y}_{1}|A)- (\bar{y}_{0}|A))p_A = 0  \text{ and } ( (\bar{y}_1|N) - (\bar{y}_{0}|N) )p_N = 0
\end{equation*}

And so, under these assumptions, the ITT is a simple function of the ITT among compliers and the proportion of compliers.

\begin{equation*}
ITT =  ( (\bar{y}_{1}|C) -  (\bar{y}_{0}|C))p_C  = ( CACE ) p_C.
\end{equation*}


## The complier average causal effect I

If we want to can learn about the the causal effect of answering the door and
having the conversation why not just compare people who answer the door to
people who do not?

\medskip

The problem with this "as-treated" or "per-protocol" comparison is that this
comparison is confounded by $x$: a simple $\bar{Y}|D=1 - \bar{Y}|D=0$ comparison
tells us about differences in the outcome due to $x$ in addition to the
difference caused by $D$. (Numbers below from some simulated data)

\begin{center}
\begin{tikzcd}[ampersand replacement=\&]
Z  \arrow[from=1-1,to=1-2] \arrow[from=1-1, to=1-4, bend left, "\text{0 (exclusion)}"] \& D  \arrow[from=1-2,to=1-4] \& \& y \\
(x_1 \ldots x_p) \arrow[from=2-1,to=1-1, "\text{-.006 (as if randomized)}"]  \arrow[from=2-1,to=1-2, ".06"] \arrow[from=2-1,to=1-4, ".48"]
\end{tikzcd}
\end{center}


## The complier average causal effect II

In actual data:

```{r cors, eval=FALSE, echo=TRUE, results="hide"}
with(dat, cor(Y, x)) ## can be any number
with(dat, cor(d, x)) ## can be any number
with(dat, cor(Z, x)) ## should be near 0
```

And we just saw that, in this design, and with these assumptions (including a
SUTVA assumption) that $ITT =  ( (\bar{y}_{1}|C) -  (\bar{y}_{0}|C))p_C  =
(CACE) p_C$, so we can define $CACE=ITT/p_C$. That is, we can learn about the
effect of answering the door without worrying about the bias from $x$ (or any
set of $x$'s).

\medskip

**VERY COOL** You can learn about the causal effect of a non-random intervention (deciding to open the door) without "controlling for" $x_1,x_2,\ldots$ in this case.

## How to calculate the ITT and CACE/LATE I

```{r simivdesign, echo=if (handout_) TRUE else if (solutions_) 1L:3L}
prob_comply <- .5
tau <- .5
N <- 100

the_pop <- declare_population(
  N = N,
  u0 = rnorm(N),
  u = ifelse(u0<=0,0,u0), ## truncated Normal
  type = sample(c("Always-Taker", "Never-Taker", "Complier", "Defier"), N,
    replace = TRUE,
    prob = c(.1, 1 - unique(prob_comply), unique(prob_comply), 0)
  )
)

##  The unobserved potential outcomes,  Y(Z=1) and Y(Z=0) relate to the observed outcome, Y, via treatment assignment and a constant additive effect of tau.
## D refers to getting a dose of feedback
d_po <- declare_potential_outcomes(
  D ~ case_when(
    Z == 0 & type %in% c("Never-Taker", "Complier") ~ 0,
    Z == 1 & type %in% c("Never-Taker", "Defier") ~ 0,
    Z == 0 & type %in% c("Always-Taker", "Defier") ~ 1,
    Z == 1 & type %in% c("Always-Taker", "Complier") ~ 1
  )
)

y_po_d <- declare_potential_outcomes(
  Y ~ tau * sd(u) * D + u, assignment_variables = "D" # c("D", "Z")
)

y_po_z <- declare_potential_outcomes(Y~(tau/2)*sd(u)*Z+u,assignment_variables=c("D","Z"))


## Treatment  assignment for  any given city is a simple fixed  proportion. It should be complete  or  urn-drawing assignment, no  t  simple or  coin-flipping assignment.
the_assign <- declare_assignment(Z=conduct_ra(N=N,m=N/2))
## the_assign <- declare_assignment(assignment_variable = "Z")

# thereveal <- declare_reveal(Y, Z)
#d_reveal <- declare_reveal(D, assignment_variable = "Z")
#y_reveal <- declare_reveal(Y, assignment_variables = c("D", "Z"))

d_reveal <- declare_measurement(D=reveal_outcomes(D~Z))
y_reveal <- declare_measurement(Y=reveal_outcomes(Y~D+Z))

base_design <- the_pop + d_po + y_po_d + y_po_z + the_assign +  d_reveal + y_reveal

dat0 <- draw_data(base_design)
```
```{asis echo=!handout_&solutions_}
...
```

Some example data (where we know all potential outcomes):

```{r showdat0}
tempdat <- dat0[1:2, -1]
names(tempdat)[5] <- "pZ"
names(tempdat) <- gsub("_", "", names(tempdat))
set.seed(123455)
sample_n(dat0,15)
#tempdat
#kableExtra::kable(tempdat, digits = 2)
```

## How to calculate the ITT and CACE/LATE II

The ITT and CACE (the parts)

```{r echo=TRUE}
itt_y <- difference_in_means(Y ~ Z, data = dat0)
itt_y
itt_d <- difference_in_means(D ~ Z, data = dat0)
itt_d
```

## How to calculate the ITT and CACE/LATE III

All together (the version dividing an unbiased estimator of ITT by an unbiased estimator of Proportion Compliers is often called Bloom's method from Bloom (1984)):^[User beware: this estimator is know *not* to be robust, particularly when faced with a "weak instrument" (weak $Z \rightarrow D$ relationship). See @bound1995piv, @imbensrosenbaum2005 and others. ("`robust`" is in reference to its SE, but it's quite questionable there too).]

```{r echo=TRUE}
cace_est <- estimatr::iv_robust(Y ~ D | Z, data = dat0)
cace_est
## Notice same as below:
coef(itt_y)[["Z"]] / coef(itt_d)[["Z"]]
```

## Variance of IV estimator

\begin{itemize}
\item Recall that there exist analytic expressions for $\Var\left[\widehat{\text{ITT}}_Y\right]$ and $\Var\left[\widehat{\text{ITT}}_D\right]$
\item  We can conservatively estimate $\Var\left[\widehat{\text{ITT}}_Y\right]$ and $\Var\left[\widehat{\text{ITT}}_D\right]$ via $\widehat{\Var}\left[\widehat{\text{ITT}}_Y\right]$ and $\widehat{\Var}\left[\widehat{\text{ITT}}_D\right]$
\item  However, in general, there is no closed-form analytic expression for the variance of a random ratio
\item  We do not have an estimator for $\Var\left[\cfrac{\widehat{\text{ITT}}_Y}{\widehat{\text{ITT}}_D}\right]$ that is known to be unbiased, consistent or conservative
\item  Bloom (1984) proposed treating $\widehat{\text{ITT}}_D$ as fixed
\item  Others use Delta method (Taylor series approximation), e.g., in \texttt{AER} or \texttt{estimatr} package in \texttt{R}
\end{itemize}

```{r child="child_files/noncompliance_attrition_estimator_performance.Rmd"}
```

## Summary of Encouragement/Complier/Dose oriented designs:

 - Analyze as you randomized: even when you don't control the dose you can
   learn something.
 - The danger of per-protocol analysis: you give up the benefits of the
   research design (i.e. randomization)
 - Variance calculations approximate (and can be untrustworthy in small
   samples, with weak instruments, and in other cases where we would worry
   about consistency (rare binary outcomes, very skewed outcomes,
   interdependence, \ldots) ).


# Hypothesis Tests about Complier causal effects

## Hypothesis Tests about Complier causal effects

\begin{itemize}
\item We can test the sharp null hypothesis no effect among all units
\item  We know by random assignment that this test
\begin{enumerate}
\item  will have a type I error probability at least as small as $\alpha$
\item  will have power greater than $\alpha$ for a class of alternative hypotheses
\end{enumerate}
\item  Under what conditions /assumptions is a test of the sharp null of no effect among all units equivalent to a test of the sharp null of no effect among Compliers?
\begin{enumerate}
\item  Exclusion restriction
\item  No Defiers
\item  Non-zero proportion of Compliers
\item Non-interference
\end{enumerate}
\end{itemize}


## Sharp null hypothesis testing example

The null hypothesis of no complier causal effect states that the individual causal effect of $\mathbf{Z}$ on $\mathbf{Y}$ is $0$ among units who are Compliers.

\medskip

Along with the exclusion restriction (i.e., that the individual causal effect is $0$ for Always Takers and Never Takers) and the assumption of no Defiers, we can "fill in" missing potential outcomes according to the null hypothesis of no complier causal effect as follows:
\begin{align*}
Y_{c,0,i} & =
\begin{cases}
Y_i - D_i \tau_i & \text{if } D_i = 1 \\
Y_i + \left(1 - D_i\right) \tau_i & \text{if } D_i = 0
\end{cases} \\
Y_{t,0,i} & =
\begin{cases}
Y_i - D_i \tau_i & \text{if } D_i = 1 \\
Y_i + \left(1 - D_i\right) \tau_i & \text{if } D_i = 0,
\end{cases}
\end{align*}
where $\tau_i = 0$ for all $i$.



## Sharp null hypothesis testing example

Imagine that our observed data is as follows:
\begin{table}[H]
\centering
    \begin{tabular}{l|l|l|l|l|l|l}
    $\mathbf{z}$ & $\mathbf{y}$ & $\mathbf{y_c}$ & $\mathbf{y_t}$ & $\mathbf{d}$ & $\mathbf{d_c}$ & $\mathbf{d_t}$ \\ \hline
    1 & 14 & ? & 14 & 0 & ? & 0 \\
    0 & 22 & 22 & ? & 0 & 0 & ? \\
    1 & 21 & ? & 21 & 1 & ? & 1 \\
    1 & 36 & ? & 36 & 1 & ? & 1 \\
    0 & 23 & 23 & ? & 0 & 0 & ? \\
    0 & 12 & 12 & ? & 1 & 1 & ? \\
    0 & 25 & 25 & ? & 1 & 1 & ? \\
    1 & 27 & ?  & 27 & 0 & ? & 0\\
    \end{tabular}
    \caption{Observed experimental data}
\end{table}

The observed Difference-in-Means test statistic, $\hat{\bar{\tau}}\left(\mathbf{Z}, \mathbf{Y}\right)$, is $16.75$. What is the distribution of that test statistic under the null hypothesis of no effects for any complier?

## Sharp null hypothesis testing example

We can represent the sharp null hypothesis of no effect for all units without
hypothesizing about non-random compliance (this is like the ITT$_Y$ in that
both can be assessed safely in a randomized experiment).

\begin{table}[H]
\centering
    \begin{tabular}{l|l|l|l|l|l|l|l}
    $\mathbf{z}$ & $\mathbf{y}$ & $\mathbf{y_c}$ & $\mathbf{y_t}$ & $\mathbf{d}$ & $\mathbf{d_c}$ & $\mathbf{d_t}$ & Principal stratum\\ \hline
    1 & 14 & ? & 14 & 0 & ? & 0 & Never Taker or Defier\\
    0 & 22 & 22 & ? & 0 & 0 & ? & Complier or Never Taker\\
    1 & 21 & ? & 21 & 1 & ? & 1 & Complier or Always Taker \\
    1 & 36 & ? & 36 & 1 & ? & 1 & Complier or Always Taker \\
    0 & 23 & 23 & ? & 0 & 0 & ? & Complier or Never Taker \\
    0 & 12 & 12 & ? & 1 & 1 & ? & Always Taker or Defier \\
    0 & 25 & 25 & ? & 1 & 1 & ? & Always Taker or Defier\\
    1 & 27 & ?  & 27 & 0 & ? & 0 & Never Taker or Defier \\
    \end{tabular}
    \caption{Sharp null of no effect for all units}
\end{table}

## Sharp null hypothesis testing example

We can represent the sharp null hypothesis of no effect for all units without
hypothesizing about non-random compliance (this is like the ITT$_Y$ in that
both can be assessed safely in a randomized experiment).

\begin{table}[H]
\centering
    \begin{tabular}{l|l|l|l|l|l|l|l}
    $\mathbf{z}$ & $\mathbf{y}$ & $\mathbf{y_c}$ & $\mathbf{y_t}$ & $\mathbf{d}$ & $\mathbf{d_c}$ & $\mathbf{d_t}$ & Principal stratum\\ \hline
    1 & 14 & 14 & 14 & 0 & ? & 0 & Never Taker or Defier\\
    0 & 22 & 22 & 22 & 0 & 0 & ? & Complier or Never Taker\\
    1 & 21 & 21 & 21 & 1 & ? & 1 & Complier or Always Taker \\
    1 & 36 & 36 & 36 & 1 & ? & 1 & Complier or Always Taker \\
    0 & 23 & 23 & 23 & 0 & 0 & ? & Complier or Never Taker \\
    0 & 12 & 12 & 12 & 1 & 1 & ? & Always Taker or Defier \\
    0 & 25 & 25 & 25 & 1 & 1 & ? & Always Taker or Defier\\
    1 & 27 & 27  & 27 & 0 & ? & 0 & Never Taker or Defier \\
    \end{tabular}
    \caption{Sharp null of no effect for all units}
\end{table}

## Sharp null hypothesis testing example


The null hypothesis of no effect among compliers under excludability (only a
complier in the treatment group can have a causal effect), no defiers and
nonzero proportion of compliers assumptions:

\begin{table}[H]
\centering
    \begin{tabular}{l|l|l|l|l|l|l|l}
    $\mathbf{z}$ & $\mathbf{y}$ & $\mathbf{y_c}$ & $\mathbf{y_t}$ & $\mathbf{d}$ & $\mathbf{d_c}$ & $\mathbf{d_t}$ & \text{Principal stratum}\\ \hline
    1 & 14 & 14 & 14 & 0 & 0 & 0 & Never Taker \sout{or Defier}\\
    0 & 22 & 22 & 22 & 0 & 0 & ? & Complier or Never Taker \\
    1 & 21 & 21 & 21 & 1 & ? & 1 & Complier or Always Taker \\
    1 & 36 & 36 & 36 & 1 & ? & 1 & Complier or Always Taker \\
    0 & 23 & 23 & 23 & 0 & 0 & ? & Complier or Never Taker\\
    0 & 12 & 12 & 12 & 1 & 1 & 1 & Always Taker \sout{or Defier} \\
    0 & 25 & 25 & 25 & 1 & 1 & 1 & Always Taker \sout{or Defier} \\
    1 & 27 & 27  & 27 & 0 & 0 & 0 & Never Taker \sout{or Defier} \\
    \end{tabular}
    \caption{Sharp null of no effect among Compliers}
    \label{tab: pot outs under null}
\end{table}

We don't need to know which of units 2  -- 5 are Compliers, only that at least one of these $4$ units is a Complier.

Excludability means that the effect must be 0 for all units who are not compliers (i.e. implying the sharp null).


## Sharp null hypothesis testing example


The null hypothesis of no effect among compliers under excludability (meaning
that only a complier in the treatment group can have a causal effect), no defiers
and nonzero proportion of Compliers assumptions:

\begin{table}[H]
\centering
    \begin{tabular}{l|l|l|l|l|l|l|l}
    $\mathbf{z}$ & $\mathbf{y}$ & $\mathbf{y_c}$ & $\mathbf{y_t}$ & $\mathbf{d}$ & $\mathbf{d_c}$ & $\mathbf{d_t}$ & \text{Principal stratum}\\ \hline
    1 & 14 & 14 & 14 & 0 & 0 & 0 & Never Taker \sout{or Defier}\\
    0 & 22 & 22 & 22 & 0 & 0 & ? & Complier or Never Taker \\
    1 & 21 & 21 & 21 & 1 & ? & 1 & Complier or Always Taker \\
    1 & 36 & 36 & 36 & 1 & ? & 1 & Complier or Always Taker \\
    0 & 23 & 23 & 23 & 0 & 0 & ? & Complier or Never Taker\\
    0 & 12 & 12 & 12 & 1 & 1 & 1 & Always Taker \sout{or Defier} \\
    0 & 25 & 25 & 25 & 1 & 1 & 1 & Always Taker \sout{or Defier} \\
    1 & 27 & 27  & 27 & 0 & 0 & 0 & Never Taker \sout{or Defier} \\
    \end{tabular}
    \caption{Sharp null of no effect among Compliers}
    \label{tab: pot outs under null 2}
\end{table}


So: a regular test of the sharp null of no effects **is also a test of the
sharp null of no effects among compliers** (under the assumptions of no
defiers, non-zero compliers, exclusion, and no interference). The fact that
$\tau_i=0$ for Never Takers and Always Takers is by assumption, not a
hypothesis.


```{r codeforabove, echo=FALSE, results="hide"}
n <- 8
n_1 <- 4

set.seed(1:5)
d_c <- rbinom(n = n, size = 1, prob = 0.3)
d_t <- rep(x = NA, times = length(d_c))
## HERE WE SATISFY THE AT LEAST ONE COMPLIER (NON-WEAK INSTRUMENT) ASSUMPTION
d_t[which(d_c != 1)] <- rbinom(n = length(which(d_c != 1)), size = 1, prob = 0.6)
## HERE WE SATISFY THE NO DEFIERS (MONOTONICITY) ASSUMPTION
d_t[which(d_c == 1)] <- rep(x = 1, times = length(which(d_c == 1)))
cbind(d_c, d_t)
prop_comp <- length(which(d_c == 0 & d_t == 1)) / n
prop_def <- length(which(d_c == 1 & d_t == 0)) / n
prop_at <- length(which(d_c == 1 & d_t == 1)) / n
prop_nt <- length(which(d_c == 0 & d_t == 0)) / n

## HERE WE SATISFY THE EXCLUSION RESTRICTION ASSUMPTION BY LETTING y_c = y_t FOR ALL ALWAYS-TAKERS AND NEVER-TAKERS AND
## WE ALSO SATISFY THE SUTVA ASSUMPTION BY LETTING ALL UNITS HAVE ONLY TWO POT OUTS
set.seed(1:5)
y_c <- round(x = rnorm(n = 8, mean = 20, sd = 10), digits = 0)
y_t_null_false <- rep(x = NA, times = n)
y_t_null_false[which(d_c == 0 & d_t == 1)] <- y_c[which(d_c == 0 & d_t == 1)] +
  round(x = rnorm(n = length(which(d_c == 0 & d_t == 1)),
                  mean = 10,
                  sd = 4),
        digits = 0)
y_t_null_false[!(d_c == 0 & d_t == 1)] <- y_c[!(d_c == 0 & d_t == 1)]
cbind(y_c, y_t_null_false)

true_data <- data.frame(y_t = y_t_null_false,
                        y_c = y_c,
                        d_t = d_t,
                        d_c = d_c,
                        tau = y_t_null_false - y_c)

true_data <- true_data %>% mutate(type = NA,
                      type = ifelse(test = d_c == 0 & d_t == 0, yes = "never_taker", no = type),
                      type = ifelse(test = d_c == 0 & d_t == 1, yes = "complier", no = type),
                      type = ifelse(test = d_c == 1 & d_t == 0, yes = "defier", no = type),
                      type = ifelse(test = d_c == 1 & d_t == 1, yes = "always_taker", no = type))
kable(true_data)

Omega <- apply(X = combn(x = 1:n,
                         m = n_1),
               MARGIN = 2,
               FUN = function(x) { as.integer(1:n %in% x) })

assign_vec_probs <- rep(x = (1/70), times = ncol(Omega))

set.seed(1:5)
obs_z <- Omega[,sample(x = 1:ncol(Omega), size = 1)]

#obs_ys <- apply(X = Omega, MARGIN = 2, FUN = function(x) { x * y_t_null_false + (1 - x) * y_c })

#obs_ds <- apply(X = Omega, MARGIN = 2, FUN = function(x) { x * d_t + (1 - x) * d_c })

obs_y <- obs_z * true_data$y_t + (1 - obs_z) * true_data$y_c
obs_d <- obs_z * true_data$d_t + (1 - obs_z) * true_data$d_c

cbind(obs_z, obs_d, obs_y)

tau <- 0

null_y_c <- obs_y - obs_d * tau
null_y_t <- obs_y + (1 - obs_d) * tau

obs_diff_means <- as.numeric((t(obs_z) %*% obs_y) / (t(obs_z) %*% obs_z) -
                               (t(1 - obs_z) %*% obs_y) / (t(1 - obs_z) %*% (1 - obs_z)))

coef(lm(formula = obs_y ~ obs_z))[["obs_z"]]

obs_null_pot_outs <- sapply(X = 1:ncol(Omega),
                            FUN = function(x) { Omega[,x] * null_y_t + (1 - Omega[,x]) * null_y_c })

null_test_stat_dist <- sapply(X = 1:ncol(Omega),
                              FUN = function(x) { mean(obs_null_pot_outs[,x][which(Omega[,x] == 1)]) -
                                  mean(obs_null_pot_outs[,x][which(Omega[,x] == 0)])})

null_test_stats_data <- data.frame(null_test_stat = null_test_stat_dist,
                                   prob = assign_vec_probs)
library(ggplot2)
null_dist_plot <- ggplot(data = null_test_stats_data,
                         mapping = aes(x = null_test_stat,
                                       y = prob)) +
  geom_bar(stat = "identity") +
  geom_vline(xintercept = obs_diff_means,
             color = "red",
             linetype = "dashed") +
  xlab(label = "Null Test Statistics") +
  ylab(label = "Probability")

ggsave(plot = null_dist_plot,
       file = here("images","null_dist_plot_iv.pdf"),
       width = 6,
       height = 4,
       units = "in",
       dpi = 600)

```

## Sharp null hypothesis testing example

\begin{figure}[H]
\includegraphics[width = \linewidth]{null_dist_plot_iv.pdf}
\caption{Distribution of the Difference-in-Means test statistic under the sharp null of no effect: under the assumptions of excludability (no effects on Always Takers and Never Takers), no defiers, at least one complier, and SUTVA, this is a test of the hypothesis of no effects on compliers.}
\end{figure}

## Summary

 - The sharp null of no effects is meaningful and can be tested in a randomized
   experiment using assignment to treatment and ignoring compliance.
 - The assumptions of excludability, no defiers, and at least one complier mean
   that we can interpret the test of the sharp null of no effects as a test of
   the sharp null of no effects on compliers: those assumptions require no
   effects among always-takers and never-takers, and there are no defiers in
   the data (again, all of this by assumption).
 - Notice: no need for approximations; weak instruments do not threaten the
   validity of the statistical inferences.

# Learning about causal effects when data are missing

## Review of core assumptions from randomized experiments

  1. Excludability: Potential outcomes depend only on assigned treatment (and not other factors)

  2. Non-interference

  3. Random assignment of treatment

## Attrition (missing data on outcomes)

- Some units may have missing data on outcomes (= units attrit) when:

  - some respondents can't be found or refuse to participate in endline data collection.

  - some records are lost.

- This is a problem when treatment affects missingness.

  - For example, units in control may be less willing to answer survey questions.
  - For example, treatment may have caused units to migrate and cannot be reached

- If we analyze the data by dropping units with missing outcomes, then we are no longer comparing similar treatment and control groups. (We have trouble analyzing as we randomized!)

- Dropping the missing observations brings us closer to per-protocol analysis and confounding.

## What can we do?

- Check whether attrition rates are similar in treatment and control groups.

- Check whether treatment and control groups have similar covariate profiles.

- Do not drop observations that are missing outcome data from your analysis.

- Analyze missingness on outcome as another outcome: could treatment have caused missing outcomes?

- When outcome data are missing we can sometimes **bound** our estimates of treatment effects.

## What can we do?

- But the best approach is to try to anticipate and prevent attrition.

   - Blind people to their treatment status.

   - Promise to deliver the treatment to the control group after the research is completed.

   - Plan ex ante to reach all subjects at endline.

   - Budget for intensive follow-up with a random sample of attriters [@gerbergreen2012] Chapter 7.

See also [Chapter on Threats to Internal Validity in Experiments](https://egap.github.io/theory_and_practice_of_field_experiments/threats-to-the-internal-validity-of-randomized-experiments.html)

## Missing data on covariates is not as problematic

- Missing **background covariates** (i.e.,variables for which values do not change as a result of treatment) for some observations is less problematic.

  - We can still learn about the causal effect of an experiment without those covariates.

  - We can also use the background covariate as planned by imputing for the missing values.

  - We can also condition on that missingness directly: we could assess causal effects for the subgroup of those missing on income and compare it to the subgroup of those not missing on income.


## Summary about Missing data and Experiments.

 - Missing outcomes or missing treatment assignment (or missing blocking
   information) are all big problems. How might those with missing outcomes
   have behaved in treatment versus control? We don't know.

 - Missing covariate information is not a problem: it is fixed, same proportion
   should be missing covariate information in both treatment and control
   conditions


## References

