---
title: |
  | Taking Stock \&
  | Linear Regression in Experiments and Observational Studies
date: '`r format(Sys.Date(), "%B %d, %Y")`'
author: |
  | ICPSR 2020 Session 2
  | Jake Bowers, Ben Hansen, Tom Leavitt
bibliography:
 - BIB/refs.bib
 - BIB/master.bib
 - BIB/misc.bib
fontsize: 10pt
geometry: margin=1in
graphics: yes
biblio-style: authoryear-comp
output:
  beamer_presentation:
    slide_level: 2
    keep_tex: true
    latex_engine: lualatex
    citation_package: biblatex
    template: icpsr.beamer
    incremental: true
    includes:
        in_header:
           - defs-all.sty
    md_extensions: +raw_attribute-tex_math_single_backslash+autolink_bare_uris+ascii_identifiers+tex_math_dollars
---


<!-- To show notes  -->
<!-- https://stackoverflow.com/questions/44906264/add-speaker-notes-to-beamer-presentations-using-rmarkdown -->

```{r setup1_env, echo=FALSE, include=FALSE}
library(here)
source(here::here("rmd_setup.R"))
```

```{r setup2_loadlibs, echo=FALSE, include=FALSE}
## We are using the renv system. So these will all be loaded from a local library directory
library(dplyr)
library(ggplot2)
library(estimatr)
library(coin)
```

## Today

 1. Agenda: Introductions, Overview and context for where we are in the
course, Statistical Inference and Causal Inference, Linear regresssion for estimation in randomized experiments, Linear regression for covariance adjustment in randomized experiments, Linear regression for covariance adjustment in observational studies ("controlling for").
 2. Random break-out rooms to work on the code for the day and/or the assignments.
 3. Questions arising from the reading or assignments or life?


# An overview of approaches to statistical inference for causal quantities


## Design Based Approach 1: Test Models of Potential Outcomes

 1. Make a guess (or model of) about $\tau_i$.
 2. Measure consistency of data with this model given the design.

\centering
  \includegraphics[width=.7\textwidth]{images/cartoonFisherNew1.pdf}

## Design Based Approach 1: Test Models of Potential Outcomes

\centering
  \includegraphics[width=\textwidth]{images/cartoonFisherNew1.pdf}

## Design Based Approach 1: Test Models of Potential Outcomes

\centering
  \includegraphics[width=\textwidth]{images/cartoonFisherNew2.pdf}

## Design Based Approach 1: Test Models of Potential Outcomes

  \centering
  \includegraphics[width=.9\textwidth]{images/cartoonFisher.pdf}

##  Design Based Approach 1: Test Models of Potential Outcomes

```{r}
smdat <- data.frame(Z = c(0, 1, 0, 1), Y = c(16, 22, 7, 4000))

tz_mean_diff <- function(z, y) {
  mean(y[z == 1]) - mean(y[z == 0])
}

tz_mean_rank_diff <- function(z, y) {
  ry <- rank(y)
  mean(ry[z == 1]) - mean(ry[z == 0])
}

newexp <- function(z) {
  sample(z)
}
```

##  Design Based Approach 1: Test Models of Potential Outcomes


```{r repexp, cache=TRUE}
set.seed(12345)
rand_dist_md <- replicate(1000, tz_mean_diff(z = newexp(smdat$Z), y = smdat$Y))
rand_dist_rank_md <- replicate(1000, tz_mean_rank_diff(z = newexp(smdat$Z), y = smdat$Y))
```

```{r calc}
obs_md <- tz_mean_diff(z = smdat$Z, y = smdat$Y)
obs_rank_md <- tz_mean_rank_diff(z = smdat$Z, y = smdat$Y)
c(obs_md, obs_rank_md)
table(rand_dist_md)
table(rand_dist_rank_md)
p_md <- mean(rand_dist_md >= obs_md)
p_rank_md <- mean(rand_dist_rank_md >= obs_rank_md)
c(p_md, p_rank_md)
```

##  Design Based Approach 1: Test Models of Potential Outcomes

```{r}
smdat$zF <- factor(smdat$Z)
smdat$rY <- rank(smdat$Y)

md_test_exact <- oneway_test(Y ~ zF, data = smdat, distribution = exact(), alternative = "less")
md_test_sim <- oneway_test(Y ~ zF,
  data = smdat, distribution = approximate(nresample = 1000),
  alternative = "less"
)
md_test_asymp <- oneway_test(Y ~ zF, data = smdat, distribution = asymptotic(), alternative = "less")

rank_md_test_exact <- oneway_test(rY ~ zF, data = smdat, distribution = exact(), alternative = "less")
rank_md_test_asymp <- oneway_test(rY ~ zF, data = smdat, distribution = asymptotic(), alternative = "less")
rank_md_test_sim <- oneway_test(rY ~ zF,
  data = smdat, distribution = approximate(nresample = 1000),
  alternative = "less"
)
```

##  Design Based Approach 1: Test Models of Potential Outcomes

```{r}
pvalue(md_test_exact)
pvalue(md_test_sim)
pvalue(md_test_asymp)
pvalue(rank_md_test_exact)
pvalue(rank_md_test_sim)
pvalue(rank_md_test_asymp)
```
##  Design Based Approach 1: Test Models of Potential Outcomes

```{r}
s_rank_md_exact <- support(rank_md_test_exact)
d_rank_md_exact <- dperm(rank_md_test_exact, s_rank_md_exact)
d_rank_md_asymp <- dperm(rank_md_test_asymp, seq(-1.6, 1.6, .1))

s_md_exact <- support(md_test_exact)
d_md_exact <- dperm(md_test_exact, s_md_exact)
d_md_asymp <- dperm(md_test_asymp, seq(-1.6, 1.6, .1))
```

```{r plotexactdists, echo=FALSE}
par(mfrow = c(1, 2))
plot(s_rank_md_exact, d_rank_md_exact, type = "h", xlim = c(-1.6, 1.6), ylim = c(0, .4), lwd = 3)
lines(seq(-1.6, 1.6, .1), d_rank_md_asymp)
plot(s_md_exact, d_md_exact, type = "h", xlim = c(-1.6, 1.6), ylim = c(0, .4), lwd = 3)
lines(seq(-1.6, 1.6, .1), d_md_asymp)
```

## Design Based Approach 2: Estimate Averages of Potential Outcomes

  1. Notice that the observed $Y_i$ are a sample from  the (small, finite) population of $(y_{i,1},y_{i,0})$.
  2. Decide to focus on the average, $\bar{\tau}$, because sample averages, $\hat{\bar{\tau}}$ are unbiased and consistent estimators of population averages.
  3. Estimate $\bar{\tau}$ with the observed difference in means.

\centering
  \includegraphics[width=.5\textwidth]{images/cartoonNeyman.pdf}

## Design Based Approach 2: Estimate Averages of Potential Outcomes

\centering
  \includegraphics[width=.9\textwidth]{images/cartoonNeyman.pdf}


## Design Based Approach 2: Estimate Averages of Potential Outcomes

```{r}

est1 <- difference_in_means(Y ~ Z, data = smdat)
est1
```

## Design Based Approach 2: Estimate Averages of Potential Outcomes

Is this estimator a good guess about the unobserved causal effect? Let's simulate to learn.

First, make up a simulated research design and underlying potential outcomes:

```{r setupdd}
library(DeclareDesign)

set.seed(12345)
smdat$x <- runif(4, min = 0, max = 1)
smdat$e0 <- round(runif(4, min = min(smdat$Y[smdat$Z == 0]), max = max(smdat$Y[smdat$Z == 0])))
smdat$y0 <- with(smdat, 3 * sd(e0) * x + e0)
smdat$y1 <- smdat$Y - smdat$y0

thepop <- declare_population(smdat[, c("y0", "y1", "x")])
theassign <- declare_assignment(m = 2)
po_fun <- function(data) {
  data$Y_Z_1 <- data$y0
  data$Y_Z_0 <- data$y1
  data
}
thepo <- declare_potential_outcomes(handler = po_fun)
thereveal <- declare_reveal(Y, Z) ## how does assignment reveal potential outcomes
thedesign <- thepop + theassign + thepo + thereveal

oneexp <- draw_data(thedesign)
```

## Design Based Approach 2: Estimate Averages of Potential Outcomes

Next specify our estimator and compare to a couple of others:

```{r}
theestimand <- declare_estimand(ATE = mean(Y_Z_1 - Y_Z_0))
theest1 <- declare_estimator(Y ~ Z,
  estimand = theestimand, model = difference_in_means,
  label = "est1: diff means1"
)
theest2 <- declare_estimator(Y ~ Z,
  estimand = theestimand, model = lm_robust,
  label = "est2: diff means2"
)
theest3 <- declare_estimator(Y ~ Z + x,
  estimand = theestimand, model = lm_robust,
  label = "est3: covadj diff means"
)

thedesign_plus_est <- thedesign + theestimand + theest1 + theest2 + theest3
theest1(oneexp)
theest2(oneexp)
theest3(oneexp)
```

## Design Based Approach 2: Estimate Averages of Potential Outcomes

Finally, repeat the design, applying our estimators and comparing to our known potential outcomes.

```{r diagnose1, warnings=FALSE, cache=TRUE}
require(DeclareDesign)
set.seed(12345)

thediagnosands <- declare_diagnosands(
  bias = mean(estimate - estimand),
  rmse = sqrt(mean((estimate - estimand)^2)),
  power = mean(p.value < .25),
  coverage = mean(estimand <= conf.high & estimand >= conf.low),
  mean_estimate = mean(estimate),
  sd_estimate = sd(estimate),
  mean_se = mean(std.error),
  mean_estimand = mean(estimand)
)

diagnosis <- diagnose_design(thedesign_plus_est,
  sims = 1000, bootstrap_sims = 0,
  diagnosands = thediagnosands
)
```


## Design Based Approach 2: Estimate Averages of Potential Outcomes

```{r}
kable(reshape_diagnosis(diagnosis)[, -c(1:2, 4)])
```

```{r simmethod, eval=FALSE, echo=FALSE}
thedesign_sims <- simulate_design(thedesign_plus_est, sims = 1000)
res <- thedesign_sims %>%
  group_by(estimator_label) %>%
  summarize(bias = mean(estimate - estimand))
```



## Model Based Approach 1: Predict Potential Outcomes}

  \smallskip
  \centering
  \includegraphics[width=.9\textwidth]{images/cartoonBayes.pdf}

## Statistical Inference and Causal Inference?

How does statistical inference help us make causal inferences?

# And third, some more context: The creation of interpretable comparisons.

## Research designs to aid the creation of interpretable comparisons

   - Randomized experiments (randomized $Z$ changes $y$) <!-- (more precision from reducing heterogeneity in $Y$) -->
     - No systematic differences between groups
     - Known reference distribution for tests of the sharp null and, in large
       samples, the weak null.
   - Instrumental variables (randomized $Z$ changes $D$ which changes $y$)
   - Natural Experiments / Discontinuities (one $X$ creates $Z$ which \ldots)
   - Semi-/Non-parametric Covariance Adjustment ($Z$ out of our control, but we can choose how to organize and observe given $X$) (matching and difference-in-differences)
   - Parametric covariance adjustment (same as above, but with a model for statistical adjustment, i.e. "controlling for")


# Now: Instruments, the ITT and CACE/LATE

## Instruments, the ITT and CACE/LATE

**Context:** An NGO invites you to help design and evaluate a program to increase donations involving in-person, door-to-door, discussions about the canvasser's own personal experiences. Building rapport with another person should increase willingness to donate, according to the NGO who had just read @broockman2016durably.

\smallskip

**Opportunity for learning about theory** You know have been interested in
understanding why and how social norms influence behavior. So, you ask to
that canvassers appeal to a descriptive social norm ("other people we have
spoken with agree") toward the end of the discussion.

\smallskip

**Design:** The NGO randomizes visits by canvassers to a set of addresses and
measures the amount pledged online in the following month.

\smallskip

**Analysis:** The NGO cares about the cost effectiveness of the approach. Did
it work? You (and the NGO should care) about **why** it worked (or did not
work).


## Defining causal effects

 - $Z_i$ is random assignment to a visit ($Z_i=1$) or not ($Z_i=0$).
 - $d_{i,Z_i=1}=1$ means that person $i$ opened the door to have a conversation when assigned a visit. Opening the door is an outcome of the treatment, is caused by the treatment.
 - $y_{i,Z_i = 1, d_{i,Z_i=1}=1}$ is the potential outcome for people who were assigned a visit and who opened the door.
 - $y_{i,Z_i = 1, d_{i,Z_i=1}=0}$ is the potential outcome for people who were assigned a visit and who did not open the door.
 - $y_{i,Z_i = 0, d_{i,Z_i=0}=1}$ is the potential outcome for people who were not assigned a visit and who opened the door.
 - $y_{i,Z_i = 0, d_{i,Z_i=0}=0}$ is the potential outcome for people who were not assigned a visit and who would not have opened the door.
 - We could also write $y_{i,Z_i = 0, d_{i,Z_i=1}=1}$ for people who were not
   assigned a visit who would have opened the door had they been assigned a
   visit etc..

In this case, $y_{i,Z_i = 0, d_{i,Z_i=1}=1} = y_{i,Z_i = 0, d_{i,Z_i=1}=0}
\equiv y_{i,Z_i=0}$ because you can't open the door unless visited.

And also $y_{i,Z_i = 1, d_{i,Z_i=1}=0} = y_{i,Z_i=0}$ because there can be no
effect of the visit if you don't open the door.

##  Defining our causal effects

People refer to an "Intent To Treat Effect" (ITT) as the (average) causal effect of
**assignment** to treatment. In our case, we can write the ITT as a weighted
sum of the responses to treatment  minus the response to assignment to control
( where $p(d_1=1)$ is the proportion of people opening their doors ).


$$ITT= ( \bar{y}_{Z=1,d_1=1} \cdot p(d_1=1) + \bar{y}_{Z=1,d_1=0} \cdot (1-p(d_1=1)) )  - \bar{y}_{Z=0}$$

\medskip

Notice that if everyone opens the door when visited then $y_{i,Z=1,d_1=1} =
y_{i,Z=1}$  so the ITT=ATE.

\medskip

Why give it a new name?

## A simulation to understand the design and analysis

Setting up the simulation and analysis before going into the field.

```{r}
library(randomizr)
set.seed(20200730)
n <- 1000
## Make a covariate, like age
xtmp <- rnorm(n, mean = 45, sd = 10)
x <- pmax(xtmp, 18)
## Make potential outcome to control
y0tmp <- round(rnorm(n, mean = 0, sd = 10)) + x / 2
y0 <- pmax(y0tmp, 0)
Z <- complete_ra(n)
## Door opening is a function of x (age) and Z (visits)
probd <- (x - min(x)) / (max(x) - min(x))
mean(probd)
d <- ifelse(Z == 1, rbinom(sum(Z == 1), prob = probd / 2, size = 1), 0)
table(Z, d, exclude = c())
mean(d[Z == 1])
```

```{r}
## No effects possible if not visited
y0d0 <- y0
y0d1 <- y0
y1d0 <- y0
taubar <- .25 * sd(y0) ## Effect of .25 sds on compliers/door openers
## The treatment also changes variance of y1d1
y1d1 <- mean(y0) + taubar + (y0 - mean(y0)) / 2
## Define observed outcome
Y <- Z * d * y1d1 + Z * (1 - d) * y1d0 + (1 - Z) * d * y0d1 + (1 - Z) * (1 - d) * y0d0
dat <- data.frame(y1d1, y1d0, y0d1, y0d0, y0, Y, Z, d, x, probd)
dat$y1 <- (1 - Z) * y0 + Z * d * y1d1 + Z * (1 - d) * y1d0
```

## A simulation to understand the design and analysis


```{r}
boxplot(Y ~ Z * d, data = dat)
```


## Learning about the ITT

First, let's learn about the effect of the policy itself, about the ITT.

```{r}
## True ITT with one-sided non-compliance
pd <- mean(dat$d[dat$Z == 1])

bary1 <- mean(dat$y1d1) * pd + mean(dat$y1d0) * (1 - pd)
bary0 <- mean(dat$y0)

trueITT <- bary1 - bary0
```


\smallskip

What evidence do we have against the claim that the policy had no effects? How
would we assess whether this testing procedure is trustworthy --- controls its
false positive error rate?

\smallskip

What is our best guess about the ITT itself? How would we assess whether or
not this estimator is biased or not?

\smallskip

What is our best guess about how our estimate of the ITT would vary from
experiment to experiment?


## Some code notes

```{r, eval=FALSE}
newExp <- function(trt) {
  sample(trt)
}
testStat <- function(outcome, trt) {
  mean(outcome[trt == 1]) - mean(outcome[trt == 0])
}
obsTestStat <- testStat(outcome = dat$Y, trt = dat$Z)
nullDist <- replicate(1000, testStat(outcome = dat$Y, trt = newExp(dat$Z)))
upperP <- mean(nullDist >= obsTestStat)
```


```{r, eval=FALSE}
## taubarhat <- with(dat, mean(Y[Z==1]) - mean(Y[Z==0]))
## or taubarhat <- coef(lm(Y~Z))[["Z"]]
library(estimatr)
taubarhat <- difference_in_means(Y ~ Z, data = dat)
## sqrt( var(Y[Z==1])/sum(Z==1) + var(Y[Z==0])/sum(Z==0) )
```


## The Complier Average Causal Effect or Local Average Treatment Effect (on the Compliers)

Now we would like to learn about the causal effect of answering the door and having the
conversation, the theoretically interesting effect. But, this comparison is confounded
by $x$ (tells us about differences in the outcome due to $x$ in addition to
the difference caused by $D$).

```{r}
with(dat, cor(Y, x))
with(dat, cor(d, x))
with(dat, cor(Z, x)) ## should be near 0
```


## Identifying the causal effect of opening the door

We want to learn about the Complier Average Causal Effect (aka the Local
Average Treatment Effect): CACE$= \bar{y_{i,Z=1,d_1=1}} -
\bar{y_{i,Z=0,d_1=1}}$


Let $p$ be the proportion of compliers ($\sum_{i=1}^N I(d_{i,1} = 1)/N$).

\begin{align*}
ITT & = \left( \bar{y}_{1,d_1=1}p + \bar{y}_{1,d_1=0}(1-p) \right)  - \bar{y}_{0} \\
    & = ( \bar{y}_{1,d_1=1}p + \bar{y}_{1,d_1=0}(1-p) )  - ( \bar{y}_{0,d_0=1}p+ \bar{y}_{0,d_0=0}(1-p) )  \\
    & = ( \bar{y}_{1,d_1=1}p - \bar{y}_{0,d_0=1}p )  +  ( \bar{y}_{1,d_1=0}(1-p)  - \bar{y}_{0,d_0=0}(1-p) )  \\
    & = ( \bar{y}_{1,d_1=1} - \bar{y}_{0,d_0=1} )p  +  ( \bar{y}_{1,d_1=0}  - \bar{y}_{0,d_0=0}) (1-p)  \\
\intertext{but $\bar{y}_{1,d_1=0} = \bar{y}_{0,d_1=0} = \bar{y}_{0,d_0=0}$ so }
  ITT  & = ( \bar{y}_{1,d_1=1} - \bar{y}_{0,d_0=1} ) p \\
  ITT  & = ( CACE ) p \\
\intertext{so we can write the CACE as ITT/p}
  ITT/p & = CACE
\end{align*}

This means that we can write the CACE as $CACE = ITT/\text{proportion of compliers}$

## Code snippets

```{r}
ITT <- with(dat, mean(Y[Z == 1]) - mean(Y[Z == 0]))
pd <- with(dat, mean(d[Z == 1]) - mean(d[Z == 0]))
ITT / pd

library(AER)
iv1 <- ivreg(Y ~ d | Z, data = dat)
```



## Summary of the Day

  - More than one way to use what we observe to reason about potential
    outcomes: testing, estimating, predicting
  - Structure of the course. Where we are now. Where we are going: away from
    randomized experiments toward observational studies, but with the statistical basis
    developed from our understanding of randomized experiments.
  - An instrument moves a dose. The effect of the dose is specific, or local,
    to those who took a dose, to those who experienced the treatment.
  - The ITT and CACE/LATE represent different causal effects. ITT tends to be
    the policy-relevant effect.
  - An "as-treated" analysis is no longer an experiment.

## References

