---
title: Statistical Inference for Causal Quantities when the Active Ingredient is not Random
date: '`r format(Sys.Date(), "%B %d, %Y")`'
author: ICPSR 2018 Session 2
bibliography:
 - refs.bib
 - BIB/master.bib
 - BIB/misc.bib
fontsize: 10pt
geometry: margin=1in
graphics: yes
biblio-style: authoryear-comp
output:
  beamer_presentation:
    slide_level: 2
    keep_tex: true
    latex_engine: xelatex
    citation_package: biblatex
    template: icpsr.beamer
    incremental: true
    includes:
        in_header:
           - defs-all.sty
---


<!-- Make this document using library(rmarkdown); render("day12.Rmd") -->


```{r include=FALSE, cache=FALSE}
# Some customization.  You can alter or delete as desired (if you know what you are doing).
# knitr settings to control how R chunks work.
rm(list=ls())

require(knitr)

## This plus size="\\scriptsize" from https://stackoverflow.com/questions/26372138/beamer-presentation-rstudio-change-font-size-for-chunk

knitr::knit_hooks$set(mysize = function(before, options, envir) {
  if (before)
    return(options$size)
})

knit_hooks$set(plotdefault = function(before, options, envir) {
    if (before) par(mar = c(3, 3, .1, .1),oma=rep(0,4),mgp=c(1.5,.5,0))
})

opts_chunk$set(
  tidy=FALSE,     # display code as typed
  echo=TRUE,
  results='markup',
  strip.white=TRUE,
  fig.path='figs/fig',
  cache=FALSE,
  highlight=TRUE,
  width.cutoff=132,
  size='\\scriptsize',
  out.width='.8\\textwidth',
  fig.retina=FALSE,
  message=FALSE,
  comment=NA,
  mysize=TRUE,
  plotdefault=TRUE)

if(!file.exists('figs')) dir.create('figs')

options(digits=4,
	scipen=8,
	width=132
	)
```

```{r eval=FALSE, include=FALSE, echo=FALSE}
## Run this only once and then not again until we want a new version from github
library('devtools')
library('withr')
with_libpaths('./lib', install_github("markmfredrickson/RItools"), 'pre')
```

```{r echo=FALSE}
library(dplyr)
library(RItools,lib.loc="./lib")
```

## Today

\begin{enumerate}
\item Agenda: Recap of the Estimation approach to Causal Inference with
Instruments, Walk through a testing based approach.
\item Reading for tomorrow from "Principled Search for designs with ignorable
assignment" on the Syllabus. DOS Chapter 1, Chapter 6 in Dunning, Hansen and
Sales 2015, Hansen and Bowers 2008. We are beginning the "Found experiments" section
of the course.
\item Questions arising from the reading or assignments or life?
%\item Recap:
\end{enumerate}

# But first: Where are we and where are we going?

## Research designs to aid the creation of interpretable comparisons

   - Randomized experiments (randomized $Z$ changes $y$) <!-- (more precision from reducing heterogeneity in $Y$) -->
     - No systematic differences between groups
     - Known reference distribution for tests of the sharp null and, in large
       samples, the weak null.
   - Instrumental variables (randomized $Z$ changes $D$ which changes $y$)
   - Natural Experiments / Discontinuities (one $X$ creates $Z$ which \ldots)
   - Semi-/Non-parametric Covariance Adjustment ($Z$ out of our control, but we can choose how to organize and observe given $X$) (matching and difference-in-differences)
   - Parametric covariance adjustment (same as above, but with a model for statistical adjustment, i.e. "controlling for")


# Now: Instruments, the ITT and CACE/LATE

## Defining causal effects for the Adams and Smith (1980) Study

 -  $Y$ & outcome ($y_{i,T}$ or $y_{i,Z_i=1}$ for potential outcome to
   treatment for person $i$, fixed)
 - $X$ & covariate/baseline variable
 - $Z$ & treatment assignment ($Z_i=1$ if assigned to a call, $Z_i=0$ if not
   assigned to a call)
 -  $D$ & treatment received ($D_i=1$ if answered phone, $D_i=0$ if person $i$
   dd not answer the phone.)

We have two causal effects: $Z \rightarrow Y$ ($\delta$, ITT, ITT$_Y$), and $Z \rightarrow D$ (GG
call this ITT$_D$).

And different types of people can react differently to the attempt to move the
dose with the instrument.

\centering
\begin{tabular}{llcc}
                       &        & \multicolumn{2}{c}{$Z=1$} \\
		       &       & $D=0$ & $D=1$ \\
		       \midrule
\multirow{2}{*}{$Z=0$} & $D=0$ & Nevertaker & Complier \\
                       & $D=1$ & Defier     & Always taker \\
		       \bottomrule
\end{tabular}

##  Defining causal effects

The $ITT=ITT_Y=delta= \bar{y}_{Z=1} - \bar{y}_{Z=0}$.

But, in this design, $\bar{y}_{Z=1}$ is split into two pieces: the outcome of
those who answered the phone and those who did not. So:

$ITT=$


 - $\delta_C = ITT_{Complier} = \bar{y}_{1,1} - \bar{y}_{0,0} | Complier$
 - $\delta_D = ITT_{Defier} = \bar{y}_{1,0} - \bar{y}_{0,1} | Defier$
 - $\delta_N = ITT_{Nevertaker} = \bar{y}_{1,0} - \bar{y}_{0,0} | Nevertaker$
 - $\delta_A = ITT_{Alwaystaker} = \bar{y}_{1,1} - \bar{y}_{0,1} | Alwaystaker$

$$ITT=\delta = ( \bar{y}_{1,1}p(d_1=1) + \bar{y}_{Z=1,d_1=0}(1-p(d_1=1)) )  - \bar{y}_{Z=0}$$

\medskip

Notice that if everyone opens the door when visited then $y_{i,Z=1,d_1=1} =
y_{i,Z=1}$  so the ITT=ATE.

\medskip

Why give it a new name?

## A simulation to understand the design and analysis

Setting up the simulation and analysis before going into the field.

```{r}
library(randomizr)
set.seed(20180730)
n <- 1000
## Make a covariate, like age
xtmp <- rnorm(n,mean=45,sd=10)
x <- pmax(xtmp,18)
## Make potential outcome to control
y0tmp <- round(rnorm(n,mean=0,sd=10)) + x/2
y0 <- pmax(y0tmp,0)
Z <- complete_ra(n)
## Door opening is a function of x (age) and Z (visits)
probd <- ( x-min(x) )/( max(x) - min(x) )
mean(probd)
d <- ifelse(Z==1, rbinom(sum(Z==1),prob=probd/2,size=1), 0)
table(Z,d,exclude=c())
mean(d[Z==1])
```

```{r}
## No effects possible if not visited
y0d0 <- y0
y0d1 <- y0
y1d0 <- y0
taubar <- .25*sd(y0) ## Effect of .25 sds on compliers/door openers
## The treatment also changes variance of y1d1
y1d1 <- mean(y0) + taubar + (y0 - mean(y0))/2
## Define observed outcome
Y <-  Z*d*y1d1 + Z*(1-d)*y1d0 + (1-Z)*d*y0d1 + (1-Z)*(1-d)*y0d0
dat <- data.frame(y1d1,y1d0,y0d1,y0d0,y0,Y,Z,d,x,probd)
dat$y1 <- (1-Z)*y0 + Z*d*y1d1 + Z*(1-d)*y1d0
```

## A simulation to understand the design and analysis


```{r}
boxplot(Y~Z*d,data=dat)
```


## Learning about the ITT

First, let's learn about the effect of the policy itself, about the ITT.

```{r}
## True ITT with one-sided non-compliance
pd <- mean(dat$d[dat$Z==1])

bary1 <- mean(dat$y1d1)*pd + mean(dat$y1d0)*(1-pd)
bary0 <- mean(dat$y0)

trueITT <- bary1 - bary0
```


\smallskip

What evidence do we have against the claim that the policy had no effects? How
would we assess whether this testing procedure is trustworthy --- controls its
false positive error rate?

\smallskip

What is our best guess about the ITT itself? How would we assess whether or
not this estimator is biased or not?

\smallskip

What is our best guess about how our estimate of the ITT would vary from
experiment to experiment?


## Some code notes

```{r, eval=FALSE}
newExp <- function(trt){ sample(trt) }
testStat <- function(outcome,trt){  mean(outcome[trt==1]) - mean(outcome[trt==0]) }
obsTestStat <- testStat(outcome=dat$Y,trt=dat$Z)
nullDist <- replicate(1000,testStat(outcome=dat$Y,trt=newExp(dat$Z)))
upperP <- mean(nullDist >= obsTestStat)
```


```{r, eval=FALSE}
##taubarhat <- with(dat, mean(Y[Z==1]) - mean(Y[Z==0]))
## or taubarhat <- coef(lm(Y~Z))[["Z"]]
library(estimatr)
taubarhat <- difference_in_means(Y~Z,data=dat)
## sqrt( var(Y[Z==1])/sum(Z==1) + var(Y[Z==0])/sum(Z==0) )
```


## The Complier Average Causal Effect or Local Average Treatment Effect (on the Compliers)

Now we would like to learn about the causal effect of answering the door and having the
conversation, the theoretically interesting effect. But, this comparison is confounded
by $x$ (tells us about differences in the outcome due to $x$ in addition to
the difference caused by $D$).

```{r}
with(dat,cor(Y,x))
with(dat,cor(d,x))
with(dat,cor(Z,x)) ## should be near 0
```


## Identifying the causal effect of opening the door

We want to learn about the Complier Average Causal Effect (aka the Local
Average Treatment Effect): CACE$= \bar{y_{i,Z=1,d_1=1}} -
\bar{y_{i,Z=0,d_1=1}}$


Let $p$ be the proportion of compliers ($\sum_{i=1}^N I(d_{i,1} = 1)/N$).

\begin{align*}
ITT & = \left( \bar{y}_{1,d_1=1}p + \bar{y}_{1,d_1=0}(1-p) \right)  - \bar{y}_{0} \\
    & = ( \bar{y}_{1,d_1=1}p + \bar{y}_{1,d_1=0}(1-p) )  - ( \bar{y}_{0,d_0=1}p+ \bar{y}_{0,d_0=0}(1-p) )  \\
    & = ( \bar{y}_{1,d_1=1}p - \bar{y}_{0,d_0=1}p )  +  ( \bar{y}_{1,d_1=0}(1-p)  - \bar{y}_{0,d_0=0}(1-p) )  \\
    & = ( \bar{y}_{1,d_1=1} - \bar{y}_{0,d_0=1} )p  +  ( \bar{y}_{1,d_1=0}  - \bar{y}_{0,d_0=0}) (1-p)  \\
\intertext{but $\bar{y}_{1,d_1=0} = \bar{y}_{0,d_1=0} = \bar{y}_{0,d_0=0}$ so }
  ITT  & = ( \bar{y}_{1,d_1=1} - \bar{y}_{0,d_0=1} ) p \\
  ITT  & = ( CACE ) p \\
\intertext{so we can write the CACE as ITT/p}
  ITT/p & = CACE
\end{align*}

This means that we can write the CACE as $CACE = ITT/\text{proportion of compliers}$

## Code snippets

```{r}
ITT <- with(dat,mean(Y[Z==1]) - mean(Y[Z==0]))
pd <- with(dat, mean(d[Z==1]) - mean(d[Z==0]))
ITT/pd

library(AER)
iv1 <- ivreg(Y~d | Z, data=dat)

```



## Summary of the Day

  - More than one way to use what we observe to reason about potential
    outcomes: testing, estimating, predicting
  - Structure of the course. Where we are now. Where we are going: away from
    randomized experiments toward observational studies, but with the statistical basis
    developed from our understanding of randomized experiments.
  - An instrument moves a dose. The effect of the dose is specific, or local,
    to those who took a dose, to those who experienced the treatment.
  - The ITT and CACE/LATE represent different causal effects. ITT tends to be
    the policy-relevant effect.
  - An "as-treated" analysis is no longer an experiment.

## References

