## Does xBalance have a controlled false positive rate here?

```{r xberror, echo=TRUE, cache=TRUE}
xbfn <- function() {
  acorn$newz <- sample(acorn$z)
  xb1 <- xBalance(newz ~ v_p2003 + v_m2003 + v_g2002 + v_p2002 + v_m2002 + v_s2001 +
    v_g2000 + v_p2000 + v_m2000 + v_s1999 + v_m1999 + v_g1998 +
    v_m1998 + v_s1998 + v_m1997 + v_s1997 + v_g1996 + v_p1996 +
    v_m1996 + v_s1996 + size,
  data = acorn,
  report = "chisquare.test"
  )
  return(xb1$overall[["p.value"]])
}

res <- replicate(1000, xbfn())
```

```{r resout, echo=TRUE}
summary(res)
mean(res <= .05)
mean(res <= .2)
```

## Does xBalance have a controlled false positive rate here?

Ex. are fewer than 5% of the p-values less than .05?

```{r}
plot(ecdf(res))
abline(0, 1)
abline(v = c(.01, .05, .1))
```


## Does the simulation based approach have a controlled false positive rate here?


```{r resdirecterror, cache=TRUE}
d2pfn <- function(z, X) {
  newz <- sample(z)

  d_dist <- replicate(1000, d_stat(sample(newz), X, ss = rep(1, nrow(X))))

  obs.d <- d_stat(newz, X, rep(1, nrow(X)))

  dps <- matrix(NA, nrow = length(obs.d), ncol = 1)
  for (i in 1:length(obs.d)) {
    dps[i, ] <- 2 * min(mean(d_dist[i, ] >= obs.d[i]), mean(d_dist[i, ] <= obs.d[i]))
  }

  invCovDDist <- solve(cov(t(d_dist)))
  obs.d2 <- d2_stat(obs.d, d_dist, invCovDDist)

  d2_dist <- apply(d_dist, 2, function(thed) {
    d2_stat(thed, theinvcov = invCovDDist)
  })

  d2_ <- mean(d2_dist >= obs.d2)

  return(d2_)
}
```

```{r doresdirect, eval=FALSE, cache=TRUE}
## This takes awhile!
resdirect <- replicate(1000, d2pfn(z = acorn$z, X = acorn[, acorncovs]))
```

```{r doresdirectparallel, eval=TRUE, cache=TRUE}
library(parallel)
resdirectlst <- mclapply(1:1000, function(i) {
  message(i)
  d2pfn(z = acorn$z, X = acorn[, acorncovs])
}, mc.cores = detectCores())
resdirect <- unlist(resdirectlst)
save(resdirect, file = "day9-resdirect.rda")
```

## Does the simulation based approach have a controlled false positive rate here?

It looks like it is a bit too high. Hmm... Maybe the simulation needs to be fixed.

```{r lazyload, echo=TRUE}
## lazyLoad("day9-AdjustmentBalance_cache/beamer/doresdirectparallel_5ec4fa8cdcbcf586138e928bc0f9fc0b")
## load("day9-resdirect.rda")
summary(resdirect)
mean(resdirect <= .05)
mean(resdirect <= .2)
```

## Does xBalance have a controlled false positive rate here?


```{r}
plot(ecdf(resdirect))
abline(0, 1)
abline(v = c(.01, .05, .1))
```

