---
title: |
  | Sensitivity Analysis: Describing the consequences of possible confounders
date: '`r format(Sys.Date(), "%B %d, %Y")`'
author: |
  | ICPSR 2026 Session 1
  | Jake Bowers \& Tom Leavitt
bibliography:
 - 'BIB/MasterBibliography.bib'
fontsize: 10pt
geometry: margin=1in
graphics: yes
biblio-style: authoryear-comp
biblatexoptions:
  - natbib=true
output:
  beamer_presentation:
    slide_level: 2
    keep_tex: true
    latex_engine: xelatex
    citation_package: biblatex
    template: styles/icpsr-beamer-template
    incremental: true
    includes:
        in_header:
           - defs-all.sty
    md_extensions: +raw_attribute-tex_math_single_backslash+autolink_bare_uris+ascii_identifiers+tex_math_dollars
    pandoc_args: [ "--csl", "chicago-author-date.csl" ]
---

```{r setup1_env, echo=FALSE, include=FALSE}
library(here)
source(here::here("rmd_setup.R"))
```

```{r setup2_loadlibs, echo=FALSE, include=FALSE}
library(tidyverse)
library(dplyr)
library(optmatch)
library(RItools)
library(estimatr)
library(senstrat)
library(sensitivitymv)
library(sensemakr)
library(blkvar)
library(propertee)
library(coin)
options(show.signif.stars = FALSE)
```

```{r setup5_fogarty, echo=FALSE, include=FALSE}
## Fogarty (2018) finely-stratified variance estimator (from Tom's
## matching-guide, R/fine_strat_var_est.R). It reads Var(tau-hat) off how the
## set-level estimates vary ACROSS sets, so it never needs a within-set (n - 1)
## variance. Conservative (>= the truth), hence valid.
fine_strat_var_est <- function(strat_ns, strat_ests) {
  N <- sum(strat_ns); B <- length(strat_ns)
  Q <- as.matrix(B * (strat_ns / N))                  # set weights (K = B)
  H <- Q %*% solve(t(Q) %*% Q) %*% t(Q)               # leverage (hat) matrix
  y <- strat_ests / sqrt(pmax(1 - diag(H), .Machine$double.eps))
  W <- diag(B * (strat_ns / N))
  as.numeric((1 / B^2) * t(y) %*% W %*% (diag(B) - H) %*% W %*% y)
}
```

## Today

1. Introductions
  - Introduction to Jake
  - Introductions all around: Name, Discipline/Focus, Home department
2. Review and Preview of the Week
3. Return to UN peacekeeping
4. **Today:Sensitivity analysis**: reasoning about how departures from the "as-if
   randomized" assumption might change our substantive conclusions. Two
   approaches to the same question --- Rosenbaum's $\Gamma$ and Cinelli and
   Hazlett's `sensemakr`.
  
# Review of Causal Inference So Far

## So far: Why and how infer? Foundations of causal inference.

A quick refresh of the first two weeks before we add to it today.

- **Statistical inference for causal effects.** A causal effect compares a
  units's two potential outcomes, $y_i(1)$ and $y_i(0)$ (under SUTVA), but only
  one is ever seen --- so we *infer* about effects rather than observe them
  directly. **How to infer?**
- **Randomization is a reasoned basis for inference.** This fact simplifies
  causal inference from **randomized experiments**. Randomization also
  **indentifies** the ATE **estimand**.
- **Two null hypotheses, two tests for statistical inference.**
   - *Sharp* (Fisher): no effect for *any* unit. Permute/shuffle assignment to
     create a reference distribution.
   - *Weak* (Neyman): the *average* effect is zero. The difference-in-means
     with a conservative variance and a Normal reference dist.
- **Estimand vs estimator.** The ATE --- the average of $y_i(1) - y_i(0)$ ---
  is the target; the difference-in-means is an unbiased guess at it.
  Unbiasedness allows the sample difference in means to be used as a test
  statistic for the Weak Null.

## So far: Beyond the RCT

- **Refinements in experiments.** Covariate adjustment for precision; the
  CACE/LATE under not-random noncompliance; the consequences of attrition
- **Observational studies** How to create a reasoned basis? How to engage
  **alternative explanations** (confounders)?
- **Recover it by design through stratification** on covariates and/or scores
  of them (propensity score, Mahalanobis scores)
   - **Check it before the outcomes, then test.** Balance (standardized
     differences, an omnibus test); then the same sharp and weak tests within
     sets.

\medskip

Questions? Comments?

## Next up: This week ideas

- **Today:** How to reason about the consequences of unobserved
  confounders/departures from the assumption of conditional ignorability?
- **Tues:** Why and how is "controlling for" in a linear regression a tool for
  causal infrence? (plus Review)
- **Wed:** Matching with continuous and/or multi-valued treatments
  ("non-bipartite matching") (plus Review)
- **Thurs:** Given longitudinal/panel data, risk-set matching (conditional
  ignorability) vs Difference-in-Differences (no anticipation, parallel trends)

\medskip

Questions? Comments?

#  The Causal Effect of UN Peacekeeping

##  Does UN peacekeeping prolong peace? \citep{gilligansergenti2008}

- **Unit**: a *post-conflict peace episode* (the period of peace after a civil
   war). E.g.\ Sierra Leone, Jan 2001--Dec 2003. There are $87$ episodes,
   1989--2003.
 - **Treatment** (`UN`): $1$ if a UN mission was *present* during the episode,
   else $0$. Here $19$ treated, $68$ control.
 - **Outcome** (`ldur`): log days of peace until conflict resumes (or
   censoring).
 - **The question**: does the presence of a UN mission *prolong* post-conflict
   peace?

\medskip

UN missions are **not randomly assigned**. Where peace is fragile for reasons
we see or do not see, the UN may be *more* likely to deploy.

## The nine covariates 

Pre-treatment characteristics thought to drive UN intervention (and to predict
the duration of peace):

 - logged war deaths (`lwdeaths`), logged previous-war duration (`lwdurat`)
 - ethnic fractionalization (`ethfrac`), logged population (`pop`)
 - logged mountainous terrain (`lmtnest`), military personnel (`milper`)
 - logged pre-war GDP per capita (`bwgdp`), Polity score (`bwplty2`)
 - `region` (E.\ Europe, Lat.\ America, Asia, Sub-Saharan Africa, N.\
   Africa / Middle East)

\medskip

The design goal: build matched sets of episodes that are
similar on these covariates, so that *within a set* a UN mission looks **as-if**
randomly assigned with repect to those observed covariates.

\medskip

Make an argument for *conditional ignorability* based on **observed covariates**.

## The design {.allowframebreaks}

```{r setup3_data_and_match, echo=TRUE, include=TRUE, cache=TRUE}
## running example:  UN peacekeeping and post-conflict peace.
peace <- readRDS(url(
  "https://raw.githubusercontent.com/tl2624/matching-guide/main/data/peace_pre_match.rds"))
covs <- c("lwdeaths", "lwdurat", "ethfrac", "pop", "lmtnest",
          "milper", "bwgdp", "bwplty2", "region")

## Estimated propensity score (logit index)
psm <- glm(reformulate(setdiff(covs, "region"), "UN"),
           family = binomial("logit"), data = peace)
peace$logit_ps <- psm$linear.predictors
pop_sd_logit <- sqrt(mean((peace$logit_ps - mean(peace$logit_ps))^2))

## A full match: rank-Mahalanobis + a 0.5-SD PS caliper + exact on region
## + calipers on ethfrac (35) and bwgdp (2); at most 4 controls per treated.
fm <- fullmatch(
  match_on(UN ~ logit_ps, caliper = 0.5 * pop_sd_logit, data = peace,
           standardization.scale = NULL, method = "euclidean") +
  match_on(reformulate(setdiff(covs, "region"), "UN"), data = peace,
           standardization.scale = NULL, method = "rank_mahalanobis") +
  match_on(UN ~ ethfrac, caliper = 35, data = peace,
           standardization.scale = NULL, method = "euclidean") +
  match_on(UN ~ bwgdp, caliper = 2, data = peace,
           standardization.scale = NULL, method = "euclidean") +
  exactMatch(UN ~ region, data = peace),
  data = peace, max.controls = 4)
peace$fm <- fm
md <- droplevels(peace[matched(fm), ])
```

## The design summary

```{r recall_design, echo=TRUE}
summary(fm, min.controls=0,max.controls=Inf)                       # structure of the matched sets
effectiveSampleSize(fm)           # equivalent number of matched pairs
```

\medskip

$13$ treated and $23$ control episodes matched into $12$ sets (a mix of $1$:$1$,
$1$:$K$, and one $2$:$1$); effective sample size $\approx 15.3$.

\medskip

Is this a good enough design? (Recall: we want an answer that does not involve
outcomes.)

## Balance (1): the omnibus test

Before we look at any outcome: does the matched design resemble a
block-randomized experiment on the covariates we balanced?

```{r balance_omnibus, echo=TRUE}
## Hansen & Bowers omnibus balance test; the "fm" row is within the matched sets
balfmla <- reformulate(covs, "UN")            # UN ~ lwdeaths + ... + region
balanceTest(update(balfmla, . ~ . + strata(fm)), data = md)$overall["fm", ]
```

\medskip

A *high* p-value ($\approx 0.95$) is the good case: across the
nine covariates jointly, the within-set imbalance is no larger than block
randomization would routinely produce. OR, The average within-set differences
in means are smaller than $\approx 95\%$ of equivalent block-randomized
experiments. 

## Balance (2): within-set differences, covariate by covariate

```{r balance_diffs, echo=TRUE}
keyvars <- setdiff(covs, "region")            # region is matched exactly
## within each set: abs(mean(treated) - mean(control)) on covariate v
wsd <- function(v){
  tapply(seq_len(nrow(md)), md$fm, function(i){
    abs(mean(md[[v]][i][md$UN[i] == 1]) - mean(md[[v]][i][md$UN[i] == 0]))})
}
## percentiles of those set-level differences, across the sets
round(t(sapply(keyvars, function(v){ quantile(wsd(v), c(0, .25, .5, .75, .90, 1)) })), 2)
```

\medskip

Columns are percentiles of the set-level differences. Does this look
substantively small? Or large? Is this good enough?

## Statistical inference: the "as-if randomized" approach

Given a matched design that (a) we can defend on substantive grounds and (b)
compares favorably with a block-randomized experiment on the *observed*
covariates, why not analyze it **as if** treatment had been randomized within
sets? "If this had been an experiment, the estimate would be $\hat{\tau}$,
$p = .03$."

Advantages:

 - No probability model of outcomes to justify.
 - No functional form relating covariates and treatment and outcomes to
   justify.
 - No population and sampling needed --- inference is about the pool of
   episodes we have.
 - The CLT enters only for convenience (a Normal reference for the weak null).
 - It already compares favorably with an experiment.

## From the design to an estimate, an SE, and a CLT

What is the average effect of UN presence on log-duration?

```{r est_se_three_ways, echo=TRUE, cache=TRUE}
## propertee design spec + ATE weights (used below)
spec <- obs_spec(UN ~ uoa(id) + block(fm), data = md)
md$atew <- ate(spec, data = md)

## (1) ATE-weighted lm_robust (HC2 standard error)
lm_ate <- lm_robust(ldur ~ UN, weights = atew, data = md)

## (2) propertee::lmitt -- design-based, ATE weights
lmitt1 <- lmitt(ldur ~ 1, specification = spec, data = md, weights = "ate")

## (3) blkvar (Pashley-Miratrix 2021 variance)
blk <- block_estimator(Yobs = ldur, Z = UN, B = fm, data = md, method = "hybrid_p")

round(c(lm_robust = coef(lm_ate)[["UN"]],
        propertee = coef(lmitt1)[["UN."]],
        blkvar    = blk$ATE_hat), 3)
```

\smallskip

The ATE is about $\hat{\tau} \approx 0.615$ (log-days). The *standard error* is
the harder part.

## The variance problem: why not just HC2? (review)

The variance of the design-based ATE is a sum over the matched sets:
\begin{equation*}
\mathrm{Var}(\hat{\tau}) = \sum_s (n_s/n)^2\, \mathrm{Var}(\hat{\tau}_s).
\end{equation*}

The textbook (Neyman) estimate of a set's variance --- the design-based SE you
learned for experiments in Gerber and Green, and what HC2 reproduces --- needs a
within-set sample variance in *each arm*:
\begin{equation*}
\widehat{\mathrm{Var}}(\hat{\tau}_s) = \frac{\hat{S}^2_{sT}}{m_s} + \frac{\hat{S}^2_{sC}}{n_s - m_s},
\end{equation*}
whose two arm variances have denominators $m_s - 1$ and $n_s - m_s - 1$.

\medskip

Every matched set here has a **single** treated or a single control unit, so one
denominator is $0$: the within-set variance is **undefined**. Finely stratified
designs break the textbook within-set estimator.

## Fogarty's fix: read the variance off the spread across sets

Estimate $\mathrm{Var}(\hat{\tau})$ from how the set-level effects $\hat{\tau}_s$
vary *across* sets --- no within-set $(n-1)$ needed, and conservative, so the
test stays valid.

```{r fogarty_se, echo=TRUE, cache=TRUE}
set_stats <- md |> dplyr::group_by(fm) |> dplyr::summarize(
  n = dplyr::n(), dim = mean(ldur[UN == 1]) - mean(ldur[UN == 0]), .groups = "drop")
ate_hat    <- with(set_stats, sum((n / sum(n)) * dim))            # 0.615
se_fogarty <- sqrt(fine_strat_var_est(set_stats$n, set_stats$dim))
round(c(HC2 = lm_ate$std.error[["UN"]], Fogarty = se_fogarty,     # HC2 ignores sets
        PashleyMir = blk$se_est), 3)                              # PM ~ Fogarty
```

\footnotesize
Pooled HC2 ($0.387$) ignores the matched sets and overstates the SE; Fogarty
($0.341$) and Pashley-Miratrix ($0.345$) condition on the design --- tighter,
still conservative. Pashley-Miratrix is a *hybrid* (the exact within-set variance
where a set is big enough, the between-set spread elsewhere); here every set has
one unit per arm, so it coincides with Fogarty.
[@fogarty2018a; @pashleymiratrix2020; @gerber2012field]
\normalsize

## Statistical inference for a causal effect {.shrink}

```{r design_pvalue, echo=TRUE}
z_obs <- ate_hat / se_fogarty               # Fogarty SE; z = 1.80
c(z = round(z_obs, 2),
  p_one_sided = round(pnorm(z_obs, lower.tail = FALSE), 3))

## And the sharp-null using 2 different test statistics and direct permutations
rank_test <- wilcox_test(ldur~factor(UN)|fm,data=md,alternative="less",
distribution=approximate(nresample=10000))
mean_test <- oneway_test(ldur~factor(UN)|fm,data=md,alternative="less",
distribution=approximate(nresample=10000))
c(rankp=pvalue(rank_test)[[1]],meanp= pvalue(mean_test)[[1]])
```

One-sided tests for the directional question --- does UN presence *prolong* peace?

 \smallskip 
 
 This is a **real but not overwhelming** rejection. Would even a little hidden
 bias overturn it? 

# Two ways to reason about the unobserved

## One question, two approaches

*How much hidden bias would it take to overturn our conclusion?* Two approaches 
answer it, each measuring hidden bias differently.

 - **Rosenbaum ($\Gamma$).** Within a matched set, how many times more likely
   (an odds ratio $\Gamma$) would one episode have to be to receive a UN
   mission --- because of an unobserved $u$ --- before our **randomization
   test** stops rejecting? Measured as: a treatment-odds ratio.
 - **Cinelli and Hazlett (partial $R^2$).** How strongly would an unobserved
   confounder have to relate to *both* treatment and outcome before our
   **regression estimate** loses significance (or reaches zero)? Measured as:
   shared variance.

\medskip

\footnotesize
(Word-collision warning: each tradition calls its threshold a "robustness
value." They are not the same number.)
\normalsize

## Sensitivity analysis as a formalized thought experiment {.fragile}

> "In an observational study, a sensitivity analysis replaces qualitative claims
> about whether unmeasured biases are present with an objective quantitative
> statement about the magnitude of bias that would need to be present to change
> the conclusions." (Rosenbaum, `sensitivitymv` manual)

\medskip

\begin{center}
\begin{tikzcd}[column sep=large,every arrow/.append style=-latex]
u \arrow[from=1-1,to=2-2, bend right, "\Gamma" near start] \arrow[from=1-1,to=2-3, bend left, "\rho" near start] \\
& Z \arrow[from=2-2,to=2-3, "\tau"] & y \\
x_1 \arrow[from=3-1,to=2-2, "{\color{red}\text{\scriptsize$\times$}}"{marking, pos=0.3}] \arrow[from=3-1,to=2-3,grey] &
x_2 \arrow[from=3-2,to=2-2, "{\color{red}\text{\scriptsize$\times$}}"{marking, pos=0.3}] \arrow[from=3-2,to=2-3,grey]
\end{tikzcd}
\end{center}

We match on $x$. The question is what an unobserved $u$ --- related to treatment
(with strength $\Gamma$) and outcome (with strength $\rho$) --- could do to the comparison.

## Where does the unobserved $u$ enter?

Rosenbaum models each unit's treatment probability
$\pi_i = \Pr(Z_i{=}1 \mid \bx_i, u_i)$ with a logit that carries an unobserved
$u_i \in [0,1]$ beside the observed $\bx_i$ --- one copy per episode:
\begin{align*}
\log\!\left(\frac{\pi_i}{1 - \pi_i}\right) &= \kappa(\bx_i) + \gamma\, u_i , \\
\log\!\left(\frac{\pi_j}{1 - \pi_j}\right) &= \kappa(\bx_j) + \gamma\, u_j .
\end{align*}

Compare two episodes by their *ratio of treatment odds*. Odds divide, so
\begin{equation*}
\frac{\pi_i / (1 - \pi_i)}{\pi_j / (1 - \pi_j)}
  = \frac{\pi_i\,(1 - \pi_j)}{\pi_j\,(1 - \pi_i)} .
\end{equation*}
A $\log$ turns this ratio into a *difference* of the two logits above --- next slide.

## Matching on $\bx$ leaves only $u$

The log of a quotient is a difference, so substitute the two logits:
\begin{align*}
\log \frac{\pi_i (1 - \pi_j)}{\pi_j (1 - \pi_i)}
  &= \log\!\frac{\pi_i}{1 - \pi_i} - \log\!\frac{\pi_j}{1 - \pi_j} \\
  &= \big(\kappa(\bx_i) + \gamma\, u_i\big) - \big(\kappa(\bx_j) + \gamma\, u_j\big) \\
  &= \underbrace{\kappa(\bx_i) - \kappa(\bx_j)}_{=\, 0 \ \text{when} \ \bx_i = \bx_j}
     \;+\; \gamma\,(u_i - u_j) .
\end{align*}
Match on $\bx$ ($\bx_i = \bx_j$) and the observed part cancels. Exponentiating leaves
the odds ratio at $e^{\gamma\,(u_i - u_j)}$, between $e^{-\gamma}$ and $e^{\gamma}$
since $u_i - u_j \in [-1, 1]$. Writing $\Gamma = e^{\gamma}$ gives the bound on the
next slide.

## Why $\Gamma$? A model for the unobserved {.fragile}

For two episodes $i$ and $j$ with the same matched-set covariates ($\bx_i =
\bx_j$), @rosenbaum2002observational bounds their ratio of treatment odds by a
single number $\Gamma$:

\begin{equation*}
\frac{1}{\Gamma} \leq \frac{\pi_i (1 - \pi_j)}{\pi_j (1 - \pi_i)} \leq \Gamma .
\end{equation*}

 - $\Gamma = 1$: within a set, both episodes are equally likely to get a UN
   mission --- exactly the as-if randomized case.
 - $\Gamma = 2$: an unobserved $u$ could make one episode **twice** as likely as
   its match to receive a mission.

We compute the **worst-case** $p$-value the test could take for each $\Gamma$,
and ask how large $\Gamma$ must be before we no longer reject.

## Why is the $p$-value a *worst case*?

- **$\Gamma = 1$: one set of probabilities.** Within a matched set, each
  episode is equally likely to be the one that received a UN mission. Those
  equal probabilities are what our randomization test has assumed all along,
  and they give a single $p$-value.
- **$\Gamma > 1$: many.** There are many different unobserved $u$, each of
  which could make some episodes more likely than their set-mates to receive a
  mission. We never see $u$ --- we know only that two set-mates can differ by
  at most a factor of $\Gamma$. Each set of probabilities allowed by that
  factor gives its own $p$-value.**The worst case is the largest of those
  $p$-values.** It is the set under which our result is least surprising: the
  episodes that had longer peace were also the ones $u$ would have made most
  likely to get a mission. Then "UN episodes lasted longer" could be $u$ at
  work, not the mission.

\medskip

If even that largest $p$ stays below $0.05$, no confounder of strength $\Gamma$
explains the result away.

## The worst case, in one matched pair

Take a $1$:$1$ set with outcomes $y_i > y_j$ and an upper-tailed test. Under the
null, which unit got treated?
\begin{equation*}
\Pr(i \text{ treated} \mid \text{one treated})
  = \frac{e^{\gamma u_i}}{e^{\gamma u_i} + e^{\gamma u_j}}
  \in \left[\frac{1}{1 + \Gamma}, \ \frac{\Gamma}{1 + \Gamma}\right] .
\end{equation*}

The upper end $\Gamma/(1+\Gamma)$ is the worst case for us: it makes the
*high-outcome* unit the one most likely to be treated even with no effect, so the
observed "treated $>$ control" is least surprising and the $p$-value is largest.
`senstrat` does this for every matched set at once --- each pushed to its
high-outcome-favoring extreme --- and returns the single upper-bound $p$ we plot
against $\Gamma$.

## The search for $\Gamma$

```{r gamma_search, echo=TRUE, cache=TRUE}
##  Only interested in positive effect (UN prolongs peace), so we test the upper tail: alternative = "greater".
pval <- function(g){ as.numeric(
  senstrat(md$ldur, z = md$UN, st = md$fm, gamma = g,
           alternative = "greater")$Result["P-value"])
           }

gammas <- seq(1, 2, by = 0.02)
pvals  <- sapply(gammas, pval)
gstar  <- uniroot(function(g){pval(g) - 0.05}, c(1, 2))$root
round(c(p_at_Gamma1 = pval(1), p_at_Gamma1.5 = pval(1.5)), 3)   # worst-case one-sided p
c(Gamma_star = round(gstar, 2))                                 # smallest Gamma with worst-case p > 0.05
```

\footnotesize

At $\Gamma = 1$ (as-if randomized) the one-sided $p \approx `r round(pval(1), 3)`$. The
smallest $\Gamma$ that pushes the worst-case $p$ above $0.05$ is $\Gamma^{*}
\approx `r round(gstar, 2)`$.

\normalsize

## The result is fragile to hidden bias

```{r gamma_plot, echo=FALSE, out.width=".6\\textwidth"}
par(mar = c(3.2, 3.2, 1, 1), mgp = c(2, .6, 0))
plot(gammas, pvals, type = "l", lwd = 2,
     xlab = expression(Gamma~"(max treatment-odds ratio from unobserved u)"),
     ylab = "Worst-case one-sided p-value", ylim = c(0, 0.2))
abline(h = 0.05, lty = 2)
abline(v = gstar, lty = 3, col = "red")
text(gstar, 0.18, labels = bquote(Gamma^"*" == .(round(gstar, 2))), pos = 4, col = "red")
```

An unobserved confounder making one episode $\approx `r round(gstar,2)`$
times more likely than its match to receive a UN mission would already be enough
to overturn the one-sided rejection.

## The choice of test statistic matters

```{r stat_helpers, echo=FALSE}
## pval_ss(sc, g): worst-case one-sided p at bias Gamma = g, scoring with sc
## gstar_ss(sc):   the Gamma where that worst-case p crosses 0.05
pval_ss  <- function(sc, g){ as.numeric(
  senstrat(sc, z = md$UN, st = md$fm, gamma = g, alternative = "greater")$Result["P-value"]) }
gstar_ss <- function(sc){ tryCatch(
  uniroot(function(g){ pval_ss(sc, g) - 0.05 }, c(1, 10))$root, error = function(e){ NA }) }
```

```{r stat_menu, echo=TRUE, cache=TRUE}
## pval_ss() wraps senstrat (as on the previous slide); gstar_ss() finds the
## Gamma where the worst-case p crosses 0.05 (both defined just above)
stats <- list(mean           = md$ldur,
              m_statistic    = mscores(md$ldur, md$UN, md$fm),
              hodges_lehmann = hodgeslehmann(md$ldur, md$UN, md$fm))
stat_tab <- round(sapply(stats, function(sc){ c(p_at_Gamma1 = pval_ss(sc, 1),
                                                Gamma_star  = gstar_ss(sc)) }), 2)
stat_tab
```

\footnotesize
The mean and Huber m-statistic reject ($p \approx 0.03$) and keep rejecting up to
$\Gamma^{*} = `r stat_tab["Gamma_star", "mean"]`$ and
$`r stat_tab["Gamma_star", "m_statistic"]`$ respectively.

\normalsize

## Two test statistics, two sensitivity curves

```{r two_curve, echo=FALSE, out.width=".68\\textwidth", cache=TRUE}
ms <- mscores(md$ldur, md$UN, md$fm)
gg <- seq(1, 2, by = 0.02)
p_mean <- sapply(gg, function(g){ pval_ss(md$ldur, g) })
p_m    <- sapply(gg, function(g){ pval_ss(ms, g) })
par(mar = c(3.2, 3.2, 1, 1), mgp = c(2, .6, 0))
plot(gg, p_mean, type = "l", lwd = 2, ylim = c(0, 0.2),
     xlab = expression(Gamma), ylab = "Worst-case one-sided p-value")
lines(gg, p_m, lwd = 2, lty = 2)
abline(h = 0.05, col = "grey50")
legend("topleft", c("mean", "m-statistic"), lwd = 2, lty = c(1, 2), bty = "n")
```

## Interpreting $\Gamma^{*}$: the amplification {.fragile}

@rosenbaum2009amplification split $\Gamma$ into the confounder's effect on
*treatment* ($\Lambda$) and on the *outcome* ($\Delta$): many $(\Lambda,
\Delta)$ pairs imply the same $\Gamma$.

```{r amplify, echo=TRUE}
g_round <- round(gstar, 2)                 # the worst-case Gamma* = 1.14, to decompose
print(g_round)
Lambda  <- c(1.3, 1.5, 2, 3, 5)            # candidate treatment-side strengths
Delta   <- unname(amplify(g_round, Lambda))   # outcome-side strength each one needs
rbind(Lambda_effect_on_Z = Lambda, Delta_effect_on_Y = round(Delta, 2))   # each column is one (Lambda, Delta) pair
```

\footnotesize
Each column is one way to split $\Gamma^{*} = `r g_round`$ into two pieces.
**$\Lambda$** (top row) is the confounder's effect on *treatment* --- the odds
ratio by which it makes an episode likelier to receive a UN mission. **$\Delta$**
(bottom row) is its effect on the *outcome* --- the odds ratio by which it makes
longer peace likelier. A confounder with both at once --- say $\Lambda = 2$ paired
with $\Delta = `r round(amplify(g_round, 2), 2)`$ --- would bias the test as much as
$\Gamma^{*} = `r g_round`$ does. Strong on one side needs less on the other; the
next slide draws the whole trade-off.
\normalsize

## The amplification curve: $\Lambda$ on treatment, $\Delta$ on outcome

A single $\Gamma^{*}$ is silent about *how* a confounder would work. The
amplification splits it into a curve of $(\Lambda, \Delta)$ pairs --- a
treatment-side strength $\Lambda$ and an outcome-side strength $\Delta$ --- that
all reproduce the same $\Gamma^{*}$.

```{r amplify_curve, echo=FALSE, out.width=".56\\textwidth"}
lam <- seq(g_round + 0.005, 8, length = 400)
par(mar = c(3.7, 3.7, 0.6, 0.6), mgp = c(2.4, .7, 0))
plot(lam, amplify(g_round, lam), type = "l", lwd = 2, xlim = c(1, 6), ylim = c(1, 4),
     xlab = expression(Lambda * " : effect on treatment (odds)"),
     ylab = expression(Delta * " : effect on outcome (odds)"))
abline(0, 1, col = "grey85", lty = 2)                       # the Lambda = Delta diagonal
abline(h = g_round, v = g_round, col = "grey70", lty = 3)   # both arms asymptote to Gamma*
sym <- g_round + sqrt(g_round^2 - 1)                        # the Lambda = Delta point
points(c(sym, 3), amplify(g_round, c(sym, 3)), pch = 19)
text(c(sym, 3), amplify(g_round, c(sym, 3)),
     labels = c("equal on both arms", "strong on treatment"), pos = 4, cex = .8)
```

\footnotesize
$\Lambda$ is the $u \to Z$ arrow ($\Gamma$ in the DAG): how much, in odds, a hidden
$u$ raises the chance of a UN mission. $\Delta$ is the $u \to y$ arrow ($\rho$): how
much $u$ raises the odds of longer peace. The same $\Gamma^{*} = `r g_round`$ comes
from $`r round(sym, 2)`\times$ on *both* arms, or $3\times$ on treatment with only
$`r round(amplify(g_round, 3), 2)`\times$ on the outcome --- more on one arm means
less is needed on the other. Each point on the curve produces exactly $\Gamma^{*}$:
a weaker confounder (nearer no effect, $\Lambda = \Delta = 1$) leaves the result
significant, a stronger one overturns it --- so the curve is the strength a real
confounder would need, on *both* arms, to undo the finding.
\normalsize

## From $\Gamma$ to $R^2$.

`sensemakr` focuses on tests of the weak null. We feed it the design ATE and the Fogarty SE:

```{r sensemakr_estimate, echo=TRUE, cache=TRUE}
## alpha = 0.10 (two-sided) == our one-sided 0.05 directional test
sens <- sensemakr(estimate  = ate_hat,
                  se        = se_fogarty,
                  treatment = "UN",
                  dof       = lm_ate$df.residual,
                  alpha     = 0.10)
sens$sensitivity_stats[, c("estimate", "se", "rv_q", "rv_qa")]
```

\footnotesize

**Robustness value** (`rv_qa` $\approx `r round(sens$sensitivity_stats$rv_qa,3)`$):
a confounder explaining only about
`r round(100*sens$sensitivity_stats$rv_qa,1)`\% of the residual variation in
*both* treatment and outcome would make the result **non-significant** (one-sided
$\alpha = 0.05$).

\normalsize

## Reading the contour plot

```{r sensemakr_plot, echo=FALSE, out.width=".68\\textwidth"}
plot(sens, sensitivity.of = "t-value")
```

\footnotesize
Axes: how much of the residual variation in treatment (x) and outcome (y) a
hypothetical confounder would explain. The contours are the *t-value* of the UN
effect; the red line is the one-sided $0.05$ significance threshold ($t \approx
1.7$). The design estimate is at the origin ($t \approx 1.8$). Where the diagonal
meets the red line is `rv_qa` $\approx `r round(sens$sensitivity_stats$rv_qa,3)`$
--- the confounder strength that ends significance.
\normalsize

# Putting it together

## Be honest about what each analysis assumes

 - The **estimate** is design-based: matched sets, as-if randomization, a CLT
   for the SE. No outcome model.
 - The **$\Gamma$ analysis** stays design-based: it perturbs the *assignment*
   probabilities and recomputes a randomization test.
 - The **`sensemakr` analysis** is a *regression* argument: Cinelli and
   Hazlett's partial-$R^2$ result is an omitted-variable-bias bound for linear
   models. Passing a matched ATE to `sensemakr` is a pragmatic approximation,
   not a design-based guarantee.

\medskip

So report it as: "*If* we treat the ATE regression as our working model, a
confounder explaining about `r round(100*sens$sensitivity_stats$rv_qa,1)`\% of the
residual variation in treatment and outcome would end the rejection" --- and show
$\Gamma^{*}$ alongside, so the same conclusion holds under both approaches.

## Where the two analyses agree

 - The UN-peacekeeping result is **real but fragile**: one-sided $p \approx
   0.034$ under as-if randomization, but $\Gamma^{*} \approx `r round(gstar,2)`$
   (Rosenbaum) and `rv_qa` $\approx `r round(sens$sensitivity_stats$rv_qa,3)`$
   (`sensemakr`) --- both tiny.
 - A *modest* unobserved confounder of UN deployment would overturn it. The
   defense has to be **substantive**: is there a feature of conflicts ---
   unmeasured here --- that strongly predicts both UN deployment and the
   duration of peace?
 - Sensitivity analysis does not say such a confounder exists. It tells us how
   hard we must argue that one does not.

## Summary

 - Formal sensitivity analysis lets us use the as-if randomized approach
   **provisionally** rather than fearfully: we ask directly how our conclusion
   would change under hidden bias.
 - **$\Gamma$** (Rosenbaum) bias as a departure from as-if randomization,
   measured as a treatment-odds ratio. Also "amplification".
 - **`sensemakr`** (Cinelli and Hazlett) give it an estimate, an SE, and read
   off the robustness value. 
- This is not "stability" or "fragility" analysis of model specification.

## References {.allowframebreaks}
