---
title: |
  | Longitudinal Designs: Risk-Set Matching vs Difference-in-Differences
date: '`r format(Sys.Date(), "%B %d, %Y")`'
author: |
  | ICPSR 2026 Session 1
  | Jake Bowers \& Tom Leavitt
bibliography:
 - 'BIB/MasterBibliography.bib'
fontsize: 10pt
geometry: margin=1in
graphics: yes
biblio-style: authoryear-comp
biblatexoptions:
  - natbib=true
output:
  beamer_presentation:
    slide_level: 2
    keep_tex: true
    latex_engine: xelatex
    citation_package: biblatex
    template: styles/icpsr-beamer-template
    incremental: true
    includes:
        in_header:
           - defs-all.sty
    md_extensions: +raw_attribute-tex_math_single_backslash+autolink_bare_uris+ascii_identifiers+tex_math_dollars
    pandoc_args: [ "--csl", "chicago-author-date.csl" ]
---

<!-- BUILD NOTE: this deck uses did, bacondecomp, HonestDiD, which live in the
     project renv. Render WITH renv active (do NOT use --no-init-file):
     Rscript -e 'rmarkdown::render("day_13_longitudinal_2026.Rmd")' -->

```{r setup1_env, echo=FALSE, include=FALSE}
library(here)
source(here::here("rmd_setup.R"))
```

```{r setup2_loadlibs, echo=FALSE, include=FALSE}
library(tidyverse)
library(dplyr)
library(estimatr)
library(optmatch)
library(RItools)
library(senstrat)        # Gamma sensitivity (Rosenbaum-Krieger sum statistic) for the matched design
library(did)             # event study / pooled DiD with clean never-treated controls
library(HonestDiD)       # sensitivity to parallel-trends violations
options(show.signif.stars = FALSE)
```

```{r build, echo=FALSE, include=FALSE, cache=TRUE}
## V-Dem again (the day-12 data). Long panel: each country-year, 1980-2010.
## Treatment = first year a country reaches democracy under V-Dem's Regimes of the
## World coding: v2x_regime in {2 (electoral), 3 (liberal)} vs {0,1} autocracy.
## Outcome = corruption (v2x_corr, 0=low ... 1=high).
WIN_LO <- 1980; WIN_HI <- 2010; POST <- 5
load(here::here("data", "vdem_democracy_transitions.rda"))
v <- vdem_sub %>% arrange(country_text_id, year) %>%
  mutate(dem = as.integer(v2x_regime %in% c(2, 3)),
         log_gdp = log(e_gdppc),
         log_gdp = ifelse(is.na(log_gdp), median(log_gdp, na.rm = TRUE), log_gdp))
fd <- v %>% filter(dem == 1) %>% group_by(country_text_id) %>%
  summarize(first_dem = min(year), .groups = "drop")
st <- v %>% group_by(country_text_id) %>%
  summarize(start_dem = dem[which.min(year)], .groups = "drop")
## keep countries autocratic at panel start (so a transition is observed) or never democratic
keep <- st %>% left_join(fd, by = "country_text_id") %>%
  filter(is.na(first_dem) | start_dem == 0) %>%
  mutate(first_treat = ifelse(is.na(first_dem), 0, first_dem))
v <- v %>% inner_join(keep %>% select(country_text_id, first_treat), by = "country_text_id")
## mean Electoral Democracy Index (polyarchy) by regime category, for the data slide
poly_by_reg <- round(tapply(v$v2x_polyarchy, v$v2x_regime, mean, na.rm = TRUE), 2)

## DiD panel: cohort = first_treat if inside the window; never-treated-in-window = 0;
## drop countries already democratic before the window (always-treated, not controls).
panel <- v %>%
  mutate(ft = case_when(first_treat == 0 ~ 0L, first_treat > WIN_HI ~ 0L,
                        first_treat < WIN_LO ~ NA_integer_, TRUE ~ as.integer(first_treat))) %>%
  filter(!is.na(ft), year >= WIN_LO, year <= WIN_HI) %>%
  mutate(first_treat = ft, country_num = as.integer(factor(country_text_id)),
         post = as.integer(first_treat > 0 & year >= first_treat))
n_never <- sum(tapply(panel$first_treat, panel$country_num, function(x) x[1] == 0))
n_cohort <- n_distinct(panel$first_treat[panel$first_treat > 0])
```

## Today

0. **Your questions.**
1. **Review (day 12):** non-bipartite matching --- does democratization reduce
   corruption? We compared non-democratic countries similar in background
   characteristics (corruption, civil society, gdp) as of 1990 in terms of
   their 2010 corruption scores. Did countries that became more democratic
   become less corrupt?
2. **Today:** treatment timing varies across units. Two designs answer the same
   longitudinal question under assumptions from **different families**:
   - **Risk-set matching:** at each transition, compare the new democracy to
     countries *still autocratic then*, alike on history (conditional
     ignorability, SUTVA).
   - **Difference-in-differences:** do not match; assume *parallel trends* and
     *no anticipation* and SUTVA.
   - Each gets a sensitivity analysis
4. **Your questions.**

\medskip

Same pre-vs-post data, same question. Different assumptions.

## The two designs side by side

\begin{center}
\small
\begin{tabular}{@{}p{2.5cm}p{4.0cm}p{2.1cm}p{1.7cm}@{}}
\toprule
design name & assumptions & check pre-outcomes & sensitivity \\
\midrule
risk-set matching & conditional ignorability & balance table & $\Gamma$ or ($\Lambda$, $\Delta$) or $R^2$ \\
difference-in-differences & parallel trends (+ no anticipation) & pre-trends & $\bar{M}$ \\
\bottomrule
\end{tabular}
\normalsize
\end{center}

\medskip

- Conditional ignorability is a claim about **probability of treatment** within set. Check with balance assessments.
- Parallel trends and no anticipation are claims about **how outcomes would have moved**. Check by comparing pre-trends.

# Review and the Staggered-Timing Problem

## The V-Dem data: what counts as "democracy" {.shrink}

Country-year panel from the Varieties of Democracy (V-Dem) project
[@vdem2025dataset], `r WIN_LO`--`r WIN_HI`. The treatment is built from
`v2x_regime`, V-Dem's **Regimes of the World** classification --- four ordered
categories:

\begin{center}
\small
\begin{tabular}{@{}clcc@{}}
\toprule
\texttt{v2x\_regime} & Regimes-of-the-World category & our coding & mean polyarchy \\
\midrule
0 & closed autocracy    & not democracy & `r poly_by_reg[["0"]]` \\
1 & electoral autocracy & not democracy & `r poly_by_reg[["1"]]` \\
2 & electoral democracy & democracy     & `r poly_by_reg[["2"]]` \\
3 & liberal democracy   & democracy     & `r poly_by_reg[["3"]]` \\
\bottomrule
\end{tabular}
\normalsize
\end{center}

\medskip

We set `dem = 1` for the two democracy rows --- `v2x_regime` 2 (electoral) and 3
(liberal) --- and 0 for the two autocracy rows (0 and 1). The cut is the move to
*electoral democracy*: de-facto multiparty, free and fair national elections.
Liberal democracy (3) adds rule of law and constraints on the executive. The line
falls where the Electoral Democracy Index jumps: mean polyarchy
`r poly_by_reg[["1"]]` just below it, `r poly_by_reg[["2"]]` just above.

\footnotesize
Outcome: `v2x_corr`, the Political Corruption Index (0 = low, 1 = high). Matching
covariates: `v2x_polyarchy` (Electoral Democracy Index), `v2x_cspart`
(civil-society participation), `v2x_rule` (rule of law), and log GDP per capita
(from `e_gdppc`). \normalsize

## Staggered transitions in V-Dem

```{r cohort_tile, echo=FALSE, out.width=".9\\textwidth", fig.width=6.5, fig.height=3.8}
## Treatment-timing plot (PanelMatch DisplayTreatment style): each country-year
## colored by democratization status, countries ordered by transition year. Drawn
## on a complete country x year grid so the staggered design reads cleanly; the
## DiD estimation still uses the real (unbalanced) panel.
cty <- panel %>% distinct(country_text_id, first_treat)
tl  <- tidyr::expand_grid(country_text_id = cty$country_text_id, year = WIN_LO:WIN_HI) %>%
  left_join(cty, by = "country_text_id") %>%
  mutate(status = factor(case_when(
           first_treat == 0   ~ "Never democratic",
           year < first_treat ~ "Autocratic (pre)",
           TRUE               ~ "Democratic (post)"),
         levels = c("Never democratic", "Autocratic (pre)", "Democratic (post)")))
ord <- cty %>% mutate(eff = ifelse(first_treat == 0, Inf, first_treat)) %>%
  arrange(eff, country_text_id)
tl %>% mutate(country_text_id = factor(country_text_id, levels = ord$country_text_id)) %>%
  ggplot(aes(year, country_text_id, fill = status)) +
  geom_tile(color = "white", linewidth = 0.05) +
  scale_fill_manual(values = c("Never democratic" = "grey88",
                               "Autocratic (pre)" = "grey60",
                               "Democratic (post)" = "steelblue"), name = "") +
  scale_x_continuous(breaks = seq(WIN_LO, WIN_HI, 5), expand = c(0, 0)) +
  labs(x = "Year", y = "countries, ordered by transition year") +
  theme_minimal(base_size = 9) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid = element_blank(), legend.position = "top")
```

\footnotesize
`r n_cohort` transition cohorts, plus
`r n_never` countries that stay autocratic across 1980--2010 --- the never-treated
pool. \normalsize

# Risk-Set Matching Over Time

## The idea

At each transition year $g$, match the newly-democratizing country to countries
**still at risk** at $g$ --- still autocratic, and not about to democratize ---
choosing controls on covariate history up to $g$
[@lietal2001; @rosenbaum2017; @rosenbaum2020book, ch. 13].

\medskip

Every comparison occurs among countries that were jointly "at risk." Within a
pair, which one democratized is the treatment; conditioning on covariate
history helps justify the as-if-random assignment style assumptions.

## Why "still at risk"

A country already democratic for years is not a counterfactual for one
democratizing now.

\medskip

So the eligible controls for cohort $g$ are countries that are still autocratic
at $g-1$ **and** do not democratize within the post-window (currently 5 years).
For today we use their covariates at $g-1$ but could use functions of covariate
history if we wanted. 

## One transition: build the risk set {.shrink}

Colombia democratizes in 1991. Its **risk set** includes the countries still
autocratic in 1990 that could be its control. Build that set directly:

```{r rs_pool_demo, echo=TRUE}
g <- 1991                                          # Colombia's transition year
## countries seen at g-1, flagged democratizer (newly) vs still-at-risk control
pool <- v %>% filter(year == g - 1) %>%
  mutate(newly = as.integer(first_treat == g)) %>%
  filter(first_treat == g |                        # the cohort-g democratizer
         first_treat == 0 |                        # never democratic in the window
         first_treat > g + POST)                   # not democratic until after the post-window
## keep only countries observed at both endpoints, g-1 and g+POST
ok <- v %>% filter(country_text_id %in% pool$country_text_id, year %in% c(g - 1, g + POST)) %>%
  count(country_text_id) %>% filter(n == 2) %>% pull(country_text_id)
pool <- pool %>% filter(country_text_id %in% ok)
unique(pool$year) ## Only 1990 countries
```

## One transition: look at the pool {.shrink}

Colombia and its five closest still-autocratic candidates on the
electoral-democracy index, covariates as of 1990:

```{r rs_pool_show, echo=TRUE}
dz <- pool$v2x_polyarchy[pool$newly == 1]          # the democratizer's polyarchy
pool %>% arrange(desc(newly), abs(v2x_polyarchy - dz)) %>%   # democratizer, then closest
  transmute(country_name, newly, v2x_polyarchy = round(v2x_polyarchy, 2),
            v2x_corr = round(v2x_corr, 2), log_gdp = round(log_gdp, 1)) %>% head(12)
```

\footnotesize
`r nrow(pool) - 1` still-autocratic countries are eligible for possible matching. \normalsize

## How wide should the polyarchy caliper be? {.shrink}

We will force pairs to be close on polyarchy. Before fixing a width, look at
the distances the caliper will act on: every treated-control polyarchy gap,
within each transition year's pool.

```{r pol_caliper, echo=TRUE}
cohorts <- sort(unique(v$first_treat[v$first_treat >= WIN_LO + 1 & v$first_treat <= WIN_HI - POST]))
pools <- map_dfr(cohorts, function(g) {              # the pool lines, run at every g
  v %>% filter(year == g - 1) %>%
    mutate(newly = as.integer(first_treat == g), cohort = g) %>%
    filter(first_treat == g | first_treat == 0 | first_treat > g + POST)
}) %>% as.data.frame()
mat <- match_on(newly ~ v2x_polyarchy, data = pools, method = "euclidean",
                within = exactMatch(newly ~ cohort, data = pools))   # blocked by cohort
summary(mat)
```

## How wide should the polyarchy caliper be? {.shrink}

```{r pol_caliper2, echo=TRUE}
dists <- findSubproblems(mat)                        # one distance matrix per cohort
gaps  <- unlist(lapply(dists, as.vector))            # every treated-control gap
round(quantile(gaps, seq(0, 1, .1)), 2)              # deciles, all cohorts pooled
nearest <- sapply(dists, function(x) { min(as.vector(x)) })  # nearest gap, per cohort
round(nearest, 2)
```

\footnotesize
A width of $0.05$ keeps the closest
`r round(100 * mean(gaps <= 0.05))`\% of the available comparisons (the 10th
percentile of gaps is `r round(quantile(gaps, .1), 2)`; the median is
`r round(median(gaps), 2)`). 

\normalsize

## Build it (1): the pool at $g$ {.shrink}

```{r rs_pool_fn, echo=TRUE}
find_pool_at <- function(g, used = character(0)) {
  pool <- v %>% filter(year == g - 1) %>%                   # <- the pool from the last slide
    mutate(newly = as.integer(first_treat == g)) %>%
    filter(first_treat == g | first_treat == 0 | first_treat > g + POST) %>%
    filter(newly == 1 | !(country_text_id %in% used))       # a control serves in one pair only
  ok <- v %>% filter(country_text_id %in% pool$country_text_id, year %in% c(g - 1, g + POST)) %>%
    count(country_text_id) %>% filter(n == 2) %>% pull(country_text_id)
  pool <- pool %>% filter(country_text_id %in% ok) %>% mutate(g = g) %>% as.data.frame()
  rownames(pool) <- pool$country_text_id
  if (sum(pool$newly) < 1 || sum(pool$newly == 0) < 1) return(NULL)  # nobody to compare
  pool
}
```

## Build it (2): match within a pool, calipers as arguments {.shrink}

```{r rs_match_fn, echo=TRUE}
mvars <- c("v2x_polyarchy", "v2x_corr", "v2x_cspart", "v2x_rule", "log_gdp")
match_pool <- function(pool, ps_width = 2, pol_width = 0.05) {
  fml  <- reformulate(mvars, response = "newly")
  dmat <- match_on(fml, data = pool, method = "rank_mahalanobis")       # covariate distance
  ps   <- caliper(match_on(glm(fml, pool, family = binomial()), data = pool),
                  width = ps_width)                         # propensity-score caliper
  pol  <- match_on(newly ~ v2x_polyarchy, data = pool, method = "euclidean")   # raw units
  total <- dmat + ps + caliper(pol, width = pol_width)  # democratizers sit just below the line
  m <- tryCatch(pairmatch(total, controls = 1, data = pool), error = function(e) NULL)
  if (is.null(m)) return(NULL)
  pool$pair <- as.character(m)                              # always character (NA-safe)
  out <- pool %>% filter(!is.na(pair)) %>%
    mutate(pair = paste0(g, "_", pair)) %>%
    select(country_text_id, country_name, pair, g, newly, all_of(mvars))
  if (nrow(out) < 2) return(NULL)                           # this cohort formed no pair
  out
}
```

## Build it (3): repeat across transitions {.shrink}

```{r usefn, echo=TRUE}
matched <- NULL; used <- character(0)        # controls already matched at earlier transitions
for (g in cohorts) {                         # transitions in time order, as they happen
  gpool <- find_pool_at(g, used)             # 1. who is at risk at g
  if (is.null(gpool)) next
  out <- suppressWarnings(match_pool(gpool, ps_width = 2, pol_width = 0.05))  # 2. match
  if (is.null(out)) next                     # no defensible pair at this transition
  used <- c(used, out$country_text_id[out$newly == 0])     # matched controls leave the pool
  matched <- bind_rows(matched, out)
}
c(pairs = n_distinct(matched$pair), treated = sum(matched$newly),
  controls = sum(1 - matched$newly))
```

```{r rs_check, echo=FALSE, include=FALSE}
## the design rule the loop above must deliver: no country serves as a control twice
stopifnot(!anyDuplicated(matched$country_text_id[matched$newly == 0]))
```

\footnotesize 

Each control country serves in **one pair only**: once matched,
it leaves the control pool for every later transition. A country used as a
control can still appear later as the *treated* side of its own pair:
eligibility already requires its transition to fall after $g + `r POST`$, past
its outcome window as a control. The analyses
below treat pairs as independent. \normalsize

## Balance: does history predict who democratized?

```{r rs_balance, echo=TRUE}
balfmla <- reformulate(c(mvars, "strata(pair)"), response = "newly")
xb <- balanceTest(balfmla, data = as.data.frame(matched))
xb$overall["pair", ]
round(xb$results[, c("Treatment", "Control", "std.diff", "p"), "pair"], 2)
```

\footnotesize
Covariates do not predict who democratized within pair. 
\normalsize

## Estimate: the difference in the differences via a change score {.shrink}

Outcome per country: the **change score** --- corruption at $g+`r POST`$ minus
at $g-1$. Estimate the effect of democratizing per pair: the
democratizer's change minus its match's change --- **the difference in the
differences**. Averaging the pair differences estimates the **ATT**: the
average effect of democratization on the matched democratizers.

```{r rs_est, echo=TRUE}
## change score: corruption at g+POST minus at g-1, for each matched country
chg <- matched %>% rowwise() %>% mutate(
  dcorr = v$v2x_corr[v$country_text_id == country_text_id & v$year == g + POST] -
          v$v2x_corr[v$country_text_id == country_text_id & v$year == g - 1]) %>% ungroup()
pd <- chg %>% group_by(pair) %>%
  summarize(diff = dcorr[newly == 1] - dcorr[newly == 0], .groups = "drop")
est_rs <- mean(pd$diff)                                   # mean within-pair difference
hl <- wilcox.test(pd$diff, conf.int = TRUE)              # Hodges-Lehmann + signed-rank p
c(mean_diff = est_rs, HL = hl$estimate[[1]],
  ci_lo = hl$conf.int[1], ci_hi = hl$conf.int[2], p = hl$p.value)
```

## Estimate: the difference in the differences {.shrink}

The newly-democratic country's corruption falls about
`r round(abs(est_rs), 2)` more than its still-autocratic match's over five years
(signed-rank $p \approx `r round(wilcox.test(pd$diff)$p.value, 3)`$).

\medskip

The same ATT by regression:

```{r rs_lmrobust, echo=TRUE}
est_lm <- lm_robust(dcorr ~ newly, fixed_effects = ~ pair, data = chg)
est_lm
```

\footnotesize
With one treated and one control per pair, the pair-fixed-effects coefficient
on `newly` *is* the mean within-pair difference --- the same
`r round(est_rs, 2)`, now with an HC2 standard error (two-sided
$p \approx `r round(est_lm$p.value[["newly"]], 3)`$, near the signed-rank
$p$). \normalsize

## Sensitivity: assess consequences of possible hidden bias

```{r rs_gamma, echo=TRUE}
## alternative = "less": the democratizer's corruption falls relative to its match.
pval_g <- function(g) as.numeric(senstrat(sc = chg$dcorr, z = chg$newly,
                     st = as.integer(factor(chg$pair)), gamma = g,   # senstrat needs integer strata
                     alternative = "less")$Result["P-value"])
gstar  <- uniroot(function(g) pval_g(g) - 0.05, c(1, 3))$root   # Gamma where p crosses 0.05
round(c(p_at_1 = pval_g(1), p_at_1.5 = pval_g(1.5), Gamma_star = gstar), 3)
```

\footnotesize
$\Gamma^{*} \approx `r round(gstar, 2)`$:
an unobserved confounder would have to give one country in a pair about
`r round(gstar, 2)`
times the odds of being the democratizer before the test stops rejecting. \normalsize

# Difference-in-Differences

## The other strategy: assume parallel trends

We can make a causal inference **without adjusting for covariates** and the conditional ignorability assumption.

\medskip

Assume that, absent democratization, the corruption
paths of democratizers and non-democratizers would have moved in **parallel** over time.
Then the democratizers' **average** change (post minus pre), minus the
non-democratizers' average change, identifies the **ATT**: the average effect
of democratization on the countries that democratized.

\medskip

The appeal: no need to find comparable countries. The cost: a claim about
counterfactual *trajectories* that we can never fully check.

## A two-period difference-in-differences {.shrink}

The simplest version: two periods, two groups. Take the post-Cold-War wave (the
countries that democratized in 1990) and the never-democratizers, and compare
corruption just before (1989) and a few years after (1995).

```{r did_2x2, echo=TRUE}
yr_pre <- 1989; yr_post <- 1995
bb <- panel %>% filter(year %in% c(yr_pre, yr_post), first_treat %in% c(0, 1990)) %>%
  group_by(country_num) %>% filter(n() == 2) %>% ungroup() %>%     
  mutate(grp = ifelse(first_treat == 1990, "democratized", "never"),
         per = ifelse(year == yr_post, "post", "pre"))
cells <- tapply(bb$v2x_corr, list(bb$grp, bb$per), mean)
did2  <- (cells["democratized","post"] - cells["democratized","pre"]) -
         (cells["never","post"]        - cells["never","pre"])
round(cells, 2)
```

\footnotesize
`cells` holds the four averages the estimate is built from --- each group's mean
corruption before (1989) and after (1995). We turn them into the DiD next.
\normalsize

## From the four cells to the estimate

`cells` is four averages: mean corruption for each group in each period. Read each
row across for its **change over time** (post minus pre); then difference the changes.

```{r did_cells_again, echo=TRUE}
round(cells[, c("pre", "post")], 2)                    # (a) four means, pre then post
changes <- cells[, "post"] - cells[, "pre"]            # (b) each group's change: post - pre
round(changes, 2)
round(unname(changes["democratized"] - changes["never"]), 2)  # (c) DiD = difference of changes
```

\medskip

- Democratizers' average change: $`r round(cells["democratized","post"],2)` - `r round(cells["democratized","pre"],2)` = `r round(cells["democratized","post"]-cells["democratized","pre"],2)`$ (roughly steady).
- Non-democratizers' average change: $`r round(cells["never","post"],2)` - `r round(cells["never","pre"],2)` = `r round(cells["never","post"]-cells["never","pre"],2)`$ (rose).
- DiD $= `r round(cells["democratized","post"]-cells["democratized","pre"],2)` - (`r round(cells["never","post"]-cells["never","pre"],2)`) = `r round(did2,2)`$: the new democracies fell about $`r round(abs(did2),2)`$ relative to the non-democratizers.

## Parallel trends as a picture

Parallel trends fills in the missing potential outcome: start at the
democratizers' 1989 level and apply the never-democratizers' observed change
($+`r round(cells["never","post"] - cells["never","pre"], 2)`$). That is the
dashed line. The estimate is the arrow: observed minus counterfactual
$= `r round(did2, 2)`$.

```{r did_pt_plot, echo=FALSE, out.width=".8\\textwidth", fig.width=6.2, fig.height=3.4}
## The 2x2 as a picture. The dashed segment is the parallel-trends assumption:
## the democratizers' counterfactual, built from their own 1989 level plus the
## never-democratizers' change. The arrow between observed and counterfactual
## 1995 corruption is the DiD estimate. Colors follow the deck: steelblue =
## democratizers, grey = never-treated reference; the counterfactual is dashed
## with an open point so the encoding survives greyscale printing.
## tapply orders columns alphabetically (post, pre): index by name, never position.
dem <- c(cells["democratized", "pre"], cells["democratized", "post"])
nev <- c(cells["never", "pre"], cells["never", "post"])
cf  <- dem[1] + (nev[2] - nev[1])       # dem's 1989 level + never's observed change
par(mar = c(2.6, 3.6, 1, 1), mgp = c(2.1, .6, 0))
plot(NA, xlim = c(yr_pre - 0.2, yr_post + 4.6),
     ylim = range(c(dem, nev, cf)) + c(-0.015, 0.02),
     xaxt = "n", xlab = "", ylab = "mean corruption (v2x_corr)")
axis(1, at = c(yr_pre, yr_post), labels = c("1989 (pre)", "1995 (post)"))
abline(v = 1990, lty = 3, col = "grey60")
lines(c(yr_pre, yr_post), nev, type = "b", pch = 19, col = "grey40", lwd = 1.6)
lines(c(yr_pre, yr_post), dem, type = "b", pch = 19, col = "steelblue", lwd = 1.6)
lines(c(yr_pre, yr_post), c(dem[1], cf), lty = 2, col = "steelblue", lwd = 1.6)
points(yr_post, cf, pch = 21, bg = "white", col = "steelblue", lwd = 1.6)
arrows(yr_post - 0.3, cf, yr_post - 0.3, dem[2], length = 0.05, code = 3)
text(yr_post - 0.45, (cf + dem[2]) / 2, paste0("DiD = ", round(did2, 2)), cex = .8, pos = 2)
text(yr_post + 0.15, nev[2] + 0.004, "never democratic (observed)", cex = .7, col = "grey40", pos = 4)
text(yr_post + 0.15, cf - 0.004, "counterfactual (parallel trends)", cex = .7, col = "steelblue", pos = 4)
text(yr_post + 0.15, dem[2], "democratized 1990 (observed)", cex = .7, col = "steelblue", pos = 4)
text(1990, max(c(dem, nev, cf)) + 0.015, "1990: the wave democratizes", cex = .7, col = "grey40", pos = 4)
```

\footnotesize
The two solid lines are all we observe; the whole design is the
claim that the blue line, absent democratization, would have run parallel to
the grey one. Whether that claim is even plausible is what more pre-periods
can address --- the event study, next. \normalsize

## Checking the assumption: the event study

A single pre/post comparison cannot show whether the trends were parallel
*before* treatment, so add pre-periods. For each cohort (group of countries), compare its average corruption
change --- from the year before it democratized (its base year) to each other
year --- against the never-treated countries' average change over the same years. The
horizontal axis below is years relative to each cohort's transition: negative
before, zero at the transition, positive after.

## Checking the assumption: the event study

```{r did_cs, echo=TRUE, cache=TRUE, message=FALSE, warning=FALSE, results='hide'}
## event study + pooled estimate, never-treated controls, all cohorts
cs <- att_gt(yname = "v2x_corr", tname = "year", idname = "country_num",
             gname = "first_treat", control_group = "nevertreated",
             base_period = "universal", allow_unbalanced_panel = TRUE, data = panel)
cs_simple <- aggte(cs, type = "simple",  na.rm = TRUE)   # pooled DiD estimate
cs_dyn    <- aggte(cs, type = "dynamic", na.rm = TRUE, min_e = -5, max_e = 5)  # leads/lags
```

```{r did_es_plot, echo=FALSE, out.width=".60\\textwidth", fig.width=6.2, fig.height=3.4}
esd <- data.frame(e = cs_dyn$egt, att = cs_dyn$att.egt, se = cs_dyn$se.egt)
esd <- esd[is.finite(esd$att) & is.finite(esd$se), ]            # drop endpoints with no SE
yl <- range(c(esd$att - 1.96*esd$se, esd$att + 1.96*esd$se), na.rm = TRUE)
par(mar = c(3.4, 3.6, 1, 1), mgp = c(2.1, .6, 0))
plot(esd$e, esd$att, type = "n", ylim = yl,
     xlab = "years since democratization", ylab = "DiD vs never-treated (corruption)")
abline(h = 0, col = "grey60"); abline(v = -0.5, lty = 3, col = "grey60")
arrows(esd$e, esd$att - 1.96*esd$se, esd$e, esd$att + 1.96*esd$se,
       length = 0.02, angle = 90, code = 3, col = "grey50")
points(esd$e, esd$att, pch = 19, col = ifelse(esd$e < 0, "darkred", "black"))
```

\footnotesize
Red = pre-transition. The pre-trends are **not** flat: would-be democracies
already had higher, *declining* corruption before the transition --- not parallel trends. The pooled estimate (all cohorts) is about
`r round(cs_simple$overall.att, 2)`. \normalsize

## What each comparison is: the 1990 cohort {.shrink}

One comparison, concretely. The 2x2 you saw --- the 1990 democratizers against
the never-treated, 1989 to 1995 --- is one of these building blocks:
$\mathrm{ATT}(1990, 1995) \approx -0.15$. Its two groups:

```{r cs_cohort90, echo=TRUE}
## treated group: the countries that democratized in 1990
panel %>% distinct(country_name, first_treat) %>%
  filter(first_treat == 1990) %>% pull(country_name)
## control group: EVERY never-treated country (the same pool for every cohort)
never <- panel %>% distinct(country_name, first_treat) %>% filter(first_treat == 0)
c(n_never_treated = nrow(never))
head(sort(never$country_name), 6)  # a few, for illustration
```

\footnotesize
CS compares all six 1990 democratizers to all `r nrow(never)` never-treated
countries at once --- no matching. The base year is 1989 ($g-1$), and **this same
never-treated pool is the control group for every cohort**. \normalsize

## Matching used different comparisons {.shrink}

Section A's risk-set matching took the same democratizers but built different
comparisons: each is paired to one still-at-risk country chosen on covariates,
and each control is used only once.

```{r rs_cohort90, echo=TRUE}
matched %>% filter(g == 1990) %>% group_by(pair) %>%
  summarize(democratizer    = country_name[newly == 1],
            matched_control = country_name[newly == 0], .groups = "drop") %>%
  select(democratizer, matched_control)
```

\footnotesize
Five pairs, not six: the German Democratic Republic reunified in 1990 and has no
1995 outcome, so it gets no matched control --- yet it is still counted in the CS
1990 cohort. The matched controls are covariate-close autocracies, each used
once. \normalsize

## The comparisons differ {.shrink}

Same democratizers, different counterfactuals:

 - **Control pool.** CS uses all `r nrow(never)` never-treated countries, the
   same set for every cohort and every year. Matching gives each democratizer
   one still-at-risk control and uses it in a single pair.
 - **Who can be a control.** CS uses never-treated countries only --- so Mexico,
   which democratizes in 1996, is *not* a CS control, yet it is Poland's matched
   control in 1990. 
 - **How controls are chosen.** CS takes all of them, unmatched --- its
   credibility rests on parallel trends. Matching keeps only countries close on
   polyarchy and the other covariates (the calipers) --- its credibility rests
   on the balance table.
 - **Timing.** CS reads the change from $g-1$ to every year $t$ (the whole
   event-time path). Matching reads one horizon, $g-1$ to $g+5$, within the pair.

## A note on staggered timing

We used only **never-treated** countries as controls --- countries carrying no
treatment effect of their own. The tempting
shortcut, two-way fixed effects with a 0/1 treatment, instead lets
*already-democratic* countries act as controls, which can distort the estimate when
effects grow or fade over time.

\medskip

Untangling that is its own literature --- the Goodman-Bacon decomposition
[@goodmanbacon2021] and estimators built to avoid the trap
[@callawaysantanna2021; @sunabraham2021; @dechaisemartin2020]. We keep today
simple by relying on never-treated controls. The reading is there when you build
staggered designs of your own.

# Sensitivity for Parallel Trends

## Passing a pre-trend test is not proof

The event study showed a pre-trend. Even when it does not, "no significant
pre-trend" judged by looking at pattern of pre-treatment CIs is weak comfort:
pre-trend tests have low power, and selecting analyses that pass them distorts
inference [@rothetal2023].

\medskip

So instead of asking "are pre-trends flat?", ask "**how badly would parallel
trends have to fail to overturn the result?**" That is HonestDiD.

## HonestDiD: bound the bias by the pre-trend {.shrink}

$\bar M$ answers a question: *how much could parallel trends fail after
treatment before our result goes away?* We cannot see the post-treatment
violation, but we can see the pre-treatment one --- how far from parallel the two
groups' trends already were before treatment. $\bar M$ lets the post-treatment
violation be at most $\bar M$ times that largest pre-treatment violation
[@rambachanroth2023]. 

```{r did_hdid, echo=TRUE, cache=TRUE, message=FALSE, warning=FALSE}
## HonestDiD needs the event-study coefficients AND their covariance matrix.
## did returns the coefficients (att.egt) but gives the covariance only through
## an influence function (n units x k coefficients); its crossproduct, scaled
## by 1/n^2, is the usual influence-function estimate of that covariance.
ifc <- cs_dyn$inf.function$dynamic.inf.func.e    # influence fn: row = unit, col = coef
Sig <- t(ifc) %*% ifc / nrow(ifc)^2              # -> covariance matrix of the coefs
npre <- sum(cs_dyn$egt < 0); npost <- sum(cs_dyn$egt >= 0)  # leads (pre), lags (post)
hd <- createSensitivityResults_relativeMagnitudes(
  betahat = cs_dyn$att.egt, sigma = Sig, numPrePeriods = npre, numPostPeriods = npost,
  Mbarvec = c(0, 0.25, 0.5),      # how much post-period drift to allow (M-bar; see below)
  l_vec   = rep(1/npost, npost))  # target = simple average of the post-period effects
hd[, c("Mbar", "lb", "ub")]       # robust CI [lb, ub] at each M-bar
```

## HonestDiD: bound the bias by the pre-trend {.shrink}

\footnotesize
$\bar M = 0$ is exact parallel trends (the usual DiD
interval); larger $\bar M$ allows more post-treatment divergence and widens the
interval. Here the interval excludes zero at $\bar M = 0$ but already covers
zero by $\bar M = 0.25$: we only have to allow the two groups' trends to differ
*after* treatment by one quarter of how much they visibly differed *before* it,
and the effect is no longer distinguishable from zero. A low breakdown, so the
DiD result is fragile. \normalsize

```{r mbar2, echo=TRUE}
hd[, c("Mbar", "lb", "ub")]       # robust CI [lb, ub] at each M-bar
```

# Putting It Together

## Same data, two designs, two verdicts

\begin{center}
\small
\begin{tabular}{@{}p{3.0cm}p{2.0cm}p{2.6cm}p{3.0cm}@{}}
\toprule
design & estimate & assumption & robustness \\
\midrule
risk-set matching & $\approx `r round(est_rs,2)`$ & conditional ignorability & $\Gamma^{*} \approx `r round(gstar, 2)`$ (small) \\
difference-in-differences & $\approx `r round(cs_simple$overall.att,2)`$ & parallel trends & breakdown $\bar M < 0.25$ (fragile) \\
\bottomrule
\end{tabular}
\normalsize
\end{center}

\medskip

The two designs agree the effect is negative and similar in size. Both are
fairly sensitive to violations in their different assumptions.

## Why the verdicts differ

Both designs face the same threat: countries on the path to democracy were
already changing before they crossed the regime line --- already scoring just
below the democracy threshold on `v2x_polyarchy`, with corruption already
falling.

\medskip

 - **Matching answers the threat in the design.** Each democratizer is paired
   with a still-autocratic country at nearly the same polyarchy score (the
   $0.05$ caliper) --- just as close to democratizing, but it did not --- and
   the balance table shows the pairs comparable on what we measured. What is
   left is what we did not measure: $\Gamma^{*} \approx `r round(gstar, 2)`$
   says a modest unmeasured within-pair difference would overturn the result.
 - **DiD leaves the threat in** and bets the two groups would have moved in
   parallel. The event study shows the bet failing --- corruption already
   falling among the would-be democracies. HonestDiD then asks how much
   post-treatment divergence the result can absorb --- very little: allowing the
   two groups' trends to differ after treatment by just a quarter of how much
   they visibly differed before it already makes the confidence interval
   include zero.

\medskip

The verdicts differ in kind: matching's weakness is a difference we cannot
see; DiD's is one we already saw.

## Built vs assumed

A matched design **builds** a comparison a reader can check: here are the pairs,
here is the balance table, here is the bias it would take to overturn us. Parallel
trends **assumes** a comparison a reader must take partly on faith: here are the
pre-trends, and here is how much they would have to continue to matter.

\medskip

Design and analysis are coupled in the first --- we choose which countries can
serve as counterfactuals and discard those we cannot defend --- and separable in
the second. Neither is always right. What both designs owe the reader is the
same: state the assumption, check what can be checked, and report the sensitivity.

## When to prefer which

 - **Few pre-treatment periods, good covariates** -> risk-set matching: condition on
   history and report $\Gamma$.
 - **Many pre-treatment periods, confounders you cannot measure** -> DiD: rely on
   trends and report HonestDiD.
 - **Either way, show the sensitivity.** A point estimate without it --- matched or
   differenced --- is half an answer.

## Summary

 - Staggered timing breaks the fixed-baseline design of day 12. Two repairs:
   move the risk set, or assume parallel trends.
 - **Risk-set matching** rebuilds the comparison at each event time, uses each
   control country in at most one pair, balances on history (the hard variable:
   polyarchy, already just below the democracy threshold), and reports $\Gamma$.
 - **DiD** assumes parallel trends; staggered TWFE needs care (Goodman-Bacon,
   Callaway--Sant'Anna), and the assumption needs HonestDiD.
 - On V-Dem both say democratization lowers corruption by a similar amount, and
   neither conclusion withstands much stress ($\Gamma^{*} \approx
   `r round(gstar, 2)`$; breakdown below $\bar M = 0.25$). The matched design
   remains the easier one to defend: balance held when we checked it, while the
   pre-trend check showed parallel trends visibly failing.
 - The week's lesson, over time: state the assumption, check it before outcomes,
   stress-test what you cannot check.

## References {.allowframebreaks}
