---
title: |
  | "Controlling For": Linear Regression Adjustment and Its Limits
date: '`r format(Sys.Date(), "%B %d, %Y")`'
author: |
  | ICPSR 2026 Session 1
  | Jake Bowers \& Tom Leavitt
bibliography:
 - 'BIB/MasterBibliography.bib'
fontsize: 10pt
geometry: margin=1in
graphics: yes
biblio-style: authoryear-comp
biblatexoptions:
  - natbib=true
output:
  beamer_presentation:
    slide_level: 2
    keep_tex: true
    latex_engine: xelatex
    citation_package: biblatex
    template: styles/icpsr-beamer-template
    incremental: true
    includes:
        in_header:
           - defs-all.sty
    md_extensions: +raw_attribute-tex_math_single_backslash+autolink_bare_uris+ascii_identifiers+tex_math_dollars
    pandoc_args: [ "--csl", "chicago-author-date.csl" ]
---

```{r setup1_env, echo=FALSE, include=FALSE}
library(here)
source(here::here("rmd_setup.R"))
```

```{r setup2_loadlibs, echo=FALSE, include=FALSE}
library(tidyverse)
library(dplyr)
library(estimatr)
library(splines)
library(optmatch)
library(olsrr)
options(show.signif.stars = FALSE)
```

```{r simdata, echo=FALSE, include=FALSE, cache=TRUE}
## A self-contained fake observational study where we KNOW the truth.
## Z (downloaded a COVID info kit) is confounded by x1: people with higher x1
## are more likely to have Z=1, AND x1 drives the outcome y0 NON-linearly
## (the -0.2*x1^2 term). So linear "controlling for" cannot fully remove the
## confounding, and the right functional form is unknown to the analyst.
N <- 100
tau <- .3
set.seed(12345)
dat <- data.frame(
  id = 1:N,
  x1 = rpois(n = N, lambda = 10),
  x2 = sample(1:6, size = N, replace = TRUE)
)
dat <- mutate(dat,
  y0 = .2 * x1 - .2 * x1^2 + .2 * (x2 < 2) + runif(n = N, min = -2 * sd(x1), max = 2 * sd(x1)),
  y0 = round(y0 + abs(min(y0)) / max(y0)),
  y0 = abs(ifelse(x1 < 3, 0, y0)),
  y1 = round(y0 + tau * sd(y0) + runif(n = N, min = -2 * tau * sd(y0), max = .5 * sd(y0))),
  x3 = rnorm(n(), mean = mean(x2), sd = sd(x2)),
  x4 = rbinom(n(), size = 1, prob = mean(x1 > 10))
)
dat$Z <- with(dat, as.numeric((.4 * sd(x1) * x1 + runif(n = N, min = -20, max = 0)) > 0))
dat <- mutate(dat, Y = Z * y1 + (1 - Z) * y0)
dat$tau <- with(dat, y1 - y0)

## The truth we are trying to recover (never observed in practice):
true_ate <- mean(dat$tau)

## Four ways of looking at the Z -> Y relationship
dim0      <- coef(lm_robust(Y ~ Z, data = dat))[["Z"]]            # unadjusted
b_x1      <- coef(lm(Y ~ Z + x1, data = dat))[["Z"]]              # +x1, linear
b_x1sq    <- coef(lm(Y ~ Z + x1 + I(x1^2), data = dat))[["Z"]]    # +x1 + x1^2
b_all     <- coef(lm(Y ~ Z + x1 + x2 + x3 + x4, data = dat))[["Z"]]
```

## Today

1. **Review:** sensitivity analysis for matched designs --- how much hidden
   bias would overturn our conclusion ($\Gamma$)? That question presumes we
   already *have* a design.
2. **Today:** the much more common alternative to a design --- "controlling
   for" covariates in a linear regression. We ask:
   - What does "controlling for" actually compute? (residualization; the
     Frisch-Waugh-Lovell theorem)
   - What three questions does it raise --- about functional form,
     extrapolation, and influential points?
   - Why are those questions answerable with **one** covariate but not with
     several?
   - When is regression adjustment a defensible tool, and what does
     stratification do instead?

\medskip

Questions? Comments?

# Review: Sensitivity Analysis for Matched Designs

## What we did Monday

We built a matched design for the UN-peacekeeping study: episodes grouped into
sets that are similar on nine covariates, so that within a set a UN mission
looks **as-if** randomly assigned.

\medskip

Then, and only then, we looked at the outcome and tested the hypothesis of no
effect. The one-sided test rejected ($p \approx 0.03$): UN presence is
associated with longer peace, by as much as a block-randomized experiment would
let us claim.

\medskip

The order matters. We made every design decision --- which covariates, which
calipers, how many controls per treated, whether balance was good enough ---
using only the covariates and treatment, **before** seeing a single outcome.

## Sensitivity analysis: the formal worry {.fragile}

> "In an observational study, a sensitivity analysis replaces qualitative
> claims about whether unmeasured biases are present with an objective
> quantitative statement about the magnitude of bias that would need to be
> present to change the conclusions." (Rosenbaum, `sensitivitymv` manual)

\medskip

\begin{center}
\begin{tikzcd}[column sep=large,every arrow/.append style=-latex]
u \arrow[from=1-1,to=2-2, bend right, "\Gamma" near start] \arrow[from=1-1,to=2-3, bend left, "\rho" near start] \\
& Z \arrow[from=2-2,to=2-3, "\tau"] & y \\
x \arrow[from=3-1,to=2-2, "{\color{red}\text{\scriptsize$\times$}}"{marking, pos=0.3}] \arrow[from=3-1,to=2-3,grey]
\end{tikzcd}
\end{center}

We matched on $\bx$. $\Gamma$ asks: how many times more likely (in treatment
odds) would one set-mate have to be to receive treatment, because of an
unobserved $u$, before the test stops rejecting?

## The Monday result, in one line

For the UN study the rejection was **real but fragile**: a confounder making one
episode only about $1.5$ times more likely than its match to receive a mission
($\Gamma^{*} \approx 1.5$) would already overturn the one-sided test. The
`sensemakr` robustness value told the same story.

\medskip

The point to carry into today: **sensitivity analysis presumes a design.** It
perturbs the assignment probabilities of a matched comparison we already built
and checked. There is one estimate, one test, and a curve describing how that
single conclusion bends under hidden bias.

\medskip

Now we turn to the tool most observational papers use *instead* of a design.

# "Regression is basically research design." Is it?

## The temptation

A colleague has data on whether people downloaded a COVID information kit (`Z`)
and how far they would travel to be tested (`Y`). The unadjusted difference is
easy:

```{r unadj, echo=TRUE}
coef(lm_robust(Y ~ Z, data = dat))[["Z"]]
```

It looks like the kit *reduces* willingness to travel. But people who download
kits differ from those who do not --- in age, income, health interest. The
common move is to put those covariates `x1, ..., x4` into the regression and
read off the coefficient on `Z`:

```{r adj1, echo=TRUE}
coef(lm(Y ~ Z + x1 + x2 + x3 + x4, data = dat))[["Z"]]
```

"We controlled for $x_1$ through $x_4$." The sign even flipped. Are we done?

## A study where we know the answer

This is simulated data, so we know each person's $y_i(1)$ and $y_i(0)$ and the
true average effect:

```{r truth, echo=TRUE}
true_ate <- mean(dat$tau)   # average of y_i(1) - y_i(0)
round(true_ate, 2)
```

\medskip

The truth is $\approx `r round(true_ate,2)`$. Here is what four standard moves
report:

```{r fourmoves, echo=FALSE}
round(c(`unadjusted`        = dim0,
        `+ x1 (linear)`     = b_x1,
        `+ x1 + x1^2`       = b_x1sq,
        `+ x1..x4 (linear)` = b_all,
        `TRUTH`             = true_ate), 2)
```

\footnotesize
Unadjusted is the wrong sign. Adjusting helps, but no specification lands on the
truth, and they disagree with each other. Which one would we report --- if we
could not see this `TRUTH` column? \normalsize

## The confounding, as a picture {.fragile}

`x1` (think: health interest) drives **both** who downloads the kit and how the
outcome behaves. And its effect on the outcome is **not** a straight line.

\begin{center}
\begin{tikzcd}[column sep=large,every arrow/.append style=-latex]
 Z \arrow[from=1-1,to=1-3, "\tau = ?"] & & Y \\
 & x_1 \arrow[from=2-2,to=1-1] \arrow[from=2-2,to=1-3, "\text{curved}"'] &
\end{tikzcd}
\end{center}

\medskip

To "control for $x_1$" we have to remove its influence on the $Z \to Y$
comparison. So the first question is concrete: **what does a linear regression
actually do when we add $x_1$?**

# What "Controlling For" Computes

## "Controlling for" is residualization (one covariate) {.fragile}

The coefficient on `Z` in `Y ~ Z + x1` is *not* a comparison at fixed `x1`. It
is what remains after removing the **linear** relationship of `x1` with `Y` and
with `Z`, then relating the two leftovers:

```{r fwl, echo=TRUE}
b_direct <- coef(lm(Y ~ Z + x1, data = dat))[["Z"]]
## residualize: strip the linear part of x1 out of Y and out of Z
eY <- resid(lm(Y ~ x1, data = dat))
eZ <- resid(lm(Z ~ x1, data = dat))
b_resid <- coef(lm(eY ~ eZ))[[2]]
c(direct = b_direct, residualized = b_resid)
```

\footnotesize
This identity is the **Frisch-Waugh-Lovell theorem** [@lovell2008;
@angrist2009mostly, "regression anatomy"]. As the FWL theorem makes precise: to
"hold $x_1$ constant" with OLS is to remove the least-squares linear projection
of $x_1$ from both $Z$ and $Y$ --- no more, no less. It does **not** compare
people with identical $x_1$. \normalsize

## What residualization looks like (one covariate)

```{r residplots, echo=FALSE, out.width=".8\\textwidth"}
lm_Y_x1 <- lm(Y ~ x1, data = dat)
lm_Z_x1 <- lm(Z ~ x1, data = dat)
dat$eY <- resid(lm_Y_x1)
dat$eZ <- resid(lm_Z_x1)
cols <- c("black", "blue")[dat$Z + 1]
pchs <- c(1, 19)[dat$Z + 1]
par(mfrow = c(2, 2), mar = c(3, 3, 1.2, 0.4), mgp = c(1.6, .5, 0), oma = rep(0, 4))
with(dat, plot(x1, Y, col = cols, pch = pchs, main = "Y on x1")); abline(lm_Y_x1)
with(dat, plot(x1, eY, col = cols, pch = pchs, ylab = "Y minus its linear fit on x1",
               main = "Residual Y")); abline(h = 0, lty = 2)
with(dat, plot(x1, Z, col = cols, pch = pchs, main = "Z on x1")); abline(lm_Z_x1)
with(dat, plot(x1, eZ, col = cols, pch = pchs, ylab = "Z minus its linear fit on x1",
               main = "Residual Z")); abline(h = 0, lty = 2)
```

\footnotesize
Blue = downloaded the kit. The adjusted estimate is the slope of `eY` on `eZ`.
With **one** covariate we can draw this and judge it. Does the relationship of
`x1` with `Y` (top left) look like the straight line we removed? \normalsize

## The payoff: with one covariate, the questions are answerable

"Controlling for $x_1$" raises three questions. With a single covariate, each
one is a plot you can actually make and defend:

 - **Functional form.** Is the `x1`--`Y` relationship the straight line OLS
   removed? Look at the scatter and a smoother. Try `x1^2`.
 - **Extrapolation.** Do treated and control overlap along `x1`, or is the
   adjustment filling an empty region?
 - **Influential points.** Are a few units swinging the slope of `eY` on `eZ`?
   Find them on the plot.

\medskip

This is the case Jake usually shows in lecture. One covariate, one axis, three
diagnostics you can look at. Now watch what happens to each question as we add
covariates.

# Three Questions, and How Each Gets Harder

## Question 1: which functional form? {.fragile}

The estimate *is* the residualization, so changing the form of a covariate
changes the estimate. With `x1` alone:

```{r ff1, echo=TRUE}
round(c(`x1`       = coef(lm(Y ~ Z + x1, data = dat))[["Z"]],
        `x1 + x1^2`= coef(lm(Y ~ Z + x1 + I(x1^2), data = dat))[["Z"]],
        `log x1`   = coef(lm(Y ~ Z + log(x1 + 1), data = dat))[["Z"]]), 2)
```

\medskip

With one covariate that is three numbers you can plot and choose between. The
skeptic's question --- "why *that* function of the covariate?" --- has a
visual answer. With four covariates, how many functions are there?

## How many ways to "control for" four covariates?

Each covariate can enter as itself, or cut into $2, 3, \ldots$ categories, in
any combination. Draw 10,000 of these specifications at random and collect the
estimate of the `Z` effect:

```{r specsim, echo=TRUE, cache=TRUE}
lmadjfn <- function() {
  covs     <- c("x1", "x2", "x3", "x4")
  somecovs <- sample(covs, size = sample(1:4, 1))            # which covariates
  ncuts    <- round(runif(length(somecovs), min = 1, max = 8))  # how to bin each
  theterms <- ifelse(ncuts == 1, somecovs,
                     paste0("cut(", somecovs, ",", ncuts, ")"))
  coef(lm(reformulate(c("Z", theterms), response = "Y"), data = dat))[["Z"]]
}
set.seed(12345)
res <- replicate(10000, lmadjfn())
round(c(summary(res), `SD` = sd(res), `pct_wrong_sign` = mean(res < 0) * 100), 2)
```

\footnotesize
Same data, same four covariates, same outcome. The estimate ranges from about
`r round(min(res),1)` to `r round(max(res),1)`; `r round(mean(res<0)*100)`\% of
these "adjustments" get the **wrong sign**. The truth is
`r round(true_ate,2)`. \normalsize

## The specification curve

```{r speccurve, echo=FALSE, out.width=".7\\textwidth"}
par(mar = c(3.4, 3.4, 1, 1), mgp = c(2, .6, 0))
plot(density(res), main = "", xlab = "Estimated effect of Z across 10,000 specifications",
     lwd = 2)
rug(res, col = rgb(0, 0, 0, 0.15))
abline(v = true_ate, col = "blue", lwd = 2)
abline(v = 0, col = "grey50", lty = 2)
text(true_ate, par("usr")[4] * 0.9, "truth", col = "blue", pos = 4)
```

\footnotesize
Each tick is one way to "control for" the four covariates. Reporting a single
number from this cloud, without saying it is one of thousands, is what
@simonsohn2015specification and @simmonsetal2011 call exploiting researcher
degrees of freedom. We have no agreed standard that picks the right tick.
\normalsize

## Question 2: extrapolation and common support

OLS returns a coefficient even where treated and control units do not overlap.
Where there is no overlap, the **functional form**, not any comparison, fills
the gap. Along one covariate you can see it:

```{r overlap, echo=FALSE, out.width=".8\\textwidth", fig.width=7, fig.height=3.4, warning=FALSE, message=FALSE}
par(mar = c(3.2, 3.2, 1.2, 1), mgp = c(1.9, .6, 0))
plot(density(dat$x1[dat$Z == 0]), lwd = 2, col = "grey40", main = "Overlap on x1",
     xlab = "x1", ylim = c(0, max(density(dat$x1)$y) * 1.2))
lines(density(dat$x1[dat$Z == 1]), lwd = 2, lty = 2)
rug(dat$x1[dat$Z == 0], col = "grey40")
rug(dat$x1[dat$Z == 1], line = 1)
legend("topright", c("control", "treated"), lwd = 2, lty = c(1, 2),
       col = c("grey40", "black"), bty = "n")
```

\footnotesize
Controls span `x1` $\approx `r min(dat$x1[dat$Z==0])`$--$`r max(dat$x1[dat$Z==0])`$,
treated $`r min(dat$x1[dat$Z==1])`$--$`r max(dat$x1[dat$Z==1])`$: the high-`x1`
treated units have **no** control match. OLS still reports an "adjusted"
comparison there, by extending the line [@kingzeng2006; @gelmanhill2006].
\normalsize

## Extrapolation you can see vs. extrapolation you cannot

With **one** covariate, "no overlap" is a gap on a number line.

\medskip

With **four** covariates, overlap means overlap in a four-dimensional cloud.
Two groups can overlap on every covariate one at a time and still share almost
no region jointly. The chance that some treated unit sits in a corner with no
nearby controls grows fast with the number of covariates --- the **curse of
dimensionality**. OLS never warns you: it returns a coefficient regardless.

\medskip

The question "am I extrapolating?" stays the same. The ability to **answer** it
by looking disappears.

## Question 3: influential points in higher dimensions

Residualizing projects the data into a new space. A point that was unremarkable
in the raw `(Z, Y)` plane can become high-leverage among the residuals. Count
how many points exceed the usual Cook's-distance flag ($> 4/n$) as we add
covariates:

```{r cooks, echo=TRUE}
cookcount <- function(fmla) {
  sum(cooks.distance(lm(fmla, data = dat)) > 4 / nrow(dat))
}
c(`Z+x1`         = cookcount(Y ~ Z + x1),
  `+x2`          = cookcount(Y ~ Z + x1 + x2),
  `+x3`          = cookcount(Y ~ Z + x1 + x2 + x3),
  `+x4`          = cookcount(Y ~ Z + x1 + x2 + x3 + x4),
  `interactions` = cookcount(Y ~ Z * x1 * x2 * x3 * x4))
```

\footnotesize
More covariates, more flagged points --- not because the data got worse, but
because the projection has more dimensions in which a point can be extreme. The
probability that *some* observation is unduly influential rises mechanically as
you add controls. \normalsize

## Cook's distance across specifications

```{r cooksgrid, echo=FALSE, out.width=".92\\linewidth", warning=FALSE, message=FALSE, cache=TRUE}
mods <- list(
  "Z + x1"              = lm(Y ~ Z + x1, data = dat),
  "Z + x1 + x2"         = lm(Y ~ Z + x1 + x2, data = dat),
  "Z + x1..x4"          = lm(Y ~ Z + x1 + x2 + x3 + x4, data = dat),
  "polynomials"         = lm(Y ~ Z + poly(x1, 3) + poly(x2, 2) + poly(x3, 4) + x4, data = dat),
  "splines + interact"  = lm(Y ~ Z + ns(x1, 3) + ns(x2, 3) * ns(x3, 3) * x4, data = dat),
  "full interactions"   = lm(Y ~ Z * x1 * x2 * x3 * x4, data = dat)
)
par(mfrow = c(2, 3), mar = c(3, 3, 2, 0.5), mgp = c(1.7, .5, 0), oma = rep(0, 4))
for (nm in names(mods)) {
  cd <- cooks.distance(mods[[nm]])
  plot(cd, type = "h", main = nm, xlab = "observation", ylab = "Cook's D",
       col = ifelse(cd > 4 / nrow(dat), "red", "grey40"))
  abline(h = 4 / nrow(dat), lty = 2)
}
```

\footnotesize
Red bars exceed $4/n$. As the specification grows richer, more single
neighborhoods can move the estimate. With one covariate you could name and
inspect them; here, which model's influential points should you worry about?
\normalsize

# Why "Controlling For" Is Not a Design

## To answer the three questions, you read the estimate

Notice how you would actually choose among the specifications. You fit one, look
at the coefficient on `Z` and its $p$-value; you try another functional form,
look again; you drop an influential point, look again. Every diagnostic decision
is made **while watching the answer**.

\medskip

That is the opposite of the matched design. There, we settled the design on
covariates alone, **before** the outcome was unblinded. Here, design and outcome
analysis are the same act, repeated.

\medskip

Choosing a specification by looking at $\hat\tau$ and its $p$-value, over and
over, is testing the same hypothesis many times and reporting once. The nominal
$p$-value no longer means what it claims [@simmonsetal2011; @young2019].

## "Regression is a fine calculator"

@berk2010 argues that regression is not a research design; @ache:05 warns that
piling in covariates without justification --- "garbage-can regression" --- can
make things worse, not better.

\medskip

> Regression is a fine calculator. But you have to tell it what to calculate.

\medskip

The formula `Y ~ Z + x1 + x2 + x3 + x4` does not supply a design. It supplies a
set of comparisons --- a particular residualization, a particular weighting, a
particular extrapolation --- that you did not choose deliberately and usually
cannot see.

## Even when it "works", which average is it?

Suppose the form is right and overlap is fine. OLS still does not return the ATE
you have in mind. It returns a **variance-weighted** average of unit- or
stratum-level effects, weighting hardest where treatment is most variable
[@aronowsamii2016; @angrist2009mostly].

\medskip

The units that dominate the estimate --- the implied "effective sample" --- need
not resemble your target population. Two analysts who both "control for the same
covariates" can be estimating effects for different, unstated sets of people.

## The problem does not stop at observational studies

The credibility revolution changed which **identification strategies** count as
credible. It did not change how most studies handle covariate adjustment
*within* a strategy.

\medskip

A difference-in-differences with covariates ("conditional DiD"), a regression
discontinuity with covariate controls, an experiment "adjusted" for baseline
imbalance --- each adds covariates to a regression and so inherits the same three
questions: which functional form, how much extrapolation, which influential
points. A good design upstream does not excuse an unexamined residualization
downstream.

# When Is Regression Adjustment Defensible?

## The narrow case for "controlling for"

Regression adjustment answers its three questions only when you can make all of
the following true, and say so:

 - **Few covariates**, so that functional form, overlap, and influence can each
   be inspected --- the one-covariate plots, repeated and defended, not waved at.
 - **Genuine common support**: treated and control occupy the same region, so
   no coefficient rests on extrapolation.
 - **A relationship you can defend as roughly linear and additive** on that
   region --- checked against a smoother, not assumed.
 - **One specification, fixed before the outcomes are analyzed.** Not chosen
   from the cloud after seeing $\hat\tau$.

\medskip

When the covariates are many and the form is unknown, these conditions fail, and
"we controlled for them" is a description of what software did, not a
justification.

## Regression as a precision tool inside a design

There is a setting where putting covariates in a regression is on firm ground:
**inside a design you already have.**

\medskip

In a randomized experiment, $Z$ is not confounded. Covariate adjustment there is
not removing confounding; it is removing noise from the outcome to gain
precision, with a known small bias that vanishes as $n$ grows [@lin2013]. The
same logic applies after matching: estimate the effect within sets you built and
balanced first.

\medskip

The distinction is the whole lecture: adjustment as a **precision** tool *within*
a design is fine; adjustment as the **device that creates the comparison** is the
move that raises the three unanswerable questions.

## Flexible models do not dissolve the problem

"Use splines, or a machine learner, and let the data pick the form." These help
with functional form, and tools like kernel regularized least squares
[@hainmuellerhazlett2013] are built for exactly that. But they trade one set of
choices for another --- basis size, penalty, tuning --- and they still:

 - extrapolate silently where there is no support, and
 - estimate while you tune, with no separation of design from outcome.

\medskip

Flexibility buys a better functional form. It does not buy a design.

# The Alternative: Stratification Holds Constant

## Compare like with like

Instead of removing a *linear* function of `x1`, compare units with *similar*
`x1`. Bin `x1` and take the difference in means inside each bin:

```{r strat, echo=TRUE}
dat$x1cut <- cut(dat$x1, breaks = c(0, 5, 8, 11, 14, 20))
stratab <- dat %>% group_by(x1cut) %>%
  summarize(nT = sum(Z), nC = sum(1 - Z),
            diff = ifelse(nT > 0 & nC > 0, mean(Y[Z == 1]) - mean(Y[Z == 0]), NA),
            .groups = "drop")
as.data.frame(stratab)
```

\footnotesize
No functional form: just within-bin differences. Look at the last row.
\normalsize

## Stratification reveals what regression hides

```{r stratest, echo=TRUE}
usable <- filter(stratab, nT > 0 & nC > 0) %>%
  mutate(nb = nT + nC, wt = nb / sum(nb))
strat_est  <- with(usable, sum(diff * wt))                 # weight by set size
no_overlap <- with(filter(stratab, nT == 0 | nC == 0), sum(nT + nC))
c(strat_estimate = round(strat_est, 2),
  truth          = round(true_ate, 2),
  units_no_overlap = no_overlap)
```

\footnotesize
The `(14,20]` bin has `r stratab$nT[stratab$x1cut=="(14,20]"]` treated and **0**
controls --- `r no_overlap` units with no comparison. Regression silently
extrapolates into that region; stratification shows the gap and sets it aside,
reporting `r round(strat_est,2)` for the population it can actually support
(truth `r round(true_ate,2)`). \normalsize

## Why stratification answers the three questions

For each problem that "controlling for" leaves open, comparing within strata
gives an answer you can state:

 - **Functional form**: none assumed. Within a bin we compare like with like, so
   no curve to choose.
 - **Extrapolation**: a bin with only treated or only control units is *visible*
   --- it drops out, and we report which units we cannot speak to.
 - **Influential points**: each unit moves only its own small set, not a slope
   fit across the whole range.

\medskip

And the order is restored: we build and balance the strata on covariates alone
(Monday's `balanceTest`), and only then look at outcomes. Design first, analysis
second.

## Honest caveat: matching is not magic

Stratification is not a guarantee of the truth; it is a way to make the
questions answerable.

 - You still choose how to form sets (cutpoints, calipers, distances), and a
   careless full match with no caliper can itself extrapolate --- which is
   exactly why Monday's UN design used calipers, exact matching on region, and a
   balance test.
 - It estimates an effect for the matched population, which may not be your
   target population --- the same weighting question regression hides, now made
   explicit.

\medskip

The gain is not certainty. The gain is that "did we adjust enough?" becomes a
question about a design we can show you, decided before we saw the outcome
[@ho:etal:07; @rosenbaum2010design].

# Summary

## What to take away

 - **"Controlling for" = residualization** (Frisch-Waugh-Lovell). It removes a
   *linear* function of the covariates; it does not hold them constant.
 - With **one** covariate, the three questions --- functional form,
   extrapolation, influential points --- are plots you can make and defend.
 - With **several** covariates the same questions remain but become
   unanswerable: thousands of forms (here, `r round(mean(res<0)*100)`\% the wrong
   sign), joint support you cannot see, influence that grows with dimension.
 - You resolve those questions by **looking at the estimate and its
   $p$-value**, repeatedly. That is not a design; it is estimating and testing
   the same hypothesis many times.
 - **Reasonable use is narrow**: few covariates, real overlap, a defensible
   form, one pre-specified specification --- or regression as a *precision* tool
   *within* a design you already have.
 - **Stratification** makes the three questions answerable and keeps design
   decisions outcome-free. Wednesday and Thursday build the designs that do this
   for continuous treatments and over time.

## References {.allowframebreaks}
