---
title: |
  | Longitudinal Designs: Risk-Set Matching vs Difference-in-Differences
date: '`r format(Sys.Date(), "%B %d, %Y")`'
author: |
  | ICPSR 2026 Session 1
  | Jake Bowers \& Tom Leavitt
bibliography:
 - 'BIB/MasterBibliography.bib'
fontsize: 10pt
geometry: margin=1in
graphics: yes
biblio-style: authoryear-comp
biblatexoptions:
  - natbib=true
output:
  beamer_presentation:
    slide_level: 2
    keep_tex: true
    latex_engine: xelatex
    citation_package: biblatex
    template: styles/icpsr-beamer-template
    incremental: true
    includes:
        in_header:
           - defs-all.sty
    md_extensions: +raw_attribute-tex_math_single_backslash+autolink_bare_uris+ascii_identifiers+tex_math_dollars
    pandoc_args: [ "--csl", "chicago-author-date.csl" ]
---

<!-- BUILD NOTE: this deck uses did, bacondecomp, HonestDiD, which live in the
     project renv. Render WITH renv active (do NOT use --no-init-file):
     Rscript -e 'rmarkdown::render("day_13_longitudinal_2026.Rmd")' -->

```{r setup1_env, echo=FALSE, include=FALSE}
library(here)
source(here::here("rmd_setup.R"))
```

```{r setup2_loadlibs, echo=FALSE, include=FALSE}
library(tidyverse)
library(dplyr)
library(estimatr)
library(optmatch)
library(RItools)
library(sensitivitymw)   # Gamma sensitivity for the matched design
library(did)             # event study / pooled DiD with clean never-treated controls
library(HonestDiD)       # sensitivity to parallel-trends violations
options(show.signif.stars = FALSE)
```

```{r build, echo=FALSE, include=FALSE, cache=TRUE}
## V-Dem again (the day-12 data). Long panel: each country-year, 1980-2010.
## Treatment = first year a country reaches electoral/liberal democracy
## (v2x_regime in {2,3}); outcome = corruption (v2x_corr, 0=low ... 1=high).
WIN_LO <- 1980; WIN_HI <- 2010; POST <- 5
load(here::here("data", "vdem_democracy_transitions.rda"))
v <- vdem_sub %>% arrange(country_text_id, year) %>%
  mutate(dem = as.integer(v2x_regime %in% c(2, 3)),
         log_gdp = log(e_gdppc),
         log_gdp = ifelse(is.na(log_gdp), median(log_gdp, na.rm = TRUE), log_gdp))
fd <- v %>% filter(dem == 1) %>% group_by(country_text_id) %>%
  summarize(first_dem = min(year), .groups = "drop")
st <- v %>% group_by(country_text_id) %>%
  summarize(start_dem = dem[which.min(year)], .groups = "drop")
## keep countries autocratic at panel start (so a transition is observed) or never democratic
keep <- st %>% left_join(fd, by = "country_text_id") %>%
  filter(is.na(first_dem) | start_dem == 0) %>%
  mutate(first_treat = ifelse(is.na(first_dem), 0, first_dem))
v <- v %>% inner_join(keep %>% select(country_text_id, first_treat), by = "country_text_id")

## DiD panel: cohort = first_treat if inside the window; never-treated-in-window = 0;
## drop countries already democratic before the window (always-treated, not controls).
panel <- v %>%
  mutate(ft = case_when(first_treat == 0 ~ 0L, first_treat > WIN_HI ~ 0L,
                        first_treat < WIN_LO ~ NA_integer_, TRUE ~ as.integer(first_treat))) %>%
  filter(!is.na(ft), year >= WIN_LO, year <= WIN_HI) %>%
  mutate(first_treat = ft, country_num = as.integer(factor(country_text_id)),
         post = as.integer(first_treat > 0 & year >= first_treat))
n_never <- sum(tapply(panel$first_treat, panel$country_num, function(x) x[1] == 0))
n_cohort <- n_distinct(panel$first_treat[panel$first_treat > 0])
```

## Today

0. **Your questions.**
1. **Review (day 12):** non-bipartite matching with a *fixed* 1990 risk set ---
   does democratization reduce corruption? We compared countries alike in 1990.
2. **Today:** treatment timing varies across units. Two designs answer the same
   longitudinal question under assumptions from **different families**:
   - **Risk-set matching:** at each transition, compare the new democracy to
     countries *still autocratic then*, alike on history (conditional
     ignorability).
   - **Difference-in-differences:** do not match; assume *parallel trends*.
   - Each gets a sensitivity analysis: $\Gamma$ for matching, HonestDiD for DiD.

\medskip

Same data, same question. The designs will agree on the rough size of the
effect and disagree on how much to believe it.

## The spine of today

\begin{center}
\small
\begin{tabular}{@{}p{2.7cm}p{4.0cm}p{2.1cm}p{1.9cm}@{}}
\toprule
design & identifying assumption & checked before $Y$ & sensitivity \\
\midrule
risk-set matching & conditional ignorability given history & balance table & $\Gamma$ \\
difference-in-differences & parallel trends (+ no anticipation) & pre-trends (weakly) & HonestDiD \\
\bottomrule
\end{tabular}
\normalsize
\end{center}

\medskip

Conditional ignorability is a claim about **who got treated**; parallel trends is
a claim about **how outcomes would have moved**. The first invites a
randomization-style argument we can check with balance; the second invites a
model we can only probe indirectly. Keep this table in mind --- we return to it at
the end.

# Review and the Staggered-Timing Problem

## What day 12 left open

Day 12 fixed the risk set at one baseline year (1990): we took the autocracies of
1990 and compared those that democratized to those that did not, matching on 1990
covariates.

\medskip

That is fine for the mechanics, but real histories are **staggered**. Countries
democratize in different years, and a country democratic for a decade is not a
counterfactual for one democratizing now. Today the comparison set moves with
time.

## Staggered transitions in V-Dem

```{r cohort_tab, echo=FALSE}
panel %>% filter(first_treat > 0) %>% distinct(country_num, first_treat) %>%
  count(first_treat, name = "n_countries") %>%
  mutate(decade = cut(first_treat, c(1979,1984,1989,1994,1999,2004,2010))) %>%
  group_by(decade) %>% summarize(cohorts = paste(first_treat, collapse=", "),
                                 countries = sum(n_countries), .groups="drop") %>%
  as.data.frame()
```

\footnotesize
`r n_cohort` transition cohorts (most in the post-Cold-War wave), plus
`r n_never` countries that stay autocratic across 1980--2010 --- the never-treated
pool. A country that democratizes in 1999 should be compared with countries still
autocratic in 1999, not with the class of 1991. \normalsize

# Risk-Set Matching Over Time

## The idea

At each transition year $g$, match the newly-democratizing country to countries
**still at risk** at $g$ --- still autocratic, and not about to democratize ---
choosing controls alike on covariate history up to $g$
[@lietal2001; @rosenbaum2017; @rosenbaum2020book, ch. 13].

\medskip

This is the day-12 design made dynamic: rebuild the risk set at each event time so
every comparison runs among countries that were jointly "at risk." Within a pair,
which one democratized is the treatment; conditioning on history is the bid for
as-if-random assignment.

## Why "still at risk"

A country already democratic for years is not a counterfactual for one
democratizing now: using it as a control lets the future pick the comparison set
(look-ahead, or immortal-time, bias).

\medskip

So the eligible controls for cohort $g$ are countries that are still autocratic at
$g-1$ **and** do not democratize within the post-window. We snapshot their
covariates at $g-1$ --- before anyone in the pair has moved.

## Build it {.allowframebreaks}

```{r rs_build, echo=TRUE}
mvars <- c("v2x_polyarchy", "v2x_corr", "v2x_cspart", "v2x_rule", "log_gdp")
## Match the cohort-g democracy to still-at-risk controls on history at g-1.
build_pair_at <- function(g) {
  pool <- v %>% filter(year == g - 1) %>%
    mutate(newly = as.integer(first_treat == g)) %>%
    filter(first_treat == g | first_treat == 0 | first_treat > g + POST)
  ## need outcome at g-1 and g+POST for everyone kept
  ok <- v %>% filter(country_text_id %in% pool$country_text_id, year %in% c(g - 1, g + POST)) %>%
    count(country_text_id) %>% filter(n == 2) %>% pull(country_text_id)
  pool <- pool %>% filter(country_text_id %in% ok) %>% as.data.frame()
  rownames(pool) <- pool$country_text_id
  if (sum(pool$newly) < 1 || sum(pool$newly == 0) < 1) return(NULL)
  fml  <- reformulate(mvars, response = "newly")
  dmat <- match_on(fml, data = pool, method = "rank_mahalanobis")       # covariate distance
  ps   <- caliper(match_on(glm(fml, pool, family = binomial()), data = pool), width = 2)
  pv   <- pool$v2x_polyarchy; names(pv) <- rownames(pool)               # caliper on the
  pcal <- ifelse(abs(outer(pv, pv, "-")) > 0.05, Inf, 0)               # cusp variable
  total <- dmat + ps + as.InfinitySparseMatrix(pcal[pool$newly == 1, pool$newly == 0, drop = FALSE])
  m <- tryCatch(pairmatch(total, controls = 1, data = pool), error = function(e) NULL)
  if (is.null(m)) return(NULL)
  pool$pair <- as.character(m)                              # always character (NA-safe)
  out <- pool %>% filter(!is.na(pair)) %>%
    mutate(pair = paste0(g, "_", pair), g = g) %>%
    select(country_text_id, country_name, pair, g, newly, all_of(mvars))
  if (nrow(out) < 2) return(NULL)                           # this cohort formed no pair
  out
}
cohorts <- sort(unique(v$first_treat[v$first_treat >= WIN_LO + 1 & v$first_treat <= WIN_HI - POST]))
matched <- suppressWarnings(map_dfr(cohorts, build_pair_at))
c(pairs = n_distinct(matched$pair), treated = sum(matched$newly),
  controls = sum(1 - matched$newly))
```

## Balance: does history predict who democratized?

```{r rs_balance, echo=TRUE}
balfmla <- reformulate(c(mvars, "strata(pair)"), response = "newly")
xb <- balanceTest(balfmla, data = as.data.frame(matched))
xb$overall["pair", ]
round(xb$results[, c("Treatment", "Control", "std.diff", "p"), "pair"], 2)
```

\footnotesize
A *high* omnibus $p$ is the good case. The hard variable is `v2x_polyarchy`: a
country on the cusp of democratizing already scores higher on it, so without the
$0.05$ caliper it will not balance. With the caliper it does --- at the cost of
dropping cohort members who have no comparable still-autocratic control. \normalsize

## Estimate: within-pair change in corruption

```{r rs_est, echo=TRUE}
## change score: corruption at g+POST minus at g-1, for each matched country
chg <- matched %>% rowwise() %>% mutate(
  dcorr = v$v2x_corr[v$country_text_id == country_text_id & v$year == g + POST] -
          v$v2x_corr[v$country_text_id == country_text_id & v$year == g - 1]) %>% ungroup()
pd <- chg %>% group_by(pair) %>%
  summarize(diff = dcorr[newly == 1] - dcorr[newly == 0], .groups = "drop")
est_rs <- mean(pd$diff)                                   # mean within-pair difference
hl <- wilcox.test(pd$diff, conf.int = TRUE)              # Hodges-Lehmann + signed-rank p
c(mean_diff = est_rs, HL = hl$estimate[[1]],
  ci_lo = hl$conf.int[1], ci_hi = hl$conf.int[2], p = hl$p.value)
```

\footnotesize
The newly-democratic country's corruption falls about
`r round(abs(est_rs), 2)` more than its still-autocratic match's over five years
(signed-rank $p \approx `r round(wilcox.test(pd$diff)$p.value, 3)`$). The change
score differences out any fixed within-pair gap --- a robustness device serving the
conditional-ignorability design, not a separate parallel-trends assumption.
\normalsize

## Sensitivity: how much hidden bias would it take?

```{r rs_gamma, echo=TRUE}
ymat <- chg %>% select(pair, newly, dcorr) %>%
  pivot_wider(names_from = newly, values_from = dcorr) %>% select(`1`, `0`) %>% as.matrix()
pval_g <- function(g) senmw(-ymat, gamma = g, method = "t")$pval   # worst-case one-sided p
gstar  <- uniroot(function(g) pval_g(g) - 0.05, c(1, 3))$root
round(c(p_at_1 = pval_g(1), p_at_1.5 = pval_g(1.5), Gamma_star = gstar), 3)
```

\footnotesize
$\Gamma^{*} \approx `r round(uniroot(function(g) senmw(-ymat, gamma=g, method="t")$pval - 0.05, c(1,3))$root, 2)`$:
an unobserved confounder would have to give one country in a pair about
`r round(uniroot(function(g) senmw(-ymat, gamma=g, method="t")$pval - 0.05, c(1,3))$root, 2)`
times the odds of being the democratizer before the test stops rejecting. Hold
this number; we compare it with the DiD's robustness at the end. \normalsize

# Difference-in-Differences

## The other strategy: assume parallel trends

Do not match on covariates. Assume that, absent democratization, the corruption
paths of democratizers and non-democratizers would have moved in **parallel**.
Then the post-minus-pre change for the treated, minus the same change for
controls, is the effect.

\medskip

The appeal: no need to find comparable countries. The cost: a claim about
counterfactual *trajectories* that we can never fully check.

## A two-period difference-in-differences {.shrink}

The simplest version: two periods, two groups. Take the post-Cold-War wave (the
countries that democratized in 1990) and the never-democratizers, and compare
corruption just before (1989) and a few years after (1995).

```{r did_2x2, echo=TRUE}
yr_pre <- 1989; yr_post <- 1995
bb <- panel %>% filter(year %in% c(yr_pre, yr_post), first_treat %in% c(0, 1990)) %>%
  group_by(country_num) %>% filter(n() == 2) %>% ungroup() %>%      # balanced 2x2
  mutate(grp = ifelse(first_treat == 1990, "democratized", "never"),
         per = ifelse(year == yr_post, "post", "pre"))
cells <- tapply(bb$v2x_corr, list(bb$grp, bb$per), mean)
did2  <- (cells["democratized","post"] - cells["democratized","pre"]) -
         (cells["never","post"]        - cells["never","pre"])
round(cells, 2)
```

\footnotesize
The estimate is the democratizers' change minus the non-democratizers' change:
$`r round(cells["democratized","post"]-cells["democratized","pre"],2)` - (`r round(cells["never","post"]-cells["never","pre"],2)`) = `r round(did2,2)`$.
Corruption rose among non-democratizers and held steady among the new democracies.
The whole estimate rests on **parallel trends**: that the two groups would have
moved together absent democratization. \normalsize

## Checking the assumption: the event study

A single pre/post comparison cannot show whether the trends were parallel
*before* treatment. Add pre-periods and look --- aligning every democratizer at its
own transition and comparing to the never-treated:

```{r did_cs, echo=TRUE, cache=TRUE, message=FALSE, warning=FALSE, results='hide'}
## event study + pooled estimate, never-treated controls, all cohorts
cs <- att_gt(yname = "v2x_corr", tname = "year", idname = "country_num",
             gname = "first_treat", control_group = "nevertreated",
             base_period = "universal", allow_unbalanced_panel = TRUE, data = panel)
cs_simple <- aggte(cs, type = "simple",  na.rm = TRUE)   # pooled DiD estimate
cs_dyn    <- aggte(cs, type = "dynamic", na.rm = TRUE, min_e = -5, max_e = 5)  # leads/lags
```

```{r did_es_plot, echo=FALSE, out.width=".60\\textwidth", fig.width=6.2, fig.height=3.4}
esd <- data.frame(e = cs_dyn$egt, att = cs_dyn$att.egt, se = cs_dyn$se.egt)
esd <- esd[is.finite(esd$att) & is.finite(esd$se), ]            # drop endpoints with no SE
yl <- range(c(esd$att - 1.96*esd$se, esd$att + 1.96*esd$se), na.rm = TRUE)
par(mar = c(3.4, 3.6, 1, 1), mgp = c(2.1, .6, 0))
plot(esd$e, esd$att, type = "n", ylim = yl,
     xlab = "years since democratization", ylab = "DiD vs never-treated (corruption)")
abline(h = 0, col = "grey60"); abline(v = -0.5, lty = 3, col = "grey60")
arrows(esd$e, esd$att - 1.96*esd$se, esd$e, esd$att + 1.96*esd$se,
       length = 0.02, angle = 90, code = 3, col = "grey50")
points(esd$e, esd$att, pch = 19, col = ifelse(esd$e < 0, "darkred", "black"))
```

\footnotesize
Red = pre-transition. The pre-trends are **not** flat: would-be democracies
already had higher, *declining* corruption before the transition --- exactly what
parallel trends rules out. The pooled estimate (all cohorts) is about
`r round(cs_simple$overall.att, 2)`. \normalsize

## A note on staggered timing

We aligned each democratizer at its own transition and used only
**never-treated** countries as controls --- the clean comparison. The tempting
shortcut, two-way fixed effects with a 0/1 treatment, instead reuses
*already-democratic* countries as controls, which can distort the estimate when
effects grow or fade over time.

\medskip

Untangling that is its own literature --- the Goodman-Bacon decomposition
[@goodmanbacon2021] and estimators built to avoid the trap
[@callawaysantanna2021; @sunabraham2021; @dechaisemartin2020]. We keep today
simple by leaning on never-treated controls; the reading is there when you build
staggered designs of your own.

# Sensitivity for Parallel Trends

## Passing a pre-trend test is not proof

The event study showed a pre-trend. Even when it does not, "no significant
pre-trend" is weak comfort: pre-trend tests have low power, and selecting analyses
that pass them distorts inference [@rothetal2023].

\medskip

So instead of asking "are pre-trends flat?", ask "**how badly would parallel
trends have to fail to overturn the result?**" That is HonestDiD.

## HonestDiD: bound the bias by the pre-trend

```{r did_hdid, echo=TRUE, cache=TRUE, message=FALSE, warning=FALSE}
## drive HonestDiD off the CS event study (handles never-treated correctly)
ifc <- cs_dyn$inf.function$dynamic.inf.func.e
Sig <- t(ifc) %*% ifc / nrow(ifc)^2
npre <- sum(cs_dyn$egt < 0); npost <- sum(cs_dyn$egt >= 0)
hd <- createSensitivityResults_relativeMagnitudes(
  betahat = cs_dyn$att.egt, sigma = Sig, numPrePeriods = npre, numPostPeriods = npost,
  Mbarvec = c(0, 0.25, 0.5), l_vec = rep(1/npost, npost))
hd[, c("Mbar", "lb", "ub")]
```

\footnotesize
$\bar M$ caps the post-period violation as a multiple of the **worst pre-period
violation** [@rambachanroth2023]. At $\bar M = 0$ the robust CI excludes zero; by
$\bar M = 0.25$ it already covers zero. The **breakdown** is below $\bar M = 0.25$:
allowing post-treatment trends to bend even a quarter as much as the visible
pre-trend is enough to erase the result. \normalsize

# Putting It Together

## Same data, two designs, two verdicts

\begin{center}
\small
\begin{tabular}{@{}p{3.0cm}p{2.0cm}p{2.6cm}p{3.0cm}@{}}
\toprule
design & estimate & assumption & robustness \\
\midrule
risk-set matching & $\approx `r round(est_rs,2)`$ & conditional ignorability & $\Gamma^{*} \approx `r round(uniroot(function(g) senmw(-ymat, gamma=g, method="t")$pval - 0.05, c(1,3))$root, 2)`$ (moderate) \\
difference-in-differences & $\approx `r round(cs_simple$overall.att,2)`$ & parallel trends & breakdown $\bar M < 0.25$ (fragile) \\
\bottomrule
\end{tabular}
\normalsize
\end{center}

\medskip

The two designs agree the effect is negative and similar in size. They disagree
on how much to trust it --- and they disagree *because* they lean on different
assumptions, each interrogated on its own terms.

## Why the verdicts differ

 - **Risk-set matching conditions on the pre-treatment trajectory** --- including
   the polyarchy cusp --- so the "they were already heading there" story is
   addressed in the design, and what is left is hidden bias, measured by $\Gamma$.
 - **DiD leaves that trajectory in** and assumes it would have continued in
   parallel. The event study shows it would not have, so the conclusion is
   fragile, measured by HonestDiD.

\medskip

Same confounding worry --- countries on the path to democracy were already
changing --- surfaces two ways: as a balance problem you fix by matching, or as a
pre-trend you bound with HonestDiD.

## Built vs assumed

A matched design **builds** a comparison a reader can check: here are the pairs,
here is the balance table, here is the bias it would take to overturn us. Parallel
trends **assumes** a comparison a reader must take partly on faith: here are the
pre-trends, and here is how much they would have to continue to matter.

\medskip

Design and analysis are coupled in the first --- we choose which countries can
serve as counterfactuals and discard those we cannot defend --- and separable in
the second. Neither is always right. What both designs owe the reader is the
same: state the assumption, check what can be checked, and report the sensitivity.

## When to prefer which

 - **Few clean pre-periods, good covariates** -> risk-set matching: condition on
   history and report $\Gamma$.
 - **Many clean pre-periods, confounders you cannot measure** -> DiD: lean on
   trends and report HonestDiD.
 - **Either way, show the sensitivity.** A point estimate without it --- matched or
   differenced --- is half an answer.

## Summary

 - Staggered timing breaks the fixed-baseline design of day 12. Two repairs:
   move the risk set, or assume parallel trends.
 - **Risk-set matching** rebuilds the comparison at each event time, balances on
   history (the cusp variable is the hard one), and reports $\Gamma$.
 - **DiD** assumes parallel trends; staggered TWFE needs care (Goodman-Bacon,
   Callaway--Sant'Anna), and the assumption needs HonestDiD.
 - On V-Dem both say democratization lowers corruption by a similar amount, but
   matching's conclusion is the more robust here --- because parallel trends is
   visibly suspect.
 - The week's lesson, over time: state the assumption, check it before outcomes,
   stress-test what you cannot check.

## References {.allowframebreaks}
