---
title: |
  | Non-bipartite Matching: Stratified Comparisons When Treatment Is Not Binary
date: '`r format(Sys.Date(), "%B %d, %Y")`'
author: |
  | ICPSR 2026 Session 1
  | Jake Bowers \& Tom Leavitt
bibliography:
 - 'BIB/MasterBibliography.bib'
fontsize: 10pt
geometry: margin=1in
graphics: yes
biblio-style: authoryear-comp
biblatexoptions:
  - natbib=true
output:
  beamer_presentation:
    slide_level: 2
    keep_tex: true
    latex_engine: xelatex
    citation_package: biblatex
    template: styles/icpsr-beamer-template
    incremental: true
    includes:
        in_header:
           - defs-all.sty
    md_extensions: +raw_attribute-tex_math_single_backslash+autolink_bare_uris+ascii_identifiers+tex_math_dollars
    pandoc_args: [ "--csl", "chicago-author-date.csl" ]
---

```{r setup1_env, echo=FALSE, include=FALSE}
library(here)
source(here::here("rmd_setup.R"))
```

```{r setup2_loadlibs, echo=FALSE, include=FALSE}
library(tidyverse)
library(dplyr)
library(estimatr)
library(RItools)
library(designmatch)
library(coin)
options(show.signif.stars = FALSE)
```

## Today

0. **Your questions.** Anything from the reading, other classes, or your own
   work?
1. **Review:** stratified designs for a *binary* treatment --- match, check
   balance against a block-randomized experiment, assess the design using
   substantive knowledge (comparing within-set differences to your own
   judgement of what is "small enough"),  test or estimate as-if randomized,
   and engage with assumptions using sensitivity analysis.
2. **Today: non-bipartite matching** what to do when the cause of interest is **not** binary.
   - The idea: pair units that are *close on covariates*, then compare the
     outcomes of the one which got *more* treatment versus the outcomes of
     the one which got *less* treatment.
   - The same workflow as before: design first (before outcomes), check
     balance, then estimate and test.

\medskip

Questions? Comments?

## Where this fits in the week

- **Mon (day 10):** sensitivity analysis --- how much hidden bias would
  overturn a matched-design conclusion?
- **Tues (day 11):** "controlling for" in a regression, and why it is not a
  design (it doesn't occur *before* and *separate from* statistical inference
  about causal effects), and why it is such a hassle or burden. How can you communicate that you **adjusted enough**?

  - How can you know that the functional form is adequate? Good fit is not the
    same as adjusted enough. 
  -  How do assess extrapolation? (Some might be ok, but how much is too much?)
  -  Did you check for overly influential points (using Cook's D for
     example) *without looking at the $p$-values or confidence intervals or
     dividing the estimates by the standard error each time*?
- **Today (day 12):** matching when the treatment is continuous or
  multi-valued.
- **Thurs (day 13):** longitudinal data --- risk-set matching *over time* and
  its contrast with difference-in-differences.

# Review: Designs for a Binary Treatment

## What we have built so far

For a binary treatment $Z \in \{0, 1\}$ we have a workflow that never looks at
the outcome until the design is settled:

 - **Stratify.** Build matched sets that are similar on covariates $\bx$
   (`optmatch::fullmatch`, distances from `match_on`, calipers, exact
   matching).
 - **Check.** Does the matched comparison resemble a block-randomized
   experiment on $\bx$? The omnibus balance test and within-set differences
   decide it --- *before* $Y$ is unblinded.
 - **Estimate and test as-if randomized.** Treat the matched sets like the
   blocks of a block-randomized experiment.
 - **Assess assumptions.** A sensitivity analysis asks how much an unobserved
   confounder (a variable that we could not match-on) would overturn the
   conclusion.

## The not-so-hidden assumption: there are two groups

Matching as we have done it is **bipartite**: two sides. The optimization pairs each
treated unit with control units, never a treated unit with another treated unit.

\medskip

But many causes we care about do not come as two groups:

 - **A degree, not a switch.** How *democratic* a country becomes; how *much*
   of a media campaign a city received [@lu:etal:2001]; what *proportion* of a
   neighborhood owns their homes; an answer to a survey question of "How many
   people in your family are likely to get the COVID vaccine?" [@rabb2022pnas].
 - **Several doses.** Low, medium, high exposure, with no natural "control."

## The move: one pool, not two {.fragile}

\textbf{Non-bipartite matching} drops the two-group structure. Start with a
single pool of units. Pair each unit with the one *most similar on covariates*,
whatever its level of treatment. Only *after* the pairs are fixed do we ask, within each
pair, which unit had more of the treatment.

# The Question: Does Democratization Reduce Corruption?

## Two stories about democracy and corruption

A long debate: does democratic governance *reduce* political corruption, or
does the association merely reflect who becomes democratic?

 - **The causal story.** Free press, competitive elections, independent courts,
   legislative oversight --- these constrain officials, so countries that
   democratize should see corruption fall.
 - **The confounded story.** Countries that democratize are not a random sample
   of autocracies. They already have stronger civil society, rising education, a
   growing middle class that demands accountability. Those same forces could
   lower corruption regardless of regime type.

\medskip

If we could randomly assign democratization to autocracies, we could separate
the two. We cannot. So we build an observational approximation.

## The data: V-Dem, autocracies in 1990

The Varieties of Democracy project rates democratic governance for nearly every
country, yearly, from expert surveys [@vdem2025dataset]. We use a curated subset,
1970--2015.

\medskip

To keep the basics clear, fix one baseline year. Take the countries that were
**autocratic in 1990** --- the pool of countries "at risk" of democratizing ---
and follow them:

 - **Covariates** ($\bx$, measured 1990): electoral-democracy index, corruption,
   freedom from physical violence, civil-society participation, rule of law,
   logged GDP per capita (and an indicator for missing GDP).
 - **Treatment** ($D$): how much each country democratized by 2005 --- a *degree*
   (change in the polyarchy index).
 - **Outcome** ($Y$): the corruption index in 2010 ($0$ = none, $1$ = high).

## The naive comparison, and why to distrust it {.allowframebreaks}

```{r vdem_build, echo=TRUE, cache=TRUE}
load(here::here("data", "vdem_democracy_transitions.rda"))
## Autocracies in 1990 = the pool. Baseline covariates from 1990.
baseline <- vdem_sub %>% filter(year == 1990, v2x_regime %in% c(0, 1)) %>%
  dplyr::select(country_name, country_text_id, regime_90 = v2x_regime,
    polyarchy_90 = v2x_polyarchy, corr_90 = v2x_corr, clphy_90 = v2x_clphy,
    cspart_90 = v2x_cspart, rule_90 = v2x_rule, gdppc_90 = e_gdppc, region = e_regionpol)
treatment <- vdem_sub %>% filter(year == 2005) %>%
  dplyr::select(country_text_id, regime_05 = v2x_regime, polyarchy_05 = v2x_polyarchy)
outcome <- vdem_sub %>% filter(year == 2010) %>%
  dplyr::select(country_text_id, corr_10 = v2x_corr)
dat <- baseline %>% inner_join(treatment, by = "country_text_id") %>%
  inner_join(outcome, by = "country_text_id") %>%
  mutate(democratized = as.integer(regime_05 %in% c(2, 3)),
    polyarchy_change = polyarchy_05 - polyarchy_90,
    gdp_missing = as.integer(is.na(gdppc_90)),
    log_gdppc_90 = log(gdppc_90),
    log_gdppc_90 = ifelse(is.na(log_gdppc_90), median(log_gdppc_90, na.rm = TRUE), log_gdppc_90))
```

## The naive comparison, and why to distrust it {.allowframebreaks}

```{r naive, echo=TRUE}
naive <- lm_robust(corr_10 ~ democratized, data = dat)
c(diff = coef(naive)[["democratized"]], p = naive$p.value[["democratized"]])
```

\medskip

Among `r nrow(dat)` autocracies in 1990, `r sum(dat$democratized)` had reached democracy by 2005 and `r sum(1 - dat$democratized)` had not. The democratizers have markedly *lower* 2010 corruption: a difference of about `r round(coef(lm_robust(corr_10 ~ democratized, data = dat))[["democratized"]], 2)` on the $0$--$1$ scale. The sd of the outcome is `r sd(dat$corr_10)` and range of the outcome is `r range(dat$corr_10)` ($p \approx `r round(lm_robust(corr_10 ~ democratized, data = dat)$p.value[["democratized"]], 3)`$).

\medskip

Taken at face value: democratization cuts corruption. But democratizers differ
from the non-democratizers on exactly the covariates $\bx$: places likely to
have low corruption choose to democratize (find it easier to democractize).

# Non-bipartite Matching: The Idea

## From two groups to one pool

Every unit can be paired with every other. Split the whole pool of $N$ units
into $N/2$ pairs so as to minimize the **total within-pair distance** on
covariates. There is no "treated group" until we look, within each formed pair,
at which unit got more of $D$ [@lu2011optimal].

\medskip

Optimal non-bipartite matching solves this partition exactly --- the same kind
of optimization behind `fullmatch`, but on one undivided set of units rather
than across two.

## Why pairs help: 

Suppose two countries were nearly identical in 1990 --- same corruption, same
civil society, same income --- and then one democratized more than the other.

\medskip

If covariates do not predict *which* one democratized more, then, within that
pair, "which country got the higher dose of democracy" looks like the flip of a coin. That is
the paired-randomized-experiment analogy.

\medskip

The **design's job**, is to make that analogy defensible: pairs
close enough on $\bx$ that covariates no longer forecast the within-pair
direction of treatment.

## A vocabulary note: "risk set"

The pool of autocracies in 1990 is a **risk set**: the units that were all "at
risk" of the event (democratizing) at a common starting point. We compare units
*within* the risk set, never a future democracy against a country that had
already been democratic for decades.

\medskip

Here the risk set is fixed at one year for simplicity. When the timing of
treatment varies --- countries democratizing in different years --- the risk set
must move with time. That is **risk-set matching**, and it is Thursday's topic
[@rosenbaum2017; @rosenbaum2020book, ch. 12].

\medskip

Today we are focusing on the cross-sectional comparison to understand
non-bipartite matching.

# Building the Design (before outcomes)

## Step 1: a distance between every pair of units

Non-bipartite matching needs a distance between *every* pair of units. Here we
summarize each country by its Mahalanobis distance from the pool's covariate
center. The distance between two countries can be how far apart those numbers
are.

```{r dist, echo=TRUE, cache=TRUE}
row.names(dat) <- dat$country_text_id
covs <- c("polyarchy_90", "corr_90", "clphy_90", "cspart_90",
          "rule_90", "log_gdppc_90", "gdp_missing")
covmat <- dat %>% dplyr::select(all_of(covs)) %>% as.data.frame()
## One Mahalanobis score per country: how "typical" on the joint covariate dist.
dat$covmh <- mahalanobis(covmat, center = colMeans(covmat), cov = cov(covmat))
## Distance between two countries = absolute difference of their scores.
mhdist <- outer(dat$covmh, dat$covmh, FUN = function(x, y) abs(x - y))
dimnames(mhdist) <- list(row.names(dat), row.names(dat))
mhdist[1:4, 1:4]
```

## Step 2: optimal non-bipartite matching

`designmatch::nmatch` partitions the pool into pairs minimizing total
within-pair distance. We add two **calipers** (`near` constraints): no pair may
bridge countries more than one unit apart on logged GDP, or more than $0.10$
apart on civil-society participation --- the rival story's central variable.
The open-source HiGHS solver finds the optimum.

```{r match, echo=TRUE, cache=TRUE, results='hide', message=FALSE, warning=FALSE}
## designmatch prefers normalized, rounded distances.
matchdist <- round(mhdist / mean(mhdist), 2)
## Calipers: within 1 unit on logged GDP AND within 0.10 on civil-society part.
nearlist <- list(covs = as.matrix(dat[, c("log_gdppc_90", "cspart_90")]),
                 pairs = c(log_gdppc_90 = 1, cspart_90 = 0.10))
solverlist <- list(name = "highs", approximate = 0, t_max = 1000, trace = 0)
set.seed(12345)
nb <- nmatch(dist_mat = matchdist, near = nearlist, subset_weight = 1, solver = solverlist)
```

## Step 2: optimal non-bipartite matching

```{r match_to_df, echo=TRUE}
## Turn nmatch output into a pair-id we can merge back onto the countries.
nmatch_to_df <- function(obj, origid) {
  d <- data.frame(pair_id = obj$group_id, match_id = c(obj$id_1, obj$id_2))
  d$country_text_id <- origid[d$match_id]; d
}
dat_m <- inner_join(dat, nmatch_to_df(nb, dat$country_text_id), by = "country_text_id") %>%
  droplevels()
c(matched = nrow(dat_m), pairs = length(unique(dat_m$pair_id)), dropped = nrow(dat) - nrow(dat_m))
```

## What the matching produced

The optimizer kept `r nrow(dat_m)` countries in
`r length(unique(dat_m$pair_id))` pairs and set aside the
`r nrow(dat) - nrow(dat_m)` it could not pair within the GDP caliper. A few
pairs:

```{r pairs_examples, echo=FALSE}
dat_m %>% arrange(pair_id) %>%
  transmute(pair = pair_id, country = country_name,
            poly_change = round(polyarchy_change, 2),
            corr_2010 = round(corr_10, 2), 
            logGDP = round(log_gdppc_90, 2)) %>%
  head(6) %>% as.data.frame()
```

\footnotesize
Each pair is two countries close on 1990 covariates. Within a pair they differ
in how much they democratized (`poly_change`). `poly_change` is the 1990--2005
change in the polyarchy index, so it can be negative (backsliding). Within a
pair, "higher dose" simply means the larger of the two changes --- the country
that moved more toward democracy, or backslid less --- regardless of sign.
\normalsize

# Checking the Design (still before outcomes)

## The trick: a pseudo-binary treatment within pairs

To borrow the balance tools from days 8--11, we manufacture a binary
treatment *inside each pair*. Rank the two countries by how much they
democratized; the larger increase gets `rank_dem = 1`, the smaller gets $0$.

```{r rank, echo=TRUE}
dat_m <- dat_m %>% group_by(pair_id) %>%
  mutate(rank_dem = rank(polyarchy_change) - 1) %>% ungroup()
table(dat_m$rank_dem)   # 0.5 marks a tie (both changed identically)
dat_notie <- dat_m %>% filter(rank_dem != 0.5)
```

```{r pairs_examples2, echo=FALSE}
dat_m %>% arrange(pair_id) %>%
  transmute(pair = pair_id, country = country_name,
            poly_change = round(polyarchy_change, 2),
            rank_dem = rank_dem,
            corr_2010 = round(corr_10, 2), 
            logGDP = round(log_gdppc_90, 2)) %>% arrange(pair,rank_dem) %>%
  head(6) %>% as.data.frame()
```
\medskip

Now "treated" means "the more-democratizing country in its pair." If the
pairing worked, covariates should not predict which country democratized more
versus less within pair. 

## Balance: does $\bx$ predict who democratized more?

```{r balance, echo=TRUE}
bal_covs <- c("corr_90", "clphy_90", "cspart_90", "rule_90", "log_gdppc_90", "gdp_missing")
balfmla  <- reformulate(bal_covs, response = "rank_dem")
xb <- balanceTest(update(balfmla, . ~ . + strata(pair_id)),
                  data = dat_notie, p.adjust.method = "holm")
xb$overall["pair_id", ]
```

\medskip

Across the covariates jointly, the within-pair imbalance is no larger than
a paired-randomized experiment would routinely produce
($p \approx `r round(xb$overall["pair_id", "p.value"], 2)`$).

## Balance, covariate by covariate

```{r balance_vars, echo=TRUE}
xbv <- data.frame(xb$results[, c("Treatment", "Control", "std.diff", "p"), "pair_id"])
round(xbv[order(xbv$p), ], 3)
```

\footnotesize
The standardized within-pair differences are all small. 
\normalsize

## What the balance check is really doing

Imagine the ideal: within each pair, a coin flip decides which country
democratizes more. Then covariates would predict the winner only as often as
chance. The balance test compares our actual design against that ideal.

\medskip

A useful exercise (and a check on the logic): replace `rank_dem` with a genuine
within-pair coin flip and rerun `balanceTest`. You should see the kind of
$p$-values the null produces --- which is what we want our real design to
resemble.

# From Design to Effect

## Outcome analysis 1: higher vs. lower democratizer

With the design fixed now we can analyze the outcome. Start with within-pair
difference in 2010 corruption between the more- and less-democratizing country.

```{r est_binary, echo=TRUE}
pair_diffs <- dat_m %>% filter(rank_dem != 0.5) %>% group_by(pair_id) %>%
  summarize(corr_diff = corr_10[rank_dem == 1] - corr_10[rank_dem == 0],
            poly_diff = polyarchy_change[rank_dem == 1] - polyarchy_change[rank_dem == 0],
            .groups = "drop")
est1 <- mean(pair_diffs$corr_diff)
est2 <- lm_robust(corr_10 ~ rank_dem, fixed_effects = ~ factor(pair_id),
                  data = dat_m, subset = rank_dem != 0.5)
c(mean_diff = est1, lm_coef = coef(est2)[["rank_dem"]],
  se = est2$std.error[["rank_dem"]], p = est2$p.value[["rank_dem"]])
```

## Outcome analysis 1: what it says

The within-pair difference is about `r round(est1, 3)` on the $0$--$1$ corruption
scale --- roughly a quarter of the naive gap, and not distinguishable from zero
($p \approx `r round(est2$p.value[["rank_dem"]], 2)`$).

\medskip

Notice that higher-vs-lower democratizers (or backsliders) differ by amounts and we are tossing that when convert the continuous differences to 0-vs-1. Next we use it.

## Outcome analysis 2: use the size of the "dose"

Ranking throws away *how much* more one country democratized. The continuous
version keeps it: relate the within-pair difference in democratization to the
within-pair difference in corruption.

```{r est_cont, echo=TRUE}
## Through the origin: no within-pair dose difference should mean no outcome diff.
est_cont <- lm_robust(corr_diff ~ poly_diff - 1, data = pair_diffs)
c(slope = coef(est_cont)[["poly_diff"]], se = est_cont$std.error[["poly_diff"]],
  p = est_cont$p.value[["poly_diff"]])
```

\medskip

The slope is negative and now distinguishable from zero
($p \approx `r round(est_cont$p.value[["poly_diff"]], 2)`$): pairs with a larger
gap in democratization show a larger drop in corruption.

\medskip

This rests on `r nrow(pair_diffs)` pairs and one design. What about influential
points? What about sensitivity analysis?

## The within-pair relationship, drawn

```{r est_cont_plot, echo=FALSE, out.width=".68\\textwidth", fig.width=6.5, fig.height=4, message=FALSE, warning=FALSE}
ggplot(pair_diffs, aes(x = poly_diff, y = corr_diff)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, formula = y ~ x - 1) +
  geom_smooth(method = "loess", se = FALSE, color = "orange",
              method.args = list(family = "symmetric", degree = 2), span = 0.5) +
  labs(x = "Within-pair difference in democratization (1990-2005)",
       y = "Within-pair difference in corruption (2010)") +
  theme_minimal(base_size = 13)
```

\footnotesize
Each point is one matched pair. Blue: the linear fit; orange: a
robust loess. \normalsize

## Naive vs. design, side by side

```{r compare_tab, echo=FALSE}
naive_d <- coef(lm_robust(corr_10 ~ democratized, data = dat))[["democratized"]]
compare <- data.frame(
  comparison = c("naive difference (dem - non-dem)",
                 "within-pair: higher vs lower",
                 "within-pair dose-response slope"),
  estimate = round(c(naive_d, est1, coef(est_cont)[["poly_diff"]]), 3))
## format = "latex" emits a booktabs tabular (not a longtable), which avoids the
## pandoc/beamer "No counter 'none' defined" error from caption-less longtables.
kable(compare, format = "latex", booktabs = TRUE, row.names = FALSE,
      col.names = c("comparison", "estimate"))
```

# What This Does and Does Not Settle

## The estimand: which countries, which average

What did we estimate? Not a population average treatment effect. We estimated the
**average within-pair difference in corruption, across the matched pairs** ---
the countries similar enough on 1990 covariates to be paired within the GDP
caliper.

\medskip

There is no fixed "treated group" here: every country is the higher-dose member
of *some* potential pair and the lower-dose member of another. 

## The unobserved, again

Balance only certifies the covariates we matched on. The confounded story warns
about forces we did *not* measure --- a reform movement, an external shock ---
that could drive both democratization and corruption.

## How much hidden bias would overturn the dose-response? {.allowframebreaks}

Monday's $\Gamma$ perturbed assignment odds in a matched *test*; the
dose-response is an OLS *slope*, so @cinellihazlett2020 ask the same question in
a regression metric --- how strongly an unobserved pair-level confounder must
relate to *both* the within-pair dose difference and the within-pair corruption
difference before the slope is no longer detectable.

```{r sens_dose, echo=TRUE, message=FALSE}
library(sensemakr)
## estimate, robust SE, and residual df from the through-origin fit (est_cont)
sens_dose <- sensemakr(estimate  = coef(est_cont)[["poly_diff"]],
                       se        = est_cont$std.error[["poly_diff"]],
                       treatment = "poly_diff",
                       dof       = est_cont$df.residual)
round(sens_dose$sensitivity_stats[, c("rv_q", "rv_qa")], 3)
```

\footnotesize
`RV_qa` is the value to report: a confounder explaining only about
`r round(100 * sens_dose$sensitivity_stats$rv_qa, 1)` percent of the residual
variation in *both* within-pair differences would push the slope back across the
$0.05$ line. That is a weak confounder --- the slope is detectable but fragile,
its $p$-value ($\approx `r round(est_cont$p.value[["poly_diff"]], 2)`$) already
at the edge. Driving the point *estimate* to zero takes much more (`RV_q`
$\approx `r round(100 * sens_dose$sensitivity_stats$rv_q)`$ percent); the gap
between the two is the sampling uncertainty `RV_qa` builds in. Whether even a
weak pair-level force like that is plausible is a substantive judgment.
\normalsize

## Other places non-bipartite matching is used

 - **Matching with doses.** @lu:etal:2001 pair communities by covariates to
   study a media campaign against drug abuse whose *intensity* varied --- the
   founding application.
 - **Descriptive norms and COVID attitudes.** @rabb2022pnas use non-bipartite
   matching with a continuous treatment in a recent published design.
 - **Perceptions of place.** @wong2012jop and @wong2025maps pair survey
   respondents to study how perceived neighborhood composition shapes attitudes.

# Summary

## What to take away

 - **Non-bipartite matching** stratifies when there is no binary treatment.
   Pair units close on covariates from a single pool, then read off the
   within-pair difference in the treatment.
 - The **workflow is the one you know**: design before outcomes, check balance
   against a (paired) randomized experiment, then estimate and test as-if
   randomized.
 - The **balance trick** is a pseudo-binary treatment --- rank the two units in a
   pair by their dose --- so the day 8--11 balance tools apply unchanged.
 - **Estimate two ways**: higher-vs-lower (does the more-democratizing country
   differ?) and the continuous dose-response (keeps the magnitude). They answer
   different questions and here they diverge.
 - **The V-Dem lesson**: the strong naive gap
   (`r round(coef(naive)[["democratized"]], 2)`) shrinks to about
   `r round(est1, 2)` --- not distinguishable from zero --- when we compare the
   higher- and lower-democratizing country in matched pairs; yet the
   dose-response, which keeps *how much* each country democratized, stays negative
   and detectable. Same design, two questions, two answers.
 - **Scope and limits**: the estimand is for the matched pairs; balance covers
   only observed covariates; sensitivity analysis ($\Gamma$) handles the rest.
 - **Tools**: `designmatch::nmatch` (with `near`/`far` calipers and a solver)
   [@lu2011optimal; @rosenbaum2020book, ch. 11].

## References {.allowframebreaks}
