---
title: |
 | Statistical Adjustment in Observational Studies,
 | Assessment of Adjustment,
 | Matched Stratification for One and Multiple Variables,
 | Tools for improving the design
date: '`r format(Sys.Date(), "%B %d, %Y")`'
author: |
  | ICPSR 2025 Session 1
  | Jake Bowers \& Ben Hansen \& Tom Leavitt
bibliography:
 - 'BIB/MasterBibliography.bib'
fontsize: 10pt
geometry: margin=1in
graphics: yes
biblio-style: authoryear-comp
biblatexoptions:
  - natbib=true
urlcolor: orange
linkcolor: orange
output:
  beamer_presentation:
    slide_level: 2
    keep_tex: true
    latex_engine: xelatex
    citation_package: biblatex
    template: styles/icpsr-beamer-template
    incremental: true
    includes:
        in_header:
           - defs-all.sty
    md_extensions: +raw_attribute-tex_math_single_backslash+autolink_bare_uris+ascii_identifiers+tex_math_dollars
    pandoc_args: [ "--csl", "chicago-author-date.csl" ]
---


<!-- To show notes  -->
<!-- https://stackoverflow.com/questions/44906264/add-speaker-notes-to-beamer-presentations-using-rmarkdown -->

```{r setup1_env, echo=FALSE, include=FALSE}
library(here)
source(here::here("rmd_setup.R"))
```

```{r setup2_loadlibs, echo=FALSE, include=FALSE}
## Load all of the libraries that we will use when we compile this file
## We are using the renv system. So these will all be loaded from a local library directory
library(dplyr)
library(ggplot2)
library(coin)
library(RItools)
library(optmatch)
```


## Today

  1. Agenda:  Continue to think about an adjustment strategy to enhance
     counterfactual causal interpretations in non-randomized studies:
      - Workflow:
         a. choose basis or covariates that need adjusting $\rightarrow$
         b. create a stratification of the data / a stratified research design (ex. `fullmatch`) $\rightarrow$
         c. choose one or more approaches to evaluate the stratification (ex. `balanceTest`) $\rightarrow$
         d. (probably) improve the stratified design / iterate
      - How to do this with one variable?
      - How to do this with  more than one variable?
      - Effective sample size, Calipers, Exact matching, Missing Data
      - Showing Cardinality matching (`designmatch`)
 2. Questions arising from the reading or assignments or life?


```{r echo=FALSE, cache=TRUE}
load(url("http://jakebowers.org/Data/meddat.rda"))
meddat <- mutate(meddat,
  HomRate03 = (HomCount2003 / Pop2003) * 1000,
  HomRate08 = (HomCount2008 / Pop2008) * 1000
)
row.names(meddat) <- meddat$nh
```

# Recap

## The optmatch workflow: The distance matrix

To minimize overall differences `optmatch` requires a matrix
of those differences (in general terms, a matrix of distances between the
treated and control units)

```{r optm1, echo=TRUE}
tmp <- meddat$nhAboveHS
names(tmp) <- rownames(meddat)
absdist <- match_on(tmp, z = meddat$nhTrt, data = meddat)
absdist[1:3, 1:3]
meddat %>%
  group_by(nhTrt) %>%
  summarize(nhAboveHS[1])
abs(meddat$nhAboveHS[meddat$nhTrt == 1][1] - meddat$nhAboveHS[meddat$nhTrt == 0][1])
```

## Created a Stratified Research Design

`fullmatch` and `pairmatch` use a distance matrix to produce strata.

```{r fm1, echo=TRUE}
fm1 <- fullmatch(absdist, data = meddat)
summary(fm1, min.controls = 0, max.controls = Inf)
table(meddat$nhTrt, fm1)
pm1 <- pairmatch(absdist, data = meddat)
summary(pm1, min.controls = 0, max.controls = Inf)
table(meddat$nhTrt, pm1, exclude = c())
```

## Created a Stratified Research Design: Controlling the algorithm

Maybe we don't want to compare any neighborhoods that differ by more than 6 pct
points in %HS and maybe we want no more than 2 treated per control in any given set:

```{r, echo=TRUE}
quantile(as.vector(absdist),seq(0,1,.1))
absdist[1:5,1:6]
newdist <- absdist + caliper(absdist,.06)
newdist[1:5,1:6]
fm2 <- fullmatch(absdist + caliper(absdist,.06), data = meddat, min.controls = .5)
stratumStructure(fm2)
summary(fm2)

summary(unlist(matched.distances(fm1,absdist)))

summary(unlist(matched.distances(fm2,absdist)))
```


## Evaluate the design: Within set differences

```{r echo=FALSE}
meddat$fm2 <- fm2
meddat$fm1 <- fm1
meddat$pm1 <- pm1
```

Differences within sets versus raw differences.

```{r echo=FALSE, out.width=".9\\textwidth"}
library(gridExtra)
bpfm1 <- ggplot(meddat, aes(x = fm1, y = nhAboveHS)) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", shape = 20, size = 3, color = "red", fill = "red")
meddat$nostrata <- rep(1, 45)
bporig <- ggplot(meddat, aes(x = nostrata, y = nhAboveHS)) +
  geom_boxplot() +
  stat_summary(
    fun = mean, geom = "point",
    shape = 20, size = 3, color = "red", fill = "red"
  )

grid.arrange(bpfm1, bporig, ncol = 2, layout_matrix = matrix(c(1, 1, 1, 1, 2), nrow = 1))
```

## Evaluate the design: Within-set differences
Differences within sets versus raw differences.

```{r echo=FALSE, out.width=".9\\textwidth"}
bpfm2 <- ggplot(meddat, aes(x = fm2, y = nhAboveHS)) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", shape = 20, size = 3, color = "red", fill = "red")

grid.arrange(bpfm2, bporig, ncol = 2, layout_matrix = matrix(c(1, 1, 1, 1, 2), nrow = 1))
```

## Evaluate the design: Within-set differences
Differences within sets versus raw differences.

```{r echo=FALSE, out.width=".9\\textwidth"}
bppm1 <- ggplot(meddat, aes(x = pm1, y = nhAboveHS)) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", shape = 20, size = 3, color = "red", fill = "red")

grid.arrange(bppm1, bporig, ncol = 2, layout_matrix = matrix(c(1, 1, 1, 1, 2), nrow = 1))
```

## Evaluate the design: Inspect within-set differences

Use what we know about Medellin neighborhoods, about Metrocable, to assess the research design:

```{r sdiffs, echo=FALSE}
rawmndiffs <- with(meddat, mean(nhAboveHS[nhTrt == 1]) - mean(nhAboveHS[nhTrt == 0]))
setdiffsfm1 <- meddat %>%
  group_by(fm1) %>%
  summarize(
    mneddiffs =
      mean(nhAboveHS[nhTrt == 1]) -
        mean(nhAboveHS[nhTrt == 0]),
    mnAboveHS = mean(nhAboveHS),
    minAboveHS = min(nhAboveHS),
    maxAboveHS = max(nhAboveHS)
  )

setdiffsfm1 %>% arrange(desc(abs(mneddiffs)))
# summary(setdiffs$mneddiffs)
# quantile(setdiffs$mneddiffs, seq(0,1,.1))
```



## Evaluate the design: Inspect within set differences

Use what we know about Medellin neighborhoods, about Metrocable, to assess the research design:

```{r echo=FALSE, warnings=FALSE}
setdiffspm1 <- meddat %>%
  group_by(pm1) %>%
  summarize(
    mneddiffs =
      mean(nhAboveHS[nhTrt == 1]) -
        mean(nhAboveHS[nhTrt == 0]),
    mnAboveHS = mean(nhAboveHS),
    minAboveHS = min(nhAboveHS),
    maxAboveHS = max(nhAboveHS)
  )

setdiffspm1 %>% arrange(desc(abs(mneddiffs)))
```


## Evaluate the design: Compare to a randomized experiment. {.allowframebreaks}

The observed within-set differences differ from those that would be expected
from a randomized experiment. Here, we compare 3 designs:

```{r xbhs2, echo=TRUE}
xbfm1 <- balanceTest(nhTrt ~ nhAboveHS + strata(fm1) + strata(fm2) + strata(pm1),
  data = meddat
)
xbfm1$results[,,]
xbfm1$overall
```


## What is balanceTest doing?:

```{r xbagain, echo=TRUE}
setmeanDiffs_fm1 <- meddat %>%
  group_by(fm1) %>%
  summarise(
    diffAboveHS = mean(nhAboveHS[nhTrt == 1]) - mean(nhAboveHS[nhTrt == 0]),
    nb = n(),
    nTb = sum(nhTrt),
    nCb = sum(1 - nhTrt),
    hwt = (2 * (nCb * nTb) / (nTb + nCb))
  )
setmeanDiffs_fm1 %>% arrange(desc(abs(diffAboveHS)))
```

## What is balanceTest doing with multiple sets/blocks?

The **test statistic** is a weighted average of the set-specific differences (same
approach as we would use to test the null in a block-randomized experiment)

```{r wtmns, echo=TRUE}
## The descriptive mean difference using block-size weights
with(setmeanDiffs_fm1, sum(diffAboveHS*nTb/sum(nTb)))
## The mean diff used as the observed value in the testing
with(setmeanDiffs_fm1, sum(diffAboveHS * hwt / sum(hwt)))
## Compare to balanceTest output
xbfm1$results[, , "fm1"]
```

## What is balanceTest doing with multiple sets/blocks?

The test statistic is a weighted average of the set-specific differences (same
approach as we would use to test the null in a block-randomized experiment)

```{r wtmns_propertee, echo=TRUE}
library(propertee)
fm1_spec <- obs_spec(nhTrt~uoa(nh03)+block(fm1),data=meddat)
## The descriptive mean difference using block-size weights
coef(lmitt(nhAboveHS~1,specification=fm1_spec,data=meddat,weights="ett"))[["nhTrt."]]
```

## What do we mean by "compare to a randomized experiment"?

Recall that $p$-values require distributions, and distributions require
processes that can be repeated (even in a Bayesian formulation). What are we
repeating in an experiment? (The assignment mechanism).

```{r d1v2, echo=TRUE}
## The strata-adjusted difference in means
dstatv3 <- function(zz, mm, ss) {
  dat <- data.frame(mm = mm, z = zz, s = ss)
  datb <- dat %>%
    group_by(s) %>%
    summarize(
      mndiff = mean(mm[z == 1]) - mean(mm[z == 0]),
      nb = n(),
      pib = mean(z),
      nbwt = nb / nrow(dat),
      hbwt0 = pib * (1 - pib) * nbwt
    )
  datb$hbwt <- datb$hbwt0 / sum(datb$hbwt0)
  adjmn <- with(datb, sum(mndiff * hbwt))
  return(adjmn)
}

## The strata-specific randomization from the experimental analogue
newexp <- function(zz, ss) {
  ## newz <- unlist(lapply(split(zz,ss),sample),ss)
  newzdat <- data.frame(zz = zz, ss = ss) %>%
    group_by(ss) %>%
    mutate(newz = sample(zz))
  return(newzdat$newz)
}
```

## What do we mean by "compare to a randomized experiment"?

Recall that $p$-values require distributions, and distributions require
processes that can be repeated (even in a Bayesian formulation). What are we
repeating in an experiment? (The assignment mechanism).

```{r xbdist, echo=TRUE, messages=FALSE, warning=FALSE}
set.seed(123455)
nulldist <- replicate(10000, with(meddat, dstatv3(zz = newexp(zz = nhTrt, ss = fm1), mm = nhAboveHS, ss = fm1)))
obsdstat <- with(meddat, dstatv3(zz = nhTrt, mm = nhAboveHS, ss = fm1))
perm_p_value <- 2 * min(mean(nulldist >= obsdstat), mean(nulldist <= obsdstat))
perm_p_value
xbfm1$results[,,"fm1"]
```


## What do we mean by "compare to a randomized experiment"?

Recall that $p$-values require distributions, and distributions require
processes that can be repeated (even in a Bayesian formulation). What are we
repeating in an experiment? (The assignment mechanism).

```{r plotxbnulldist, out.width=".6\\textwidth"}
plot(density(nulldist))
curve(dnorm(x,sd=sd(nulldist)),from=min(nulldist),to=max(nulldist),col="blue",add=TRUE)
rug(nulldist)
abline(v = obsdstat)
```

## Matching on the propensity score

```{r echo=TRUE}
theglm <- glm(nhTrt ~ nhPopD + nhAboveHS + HomRate03, data = meddat, family = binomial(link = "logit"))
psdist <- match_on(theglm, data = meddat)
psdist[1:4, 1:4]
fmPs <- fullmatch(psdist, data = meddat)
summary(fmPs, min.controls = 0, max.controls = Inf)
```

## Why a propensity score? {.fragile}

1. The Mahanobis distance weights all covariates equally (and the rank based
   version especially tries to do this). But, maybe not all covariates matter
   equally for $Z$. Or maybe we care most about covariates that drive the
   selection or other process by which units get $Z=1$ versus $Z=0$.


\begin{center}
\begin{tikzcd}[column sep=large,every arrow/.append style=-latex]
& Z  \arrow[from=1-2,to=1-3, "\tau"] &  y \\
x_1 \arrow[from=2-1,to=1-2, "\beta_1" ] \arrow[from=2-1,to=1-3,grey] &
x_2 \arrow[from=2-2,to=1-2, "\beta_2" ] \arrow[from=2-2,to=1-3,grey] &
\ldots &
x_p \arrow[from=2-4,to=1-2,  "\beta_p" near start ] \arrow[from=2-4,to=1-3,grey]
\end{tikzcd}
\end{center}

2. @rosenbaum:rubi:1983 and @rosenbaum:rubi:1984a show that if we knew the true
   propensity score (ex. we knew the true model and the correct covariates)
   then we could stratify on the p-score and have adjusted for all of the
   covariates. (In practice, we don't know either. But the propensity score
   often performs well as an ingredient in a matched design)


## Matching on the propensity score: What do the distance matrix entries mean?

`optmatch` creates a scaled propensity score distance by default --- scaling by,
roughly, the pooled median absolute deviation of the covariate (or here, the
propensity score). So, the distance matrix entries are like standard deviations
--- standardized scores.


```{r echo=TRUE}
psdist[1:4, 1:4]
```

## Matching on the propensity score: What do the distance matrix entries mean? {.allowframebreaks}

What do those distances mean? (They are standardized absolute differences.)

```{r echo=TRUE}
thepscore <- theglm$linear.predictor
thepscore01 <- predict(theglm, type = "response")

## Calculate the simple absolute distances in propensity score
simpdist <- outer(thepscore, thepscore, function(x, y) {
  abs(x - y)
})
simpdist["101", c("401", "402", "403")]
## Notice that these are not the same
psdist["101", c("401", "402", "403")]
## But they are perfectly linearly related
cor(as.vector(simpdist["101",colnames(psdist)]),as.vector(psdist["101",]))
## Notice that they have a fixed relationship
simpdist["101", c("401", "402", "403")] / psdist["101", c("401", "402", "403")]
```

Optmatch scales the scores so that we can interpret them as standardized differences:

```{r}
## Median Absolute Deviations
mad(thepscore[meddat$nhTrt == 1],na.rm=TRUE)
mad(thepscore[meddat$nhTrt == 0], na.rm=TRUE)
## Pooled median absolute deviation (average MAD between the treated and control groups)
(mad(thepscore[meddat$nhTrt == 1],na.rm=TRUE) + mad(thepscore[meddat$nhTrt == 0],na.rm=TRUE)) / 2
pooledmad <- (mad(thepscore[meddat$nhTrt == 1],na.rm=TRUE) + mad(thepscore[meddat$nhTrt == 0],na.rm=TRUE)) / 2
## We can see the actual R function here: optmatch:::match_on_szn_scale
optmatch:::match_on_szn_scale(thepscore, trtgrp = meddat$nhTrt)
simpdist["101", c("401", "402", "403")] / 1.569
## Same as:
psdist["101", c("401", "402", "403")]
```

## Inspecting the Design from a substantive perspective

Do any of these sets look like the differences in baseline homicide rates are
too big within the set?

```{r}
absHomMatch <- fullmatch(absdist, data = meddat)
meddat$absHomMatch <- absHomMatch
setdiffsHM <- meddat %>%
  group_by(absHomMatch) %>%
  summarize(
    mndiffs =
      mean(HomRate03[nhTrt == 1]) -
        mean(HomRate03[nhTrt == 0]),
    mnHomRate03 = mean(HomRate03),
    minHomRate03 = min(HomRate03),
    maxHomRate03 = max(HomRate03)
  )
setdiffsHM
```



# Information and Balance: Matching structure and effective sample size


## Setup

Make some different stratifications:

```{r setupmatches, echo=TRUE}

balfmla <- nhTrt ~ nhClass + nhQP03 + nhPV03 +
    nhTP03 + nhBI03 + nhCE03 + nhNB03 + nhAgeYoung +
    nhAgeMid + nhMarDom + nhOwn +  nhEmp +
    nhAboveHS + nhHS + HomRate03

thebglm <- arm::bayesglm(balfmla, data = meddat, family = binomial(link = "logit"))

mhdist <- match_on(balfmla, data = meddat)
psdist <- match_on(thebglm, data = meddat)
## This is just standardized and centered
hrdist1 <- match_on(nhTrt ~ HomRate03, data = meddat)

## Distance in terms of homicide rate itself
tmp <- meddat$HomRate03
names(tmp) <- rownames(meddat)
hrdist2 <- match_on(tmp, z = meddat$nhTrt, data = meddat)

psCal <- quantile(as.vector(psdist), .9)
mhCal <- quantile(as.vector(mhdist), .9)
hrCal <- quantile(as.vector(hrdist2), .9)

matchdist <- psdist + caliper(psdist, psCal) + caliper(mhdist, mhCal) + caliper(hrdist2, 2)
fm0 <- fullmatch(matchdist, data = meddat)
summary(fm0,min.controls=0,max.controls=Inf,propensity.model=thebglm)
fm1 <- fullmatch(matchdist, data = meddat, min.controls = 1)
summary(fm1, min.controls = 0, max.controls = Inf, propensity.model = thebglm)
## Notice that mean.controls=22/23 drops observations.
fm2a <- fullmatch(matchdist, data = meddat, min.controls = .5, mean.controls = 23 / 22)
summary(fm2a, min.controls = 0, max.controls = Inf, propensity.model = thebglm)
fm3 <- fullmatch(matchdist, data = meddat, mean.controls = .9)
summary(fm3, min.controls = 0, max.controls = Inf, propensity.model = thebglm)

## The larger the differences in psdist, mhdist, and hrdist, the worse the
## matches (by maxdist). This is a penalized match rather than a caliper version.

maxdist <- max(matchdist[!is.infinite(matchdist)])

psdist01 <- psdist / max(as.matrix(psdist))
mhdist01 <- (mhdist - min(as.matrix(mhdist))) / (max(as.matrix(mhdist)) - min(as.matrix(mhdist)))
hrdist201 <- (hrdist2 - min(as.matrix(hrdist2))) / (max(as.matrix(hrdist2)) - min(as.matrix(hrdist2)))

summary(as.vector(psdist01))
summary(as.vector(mhdist01))
summary(as.vector(hrdist201))

matchdistPen <- psdist + psdist01 * maxdist + mhdist01 * maxdist + hrdist201 * maxdist
fm4 <- fullmatch(matchdistPen, data = meddat)
summary(fm4, min.controls = 0, max.controls = Inf, propensity.model = thebglm)

```


<!--
## Showing matches

Fullmatching offers us more choices:

\centering
```{r out.width=".8\\textwidth", echo=FALSE, eval=FALSE}
## perhaps try this https://briatte.github.io/ggnet/#example-2-bipartite-network next time
library(igraph)
blah0 <- outer(fm0, fm0, FUN = function(x, y) {
  as.numeric(x == y)
})
blah1 <- outer(fm1, fm1, FUN = function(x, y) {
  as.numeric(x == y)
})
blah2 <- outer(fm2a, fm2a, FUN = function(x, y) {
  as.numeric(x == y)
})
blah2a <- outer(fm3, fm3, FUN = function(x, y) {
  as.numeric(x == y)
})
par(mfrow = c(2, 2), mar = c(3, 3, 3, 1))
plot(graph_from_adjacency_matrix(blah0, mode = "undirected", diag = FALSE),
  vertex.color = c("white", "green")[meddat$nhTrt + 1], main = "Min Ctrls=0, Max Ctrls=Inf"
)
plot(graph_from_adjacency_matrix(blah1, mode = "undirected", diag = FALSE),
  vertex.color = c("white", "green")[meddat$nhTrt + 1], main = "Min Ctrls=1, Max Ctrls=Inf"
)
plot(graph_from_adjacency_matrix(blah2, mode = "undirected", diag = FALSE),
  vertex.color = c("white", "green")[meddat$nhTrt + 1], main = "Penalties,Min Ctrls=.5, Mean Ctrls=23/22"
)
plot(graph_from_adjacency_matrix(blah2a, mode = "undirected", diag = FALSE),
  vertex.color = c("white", "green")[meddat$nhTrt + 1], main = "Pen V 2,MinCtrls=.5, MeanCtrls=23/22"
)
```
-->

##  Tracking effective sample size

In 2-sample comparisons, total sample size can be a misleading as a measure of information content.  Example:
\begin{itemize}
\item say $Y$ has same variance, $\sigma^{2}$,in the Tx and the Ctl population.
\item Ben H. samples 10 Tx and 40 Ctls, and
\item Jake B. samples 25 Tx and 25 Ctls
\end{itemize}
--- so that total sample sizes are the same.  However,

\begin{eqnarray*}
  V_{BH}(\bar{y}_{t} - \bar{y}_{c}) &=& \frac{\sigma^{2}}{10} + \frac{\sigma^{2}}{40}=.125\sigma^{2}\mbox{;}\\
  V_{JB}(\bar{y}_{t} - \bar{y}_{c}) &=& \frac{\sigma^{2}}{25} + \frac{\sigma^{2}}{25}=.08\sigma^{2}.\\
\end{eqnarray*}

Similarly, a matched triple is roughly $[(\sigma^{2}/1 + \sigma^{2}/2)/(\sigma^{2}/1 + \sigma^{2}/1)]^{-1}= 1.33$ times as informative as a matched pair.

## Details

Use pooled 2-sample t statistic SE formula to compare 1-1 vs 1-2 matched sets' contribution to variance:

$$
\begin{array}{c|c}
  \atob{1}{1} & \atob{1}{2} \\
M^{-2}\sum_{m=1}^{M} (\sigma^{2}/1 + \sigma^{2}/1) & M^{-2}\sum_{m=1}^{M} (\sigma^{2}/1 + \sigma^{2}/2) \\
\frac{2\sigma^{2}}{M} & \frac{1.5\sigma^{2}}{M} \\
\end{array}
$$

So 20 matched pairs is comparable to 15 matched triples.

(Correspondingly, h-mean of $n_{t},n_{c}$ for a pair is 1, while for a triple it's $[(1/1 + 1/2)/2]^{-1}=4/3$.)

The variance of the `Z`-coeff in `y~Z + match` is
$$
 \frac{2 \sigma^{2}}{\sum_{s} h_{s}}, \hspace{3em} h_{s} = \left( \frac{n_{ts}^{-1} + n_{cs}^{-1} }{2}  \right)^{-1} ,
$$

assuming the OLS model and homoskedastic errors.  (This is b/c the anova formulation is equivalent to harmonic-mean weighting, under which $V(\sum_{s}w_{s}(\bar{v}_{ts} - \bar v_{cs})) = \sum_{s} w_{s}^{2}(n_{ts}^{-1} + n_{cs}^{-1}) \sigma^{2} = \sigma^{2} \sum_{s} w_{s}^{2} 2 h_{s}^{-1} = 2\sigma^{2} \sum_{s}w_{s}/\sum_{s}h_{s} = 2\sigma^{2}/\sum_{s} h_{s}$.)

For matched pairs, of course, $h_{s}=1$.  Harmonic mean of 1, 2 is $4/3$. Etc.


## Matching so as to maximize effective sample size {.allowframebreaks}

```{r echo=TRUE, tidy=FALSE}
stratumStructure(fm1)
stratumStructure(fm0)
effectiveSampleSize(fm1)
effectiveSampleSize(fm0)
meddat$fm1 <- fm1
meddat$fm0 <- fm0
wtsfm1 <- meddat %>%
  filter(!is.na(fm1)) %>%
  group_by(fm1) %>%
  summarise(
    nb = n(),
    nTb = sum(nhTrt),
    nCb = nb - nTb,
    hwt = (2 * (nCb * nTb) / (nTb + nCb))
  )
wtsfm1
sum(wtsfm1$hwt)
stratumStructure(fm1)
mean(unlist(matched.distances(fm1, matchdist)))
```


## Matching so as to maximize effective sample size {.allowframebreaks}


```{r echo=TRUE, tidy=FALSE}
wtsfm0 <- meddat %>%
  filter(!is.na(fm0)) %>%
  group_by(fm0) %>%
  summarise(
    nb = n(),
    nTb = sum(nhTrt), nCb = nb - nTb,
    hwt = (2 * (nCb * nTb) / (nTb + nCb))
  )
wtsfm0
sum(wtsfm0$hwt)
stratumStructure(fm0)
mean(unlist(matched.distances(fm0, matchdist)))
```

Notice for `fm0` we weight the 1-to-9 match by `2*(1 * 9)/(9+1)=1.8` and the 5-to-1 match by `2*(5*1)/(5+1)=1.67`.

In a pairmatch all of the sets have weight `2*(1*1)/(1+1) = 2/2=1`.

```{r}
pm1 <- pairmatch(matchdist, data = meddat, remove.unmatchables = TRUE)
summary(pm1, propensity.model = thebglm)
mean(unlist(matched.distances(pm1, matchdist)))
effectiveSampleSize(pm1)
```

## Why does it matter? {.allowframebreaks}

Or see here: Higher effective sample size $\rightarrow$ lower standard error.

```{r echo=TRUE}
stratumStructure(fm1)
stratumStructure(fm2a)
effectiveSampleSize(fm1)
effectiveSampleSize(fm2a)
library(estimatr)
library(propertee)

meddat$fm2a <- fm2a

fm2a_spec <-obs_spec(nhTrt~uoa(nh03)+block(fm2a),
  data=droplevels(meddat[!is.na(meddat$fm2a),]))

specification_table(fm2a_spec,"treatment","blocks")

fm2a_ett_weights<-ett(fm2a_spec,data=droplevels(meddat[!is.na(meddat$fm2a),]))

lm_fm2a <- lm_robust(HomRate08~nhTrt,
  data=droplevels(meddat[!is.na(meddat$fm2a),]),
  weights=fm2a_ett_weights,
  se_type="HC2"
)


fm1_spec <-obs_spec(nhTrt~uoa(nh03)+block(fm1),
  data=droplevels(meddat[!is.na(meddat$fm1),]))

specification_table(fm1_spec,"treatment","blocks")

fm1_ett_weights<-ett(fm1_spec,data=droplevels(meddat[!is.na(meddat$fm1),]))

lm_fm1 <- lm_robust(HomRate08~nhTrt,
  data=droplevels(meddat[!is.na(meddat$fm1),]),
  weights=fm1_ett_weights,
  se_type="HC2"
)

lm_fm1$std.error
lm_fm2a$std.error


```



## Decision Points in Creating Matched Designs

 - Which covariates and their scaling and coding. (For example, exclude
   covariates with no variation!)
 - Which distance matrices (scalar distances for one or two important
   variables, Mahalanobis distances (rank  transformed or not), Propensity
   distances (using linear predictors)).
 - (Possibly) which calipers (and how many, if any, observations to drop. Note
   about ATT as a random quantity and ATE/ACE as fixed.)
 - (Possibly) which exact matching or strata
 - (Possibly) which structure of sets (how many treated per control, how many
   controls per treated)
 - Which remaining differences are tolerable from a substantive perspective?
 - How well does the resulting research design compare to an equivalent
   block-randomized study (`balanceTest`)?
 - (Possibly) How much statistical power does this design provide for the
   quantity of interest?
 - Other questions to ask about a research design aiming to help clarify
   comparisons.


# The separation problem in Logistic Regression

## What about using many covariates? The separation problem in logistic regression {.shrink}

What if we want to match on more than two covariates? Let's step through the following to discover a problem with logistic regression when the number of covariates is large relative to the size of the dataset.

```{r echo=TRUE}
library(splines)
library(arm)
thecovs <- unique(c(names(meddat)[c(5:7,9:24)],"HomRate03"))
balfmla<-reformulate(thecovs,response="nhTrt")
psfmla <- update(balfmla,.~.+ns(HomRate03,2)+ns(nhPopD,2)+ns(nhHS,2))
glm0 <- glm(balfmla,data=meddat,family=binomial(link="logit"))
glm1 <- glm(psfmla,data=meddat,family=binomial(link="logit"))
bayesglm0 <- bayesglm(balfmla,data=meddat,family=binomial(link="logit"))
bayesglm1 <- bayesglm(psfmla,data=meddat,family=binomial(link="logit"))
psg1 <- predict(glm1,type="response")
psg0 <- predict(glm0,type="response")
psb1 <- predict(bayesglm1,type="response")
psb0 <- predict(bayesglm0,type="response")
```


## The separation problem

Logistic regression is excellent at discriminating between groups \ldots often
**too excellent** for us \autocite{gelman2008weakly}. First evidence of this is
big and/or missing coefficients in the propensity score model. See the
coefficients below (recall that we are predicting `nhTrt` with these covariates
in those models):

```{r echo=FALSE}
thecoefs <- rbind(glm0=coef(glm0)[1:20],
      glm1=coef(glm1)[1:20],
      bayesglm0=coef(bayesglm0)[1:20],
      bayesglm1=coef(bayesglm1)[1:20]
      )
thecoefs[,1:5]
```

## The separation problem

```{r, echo=FALSE, out.width=".9\\textwidth"}
par(mfrow=c(1,2))
matplot(t(thecoefs),axes=FALSE)
axis(2)
axis(1,at=0:19,labels=colnames(thecoefs),las=2)

matplot(t(thecoefs),axes=FALSE,ylim=c(-15,10))
axis(2)
axis(1,at=0:19,labels=colnames(thecoefs),las=2)

legend("topright",col=1:4,lty=1:4,legend=c("glm0","glm1","bayesglm0","bayesglm1"))
```


## The separation problem in logistic regression

So, if we are interested in using the propensity score to compare observations in regards the multi-dimensional space of many covariates, we would probably prefer a dimensional reduction model like `bayesglm` over `glm`.

```{r out.width=".9\\textwidth", echo=FALSE}
par(mfrow=c(2,2),mar=c(3,3,2,.1))
boxplot(psg0~meddat$nhTrt,main=paste("Logit",length(coef(glm0))," parms",sep=" "))
stripchart(psg0~meddat$nhTrt,vertical=TRUE,add=TRUE)
boxplot(psg1~meddat$nhTrt,main=paste("Logit",length(coef(glm1))," parms",sep=" "))
stripchart(psg1~meddat$nhTrt,vertical=TRUE,add=TRUE)
boxplot(psb0~meddat$nhTrt,main=paste("Shrinkage Logit",length(coef(bayesglm0))," parms",sep=" "))
stripchart(psb0~meddat$nhTrt,vertical=TRUE,add=TRUE)
boxplot(psb1~meddat$nhTrt,main=paste("Shrinkage Logit",length(coef(bayesglm1))," parms",sep=" "))
stripchart(psb1~meddat$nhTrt,vertical=TRUE,add=TRUE)
```

# Major Matching modes: Briefly, Greed versus Optimal Matching

##  Optimal (communitarian) vs greedy (individualistic) matching {.fragile}

Compare the greedy to optimal matched designs:

\begin{center}
    \begin{tabular}{l|cccc}
      & \multicolumn{4}{c}{Illustrator} \\
Writer & Mo& John & Terry  & Pat \\ \hline
Ben    & 0 & 1 & 1 & 10 \\
Jake   & 10& 0 & 10 & 10 \\
Tom    &  1& 1 & 20 & $\infty$ \\ \hline
    \end{tabular}
  \end{center}

##  Optimal (communitarian) vs greedy (individualistic) matching {.fragile}

Compare the greedy to optimal matched designs:

\begin{center}
    \begin{tabular}{l|cccc}
      & \multicolumn{4}{c}{Illustrator} \\
Writer & Mo& John & Terry  & Pat \\ \hline
Ben    & 0 & 1 & 1 & 10 \\
Jake   & 10& 0 & 10 & 10 \\
Tom    &  1& 1 & 20 & $\infty$ \\ \hline
    \end{tabular}
  \end{center}


Greedy match without replacement has mean distance (0+0+20)/3=6.67. The optimal
match keeps all the obs, and has mean distance (1+0+1)/3=.67.

```{r simpmatch, echo=TRUE, warning=FALSE, messages=FALSE}
bookmat <- matrix(c(0, 1, 1, 10, 10, 0, 10, 10, 1, 1, 20, 100 / 0), nrow = 3, byrow = TRUE)
dimnames(bookmat) <- list(c("Ben", "Jake", "Tom"), c("Mo", "John", "Terry", "Pat"))
pairmatch(bookmat)
fullmatch(bookmat)
```

\note{
*Greedy:* Ben-Mo (0) , Jake-John (0), Tom-Terry (20)
*Optimal:* Ben-Terry (1), Jake-John (0), Tom-Mo (1)
}


# Matching Tricks of the Trade: Calipers, Exact Matching, Missing Data

## Calipers {.allowframebreaks}

The optmatch package allows calipers (which forbids certain pairs from being matched).^[You can implement penalties by hand.] Here, for example, we forbid comparisons which differ by more than 2 propensity score standardized distances.

```{r echo=TRUE}
## First inspect the distance matrix itself: how are the distances distributed?
quantile(as.vector(psdist), seq(0, 1, .1))
```

## Calipers {.shrink}


```{r echo=TRUE}
## Next, apply a caliper (setting entries to Infinite)
psdistCal <- psdist + caliper(psdist, 2)
as.matrix(psdist)[5:10, 5:10]
as.matrix(psdistCal)[5:10, 5:10]
```

## Calipers {.allowframebreaks}

 Here, for example, we forbid comparisons which differ by more than 2 standard
 deviations on the propensity score. (Notice that we also use the
 `propensity.model` option to `summary` here to get a quick look at the balance
 test:)

```{r echo=TRUE}
fmCal1 <- fullmatch(psdist + caliper(psdist, 2), data = meddat, tol = .00001)
summary(fmCal1, min.controls = 0, max.controls = Inf, propensity.model = theglm)
pmCal1 <- pairmatch(psdist + caliper(psdist, 2), data = meddat, remove.unmatchables = TRUE)
summary(pmCal1, propensity.model = theglm)
```

## Calipers {.allowframebreaks}


Another example: We may want to match on propensity distance but disallow any
pairs with extreme mahalnobis distance and/or extreme differences in baseline
homicide rates (here using many covariates all together).


```{r echo=TRUE}
## Create an R formula object from vectors of variable names
balfmla <- reformulate(c("nhPopD", "nhAboveHS"), response = "nhTrt")
## Create a mahalanobis distance matrix (of rank transformed data)
mhdist <- match_on(balfmla, data = meddat, method = "rank_mahalanobis")
## Now make a matrix recording absolute differences between neighborhoods in
## terms of baseline homicide rate
tmpHom03 <- meddat$HomRate03
names(tmpHom03) <- rownames(meddat)
absdist <- match_on(tmpHom03, z = meddat$nhTrt, data = meddat)
absdist[1:3, 1:3]
quantile(as.vector(absdist), seq(0, 1, .1))
quantile(as.vector(mhdist), seq(0, 1, .1))
```

## Calipers

Another example: We may want to match on propensity distance but disallow any pairs with extreme mahalnobis distance and/or extreme differences in baseline homicide rates (here using many covariates all together).

```{r echo=TRUE}
## Now create a new distance matrix using two calipers:
distCal <- psdist + caliper(mhdist, 3) + caliper(absdist, 2)
as.matrix(distCal)[5:10, 5:10]
## Compare to:
as.matrix(mhdist)[5:10, 5:10]
```

## Calipers

Now, use this new matrix for the creation of stratified designs --- but possibly excluding some units (also showing here the `tol` argument. The version with the tighter tolerance produces a solution with smaller overall distances)

```{r echo=TRUE}
fmCal2a <- fullmatch(distCal, data = meddat, tol = .1)
summary(fmCal2a, min.controls = 0, max.controls = Inf, propensity.model = theglm)
fmCal2b <- fullmatch(distCal, data = meddat, tol = .00001)
summary(fmCal2b, min.controls = 0, max.controls = Inf, propensity.model = theglm)

meddat$fmCal2a <- fmCal2a
meddat$fmCal2b <- fmCal2b

fmCal2a_dists <- matched.distances(fmCal2a, distCal)
fmCal2b_dists <- matched.distances(fmCal2b, distCal)

mean(unlist(fmCal2a_dists))
mean(unlist(fmCal2b_dists))
```

## Exact Matching

We often have covariates that are categorical/nominal and for which we really care about strong balance. One approach to solve this problem is match **exactly** on one or more of such covariates. If `fullmatch` or `match_on` is going slow, this is also an approach to speed things up.

```{r echo=FALSE}
meddat$classLowHi <- ifelse(meddat$nhClass %in% c(2, 3), "hi", "lo")
```

```{r echo=TRUE}
dist2 <- psdist + exactMatch(nhTrt ~ classLowHi, data = meddat)
## or mhdist <- match_on(balfmla,within=exactMatch(nhTrt~classLowHi,data=meddat),data=meddat,method="rank_mahalanobis")
## or fmEx1 <- fullmatch(update(balfmla,.~.+strata(classLowHi)),data=meddat,method="rank_mahalanobis")
fmEx1 <- fullmatch(dist2, data = meddat, tol = .00001)
fmEx1 <- fullmatch(dist2, data = meddat, tol = .00001, min.controls=c("hi"=1,"lo"=.5))
summary(fmEx1, min.controls = 0, max.controls = Inf, propensity.model = theglm)
print(fmEx1, grouped = T)
meddat$fmEx1 <- fmEx1
```

## Exact Matching

Showing treatment and control by matched set (including the variable defining
exactly matched sets)

```{r echo=TRUE}
ftable(Class = meddat$classLowHi, Trt = meddat$nhTrt, fmEx1, col.vars = c("Class", "Trt"))
```

## Missing data and matching

What if `nhPopD` had some missing data?

```{r md1, echo=TRUE}
set.seed(12345)
meddat$nhPopD[sample(1:45, 10)] <- NA
summary(meddat$nhPopD)
xb0 <- balanceTest(nhTrt ~ nhPopD + nhAboveHS + HomRate03, data = meddat)
```

We would want to compare units who are equally likely to have `nhPopD` missing. So, we create a new variable:

```{r md2, echo=TRUE}
newdat <- fill.NAs(meddat[, c("nhAboveHS", "nhPopD")])
head(newdat)
stopifnot(all.equal(row.names(newdat), row.names(meddat)))
```

## Missing data and matching

What if `nhPopD` had some missing data?
We would want to compare units who are equally likely to have `nhPopD` missing. So, we create a new variable:

```{r echo=TRUE}
newdat <- cbind(newdat, meddat[, c("nhTrt", "HomRate08", "HomRate03")])
head(newdat)
```

## Missing data and matching

And we include that variable in our balance testing and matching:

```{r echo=TRUE}
theglm <- arm::bayesglm(nhTrt ~ nhAboveHS + nhPopD + nhPopD.NA + HomRate03, data = newdat)
psdist <- match_on(theglm, data = meddat)
maxCaliper(theglm$linear.predictor, z = newdat$nhTrt, widths = c(.1, .5, 1))
balfmla <- formula(theglm)
fm0 <- fullmatch(psdist + caliper(psdist, 2), data = newdat)
summary(fm0, min.controls = 0, max.controls = Inf, propensity.model = theglm)
summary(unlist(matched.distances(fm0, psdist)))
newdat$fm0 <- fm0
xb0 <- balanceTest(update(balfmla,.~.+strata(fm0)), data = newdat)
xb0$overall
xb0$results[,,]
```

## Missing Data and Matching

So:

 1. Missing data on covariates is not a big problem --- such data reveals
    information to us about the units pre-treatment, so we just stratify on it.
    We treat missing data as just another covariate.
 2. Missing data on treatment assignment or the outcome is a bigger problem: we
    will tend to use bounds to report on the range of possible answers in such
    cases.



## Cardinality Matching Example {.allowframebreaks}

Another approach to matching combines different constraints --- attempting to,
for example minimize the sum of distances between units within set while also
maximizing the number of units in the design. See the citations in the
`designmatch` package for papers explaining and applying these ideas.

```{r designmatchsetup, echo=TRUE, eval=TRUE, waring=FALSE, message=FALSE, cache=FALSE}
library(designmatch)
## library(gurobi) ## to make designmatch work faster on bigger datasets you'd want to install this

balfmla <- nhTrt ~ nhClass + nhQP03 + nhPV03 +
    nhTP03 + nhBI03 + nhCE03 + nhNB03 + nhAgeYoung +
    nhAgeMid + nhMarDom + nhOwn +  nhEmp +
    nhAboveHS + nhHS + HomRate03

thebglm <- arm::bayesglm(balfmla, data = meddat, family = binomial(link = "logit"))
## Let's match on the propensity score
meddat$pscore <- thebglm$linear.predictors
## designmatch needs the observations on the data to be in order of treatment assignment
meddat_new <- meddat[order(meddat$nhTrt, decreasing = TRUE), ]
z <- as.vector(meddat_new$nhTrt)
Xmat <- model.matrix(~pscore, data = meddat_new)
thedistmat <- distmat(z, Xmat, digits = 2)
psdist_new <- match_on(nhTrt ~ pscore, data = meddat_new)
distmat <- as.matrix(psdist)
## bmatch prefers distance matrices without so many decimal places
distmat_scaled <- round(distmat / mean(distmat), 2)
dimnames(distmat_scaled) <- dimnames(thedistmat)
## Minimize distances but keep HomRate03 diffs below 2 and nhAboveHS  differences below .3
nearlist <- list(covs = as.matrix(meddat_new[,c("HomRate03","nhAboveHS")]),
                 pairs = c(HomRate03=2,nhAboveHS=.3))
```

## Cardinality Matching Example {.allowframebreaks}


```{r do_bmatch, echo=TRUE}
## solverlist <- list(name='gurobi',approximate=0,t_max=2000,trace=1)
solverlist <- list(name = "highs", approximate = 0, t_max = 1000)
res <- bmatch(
  t_ind = z,
  dist_mat = thedistmat,
  near = nearlist,
  solver = solverlist,
  subset_weight = 1
)

```

```{r}
#' Convert the output into a factor variable for use in analysis
nmatch_to_df <- function(obj, origid) {
  ## We want a factor that we can merge onto our
  ## existing dataset. Here returning a data.frame so that
  ## we can merge --- seems less error prone than using
  ## rownames even if it is slower.
  matchesdat <- data.frame(
    bm = obj$group_id,
    match_id = c(obj$t_id, obj$c_id)
  )
  matchesdat$id <- origid[matchesdat$match_id]
  return(matchesdat)
}

## The nmatch_to_df function creates a column labeled "bm" which contains
## indicators of match/pair membership
res_df <- nmatch_to_df(res, origid = meddat$nh)
res_df$nh <- res_df$id
meddat2 <- left_join(meddat, res_df, by = "nh")
meddat2 <- droplevels(meddat2)
meddat2$dm1 <- meddat2$bm
row.names(meddat2) <- row.names(meddat)
```


```{r designmatcheval}
## Compare to optmatch pairmatch
pm1 <- pairmatch(psdist, data = meddat2)
meddat2$pm1 <- pm1

bt_fmla <- update(balfmla,.~.+strata(dm1) + strata(pm1) )

xbdm <- balanceTest( bt_fmla , data = meddat2, p.adjust.method="hommel")
xbdm$overall

xbdm$results[, "std.diff", ]
xbdm$results[, "p", ]
```


## Summary:

 - Statistical adjustment with linear regression models is hard to justify.
 - Stratification via matching is easier to justify and assess (and describe).
 - We can build a stratified research design allowing an algorithm to do some
   work, and bringing in our substantive expertise and rhetorical strategies
   for other work (choosing the covariates, deciding which covariates should be
   closely matched-on and which not so closely).
 - Matching solves the problem of making comparisons that are transparent
   (Q: "Did you adjust enough for X?" A: "Here is some evidence about
   how well I did.")
 - You can adjust for one variable or more than one (if more than one, you
   need to choose one or more methods for reducing many columns to one
   column).
 - The workflow involves the creation of a distance matrix, asking an
   algorithm to find the best configuration of sets that minimize the
   distances within set, and checking balance. (Eventually, it will also be
   concerned about the effective sample size.)
 - We can restrict certain matches --- using calipers or exact-matching, using
   different distance matrices.
 - We can match on missing values on covariates.
 - Some choices of score creation raise other questions
   and problems (ex. Mahalanobis versus Rank-Based Mahalanobis distances;
   Propensity Score creation without overfitting.)


## References

