---
title: |
 | Sensitivity Analysis
date: '`r format(Sys.Date(), "%B %d, %Y")`'
author: |
  | ICPSR 2026 Session 1
  | Jake Bowers \& Ben Hansen \& Tom Leavitt
bibliography:
 - 'BIB/MasterBibliography.bib'
fontsize: 10pt
geometry: margin=1in
graphics: yes
biblio-style: authoryear-comp
biblatexoptions:
  - natbib=true
output:
  beamer_presentation:
    slide_level: 2
    keep_tex: true
    latex_engine: xelatex
    citation_package: biblatex
    template: styles/icpsr-beamer-template
    incremental: true
    includes:
        in_header:
           - defs-all.sty
    md_extensions: +raw_attribute-tex_math_single_backslash+autolink_bare_uris+ascii_identifiers+tex_math_dollars
    pandoc_args: [ "--csl", "chicago-author-date.csl" ]
---

```{r setup1_env, echo=FALSE, include=FALSE}
library(here)
source(here::here("rmd_setup.R"))
```

```{r setup2_loadlibs, echo=FALSE, include=FALSE}
library(tidyverse)
library(dplyr)
library(optmatch)
library(estimatr)
library(senstrat)
library(sensitivitymv)
library(sensitivityfull)
library(sensemakr)
library(blkvar)
library(propertee)
options(show.signif.stars = FALSE)
```

```{r setup3_data_and_match, echo=FALSE, include=FALSE, cache=TRUE}
## Tom's running example (June 24--26): UN peacekeeping and post-conflict peace.
## We RECALL the design Tom built; we do not re-derive matching here.
peace <- readRDS(url(
  "https://raw.githubusercontent.com/tl2624/matching-guide/main/data/peace_pre_match.rds"))
covs <- c("lwdeaths", "lwdurat", "ethfrac", "pop", "lmtnest",
          "milper", "bwgdp", "bwplty2", "region")

## Estimated propensity score (logit index), as in Tom's deck
psm <- glm(reformulate(setdiff(covs, "region"), "UN"),
           family = binomial("logit"), data = peace)
peace$logit_ps <- psm$linear.predictors
pop_sd_logit <- sqrt(mean((peace$logit_ps - mean(peace$logit_ps))^2))

## Tom's full match: rank-Mahalanobis + a 0.5-SD PS caliper + exact on region
## + calipers on ethfrac (35) and bwgdp (2); at most 4 controls per treated.
fm <- fullmatch(
  match_on(UN ~ logit_ps, caliper = 0.5 * pop_sd_logit, data = peace,
           standardization.scale = NULL, method = "euclidean") +
  match_on(reformulate(setdiff(covs, "region"), "UN"), data = peace,
           standardization.scale = NULL, method = "rank_mahalanobis") +
  match_on(UN ~ ethfrac, caliper = 35, data = peace,
           standardization.scale = NULL, method = "euclidean") +
  match_on(UN ~ bwgdp, caliper = 2, data = peace,
           standardization.scale = NULL, method = "euclidean") +
  exactMatch(UN ~ region, data = peace),
  data = peace, max.controls = 4)
peace$fm <- fm
md <- droplevels(peace[matched(fm), ])

## propertee design spec + ATE weights (used below)
spec <- obs_spec(UN ~ uoa(id) + block(fm), data = md)
md$atew <- ate(spec, data = md)
```

## Today

 1. Return to Tom's question and design (UN peacekeeping).
 2. **Sensitivity analysis**: reasoning about how departures from the "as-if
    randomized" assumption might change our substantive conclusions. (This is
    not "stability" or "fragility" analysis of model specification.)
 3. Two currencies for the same question --- Rosenbaum's $\Gamma$ for the idea,
    Cinelli and Hazlett's `sensemakr` for the tool you will use in practice.

# The substantive question

## Does UN peacekeeping prolong peace? \citep{gilligansergenti2008}

 - **Unit**: a *post-conflict peace episode* (the period of peace after a civil
   war). E.g.\ Sierra Leone, Jan 2001--Dec 2003. There are $87$ episodes,
   1989--2003.
 - **Treatment** (`UN`): $1$ if a UN mission was *present* during the episode,
   else $0$. Here $19$ treated, $68$ control.
 - **Outcome** (`ldur`): log days of peace until conflict resumes (or
   censoring).
 - **The question**: does the presence of a UN mission *prolong* post-conflict
   peace?

\medskip

UN missions are **not randomly assigned**. Where peace is fragile for reasons
we do not see, the UN may be *more* likely to deploy --- exactly the kind of
hidden selection sensitivity analysis is built to reason about.

## The nine covariates Tom matched on

Pre-treatment characteristics thought to drive UN intervention (and to predict
the duration of peace):

 - logged war deaths (`lwdeaths`), logged previous-war duration (`lwdurat`)
 - ethnic fractionalization (`ethfrac`), logged population (`pop`)
 - logged mountainous terrain (`lmtnest`), military personnel (`milper`)
 - logged pre-war GDP per capita (`bwgdp`), Polity score (`bwplty2`)
 - `region` (E.\ Europe, Lat.\ America, Asia, Sub-Saharan Africa, N.\
   Africa / Middle East)

\medskip

The design goal (Tom, June 24--26): build matched sets of episodes that are
similar on these covariates, so that *within a set* a UN mission looks **as-if**
randomly assigned.

# But first, review

## The "as-if randomized" approach

Given a matched design that (a) we can defend on substantive grounds and (b)
compares favorably with a block-randomized experiment on the *observed*
covariates, why not analyze it **as if** treatment had been randomized within
sets? "If this had been an experiment, the estimate would be $\hat{\tau}$,
$p = .03$."

Advantages (Tom's whole second week):

 - No probability model of outcomes to justify (no $Y \sim
   \text{Poisson}(\lambda)$).
 - No functional form for covariates and treatment to justify.
 - No population and sampling needed --- inference is about the pool of episodes
   we have.
 - The CLT enters only for convenience (a Normal reference for the weak null).

## The disadvantage that sensitivity analysis confronts

**This is not a randomized experiment.** A high $p$-value from an omnibus
balance test tells us our comparison carries about as little bias *from the
covariates we balanced* as a block-randomized experiment would. It says
**nothing** about *unobserved* covariates that a real randomization would also
have balanced.

\medskip

All non-randomized studies carry some bias from unobserved confounders. The
question is **how big** that bias would have to be before our conclusion
changes.

## Recall the design Tom built

```{r recall_design, echo=TRUE}
summary(fm)                       # structure of the matched sets
effectiveSampleSize(fm)           # equivalent number of matched pairs
```

\footnotesize
$13$ treated and $23$ control episodes matched into $12$ sets (a mix of $1$:$1$,
$1$:$K$, and one $2$:$1$); effective sample size $\approx 15.3$. This is the
design we now treat as-if randomized.
\normalsize

## From the design to an estimate, an SE, and a CLT

Sensitivity analysis needs a result to be sensitive *about*. Tom's weak-null
analysis gives the average effect of UN presence on log-duration --- and it
does not matter which design-based estimator we use to get it:

```{r est_se_three_ways, echo=TRUE, cache=TRUE}
## (1) ATE-weighted lm_robust (HC2 standard error)
lm_ate <- lm_robust(ldur ~ UN, weights = atew, data = md)

## (2) propertee::lmitt -- design-based, ATE weights
lmitt1 <- lmitt(ldur ~ 1, specification = spec, data = md, weights = "ate")

## (3) blkvar / Fogarty between-set variance -- Tom's June-26 tool
blk <- block_estimator(Yobs = ldur, Z = UN, B = fm, data = md, method = "hybrid_p")

round(c(lm_robust  = coef(lm_ate)[["UN"]],
        propertee  = coef(lmitt1)[["UN."]],
        blkvar     = blk$ATE_hat), 3)
round(c(se_lm_robust = lm_ate$std.error[["UN"]],
        se_blkvar    = blk$se_est), 3)
```

\footnotesize
The ATE is about $\hat{\tau} \approx 0.61$ (log-days), SE $\approx 0.34$--$0.39$.
(Plain `lm_robust` with set fixed effects and *no* weights returns $0.673$
instead --- the harmonic-mean--weighted estimand from Tom's "two weightings"
slide; here we want the ATE, so we use ATE weights.)
\normalsize

## A borderline result --- before we even add hidden bias

```{r design_pvalue, echo=TRUE}
z_obs <- blk$ATE_hat / blk$se_est           # design-based, Fogarty SE
c(z = round(z_obs, 2),
  p_one_sided = round(pnorm(z_obs, lower.tail = FALSE), 3))
```

 - One-sided $p \approx 0.03$--$0.06$ depending on the variance estimator
   (Fogarty $\approx 0.034$; HC2 $\approx 0.056$).
 - The directional question --- does UN presence *prolong* peace? --- makes the
   one-sided test the relevant one.
 - This is a **real but not overwhelming** rejection. Would even a little hidden
   bias overturn it?

# Two ways to reason about the unobserved

## One question, two currencies

*How much hidden bias would it take to overturn our conclusion?* Two traditions
answer in two units.

 - **Rosenbaum ($\Gamma$).** Within a matched set, how many times more likely
   (an odds ratio $\Gamma$) would one episode have to be to receive a UN
   mission --- because of an unobserved $u$ --- before our **randomization
   test** stops rejecting? Currency: a treatment-odds ratio.
 - **Cinelli and Hazlett (partial $R^2$).** How strongly would an unobserved
   confounder have to relate to *both* treatment and outcome before our
   **regression estimate** loses significance (or reaches zero)? Currency:
   shared variance.

\medskip

Same logic, different units. We learn the idea with $\Gamma$ and report it with
`sensemakr`, because you will estimate an effect and pass it to `sensemakr`.

\footnotesize
(Word-collision warning: each tradition calls its threshold a "robustness
value." They are not the same number.)
\normalsize

# Reasoning with $\Gamma$ (the general idea)

## Sensitivity analysis as a formalized thought experiment {.fragile}

> "In an observational study, a sensitivity analysis replaces qualitative claims
> about whether unmeasured biases are present with an objective quantitative
> statement about the magnitude of bias that would need to be present to change
> the conclusions." (Rosenbaum, `sensitivitymv` manual)

\medskip

\begin{center}
\begin{tikzcd}[column sep=large,every arrow/.append style=-latex]
u \arrow[from=1-1,to=2-2, bend right, "\Gamma" near start] \arrow[from=1-1,to=2-3, bend left, "\rho" near start] \\
& Z \arrow[from=2-2,to=2-3, "\tau"] & y \\
x_1 \arrow[from=3-1,to=2-2] \arrow[from=3-1,to=2-3,grey] &
x_2 \arrow[from=3-2,to=2-2] \arrow[from=3-2,to=2-3,grey]
\end{tikzcd}
\end{center}

We match on $x$. The question is what an unobserved $u$ --- related to treatment
($\Gamma$) and outcome ($\rho$) --- could do to the comparison.

## Why $\Gamma$? A model for the unobserved {.fragile}

For two episodes $i$ and $j$ with the same matched-set covariates ($\bx_i =
\bx_j$), @rosenbaum2002observational bounds their ratio of treatment odds by a
single number $\Gamma$:

\begin{equation*}
\frac{1}{\Gamma} \;\leq\; \frac{\pi_i (1 - \pi_j)}{\pi_j (1 - \pi_i)} \;\leq\; \Gamma .
\end{equation*}

 - $\Gamma = 1$: within a set, both episodes are equally likely to get a UN
   mission --- exactly the as-if randomized benchmark.
 - $\Gamma = 2$: an unobserved $u$ could make one episode **twice** as likely as
   its match to receive a mission.

We compute the **worst-case** $p$-value the test could take for each $\Gamma$,
and ask how large $\Gamma$ must be before we no longer reject.

## The search for $\Gamma$ on Tom's design

```{r gamma_search, echo=TRUE, cache=TRUE}
## senstrat handles full matches directly. Effect is POSITIVE
## (UN prolongs peace), so we test the upper tail: alternative = "greater".
pval <- function(g) as.numeric(
  senstrat(md$ldur, z = md$UN, st = md$fm, gamma = g,
           alternative = "greater")$Result["P-value"])

gammas <- seq(1, 2, by = 0.02)
pvals  <- sapply(gammas, pval)
gstar  <- uniroot(function(g) pval(g) - 0.05, c(1, 2))$root
round(c(p_at_Gamma1 = pval(1), p_at_Gamma1.5 = pval(1.5), Gamma_star = gstar), 3)
```

\footnotesize
At $\Gamma = 1$ (as-if randomized) the one-sided $p \approx 0.034$, matching
Tom's June-26 result. The smallest $\Gamma$ that pushes the worst-case $p$ above
$0.05$ is $\Gamma^{*} \approx `r round(gstar,2)`$.
\normalsize

## The result is fragile to hidden bias

```{r gamma_plot, echo=FALSE, out.width=".7\\textwidth"}
par(mar = c(3.2, 3.2, 1, 1), mgp = c(2, .6, 0))
plot(gammas, pvals, type = "l", lwd = 2,
     xlab = expression(Gamma~"(max treatment-odds ratio from unobserved u)"),
     ylab = "Worst-case one-sided p-value", ylim = c(0, 0.2))
abline(h = 0.05, lty = 2)
abline(v = gstar, lty = 3, col = "red")
text(gstar, 0.18, labels = bquote(Gamma^"*" == .(round(gstar, 2))), pos = 4, col = "red")
```

\footnotesize
An unobserved confounder making one episode only $\approx `r round(gstar,2)`$
times more likely than its match to receive a UN mission would already be enough
to overturn the one-sided rejection. **Not much hidden bias is required.**
\normalsize

## Interpreting $\Gamma^{*}$: the amplification {.fragile}

@rosenbaum2009amplification split $\Gamma$ into the confounder's effect on
*treatment* ($\Lambda$) and on the *outcome* ($\Delta$): many $(\Lambda,
\Delta)$ pairs imply the same $\Gamma$.

```{r amplify, echo=TRUE}
g_round <- round(gstar, 1)                       # a round Gamma to decompose
amp <- amplify(g_round, lambda = seq(g_round + 0.1, 4, length = 50))
head(round(amp, 2), 3)                           # (Lambda -> Delta) pairs
```

\footnotesize
Read a column as: "a confounder that raised the odds of UN deployment by
$\Lambda$ *and* the odds of longer peace by $\Delta$ would matter as much as
$\Gamma = `r g_round`$." It forces the naming question: *what* unobserved feature
of a conflict could do that?
\normalsize

# From $\Gamma$ to `sensemakr`

## All `sensemakr` needs: an estimate, an SE, and a CLT

`sensemakr` does not need the matched sets or the randomization distribution ---
only a point estimate, its standard error, and the Normal (CLT) reference we
already use for the weak null. So we feed it the design-based ATE from any of
the estimators above:

```{r sensemakr_estimate, echo=TRUE, cache=TRUE}
sens <- sensemakr(estimate  = coef(lm_ate)[["UN"]],
                  se        = lm_ate$std.error[["UN"]],
                  treatment = "UN",
                  dof       = lm_ate$df.residual)
sens$sensitivity_stats[, c("estimate", "se", "rv_q")]
```

\footnotesize
**Robustness value** (`rv_q` $\approx `r round(sens$sensitivity_stats$rv_q,2)`$):
a confounder explaining about `r round(100*sens$sensitivity_stats$rv_q)`\% of the
residual variation in *both* treatment and outcome would drive the point
estimate to zero. Small RV = fragile.
\normalsize

## Benchmarking against the covariates we *did* observe

Calibrate the hypothetical confounder against the covariates we did measure. Fit
the covariate-adjusted model and ask `sensemakr` to benchmark.

```{r sensemakr_model, echo=TRUE, cache=TRUE}
ncovs <- setdiff(covs, "region")
mod   <- lm(reformulate(c("UN", ncovs), "ldur"), data = md, weights = atew)
sens2 <- sensemakr(model = mod, treatment = "UN",
                   benchmark_covariates = c("bwgdp", "milper"), kd = 1:2)
sens2$bounds[, c("bound_label", "r2dz.x", "adjusted_estimate", "adjusted_t")]
```

\footnotesize
A confounder even **twice as strong as GDP** (`2x bwgdp`) leaves the estimate
positive ($\approx 0.63$) and $t \approx 1.6$. The observed covariates are *weak*
confounders: nothing we measured is nearly strong enough to overturn the result
on its own.
\normalsize

## The contour plot, in one picture

```{r sensemakr_plot, echo=FALSE, out.width=".7\\textwidth"}
plot(sens2)
```

\footnotesize
Axes: a confounder's partial $R^2$ with treatment (x) and outcome (y). Contours:
the adjusted estimate. The labelled points are "as strong as GDP / military
personnel" (and twice as strong). The unadjusted estimate is at the origin.
\normalsize

# Putting it together

## Be honest about what each analysis assumes

 - The **estimate** is design-based: matched sets, as-if randomization, a CLT
   for the SE. No outcome model --- this is Tom's framework, unchanged.
 - The **$\Gamma$ analysis** stays design-based: it perturbs the *assignment*
   probabilities and recomputes a randomization test.
 - The **`sensemakr` analysis** is a *regression* argument: Cinelli and
   Hazlett's partial-$R^2$ result is an omitted-variable-bias bound for linear
   models. Passing a matched ATE to `sensemakr` is a pragmatic approximation,
   not a design-based guarantee.

\medskip

So report it as: "*If* we treat the ATE regression as our working model, a
confounder with partial $R^2 \approx$ RV on treatment and outcome would overturn
us" --- and show the $\Gamma$ alongside, so the same conclusion appears in both
currencies.

## Where the two analyses agree

 - The UN-peacekeeping result is **real but fragile**: one-sided $p \approx
   0.034$ under as-if randomization, but $\Gamma^{*} \approx `r round(gstar,2)`$
   and RV $\approx `r round(sens$sensitivity_stats$rv_q,2)`$.
 - A *modest* unobserved confounder of UN deployment would overturn it. The
   defense has to be **substantive**: is there a feature of conflicts ---
   unmeasured here --- that strongly predicts both UN deployment and the
   duration of peace?
 - Sensitivity analysis does not say such a confounder exists. It tells us how
   hard we must argue that one does not.

## Summary

 - Formal sensitivity analysis lets us use the as-if randomized approach
   **provisionally** rather than fearfully: we ask directly how our conclusion
   would change under hidden bias.
 - **$\Gamma$** (Rosenbaum) builds the intuition: bias as a departure from
   as-if randomization, measured as a treatment-odds ratio.
 - **`sensemakr`** (Cinelli and Hazlett) is the tool you will use after
   estimating an ATE with `lm_robust` / `propertee` / `blkvar`: give it an
   estimate, an SE, and a CLT, read off the robustness value, and benchmark
   against your observed covariates.
 - See @rabb2022pnas for an application in this style, and @hazlett2020 ("Angry
   or Weary?") for the regression-based approach in a violence-and-attitudes
   study.

## References {.allowframebreaks}
