
## Capturing perceptions

White, Engish-speaking, Canadian respondents' reports about "visible minorities" in their hand drawn "local communities".

\centering
```{r echo=FALSE}
par(mfrow=c(1,2))
with(canadamapdat, scatter.smooth(vm.da, vm.community.norm2,
  col = "gray", ylab="Perceptions",xlab="Census Neighborhood (DA)",
  xlim = c(0, 1), ylim = c(0, 1), lpars = list(lwd = 2)
))
with(canadamapdat, scatter.smooth(vm.csd, vm.community.norm2,
  col = "gray", ylab="Perceptions",xlab="Census Municipality (CSD)",
  xlim = c(0, 1), ylim = c(0, 1), lpars = list(lwd = 2)
))
##summary(canadamapdat$vm.community.norm2)
```

## Codebook: Mainly for Rmd file

The variables are: age in years, income as a scale, sex in categories, a
social.capital scale coded to run 0 to 1, country of ancestry in categories,
csd.pop is population of the Census Subdivision (like a municipality), vm.csd
is 2006 proportion visible minority in the CSD, vm.da is proportion visible
minority in the Census Dissemination Area (a small area containing 400--700
persons), and vm.community.norm2 is the proportion of visible minorities
reported by respondents in their map of their local community,
community_area_km is the area within their drawing in square km.

## How to make the case for perceptions?

If we could randomly assign different perceptions to people, we could claim
that differences of perceptions matter (above and beyond and independent of
objective characteristics of the context).

\medskip

What is an observational design that would do this? Match people on objective
context (and maybe covariates) who differ in perceptions.

\medskip

But objective context is continuous not binary: rather than matching $m$ "treated"
to $n-m$ "controls", we want to compare all $n$ with all $n$ respondents.

```{r echo=FALSE}
## Exclude people who did not offer a perception or an outcome
wrkdat0<-canadamapdat[!is.na(canadamapdat$vm.community.norm2) &
		      !is.na(canadamapdat$social.capital01),]
## Take a random sample so that the lecture compiles
set.seed(12345)
wrkdat <- droplevels(sample_n(wrkdat0,500))
wrkdat$vmdaPct <- wrkdat$vm.da * 100 ## express in pct
```

## Create $n \times n$ distance matrices

Our main design compares white, English-speaking, Canadians with similar
neighborhood proportions of visible minorities (as measured by the Canadian Census in 2006).

```{r echo=TRUE}
scalar.dist<-function(v){
	## Utility function to make n x n abs dist matrices
  outer(v, v, FUN = function(x, y) {
    abs(x - y)
  })
}

vmdaDist<-round(scalar.dist(wrkdat$vmdaPct),1)
dimnames(vmdaDist)<-list(row.names(wrkdat), row.names(wrkdat))
## The nbpmatching way (Mahalanobis \equiv standardized in one dimension) takes a while:
##obj.com.dist.mat2<-distancematrix(gendistance(wrkdat[,"vmdaPct",drop=FALSE]))
## compare to tmp<-scalar.dist(wrkdat$vmdaPct/sd(wrkdat$vmdaPct))
wrkdat$vmdaPct[1:4]
diff(wrkdat$vmdaPct[1:4])
vmdaDist[1:4,1:4]
```

## Non-bipartite match

```{r nbp1, echo=TRUE, cache=TRUE}

canada_nearlist <- list(covs = as.matrix(wrkdat[, c("csd.pop","community_area_km")]),
        pairs=c(csd.pop=100000,community_area_km=5))

## Try not to match two people with the same perceptions --- that doesn't add anything to our analysis
canada_farlist <- list(covs = as.matrix(wrkdat[, "vm.community.norm2"]),
    pairs = c(vm.community.norm2 = .1))

canada_pairs <- nmatch(
  dist_mat = vmdaDist,
  near = canada_nearlist,
  far = canada_farlist,
  subset_weight = 1,
  solver = solverlist
)

## Version using nonbimatch
## vmdaDistMat <- distancematrix(vmdaDist)
## nbp1match<-nonbimatch(vmdaDistMat)
## nbp1<-get.sets(nbp1match$matches,remove.unpaired=TRUE)
wrkdat$id <- row.names(wrkdat)
canada_pairs_df <- nmatch_to_df(canada_pairs,origid=wrkdat$id)
nrow(canada_pairs_df)

## So, in matched set 1 (bm==1) we see two neighborhoods:
canada_pairs_df %>% filter(bm==1)

# The nmatch_to_df function creates a column labeled "bm" which contains
wrkdat2 <- inner_join(wrkdat, canada_pairs_df, by = "id")
wrkdat2 <- droplevels(wrkdat2)
stopifnot(nrow(wrkdat2) == nrow(canada_pairs_df))

## Number of matches:
# wrkdat2$bm is the matched set indicator.
stopifnot(length(unique(wrkdat2$bm)) == nrow(wrkdat2) / 2)
nrow(canada_pairs_df)
nrow(wrkdat2)
## Notice some observations were not matched:
nrow(wrkdat)
wrkdat2$nbp1 <- wrkdat2$bm

##wrkdat[names(nbp1),"nbp1"]<-nbp1
##nbp1[1:5]
##table(is.na(wrkdat$nbp1)) ## recall the "ghost message"
```

## Inspect the solution

```{r nbpsol, echo=TRUE }
wrkdat2[order(wrkdat2$nbp1),c("nbp1","vmdaPct","vm.community.norm2")][1:6,]
## table(wrkdat2$nbp1)
nbp1vmdiffs <- tapply(wrkdat2$vmdaPct, wrkdat2$nbp1, function(x) {
  abs(diff(x))
})
nbp1percdiffs <- tapply(wrkdat2$vm.community.norm2, wrkdat2$nbp1, function(x) {
  abs(diff(x))
})
summary(nbp1vmdiffs)
summary(nbp1percdiffs)
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
source(url("http://jakebowers.org/Matching/nonbimatchingfunctions.R"))
```

## Inspect the solution

\centering
```{r out.width=".8\\textwidth"}
nbmplot(wrkdat2,
  yvar = "vmdaPct", xvar = "vm.community.norm2", strata = "nbp1", points = FALSE,
  ylim = range(wrkdat2$vmdaPct)
)
```

## Assess balance

No treatment and control groups to compare. But we can still compare the **relationships** between the adjusted variable (`vmdaPct`) and other covariates conditional on pair. Here using `xBalance` because it can handle continuous treatments.

```{r balnbp1, cache=TRUE }
thecovs <- c(
  "age", "income.coded", "education", "x.years", "sex",
  "csd.pop", "vm.csd", "community_area_km"
)
balfmla<-reformulate(thecovs,response="vmdaPct")
xb1<-xBalance(balfmla,strata=list(unstrat=NULL,nbp1=~nbp1), report="all",data=wrkdat2)
xb1$overall
xb1$results[,c("z","p"),"nbp1"]
```

## Assess balance: Approach with higher-vs-lower

No treatment and control groups to compare. But we can still compare the
**relationships** between which person is higher versus lower on the adjusted
variable (`vmdaPct`) and other covariates conditional on pair.

```{r echo=FALSE}
rank.pairs<-function (x, block) { ## Identify the low and high subj in each pair
	unsplit(lapply(split(x, block), function(x) {
			       rank(x)
				 }), block)
}
```

```{r balnbp1_ranked, cache=TRUE }
wrkdat2$id <- row.names(wrkdat2)
wrkdat2 <- wrkdat2 %>% group_by(nbp1) %>%
    mutate(vmdaPct_ranked=rank(vmdaPct,ties="random")-1)
wrkdat2 <- as.data.frame(wrkdat2)
row.names(wrkdat2) <- wrkdat2$id

wrkdat2 %>% arrange(nbp1) %>% dplyr::select(nbp1,vmdaPct,vmdaPct_ranked) %>% head()

thecovs <- c(
  "age", "income.coded", "education", "x.years", "sex",
  "csd.pop", "vm.csd", "community_area_km"
)
balfmla_ranked<-reformulate(thecovs,response="vmdaPct_ranked")
xb1_ranked<-balanceTest(update(balfmla_ranked,.~.+strata(nbp1)),data=wrkdat2,p.adjust="none")
xb1_ranked$overall
xb1_ranked$results[,,"nbp1"]
```

## Strength of the treatment

The difference in "treatment" within sets varies --- and so we expect the size
of the effect to vary. For example, consider the ratio of  objective context
differences to  perceived context differences:

```{r treatmentstr, echo=TRUE}
summary(nbp1vmdiffs)
summary(nbp1percdiffs)
percDist <- scalar.dist(wrkdat2$vm.community.norm2*100)
da <- vmdaDist[1:5,1:5]
perc <- percDist[1:5,1:5]
da/perc
```


```{r summarizenbp}

## Size of causal driver differences: bigger is better for statistical power later
perc_diffs_by_nbp1 <- wrkdat2 %>% filter(!is.na(nbp1)) %>% group_by(nbp1) %>% summarize(perc_diff=diff(vm.community.norm2))

## Notice no pairs with 0 difference in the designmatch result
summary(abs(perc_diffs_by_nbp1$perc_diff))

```



## Assess hypotheses about effects

Test the hypothesis of no relationship between perceptions as measured by
`vm.community.norm2` and `social capital`.

```{r eval=TRUE,echo=TRUE}
library(coin)
wrkdat2$nbp1F <- factor(wrkdat2$nbp1)
test1<-independence_test(social.capital01~vm.community.norm2|nbp1F,data=wrkdat2[!is.na(wrkdat2$nbp1F),])
test1
```


## Describe the differences within pairs

Does the person who perceives more visible minorities in their community tend
to be higher (or lower) in `social.capital` than the other person in the pair?


```{r}
wrkdat2$scRank<-with(wrkdat2,rank.pairs(social.capital01,nbp1))
wrkdat2$vmCRank<-with(wrkdat2,rank.pairs(vm.community.norm2,nbp1))
wrkdat2[order(wrkdat2$nbp1),c("nbp1","social.capital01","scRank","vm.community.norm2","vmCRank")][1:6,]
with(wrkdat2,tapply(scRank,vmCRank,mean))
```

## Summarize mean differences within pairs

If perceptions matters for social capital then we would expect pairs differing
greatly in subjective context to display greater differences in social capital
than pairs that differ a little.


```{r echo=FALSE,results="hide"}
## By default, this rescales each observation to be the distance from the group mean.
align.by.block<-function (x, block, fn = mean, thenames=NULL) {
	newx<-unsplit(lapply(split(x, block), function(x) {
				     x - fn(x) 
				 }), block)
	if (!is.null(names)) {
		names(newx) <- thenames
	}
	return(newx)
}
```

```{r}
wrkdat2$scMD <- with(wrkdat2, align.by.block(social.capital01, nbp1))
wrkdat2$vmcn2MD <- with(wrkdat2, align.by.block(vm.community.norm2, nbp1))
wrkdat2[order(wrkdat2$nbp1),c("social.capital01","scMD","vm.community.norm2","vmcn2MD","nbp1")][1:4,]
## notice that aligning or pair-mean-centering the data preserves the within
## set relationships
## summary(tapply(wrkdat2$scMD,wrkdat2$nbp1,function(x){ abs(diff(x)) }))
## summary(tapply(wrkdat2$social.capital01,wrkdat2$nbp1,function(x){ abs(diff(x)) }))
lm1 <- lm_robust(scMD ~ vmcn2MD, data = wrkdat2[!is.na(wrkdat2$nbp1), ])
lm1
lm1_fe <- lm_robust(social.capital01~vm.community.norm2,fixed_effects=~nbp1,data=wrkdat2[!is.na(wrkdat2$nbp1), ])
lm1_fe
##library(fixest) ## for more speed with strata-by-strata estimation
```
## Summarize mean differences within pairs

```{r warning=FALSE,cache=TRUE}
lm2 <- lm_robust(scMD~vmcn2MD,data=wrkdat2[!is.na(wrkdat2$nbp1),])
lm2
lm3 <- lm_robust(social.capital01 ~ vm.community.norm2, fixed_effects = ~nbp1, data = wrkdat2, subset = !is.na(wrkdat2$nbp1))
lm3
table(wrkdat2$vmCRank,exclude=c())
lm4 <- lm_robust(social.capital01 ~ I(vmCRank - 1), fixed_effects = ~nbp1, data = wrkdat2, subset = !is.na(wrkdat2$nbp1))
lm4
```

## Summarize mean differences within pairs

If perceptions matter for social capital above and beyond objective context
then we would expect pairs differing greatly in subjective context to display
greater differences in social capital than pairs that differ a little.

```{r}
lm2
lm3
pairdiffs <- wrkdat2 %>%
  filter(!is.na(vmCRank) & !is.na(social.capital01) & !is.na(nbp1)) %>%
  group_by(vmCRank) %>%
  summarize(mnsc = mean(social.capital01))
wrkdat2[order(wrkdat2$nbp1),c("social.capital01","scRank","scMD","vm.community.norm2","vmcn2MD","vmCRank","nbp1")][1:4,]
lm4
```

## Summarize mean differences within pairs

```{r}
summary(wrkdat2$vmcn2MD)
summary(wrkdat2$scMD)
```

Within matched pair, the person who perceives more visible minorities within set tends to report
lower social capital than the person who perceives fewer visible minorities
within set.

\medskip

The largest difference is about `r round(max(wrkdat2$vmcn2MD,na.rm=TRUE),2)`. 

The model predicts that social capital would differ by about `r
coef(lm1)[[2]]*.4` for such a difference. This is about `r
coef(lm1)[[2]]*.4/sd(wrkdat2$scMD,na.rm=TRUE)` of a standard deviation of the
social capital scale. Or about `r
coef(lm1)[[2]]*.4/abs(diff(range(wrkdat2$scMD,na.rm=TRUE)))` of the range.


## Summarize mean differences within pairs

Here is a look at the within-pair differences in perceptions of visible minorities as well as social capital.

```{r smoothplot, out.width=".7\\textwidth", echo=FALSE}
with(wrkdat2,scatter.smooth(vmcn2MD,scMD,span=.3,cex=.7,col="gray",pch=19,lpars=list(lwd=2)))
abline(h=0,lwd=.5)
```


## Summary of matching without groups

 - Workflow in general is the same as matching with groups (covariates,
   distance matrices, optimization to select a stratification, assessment of
   the stratification by comparison to an experiment)
 - Estimation is more flexible --- could look simply at "higher versus lower"
   within  pair, or could average over scores.

## Another estimation approach

\autocite{smith:1997} presents a multi-level modelling approach to taking
matched sets into account. The weights implied here are a bit different from
the weights that we've discussed before (although with pairs they might be more
or less the same). What is the data model? What additional assumptions are involved
here?

```{r lmer, cache=TRUE, message=FALSE, warning=FALSE}
library(lme4)
wrkdat2$vmCbi<-ave(wrkdat2$vm.community.norm2,wrkdat2$nbp1)
lmer1<-lmer(social.capital01~vm.community.norm2+vmCbi+(1|nbp1),data=wrkdat2)
confint(lmer1)["vm.community.norm2",]
## Notice that we may have less information than we thought (some clustered of observation by DA).
table(table(wrkdat2$dauid))
## So maybe:
lmer2<-lmer(social.capital01~vm.community.norm2+vmCbi+(1|nbp1)+(1|dauid),data=wrkdat2)
confint(lmer2)["vm.community.norm2",]
```

