---
title: |
  | Non-bipartite Matching: Stratified Comparisons When Treatment Is Not Binary
date: '`r format(Sys.Date(), "%B %d, %Y")`'
author: |
  | ICPSR 2026 Session 1
  | Jake Bowers \& Tom Leavitt
bibliography:
 - 'BIB/MasterBibliography.bib'
fontsize: 10pt
geometry: margin=1in
graphics: yes
biblio-style: authoryear-comp
biblatexoptions:
  - natbib=true
output:
  beamer_presentation:
    slide_level: 2
    keep_tex: true
    latex_engine: xelatex
    citation_package: biblatex
    template: styles/icpsr-beamer-template
    incremental: true
    includes:
        in_header:
           - defs-all.sty
    md_extensions: +raw_attribute-tex_math_single_backslash+autolink_bare_uris+ascii_identifiers+tex_math_dollars
    pandoc_args: [ "--csl", "chicago-author-date.csl" ]
---

```{r setup1_env, echo=FALSE, include=FALSE}
library(here)
source(here::here("rmd_setup.R"))
```

```{r setup2_loadlibs, echo=FALSE, include=FALSE}
library(tidyverse)
library(dplyr)
library(estimatr)
library(RItools)
library(designmatch)
library(coin)
options(show.signif.stars = FALSE)
```

## Today

0. **Your questions.** Anything from the reading, other classes, or your own
   work?
1. **Review (days 8--11):** stratified designs for a *binary* treatment ---
   match, check balance against a block-randomized experiment, then test or
   estimate as-if randomized, and stress-test with sensitivity analysis.
2. **Today:** what to do when the cause of interest is **not** binary --- a
   degree of democratization, a dose of a campaign, a proportion. We build
   stratified comparisons without ever sorting units into "treated" and
   "control." This is **non-bipartite matching**.
   - The idea: pair units that are *close on covariates*, then read off which
     one got more treatment.
   - The same workflow as before: design first (before outcomes), check
     balance, then estimate and test.

\medskip

Questions? Comments?

## Where this sits in the week

- **Mon (day 10):** sensitivity analysis --- how much hidden bias would
  overturn a matched-design conclusion?
- **Tues (day 11):** "controlling for" in a regression, and why it is not a
  design.
- **Today (day 12):** matching when the treatment is continuous or
  multi-valued.
- **Thurs (day 13):** longitudinal data --- risk-set matching *over time* and
  its contrast with difference-in-differences.

\medskip

Today is the basics. Thursday we let the comparison set change with time.

# Review: Designs for a Binary Treatment

## What we have built so far

For a binary treatment $Z \in \{0, 1\}$ we have a workflow that never looks at
the outcome until the design is settled:

 - **Stratify.** Build matched sets that are similar on covariates $\bx$
   (`optmatch::fullmatch`, distances from `match_on`, calipers, exact matching).
 - **Check.** Does the matched comparison resemble a block-randomized
   experiment on $\bx$? The omnibus balance test and within-set differences
   decide it --- *before* $Y$ is unblinded.
 - **Estimate and test as-if randomized.** Treat the matched sets like the
   blocks of a block-randomized experiment.
 - **Stress-test.** A sensitivity analysis ($\Gamma$) asks how much hidden bias
   would overturn the conclusion.

\medskip

Every step from days 8--11 assumed two groups: a set of treated units and a set
of control units to pair them with.

## The hidden assumption: there are two groups

Matching as we have done it is **bipartite**: two sides --- treated and control
--- and we draw edges only *between* the sides. The optimization pairs each
treated unit with control units, never a treated unit with another treated unit.

\medskip

But many causes we care about do not come as two groups:

 - **A degree, not a switch.** How *democratic* a country becomes; how *much* of
   a media campaign a city received [@lu:etal:2001]; what *proportion* of a
   neighborhood owns their homes.
 - **Several doses.** Low, medium, high exposure, with no natural "control."

\medskip

We could dichotomize --- call the top third "treated" and the bottom third
"control" --- but that discards the middle and throws away the gradation that
carries the information.

## The move: one pool, not two {.fragile}

\textbf{Non-bipartite matching} drops the two-group structure. Start with a
single pool of units. Pair each unit with the one *most similar on covariates*,
wherever it sits on the treatment. Only *after* the pairs are fixed do we ask, within each
pair, which unit had more of the treatment.

\medskip

\begin{center}
\begin{tikzcd}[column sep=large,every arrow/.append style=-latex]
 D \arrow[from=1-1,to=1-3, "\tau = ?"] & & Y \\
 & x \arrow[from=2-2,to=1-1] \arrow[from=2-2,to=1-3] &
\end{tikzcd}
\end{center}

\medskip

$D$ is the treatment (now a degree, not a switch), $Y$ the outcome, $\bx$ the
covariates that drive both. Pairing on $\bx$ is meant to break the
$\bx \rightarrow D$ relationship within pairs, just as stratifying did for a
binary $Z$.

# The Question: Does Democratization Reduce Corruption?

## Two stories about democracy and corruption

A long debate: does democratic governance *reduce* political corruption, or
does the association merely reflect who becomes democratic?

 - **The causal story.** Free press, competitive elections, independent courts,
   legislative oversight --- these constrain officials, so countries that
   democratize should see corruption fall.
 - **The confounded story.** Countries that democratize are not a random sample
   of autocracies. They already have stronger civil society, rising education, a
   growing middle class that demands accountability. Those same forces could
   lower corruption regardless of regime type.

\medskip

If we could randomly assign democratization to autocracies, we could separate
the two. We cannot. So we build an observational approximation.

## The data: V-Dem, autocracies in 1990

The Varieties of Democracy project rates democratic governance for nearly every
country, yearly, from expert surveys [@vdem2025dataset]. We use a curated subset,
1970--2015.

\medskip

To keep the basics clear, fix one baseline year. Take the countries that were
**autocratic in 1990** --- the pool of countries "at risk" of democratizing ---
and follow them:

 - **Covariates** ($\bx$, measured 1990): electoral-democracy index, corruption,
   freedom from physical violence, civil-society participation, rule of law,
   logged GDP per capita (and an indicator for missing GDP).
 - **Treatment** ($D$): how much each country democratized by 2005 --- a *degree*
   (change in the polyarchy index), which we can also coarsen to "reached
   democracy: yes/no."
 - **Outcome** ($Y$): the corruption index in 2010 ($0$ = none, $1$ = high).

## The naive comparison, and why to distrust it

```{r vdem_build, echo=TRUE, cache=TRUE}
load(here::here("data", "vdem_democracy_transitions.rda"))
## Autocracies in 1990 = the pool. Baseline covariates from 1990.
baseline <- vdem_sub %>% filter(year == 1990, v2x_regime %in% c(0, 1)) %>%
  dplyr::select(country_name, country_text_id, regime_90 = v2x_regime,
    polyarchy_90 = v2x_polyarchy, corr_90 = v2x_corr, clphy_90 = v2x_clphy,
    cspart_90 = v2x_cspart, rule_90 = v2x_rule, gdppc_90 = e_gdppc, region = e_regionpol)
treatment <- vdem_sub %>% filter(year == 2005) %>%
  dplyr::select(country_text_id, regime_05 = v2x_regime, polyarchy_05 = v2x_polyarchy)
outcome <- vdem_sub %>% filter(year == 2010) %>%
  dplyr::select(country_text_id, corr_10 = v2x_corr)
dat <- baseline %>% inner_join(treatment, by = "country_text_id") %>%
  inner_join(outcome, by = "country_text_id") %>%
  mutate(democratized = as.integer(regime_05 %in% c(2, 3)),
    polyarchy_change = polyarchy_05 - polyarchy_90,
    gdp_missing = as.integer(is.na(gdppc_90)),
    log_gdppc_90 = log(gdppc_90),
    log_gdppc_90 = ifelse(is.na(log_gdppc_90), median(log_gdppc_90, na.rm = TRUE), log_gdppc_90))
```

```{r naive, echo=TRUE}
naive <- lm_robust(corr_10 ~ democratized, data = dat)
c(diff = coef(naive)[["democratized"]], p = naive$p.value[["democratized"]])
```

## The naive comparison, and why to distrust it

Among `r nrow(dat)` autocracies in 1990, `r sum(dat$democratized)` had reached
democracy by 2005 and `r sum(1 - dat$democratized)` had not. The democratizers
have markedly *lower* 2010 corruption: a difference of about
`r round(coef(lm_robust(corr_10 ~ democratized, data = dat))[["democratized"]], 2)`
on the $0$--$1$ scale ($p \approx `r round(lm_robust(corr_10 ~ democratized, data = dat)$p.value[["democratized"]], 3)`$).

\medskip

Taken at face value: democratization cuts corruption. But this is the
confounded story's natural prediction too. The democratizers differ from the
non-democratizers on exactly the covariates $\bx$ that the second story names.
We need to compare like with like.

# Non-bipartite Matching: The Idea

## From two groups to one pool

\textbf{Bipartite} (days 8--11): a treated side and a control side; the
optimizer draws edges only across the divide and pairs each treated unit with
control(s) to minimize the total treated-to-control distance.

\medskip

\textbf{Non-bipartite}: no divide. Every unit can be paired with every other.
Split the whole pool of $N$ units into $N/2$ pairs so as to minimize the
**total within-pair distance** on covariates. There is no "treated group" until
we look, within each formed pair, at which unit got more of $D$
[@lu2011optimal].

\medskip

Optimal non-bipartite matching solves this partition exactly --- the same kind
of optimization behind `fullmatch`, but on one undivided set of units rather
than across two.

## Why pairs help: the as-if-randomized analogy

Suppose two countries were nearly identical in 1990 --- same corruption, same
civil society, same income --- and then one democratized more than the other.

\medskip

If covariates do not predict *which* one democratized more, then, within that
pair, "which country got the higher dose" looks like the flip of a coin. That is
the paired-randomized-experiment analogy: hold $\bx$ fixed by pairing, and let
the difference in $D$ stand in for an assignment we did not control.

\medskip

The design's job, before any outcome, is to make that analogy defensible: pairs
close enough on $\bx$ that covariates no longer forecast the within-pair
direction of treatment.

## A vocabulary note: "risk set"

The pool of autocracies in 1990 is a **risk set**: the units that were all "at
risk" of the event (democratizing) at a common starting point. We compare units
*within* the risk set, never a future democracy against a country that had
already been democratic for decades.

\medskip

Here the risk set is fixed at one year for simplicity. When the timing of
treatment varies --- countries democratizing in different years --- the risk set
must move with time. That is **risk-set matching**, and it is Thursday's topic
[@rosenbaum2017; @rosenbaum2020book, ch. 12].

# Building the Design (before outcomes)

## Step 1: a distance between every pair of units

Bipartite matching needs distances from treated units to control units.
Non-bipartite matching needs a distance between *every* pair. Here we summarize
each country by its Mahalanobis distance from the pool's covariate center --- one
number per country --- and let the distance between two countries be how far apart
those numbers are.

```{r dist, echo=TRUE, cache=TRUE}
row.names(dat) <- dat$country_text_id
covs <- c("polyarchy_90", "corr_90", "clphy_90", "cspart_90",
          "rule_90", "log_gdppc_90", "gdp_missing")
covmat <- dat %>% dplyr::select(all_of(covs)) %>% as.data.frame()
## One Mahalanobis score per country: how "typical" on the joint covariate dist.
dat$covmh <- mahalanobis(covmat, center = colMeans(covmat), cov = cov(covmat))
## Distance between two countries = absolute difference of their scores.
mhdist <- outer(dat$covmh, dat$covmh, FUN = function(x, y) abs(x - y))
dimnames(mhdist) <- list(row.names(dat), row.names(dat))
mhdist[1:4, 1:4]
```

## Step 2: optimal non-bipartite matching

`designmatch::nmatch` partitions the pool into pairs minimizing total within-pair
distance. We add two **calipers** (`near` constraints): no pair may bridge
countries more than one unit apart on logged GDP, or more than $0.10$ apart on
civil-society participation --- the confounded story's central variable, which we
refuse to leave imbalanced. The open-source HiGHS solver finds the optimum.

```{r match, echo=TRUE, cache=TRUE, results='hide', message=FALSE, warning=FALSE}
## designmatch prefers normalized, rounded distances.
matchdist <- round(mhdist / mean(mhdist), 2)
## Calipers: within 1 unit on logged GDP AND within 0.10 on civil-society part.
nearlist <- list(covs = as.matrix(dat[, c("log_gdppc_90", "cspart_90")]),
                 pairs = c(log_gdppc_90 = 1, cspart_90 = 0.10))
solverlist <- list(name = "highs", approximate = 0, t_max = 1000, trace = 0)
set.seed(12345)
nb <- nmatch(dist_mat = matchdist, near = nearlist, subset_weight = 1, solver = solverlist)
```

```{r match_to_df, echo=TRUE}
## Turn nmatch output into a pair-id we can merge back onto the countries.
nmatch_to_df <- function(obj, origid) {
  d <- data.frame(pair_id = obj$group_id, match_id = c(obj$id_1, obj$id_2))
  d$country_text_id <- origid[d$match_id]; d
}
dat_m <- inner_join(dat, nmatch_to_df(nb, dat$country_text_id), by = "country_text_id") %>%
  droplevels()
c(matched = nrow(dat_m), pairs = length(unique(dat_m$pair_id)), dropped = nrow(dat) - nrow(dat_m))
```

## What the matching produced

The optimizer kept `r nrow(dat_m)` countries in
`r length(unique(dat_m$pair_id))` pairs and set aside the
`r nrow(dat) - nrow(dat_m)` it could not pair within the GDP caliper. A few
pairs:

```{r pairs_examples, echo=FALSE}
dat_m %>% arrange(pair_id) %>%
  transmute(pair = pair_id, country = country_name,
            poly_change = round(polyarchy_change, 2),
            corr_2010 = round(corr_10, 2), logGDP = round(log_gdppc_90, 2)) %>%
  head(6) %>% as.data.frame()
```

\footnotesize
Each pair is two countries close on 1990 covariates. Within a pair they differ
in how much they democratized (`poly_change`) --- that difference is the
treatment contrast we will use. \normalsize

# Checking the Design (still before outcomes)

## The trick: a pseudo-binary treatment within pairs

To borrow the balance machinery from days 8--11, we manufacture a binary
treatment *inside each pair*. Rank the two countries by how much they
democratized; the larger increase gets `rank_dem = 1`, the smaller gets $0$.

```{r rank, echo=TRUE}
dat_m <- dat_m %>% group_by(pair_id) %>%
  mutate(rank_dem = rank(polyarchy_change) - 1) %>% ungroup()
table(dat_m$rank_dem)   # 0.5 marks a tie (both changed identically)
dat_notie <- dat_m %>% filter(rank_dem != 0.5)
```

\medskip

Now "treated" means "the more-democratizing country in its pair." If the pairing
worked, covariates should not predict which country that is --- exactly the
balance question we already know how to ask.

## Balance: does $\bx$ predict who democratized more?

```{r balance, echo=TRUE}
bal_covs <- c("corr_90", "clphy_90", "cspart_90", "rule_90", "log_gdppc_90", "gdp_missing")
balfmla  <- reformulate(bal_covs, response = "rank_dem")
xb <- balanceTest(update(balfmla, . ~ . + strata(pair_id)),
                  data = dat_notie, p.adjust.method = "holm")
xb$overall["pair_id", ]
```

\medskip

The omnibus test is the same one from day 10: a *high* $p$-value is the good
case. Across the covariates jointly, the within-pair imbalance is no larger than
a paired-randomized experiment would routinely produce
($p \approx `r round(xb$overall["pair_id", "p.value"], 2)`$).

## Balance, covariate by covariate

```{r balance_vars, echo=TRUE}
xbv <- data.frame(xb$results[, c("Treatment", "Control", "std.diff", "p"), "pair_id"])
round(xbv[order(xbv$p), ], 3)
```

\footnotesize
The standardized within-pair differences are all small. Because we calipered on
civil-society participation (`cspart_90`) --- the confounded story's central
variable --- it now balances closely rather than tilting toward the
more-democratizing country. The largest remaining difference is baseline
corruption (`corr_90`), still well inside what a paired experiment produces (the
omnibus $p$ above). The design is settled; now we unblind the outcome.
\normalsize

## What the balance check is really doing

Imagine the ideal: within each pair, a coin flip decides which country
democratizes more. Then covariates would predict the winner only as often as
chance. The balance test compares our actual design against that ideal.

\medskip

A useful exercise (and a check on the logic): replace `rank_dem` with a genuine
within-pair coin flip and rerun `balanceTest`. You should see the kind of
$p$-values the null produces --- which is what we want our real design to
resemble. We settle all of this *before* touching 2010 corruption.

# From Design to Effect

## Outcome analysis 1: higher vs. lower democratizer

With the design fixed, unblind the outcome. The most transparent estimate is the
mean, across pairs, of the within-pair difference in 2010 corruption between the
more- and less-democratizing country.

```{r est_binary, echo=TRUE}
pair_diffs <- dat_m %>% filter(rank_dem != 0.5) %>% group_by(pair_id) %>%
  summarize(corr_diff = corr_10[rank_dem == 1] - corr_10[rank_dem == 0],
            poly_diff = polyarchy_change[rank_dem == 1] - polyarchy_change[rank_dem == 0],
            .groups = "drop")
est1 <- mean(pair_diffs$corr_diff)
## Same number, as a within-pair (fixed-effects) regression, with an HC2 SE:
est2 <- lm_robust(corr_10 ~ rank_dem, fixed_effects = ~ factor(pair_id),
                  data = dat_m, subset = rank_dem != 0.5)
c(mean_diff = est1, lm_coef = coef(est2)[["rank_dem"]],
  se = est2$std.error[["rank_dem"]], p = est2$p.value[["rank_dem"]])
```

## Outcome analysis 1: what it says

The within-pair difference is about `r round(est1, 3)` on the $0$--$1$ corruption
scale --- roughly a quarter of the naive gap, and not distinguishable from zero
($p \approx `r round(est2$p.value[["rank_dem"]], 2)`$).

\medskip

Compare with the naive difference of `r round(coef(naive)[["democratized"]], 2)`.
Comparing countries that looked alike in 1990, and balanced on civil society,
shrinks the strong "democratization lowers corruption" gap to a small, uncertain
number. Much of the naive association was the confounded story: democratizers
were already on a different path.

\medskip

The estimators agree by construction --- the mean of within-pair differences *is*
the `rank_dem` coefficient in a pair-fixed-effects regression. But comparing the
higher- against the lower-democratizer discards *how much* more one country
democratized. Next we use it.

## Outcome analysis 2: use the size of the dose

Ranking throws away *how much* more one country democratized. The continuous
version keeps it: relate the within-pair difference in democratization to the
within-pair difference in corruption.

```{r est_cont, echo=TRUE}
## Through the origin: no within-pair dose difference should mean no outcome diff.
est_cont <- lm_robust(corr_diff ~ poly_diff - 1, data = pair_diffs)
c(slope = coef(est_cont)[["poly_diff"]], se = est_cont$std.error[["poly_diff"]],
  p = est_cont$p.value[["poly_diff"]])
```

\medskip

The slope is negative and now distinguishable from zero
($p \approx `r round(est_cont$p.value[["poly_diff"]], 2)`$): pairs with a larger
gap in democratization show a larger drop in corruption. Keeping the dose
magnitude recovers a signal the higher-vs-lower comparison diluted --- in a pair
where one country barely out-democratized the other, "higher vs. lower" is a tiny
treatment, and lumping it with large-gap pairs washes the effect out.

\medskip

Read it cautiously: this rests on `r nrow(pair_diffs)` pairs and one design, and a
through-origin slope is sensitive to a few points. It is a reason to run the
sensitivity tools from Monday, not a settled magnitude.

## The within-pair relationship, drawn

```{r est_cont_plot, echo=FALSE, out.width=".68\\textwidth", fig.width=6.5, fig.height=4, message=FALSE, warning=FALSE}
ggplot(pair_diffs, aes(x = poly_diff, y = corr_diff)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, formula = y ~ x - 1) +
  geom_smooth(method = "loess", se = FALSE, color = "orange",
              method.args = list(family = "symmetric", degree = 1), span = 0.9) +
  labs(x = "Within-pair difference in democratization (1990-2005)",
       y = "Within-pair difference in corruption (2010)") +
  theme_minimal(base_size = 13)
```

\footnotesize
Each point is one matched pair. Blue: the through-origin linear fit; orange: a
robust loess. Both tilt negative --- the dose-response --- but the cloud is wide,
so the slope rests on a modest number of pairs. \normalsize

## Naive vs. design, side by side

```{r compare_tab, echo=FALSE}
naive_d <- coef(lm_robust(corr_10 ~ democratized, data = dat))[["democratized"]]
round(c(`naive difference (dem - non-dem)` = naive_d,
        `within-pair: higher vs lower democratizer` = est1,
        `within-pair dose-response slope` = coef(est_cont)[["poly_diff"]]), 3)
```

\footnotesize
The naive and higher-vs-lower numbers share a scale (a difference in the
corruption level): the naive leans on countries that were never comparable, and
matching shrinks it by about three-quarters --- Tuesday's lesson, reached by
stratifying instead of residualizing. The dose-response slope is a *different*
scale (change in corruption per unit of within-pair democratization difference)
and a different question: not "is the more-democratic country less corrupt?" but
"does more democratization go with less corruption?" The answers diverge because
the higher-vs-lower comparison ignores dose size. \normalsize

# What This Does and Does Not Settle

## The estimand: which countries, which average

What did we estimate? Not a population average treatment effect. We estimated the
**average within-pair difference in corruption, across the matched pairs** ---
the countries similar enough on 1990 covariates to be paired within the GDP
caliper.

\medskip

There is no fixed "treated group" here: every country is the higher-dose member
of *some* potential pair and the lower-dose member of another. The matched
comparison speaks to the units in the overlap --- the same restriction we met as
the difference between an effect for the matched sample and an effect for the
whole population. Countries we dropped, or that anchor an extreme of the dose,
are outside what this design can support.

## The unobserved, again

Balance only certifies the covariates we matched on. The confounded story warns
about forces we did *not* measure --- a reform movement, an external shock ---
that could drive both democratization and corruption.

\medskip

That is exactly Monday's question, and the same tools apply: a sensitivity
analysis ($\Gamma$) asks how strong such an unobserved confounder would have to
be, within pairs, to overturn the conclusion. With a near-zero point estimate
here, the more useful sensitivity question runs the other way: how confident can
we be that the effect is *small*, not that it is nonzero?

## What changes Thursday

We fixed the risk set at 1990. Real histories are staggered: Colombia
democratized around 1991, Indonesia around 1999, Georgia around 2004. A country
that democratized in 1999 should be compared with countries *still autocratic in
1999*, not with ones already democratic for a decade.

\medskip

Letting the risk set move with time is **risk-set matching**
[@rosenbaum2017; @rosenbaum2020book, ch. 12]. Thursday we build it, and set it
against **difference-in-differences**, which answers the same longitudinal
question under different assumptions (no anticipation, parallel trends) --- and
how to bound the conclusion when parallel trends holds only approximately
(HonestDiD).

## Other places non-bipartite matching earns its place

The same machinery, different substance:

 - **Matching with doses.** @lu:etal:2001 pair communities by covariates to
   study a media campaign against drug abuse whose *intensity* varied --- the
   founding application.
 - **Descriptive norms and COVID attitudes.** @rabb2022pnas use non-bipartite
   matching with a continuous treatment in a recent published design.
 - **Perceptions of place.** @wong2012jop and @wong2025maps pair survey
   respondents to study how perceived neighborhood composition shapes attitudes.

\medskip

Whenever the cause is a degree rather than a switch, pairing on covariates lets
you keep the gradation instead of dichotomizing it away.

# Summary

## What to take away

 - **Non-bipartite matching** stratifies when there is no binary treatment.
   Pair units close on covariates from a single pool, then read off the
   within-pair difference in the treatment.
 - The **workflow is the one you know**: design before outcomes, check balance
   against a (paired) randomized experiment, then estimate and test as-if
   randomized.
 - The **balance trick** is a pseudo-binary treatment --- rank the two units in a
   pair by their dose --- so the day 8--11 balance tools apply unchanged.
 - **Estimate two ways**: higher-vs-lower (does the more-democratizing country
   differ?) and the continuous dose-response (keeps the magnitude). They answer
   different questions and here they diverge.
 - **The V-Dem lesson**: the strong naive gap
   (`r round(coef(naive)[["democratized"]], 2)`) shrinks to about
   `r round(est1, 2)` --- not distinguishable from zero --- when we compare the
   higher- and lower-democratizing country in matched pairs; yet the
   dose-response, which keeps *how much* each country democratized, stays negative
   and detectable. Same design, two questions, two answers.
 - **Scope and limits**: the estimand is for the matched pairs; balance covers
   only observed covariates; sensitivity analysis ($\Gamma$) handles the rest.
 - **Tools**: `designmatch::nmatch` (with `near`/`far` calipers and a solver);
   the classic `nbpMatching` package implements the same optimization
   [@lu2011optimal; @rosenbaum2020book, ch. 11].

## References {.allowframebreaks}
