---
title: |
  | Longitudinal Designs: Risk-Set Matching vs Difference-in-Differences
date: '`r format(Sys.Date(), "%B %d, %Y")`'
author: |
  | ICPSR 2026 Session 1
  | Jake Bowers \& Tom Leavitt
bibliography:
 - 'BIB/MasterBibliography.bib'
fontsize: 10pt
geometry: margin=1in
graphics: yes
biblio-style: authoryear-comp
biblatexoptions:
  - natbib=true
output:
  beamer_presentation:
    slide_level: 2
    keep_tex: true
    latex_engine: xelatex
    citation_package: biblatex
    template: styles/icpsr-beamer-template
    incremental: true
    includes:
        in_header:
           - defs-all.sty
    md_extensions: +raw_attribute-tex_math_single_backslash+autolink_bare_uris+ascii_identifiers+tex_math_dollars
    pandoc_args: [ "--csl", "chicago-author-date.csl" ]
---

<!--
DRAFT SKELETON. See day_13_longitudinal_2026_PLAN.md for the full plan and TODOs.
Heavy code chunks are eval=FALSE until: (1) HonestDiD is installed, (2) DiD bib
keys are added to BIB/MasterBibliography.bib, (3) the V-Dem long panel + risk-set
data prep is written. This file compiles as-is (prose + non-evaluated code).
-->

```{r setup1_env, echo=FALSE, include=FALSE}
library(here)
source(here::here("rmd_setup.R"))
```

```{r setup2_loadlibs, echo=FALSE, include=FALSE}
library(tidyverse)
library(dplyr)
library(estimatr)
library(RItools)
## TODO day 13: did, fixest, bacondecomp (installed); HonestDiD (install first).
options(show.signif.stars = FALSE)
```

## Today

0. **Your questions.**
1. **Review (day 12):** non-bipartite matching with a *fixed* 1990 risk set ---
   does democratization reduce corruption? We compared countries that looked
   alike in 1990 and found the strong naive association nearly vanished.
2. **Today:** treatment timing varies across units. Two strategies, two
   assumption families, a sensitivity analysis for each.
   - **Risk-set matching:** compare each newly-treated unit to units *still at
     risk* at that time, similar on history (conditional ignorability).
   - **Difference-in-differences:** do not match; assume *parallel trends*.
   - Stress-test each: Gamma for matching, HonestDiD for DiD.

# Review and the Staggered-Timing Problem

## What day 12 left open

Day 12 fixed the risk set at one baseline year (1990). That is fine for teaching
the mechanics, but real histories are staggered: countries democratize in
different years, and a country already democratic for a decade is not a
counterfactual for one democratizing now.

\medskip

Today we let the comparison set move with time.

## The staggered transitions, drawn

<!-- TODO: timeline of a handful of V-Dem countries and their transition years
     (reuse transition_years from exploration_5). One figure. -->

A country that democratized in 1999 should be compared with countries still
autocratic in 1999 --- not with ones that transitioned in 1991.

# Risk-Set Matching Over Time

## The idea

At each transition year $t$, the newly-democratizing country is matched to
countries **still autocratic at $t$**, chosen to be similar on covariate history
up to $t$ [@rosenbaum2017; @rosenbaum2020book; @lietal2001].

\medskip

This is the day-12 design made dynamic: the risk set is rebuilt at each event
time so comparisons only ever run among units that were jointly "at risk."

## Why "still at risk" matters

Using an already-democratic country as a control looks ahead in time: it lets the
future determine the comparison set. Restricting to still-at-risk units removes
that look-ahead (immortal-time) bias.

## Build it

```{r riskset_build, echo=TRUE, eval=FALSE}
## TODO. Loop over transition years; within each, form the risk set of
## still-autocratic countries and match the transitioner to controls on
## lagged covariate history (t-1). Stack matched sets with a set id.
## See day_13_longitudinal_2026_PLAN.md for the sketch.
```

## Balance and estimate

```{r riskset_estimate, echo=TRUE, eval=FALSE}
## TODO. Balance within the stacked design (day-12 rank trick + balanceTest).
## Estimate post-transition corruption; event-study view.
## Sensitivity: reuse the day-10 Gamma machinery (senstrat).
```

# Difference-in-Differences

## The other strategy: assume parallel trends

Instead of matching on covariates, assume that --- absent treatment --- treated
and control units would have followed **parallel** corruption paths. Then the
post-minus-pre change for the treated, minus the same change for controls,
identifies the effect.

## Two-period DiD: the clean case

```{r did_2x2, echo=TRUE, eval=FALSE}
## TODO. Pick one cohort (one transition year). 2x2 DiD via lm/fixest.
## Show the pre-trend (event-study leads) as the partial, falsifiable check.
```

## Staggered DiD and the TWFE trap

With staggered timing, two-way fixed effects can put **negative weights** on some
comparisons and use **already-treated** units as controls ("forbidden
comparisons") [@goodmanbacon2021; @dechaisemartin2020].

```{r bacon, echo=TRUE, eval=FALSE}
## TODO. Goodman-Bacon decomposition (bacondecomp) on the V-Dem panel.
```

## An estimator that avoids the trap

Callaway--Sant'Anna group-time ATTs, aggregated to an event study, use only
not-yet-treated (or never-treated) controls [@callawaysantanna2021].

```{r cs, echo=TRUE, eval=FALSE}
library(did)
## TODO. att_gt(yname="v2x_corr", tname="year", idname="country_id",
##              gname="first_treat", control_group="notyettreated", data=panel_long)
## aggte(..., type="dynamic"); ggdid().
```

# Sensitivity for Parallel Trends

## Passing a pre-trend test is not proof

Parallel trends never holds exactly, and pre-trend tests have low power: failing
to reject a pre-trend is weak evidence, not a clean bill of health
[@rothetal2023].

## HonestDiD

Bound the post-treatment bias by the magnitude of pre-trend violations and report
**robust confidence sets** [@rambachanroth2023].

 - **Smoothness ($M$):** how much the underlying trend may bend per period.
 - **Relative magnitudes ($\bar{M}$):** post-period violation at most $\bar{M}$
   times the largest pre-period violation.

```{r honestdid, echo=TRUE, eval=FALSE}
## TODO. Feed the event-study betas + (clustered) vcov to HonestDiD.
library(HonestDiD)
hd <- createSensitivityResults_relativeMagnitudes(
  betahat = coef(es), sigma = vcov(es),
  numPrePeriods = nPre, numPostPeriods = nPost,
  Mbarvec = seq(0, 1, by = 0.1), l_vec = rep(1 / nPost, nPost))
## Robust CI vs Mbar; breakdown Mbar* = smallest Mbar whose CI covers 0
## (the DiD analog of Gamma-star).
```

# Putting It Together

## Same question, two designs

\begin{center}
\small
\begin{tabular}{@{}p{2.6cm}p{4.0cm}p{2.2cm}p{1.8cm}@{}}
\toprule
design & identifying assumption & pre-outcome check & sensitivity \\
\midrule
risk-set matching & conditional ignorability given history & balance & $\Gamma$ \\
difference-in-differences & parallel trends (+ no anticipation) & pre-trends (weak) & HonestDiD \\
\bottomrule
\end{tabular}
\normalsize
\end{center}

\medskip

Each design replaces an untestable assumption with one you can state, check
before outcomes, and stress-test. That is the week's lesson, now over time.

## When to prefer which

 - **Few pre-periods, good covariates** -> risk-set matching.
 - **Many clean pre-periods, covariates you cannot measure** -> DiD with a
   HonestDiD sensitivity analysis.

\medskip

The deeper contrast: a matched design **builds** a null world a reader can check;
parallel trends **assumes** one a reader must take on faith.

## Summary

 - Staggered timing breaks the fixed-baseline design of day 12.
 - Risk-set matching rebuilds the comparison set at each event time.
 - DiD trades covariate matching for a parallel-trends assumption; staggered TWFE
   needs modern estimators, and parallel trends needs HonestDiD.
 - Both are designs: assumption stated, checked, and stress-tested.

## References {.allowframebreaks}
