\DocumentMetadata{
  lang = en,
  tagging = on,
  tagging-setup = {
    math/setup = mathml-SE,
    table/header-rows = {1}
  },
pdfstandard = {ua-2,a-4},
}
\documentclass[12pt]{article}

% Use Unicode fonts with LuaLaTeX/XeLaTeX.
% Load by filename (resolved via the texmf tree by kpathsea) rather than by
% system font name, so the document compiles under XeLaTeX and LuaLaTeX even
% when the Libertinus fonts are not registered with the OS / fontconfig.
\usepackage{fontspec}
\setmainfont{LibertinusSerif}[
  Extension    = .otf,
  UprightFont  = *-Regular,
  ItalicFont   = *-Italic,
  BoldFont     = *-Bold,
  BoldItalicFont = *-BoldItalic,
]
\setsansfont{LibertinusSans}[
  Extension    = .otf,
  UprightFont  = *-Regular,
  ItalicFont   = *-Italic,
  BoldFont     = *-Bold,
]
\setmonofont{LibertinusMono}[
  Extension    = .otf,
  UprightFont  = *-Regular,
]

% NO fontenc and NO libertinus package in this setup
% \usepackage[T1]{fontenc}   % <- remove
% \usepackage{libertinus}    % <- remove
\usepackage{unicode-math}
\setmathfont{LibertinusMath-Regular.otf}
\usepackage[letterpaper, top=.5in, bottom=.5in, right=1in, left=1in,includehead,includefoot]{geometry}
% from fancyhdr error messages
\setlength{\headheight}{13.59999pt}
\addtolength{\topmargin}{-1.59999pt}

\usepackage{graphicx}
\usepackage{multirow}
%\usepackage{natbib}
%\usepackage{fullcite} %% to do bib in syllabus
%\newcommand{\fullcite}[1]{\begin{verse} \fullcite{#1}. \end{verse}}

% To debug bib error messages
%\DeclareUnicodeCharacter{0301}{*************************************}
\usepackage[hidelinks]{hyperref}
\hypersetup{
  pdftitle  = {Causal Inference for the Social Sciences},
  pdfauthor = {Jake Bowers, Thomas Leavitt}
}
\usepackage[style=authoryear, backend=biber, maxbibnames=99, maxcitenames=99, uniquelist=false]{biblatex}
\addbibresource{bib_2027.bib}

\usepackage{parskip}
\usepackage{bookmark}
\usepackage{rotating}
\usepackage{pdflscape}

\usepackage{setspace}
\setcounter{secnumdepth}{4}

\usepackage{fancyhdr}
\fancypagestyle{myfancy}{%
  \fancyhf{}
  % \fancyhead[R]{\small{Page~\thepage}}
  \fancyfoot[R]{\footnotesize{Version~of~\today -- Causal Inference 1 -- ICPSR 2027-- \thepage}}
  % \fancyfoot[R]{\small{\today -- Jake Bowers}}
  \renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}}

\usepackage{soul}
% titlesec 2.17 is incompatible with the tagged-PDF sectioning installed by
% \DocumentMetadata{tagging=on}: it patches the legacy \@startsection machinery,
% which the tagging code has replaced, so every \section/\subsection fails with
% "Package titlesec Error: No format for this command" plus an "Illegal unit of
% measure" cascade. We previously loaded it only for compact spacing
% (options [compact,nobottomtitles*]; no \titleformat/\titlespacing were used).
% Reproduce the compact spacing here through the kernel's own \@startsection,
% which the tagging code understands, so headings stay tagged and compile clean.
% \usepackage[compact,nobottomtitles*]{titlesec}
\makeatletter
\renewcommand\section{\@startsection{section}{1}{\z@}%
  {-2.0ex \@plus -0.5ex \@minus -.2ex}%
  {1.0ex \@plus .2ex}%
  {\normalfont\Large\bfseries}}
\renewcommand\subsection{\@startsection{subsection}{2}{\z@}%
  {-1.5ex \@plus -0.5ex \@minus -.2ex}%
  {0.6ex \@plus .2ex}%
  {\normalfont\large\bfseries}}
\renewcommand\subsubsection{\@startsection{subsubsection}{3}{\z@}%
  {-1.2ex \@plus -0.5ex \@minus -.2ex}%
  {0.4ex \@plus .2ex}%
  {\normalfont\normalsize\bfseries}}
\makeatother

\title{Causal Inference for the Social Sciences}

\input{courseedition} %misnomer, as of now file only carries date

\author{%
  Jake Bowers\textsuperscript{1} \and
  Thomas (Tom) Leavitt\textsuperscript{2}
}

\begin{document}
\pagestyle{myfancy}
\maketitle

\begin{center}\small
  \textsuperscript{1}Departments of Political Science and Statistics,
  University of Illinois at Urbana–Champaign; \texttt{jwbowers@illinois.edu}\\
  \textsuperscript{2}Marxe School of Public and International Affairs,
  Baruch College, City University of New York (CUNY);
  \texttt{thomas.leavitt@baruch.cuny.edu}
\end{center}

% CONFIRM FOR 2027: teaching assistant name/affiliation and meeting room carried
% over from 2026; update once ICPSR assigns them.
\begin{center}\small
  Teaching Assistant: Alice Malmberg, Department of Political Science,
  University of California, Davis; \texttt{ammalmberg@ucdavis.edu}
\end{center}

\begin{center}\small
  Class meets in the North Quadrangle Residential and Academic Complex (NQ),
  Room~1110 (NQ~1110).
\end{center}

\begin{abstract}
\noindent This course introduces methods and concepts used to infer causal effects from comparisons of intervention and control groups. We'll use the potential outcomes framework of causality to show how a study's research design provides a foundation for estimation and testing. We focus, first, on properties of estimators and tests in randomized experiments, e.g., unbiasedness, consistency, controlled error rates. We then turn to research designs that are either partially controlled (e.g., experiments with noncompliance and/or attrition) or uncontrolled (e.g., observational studies). For observational studies, we focus primarily on matching methods implemented via \texttt{optmatch} and related packages in \texttt{R} --- from matched pairs and full matching for a binary treatment, to non-bipartite matching for continuous or multivalued treatments, to risk-set matching for panel data, which we compare with difference-in-differences. Finally, we turn to sensitivity analysis --- namely, how to assess how inferences would change should certain assumptions about the research design be false. Examples throughout the course are drawn from economics, political science, public health, and sociology.

We assume familiarity with linear algebra and strong knowledge of
statistical concepts, such as sampling distributions, statistical
inference, and hypothesis testing. Demonstrations, examples and
assignments make extensive use of \texttt{R}.
\end{abstract}

% \section{Meetings}

%\input{meetings}

\clearpage

%\nobibliography*

\section*{Overview}

We may all warn our freshmen that association is not causation, but inferring causation has always been a central aim both for statisticians and for their collaborators. Until recently, however, inference of causation from statistical evidence depended on murky, scarcely attainable requirements; in practice, the weight of causal arguments was largely determined by the scientific authority of the people making them.

Requirements for causal inference become more clear when they are framed in terms of \emph{potential outcomes}.  This was first done by Neyman, who in the 1920s used potential outcomes to model agricultural experiments.  Fisher independently proposed a related but distinct, ultimately more influential, analysis of experiments in 1935, and a rich strain of causal analysis developed among his intellectual progeny. It clarified the differing requirements for causal inference with experiments and with observational data, isolating the distinct contributions required of the statistician and of his disciplinary collaborators; generated more satisfying methods with which to address potential confounding due to measured variables; qualitatively and quantitatively advanced our grasp of unmeasured confounding and its potential ramifications; furnished statistical methods with which to eke more out of the strongest study designs, under fewer assumptions; and articulated principles with which to understand study designs as a spectrum, rather than a dichotomy between ``good" experiments and ``bad" observational studies. Understanding the methods and outlook of the school founded by Fisher's student W. G. Cochran will be the central task of this course.

The course begins by applying the Fisher and Neyman-Rubin approaches to statistical inference for counterfactual causal effects to randomized experiments, touching on considerations specific to clustered treatment assignment, ``small'' sample sizes and treatment effect heterogeneity. The next segment addresses conceptual and methodological challenges of applying the same models to analysis of non-experimental data. This course segment covers ignorability, selection, ``common support,'' covariate balance, paired comparisons, optimal matching and propensity scores. With these foundations in place, the course then turns to sensitivity analysis --- namely, how to make principled assessments of how inferences would change should crucial assumptions be false. Over the course's three weeks, the course becomes progressively less conceptual and more applied with increasing emphasis on computing strategies in R.

\section*{Administrative}

\subsection*{Textbooks}

The main texts for the course are

\begin{verse} \fullcite{rosenbaum2017} \end{verse}

\begin{verse} \fullcite{rosenbaum2010} \end{verse}

\begin{verse} \fullcite{rosenbaum2002a} \end{verse}

These three textbooks are presented in varying difficulty and we will draw from all three. Although we won't follow these books closely, their goals and methods align with the course's, and they will be useful as references and supplements.

Other texts that we draw on include

\begin{verse} \fullcite{gerbergreen2012}  \end{verse}

\begin{verse} \fullcite{imbensrubin2015} \end{verse}

\begin{verse} \fullcite{ding2024} \end{verse}

\begin{verse} \fullcite{rosenbaum2025} \end{verse}

% If you'll be using experiments, propensity scores, regression
% discontinuity, or instrumental variables in your work, then each of
% these is a worthwhile investment.

%Several other graduate-level monographs focus or touch
%on causal inference.  Texts that we've found to contain helpful
%discussion include:

%\begin{verse}

%\fullcite{angristpischke2008}

%\fullcite{aronowmiller2019}

%\fullcite{berk2004}

%\fullcite{gelmanhill2006}

%\fullcite{morganwinship2015}

%\fullcite{murnanewillett2010}

%\fullcite{dunning2012}

%\end{verse}

Other readings will be assigned and distributed electronically.

If you're new to \texttt{R}, we suggest getting a hold of:

\begin{verse} \fullcite{fox2016}  \end{verse}

\begin{verse} \fullcite{wickhamgrolemund2017} \end{verse}

\texttt{R} software will be required for several specific segments of the course. With some independent effort, students not familiar with \texttt{R} in advance should be able to learn enough \texttt{R} during the course to complete these assignments. We also recommend some work with \texttt{R} --- for example, via working through some online \texttt{R} courses --- before the course for students who have never used it before.

\subsection*{Assignments}
%Computer or pencil and paper exercises will be assigned periodically and collected on Mondays and on Thursdays at the beginning of class.
Assignments are due each Friday, at the beginning of
class. Parts of the assignment will be given at the beginning of the week, but other parts will be given during class, over the course of the week.  Late homework will not be accepted without cause (or prior arrangement with a teaching assistant).

Participation is expected. It can take various forms:
\begin{enumerate}
\item Doing in-class exercises and discussing them with your peers;
\item From time to time, making a clarification or raising a clarifying question;
\item \label{it:part0} Contributing to in-class discussions;
%\item \label{it:part0} Using a github pull request to suggest a clarification or other enhancement to a course slide or worksheet.
\item \label{it:part1} Drop by one of the professor's office hours to share a point that you \textit{and at least one classmate} would like to have clarified or amplified, or to point out a connection to your field;
%\item \label{it:part2} Give a 5-10 minute in-class presentation of a paper in your field that uses methods or designs we're discussing in the course.
\end{enumerate}
If you are taking the course for a grade, make a point of doing at
least one of \ref{it:part0} and \ref{it:part1}. %and \ref{it:part2}.  There'll be an electronic sign-up for \ref{it:part2}.

% Course participants will be expected either to give a presentation or to submit a short paper about assigned readings related to their fields of study.  Presentations will be given alone or in pairs, and last 10-20 minutes; short papers should be 5-8 pages long.  Students must decide no later than the end of the second week whether they’ll present or submit a paper, and on what topic; presentations will be given at times designated by the instructor, whereas papers must be submitted by the beginning of the fourth week.

\newpage
\begin{landscape}
\section*{Course Schedule}\label{sec:schedule}

The schedule appears below; the Course content section gives fuller readings for each topic. The Additional topics section collects further readings --- on quasi-experimental designs, interference, weighting, synthetic control, and more --- that we draw on selectively and that students may pursue on their own.

\begingroup
\setlength{\tabcolsep}{4pt}
\renewcommand{\arraystretch}{0.9}
\begin{center}
{\small\textbf{Table 1:} Course schedule by date, instructor, topic, readings, and application}
\par\vspace{2pt}
\resizebox{7.4in}{!}{\begin{tabular}{|l|l|l|l|l|}
\hline
\textbf{Date} & \textbf{Instructor} & \textbf{Topic} & \textbf{Required readings} & \textbf{Application} \\ \hline
Mon, June 14 & Tom & \begin{tabular}{@{}l@{}} Introduction: random allocation, \\ potential outcomes, and Fisher's exact test \end{tabular} & \begin{tabular}{@{}c@{}} Holland (1986, \S~1--4) \\ Rosenbaum (2017, Chapter 2) \end{tabular} & \multirow{4}{*}{Arceneaux (2005)} \\ \cline{1-4}
Tues, June 15 & Tom & \begin{tabular}{@{}l@{}} The sharp (Fisherian) framework: testing \\ sharp null hypotheses and constant-effect \\ models \end{tabular} & \begin{tabular}{@{}c@{}} Rosenbaum (2017, Chapter 3) \\ Fisher (1935, \S~5--10) \end{tabular} & ~ \\ \cline{1-4}
  Wed, June 16 & Tom & \begin{tabular}{@{}l@{}} The weak (Neyman) framework I: average \\ effects and the variance of the \\ difference-in-means \end{tabular} & \begin{tabular}{@{}c@{}} Gerber and Green (2012, Ch.~2, pp.~51--61) \\ Aronow and Middleton (2013) \\ Imbens and Rubin (2015, pp.~87--98) \end{tabular} & ~ \\ \cline{1-4}
Thurs, June 17 & Tom & \begin{tabular}{@{}l@{}} The weak framework II: conservative \\ variance, hypothesis testing, and \\ confidence intervals \end{tabular} & \begin{tabular}{@{}c@{}} Gerber and Green (2012, Ch.~3) \\ Freedman et al.\ (1998) \end{tabular} & ~ \\ \hline
  Fri, June 18 (HW 1 due) & -- & No Class (see calendar note) & & ~ \\ \hline
Mon, June 21 & Tom & \begin{tabular}{@{}l@{}} Covariance adjustment: blocking, gain \\ scores, and regression adjustment of \\ experiments \end{tabular} & \begin{tabular}{@{}c@{}} Gerber and Green (2012, Ch.~4) \\ Rosenbaum (2002); Lin (2013) \end{tabular} & \multirow{2}{*}{\begin{tabular}{@{}c@{}} Albertson and \\ Lawrence (2009) \end{tabular}} \\ \cline{1-4}
Tues, June 22 & Tom & \begin{tabular}{@{}l@{}} Noncompliance and attrition: the CACE, \\ instrumental variables, and bounds \end{tabular} & \begin{tabular}{@{}c@{}} Gerber and Green (2012, Ch.~5--7) \\ Rosenbaum (1996); Rosenbaum (2010, \S~5.3) \end{tabular} & ~ \\ \hline
Wed, June 23 & Tom & \begin{tabular}{@{}l@{}} Observational studies and building a \\ matched design: distances and forming \\ matches \end{tabular} & \begin{tabular}{@{}c@{}} Leavitt and Miratrix (2026, \S~1--2.2.4) \\ Rosenbaum (2017, pp.~65--96) \\ Hansen (2011) \end{tabular} & \multirow{5}{*}{\begin{tabular}{@{}c@{}} Gilligan and \\ Sergenti (2008) \end{tabular}} \\ \cline{1-4}
  Thurs, June 24 & Tom & \begin{tabular}{@{}l@{}} Evaluating a matched design: balance \\ and effective sample size \end{tabular} & \begin{tabular}{@{}c@{}} Leavitt and Miratrix (2026, \S~2.2.5) \\ Hansen and Bowers (2008) \end{tabular} & ~ \\ \cline{1-4}
Fri, June 25 (HW 2 due) & Tom & \begin{tabular}{@{}l@{}} Inference under as-if randomization: \\ the sharp and weak frameworks \end{tabular} & \begin{tabular}{@{}c@{}} Leavitt and Miratrix (2026, \S~2.3) \\ Fogarty (2018); Pashley and Miratrix (2020) \end{tabular} & ~ \\ \cline{1-4}
Mon, June 28 & Jake & \begin{tabular}{@{}l@{}} Sensitivity analysis: hidden bias in a \\ matched design ($\Gamma$, amplification, \\ regression-based bounds) \end{tabular} & \begin{tabular}{@{}c@{}} Leavitt and Miratrix (2026, \S~2.4) \\ Rosenbaum (2017, Ch.~9); Rosenbaum (2018) \\ Cinelli and Hazlett (2020) \end{tabular} & ~ \\ \cline{1-4}
Tues, June 29 & Jake & \begin{tabular}{@{}l@{}} Why ``controlling for'' is not a research \\ design: specification curves, extrapolation, \\ and influential points \end{tabular} & \begin{tabular}{@{}c@{}} Gelman and Hill (2006, \S~9.0--9.2) \\ Berk (2010); Fox (2016, Ch.~11, 19) \end{tabular} & ~ \\ \hline
Wed, June 30 & Jake & \begin{tabular}{@{}l@{}} Non-bipartite matching: continuous and \\ multivalued treatments, and dose-response \end{tabular} & \begin{tabular}{@{}c@{}} Lu et al.\ (2001); Lu et al.\ (2011) \\ Rosenbaum (2010, pp.~207--221) \end{tabular} & \multirow{2}{*}{\begin{tabular}{@{}c@{}} V-Dem: democra- \\ tization and \\ corruption \end{tabular}} \\ \cline{1-4}
Thurs, July 1 (HW 3 due) & Jake & \begin{tabular}{@{}l@{}} Longitudinal designs: risk-set matching \\ versus difference-in-differences; parallel \\ trends and HonestDiD \end{tabular} & \begin{tabular}{@{}c@{}} Ham and Miratrix (2024) \\ Rambachan and Roth (2023) \\ Rosenbaum (2017, pp.~162--167) \end{tabular} & ~ \\ \hline
\end{tabular}}
\end{center}
\endgroup

\end{landscape}
\newpage
\section*{Course content}

\section{Randomized experiments}

\subsection{Introduction: Random allocation, potential outcomes and Fisher's exact test}

\paragraph*{Required}

\begin{verse} \fullcite{holland1986}, Sections~1 -- 4. \end{verse}

\begin{verse} Chapter 2 of \fullcite{rosenbaum2017}. \end{verse}

\paragraph*{Recommended}

\begin{verse} Section~1.2, ``Experimentation defined,'' of \fullcite{kinderpalfrey1993}. (Particularly pp. 5 -- 10.) \end{verse}

\begin{verse} Chapter~1, ``Introduction,'' of \fullcite{fisher1935a} and pages 131 -- 135 of \fullcite{box1978} for historical context. \end{verse}

\begin{verse} \fullcite{neyman1923} \end{verse}

\begin{verse} \fullcite{rubin1974} \end{verse}

\begin{verse} \fullcite{bowersleavitt2020} \end{verse}

\subsection{The sharp framework: Fisherian inference for causal effects}

\subsection*{Application}

\begin{verse} \fullcite{arceneaux2005} \end{verse}

\paragraph*{Required}

\begin{verse} Chapter 3 of \fullcite{rosenbaum2017}. \end{verse}

\begin{verse} Sections~5 -- 10 (pp.~11 -- 19) of \fullcite{fisher1935a}. \end{verse}

\paragraph*{Recommended}

\begin{verse} Pages 27 -- 49 of \fullcite{rosenbaum2002a} \end{verse}

\begin{verse} Chapter 2 of \fullcite{rosenbaum2010}. \end{verse}

\begin{verse} \fullcite{caugheyetal2023} \end{verse}

\begin{verse} \fullcite{kimetal2026} \end{verse}

\begin{verse} \fullcite{gelman2003} \end{verse}

\subsection{The weak framework I: Estimating average effects and the variance of the difference-in-means}

\paragraph*{Required}

\begin{verse} Chapter~2 and pp. 51 -- 61 of \fullcite{gerbergreen2012} \end{verse}
\begin{verse} \fullcite{aronowmiddleton2013} \end{verse}
\begin{verse} \fullcite{middletonaronow2015} \end{verse}

\begin{verse} Endnote spanning pages A-32 and 33,
  \fullcite{freedmanpisanipurves1998}.  (This can be read as a pr{\'e}cis
  of: \fullcite{neyman1923}.)
\end{verse}

\begin{verse} Chapter 6, pp. 87 -- 98 of \fullcite{imbensrubin2015} \end{verse}

\subsection{The weak framework II: Conservative variance estimation and hypothesis testing}

\paragraph*{Required}

\begin{verse}
  Chapter~3 of \fullcite{gerbergreen2012}.
\end{verse}

\paragraph*{Recommended}

\begin{verse} Chapter~6 of \fullcite{dunning2012} \end{verse}

\begin{verse} \fullcite{liding2017} \end{verse}

\begin{verse} \fullcite{ding2017a} \end{verse}

\begin{verse} \fullcite{aronowetal2014} \end{verse}

\begin{verse} \fullcite{robins1988} \end{verse}

\begin{verse} \fullcite{harshawetal2026} \end{verse}

\subsection{Covariance adjustment (sharp and weak frameworks)}

\paragraph*{Required}

\begin{verse} Chapter~4 of \fullcite{gerbergreen2012}. \end{verse}

\begin{verse} \fullcite{rosenbaum2002c}. \end{verse}

\begin{verse} \fullcite{lin2013} \end{verse}

\paragraph*{Recommended}

\begin{verse} \fullcite{miratrixetal2013} \end{verse}

\begin{verse} \fullcite{freedman2008a} \end{verse}

\begin{verse} \fullcite{freedman2008b} \end{verse}

\begin{verse} \fullcite{freedman2008c} \end{verse}

\begin{verse} \fullcite{samiiaronow2012} \end{verse}

\begin{verse} \fullcite{aronowsamii2016} \end{verse}

\begin{verse} \fullcite{abadieetal2020} \end{verse}

\begin{verse} \fullcite{fogarty2018} \end{verse}

\begin{verse} \fullcite{guobasse2023} \end{verse}

\begin{verse} \fullcite{cohenfogarty2023} \end{verse}

\begin{verse} \fullcite{changetal2024} \end{verse}

\subsection{Noncompliance and attrition}

\subsection*{Application}

\begin{verse} \fullcite{albertsonlawrence2009} \end{verse}

\subsubsection{Noncompliance and instrumental variables}

\paragraph*{Required}

\begin{verse} Chapters 5 and 6 of \fullcite{gerbergreen2012}. \end{verse}

\begin{verse}
  Section 5.3, ``Instruments,'' of \fullcite{rosenbaum2010} \end{verse}

\begin{verse}
\fullcite{rosenbaum1996} \end{verse}

\paragraph*{Recommended}

\begin{verse} Section 2.3 of \fullcite{rosenbaum2002c} \end{verse}

\begin{verse} \fullcite{angristetal1996} \end{verse}

\begin{verse} \fullcite{imbensrosenbaum2005} \end{verse}

\begin{verse} \fullcite{kangpeckkeele2018} \end{verse}

\begin{verse} \fullcite{hansenbowers2008} \end{verse}

\begin{verse} \fullcite{aronowetal2025} \end{verse}

\begin{verse} \fullcite{pashley2022} \end{verse}

\begin{verse} \fullcite{pashleyetal2024} \end{verse}

\begin{verse} \fullcite{blackwellpashley2021} \end{verse}

\subsubsection{Attrition, or missing outcomes}

\paragraph*{Required}

\begin{verse} Chapter 7 of \fullcite{gerbergreen2012} \end{verse}

\paragraph*{Recommended}

\begin{verse} \fullcite{lee2009} \end{verse}

\begin{verse} \fullcite{aronowetal2019} \end{verse}

\begin{verse} \fullcite{horowitzmanski2000} \end{verse}

\begin{verse} \fullcite{coppocketal2017} \end{verse}

\begin{verse} \fullcite{lietal2025} \end{verse}

\begin{verse} \fullcite{changyu2026} \end{verse}

\section{Observational studies}

\subsection{Observational studies and as-if randomization}

\paragraph*{Required}

\begin{verse} Section~1, ``Design-Based Foundations of the Matching Pipeline,'' of \fullcite{leavittmiratrix2026} \end{verse}

\begin{verse} Chapter 5 of \fullcite{rosenbaum2017} \end{verse}

\begin{verse} Sections 9.0 -- 9.2 (especially discussion of interpolation and extrapolation) of \fullcite{gelmanhill2006} \end{verse}

\begin{verse} \fullcite{berk2010} \end{verse}

\paragraph*{Recommended}

\begin{verse} \fullcite{bindrubin2019} \end{verse}

\begin{verse} \fullcite{rubin1977} \end{verse}

\begin{verse} \fullcite{cochran1965} \end{verse}

\begin{verse} \fullcite{achen2002} \end{verse}

\begin{verse} Chapters 11 and 19 (on overly influential points) of \fullcite{fox2016} \end{verse}

\subsection{Building a matched design: Measuring similarity and forming matches}

\subsection*{Application}

\begin{verse} \fullcite{gilligansergenti2008} \end{verse}

\begin{verse} \fullcite{cerdaetal2012} \end{verse}

\paragraph*{Required}

\begin{verse} Sections~2.1 -- 2.2.4 (running example through forming the matches) of \fullcite{leavittmiratrix2026} \end{verse}

\begin{verse} Pages 65 -- 96 of \fullcite{rosenbaum2017} \end{verse}

\begin{verse} Chapters 7 -- 8 of \fullcite{rosenbaum2010} \end{verse}

\begin{verse} \fullcite{hansen2011} \end{verse}

\paragraph*{Recommended}

\begin{verse} \fullcite{hansen2004} \end{verse}

\begin{verse} \fullcite{hansenklopfer2006} \end{verse}

\begin{verse} \fullcite{gurosenbaum1993} \end{verse}

\begin{verse} \fullcite{rubinwaterman2006} \end{verse}

\begin{verse} \fullcite{hansen2008a} \end{verse}

\begin{verse} \fullcite{saleshansenrowan2018} \end{verse}

\begin{verse} \fullcite{rosenbaum2020} \end{verse}

\begin{verse} Chapter 3 of \fullcite{rosenbaum2002a}, specifically Sections 3.1 -- 3.2 and 3.4 -- 3.5. \end{verse}

\begin{verse} \fullcite{rosenbaum2001d} \end{verse}

\begin{verse} \fullcite{bifulco2012} \end{verse}

\begin{verse} \fullcite{arceneauxetal2010} \end{verse}

\begin{verse} \fullcite{rubin1979} \end{verse}

\begin{verse} \fullcite{robinsetal2000a} \end{verse}

\begin{verse} \fullcite{hoetal2007} \end{verse}

\begin{verse} Chapter 13 of \fullcite{rosenbaum2010} \end{verse}

\begin{verse} \fullcite{rosenbaumrubin1985} \end{verse}

\subsection{Evaluating a matched design: Balance and effective sample size}

\paragraph*{Required}

\begin{verse} Section~2.2.5 (balance and effective sample size) of \fullcite{leavittmiratrix2026} \end{verse}

\begin{verse} \fullcite{hansenbowers2008} \end{verse}

\paragraph*{Recommended}

\begin{verse} \fullcite{austin2009} \end{verse}

\begin{verse} \fullcite{stuart2010} \end{verse}

\begin{verse} \fullcite{rubin2001} \end{verse}

\begin{verse} \fullcite{austin2008} \end{verse}

\begin{verse} \fullcite{hansen2008b} \end{verse}

\begin{verse} \fullcite{holm1979} \end{verse}

\begin{verse} \fullcite{imai2008a} \end{verse}

\begin{verse} \fullcite{imaietal2008} \end{verse}

\subsection{Inference under as-if randomization I: The sharp framework}

\paragraph*{Required}

\begin{verse} Sections~2.3 -- 2.3.1 (inference under the sharp framework) of \fullcite{leavittmiratrix2026} \end{verse}

\begin{verse} Chapter 3 of \fullcite{rosenbaum2017} \end{verse}

\paragraph*{Recommended}

\begin{verse} \fullcite{pashleyetal2021} \end{verse}

\subsection{Inference under as-if randomization II: The weak framework}

\paragraph*{Required}

\begin{verse} Section~2.3.2 (inference under the weak framework) of \fullcite{leavittmiratrix2026} \end{verse}

\begin{verse} Pages 71 -- 79 of \fullcite{gerbergreen2012} \end{verse}

\begin{verse} \fullcite{fogarty2018a} \end{verse}

\begin{verse} \fullcite{pashleymiratrix2020} \end{verse}

\paragraph*{Recommended}

\begin{verse} \fullcite{lietal2018a} \end{verse}

\begin{verse} \fullcite{pashleyetal2021} \end{verse}

\subsection{Why regression adjustment is not a research design}

The last week contrasts the matched design with the reflex of ``controlling for'' covariates in a regression. We ask three questions a regression cannot answer from the estimate alone: which functional form, whether the comparison extrapolates beyond the data, and whether a few points drive the result.

\paragraph*{Required}

\begin{verse} Sections 9.0 -- 9.2 (interpolation versus extrapolation) of \fullcite{gelmanhill2006} \end{verse}

\begin{verse} \fullcite{berk2010} \end{verse}

\begin{verse} Chapters 11 and 19 (overly influential points) of \fullcite{fox2016} \end{verse}

\paragraph*{Recommended}

\begin{verse} \fullcite{cinellihazlett2020} \end{verse}

\begin{verse} \fullcite{aronowsamii2016} \end{verse}

\subsection{Non-bipartite matching: continuous and multivalued treatments}

When treatment is not binary --- a dose, a score, a degree of exposure --- we pair units from a single pool rather than matching treated to control. The pair-level dose difference then plays the role the treatment indicator played before.

\paragraph*{Required}

\begin{verse} \fullcite{luetal2001} \end{verse}

\begin{verse} \fullcite{luetal2011} \end{verse}

\begin{verse} Pages 207 -- 221 of \fullcite{rosenbaum2010} \end{verse}

Section~\ref{sec:nbp} lists further work on non-bipartite, risk-set, cardinality, fine-balance, and other specialized matching.

\subsection{Longitudinal designs: risk-set matching and difference-in-differences}\label{sec:long}

The final day sets two strategies for panel data side by side. Risk-set matching pairs a unit that changes treatment at time $g$ with a unit still untreated at $g$, extending the matched-pair logic across time. Difference-in-differences instead assumes parallel trends and reads the effect off the change in changes. We test the parallel-trends assumption and then bound the bias it leaves with HonestDiD.

\paragraph*{Required}

\begin{verse} \fullcite{hammiratrix2024} \end{verse}

\begin{verse} Pages 162 -- 167 of \fullcite{rosenbaum2017} \end{verse}

\begin{verse} \fullcite{rambachanroth2023} \end{verse}

\paragraph*{Recommended}

\begin{verse} \fullcite{callawaysantanna2021} \end{verse}

\begin{verse} \fullcite{rothetal2023} \end{verse}

Section~\ref{sec:did} lists the fuller difference-in-differences reading, including staggered adoption and pre-trend testing.

\section{Sensitivity analysis}

We treat sensitivity analysis in one pass on a full matched design (UN peacekeeping and post-conflict peace), rather than as two separate days. We ask how large a departure from as-if randomization --- how much hidden bias from an unobserved confounder --- it would take to overturn the conclusion, under both the sharp and the weak frameworks. We then read the same worst case through a regression lens (the $R^2$ that an unobserved confounder would need with treatment and with outcome), which connects Rosenbaum's $\Gamma$ to the regression-based bounds of \textcite{cinellihazlett2020}; see Section~\ref{sec:regsens}.

\subsection*{Application}

\begin{verse} \fullcite{gilligansergenti2008} \end{verse}

\subsection{Sensitivity analysis I: The sharp framework}

\paragraph*{Required}

\begin{verse} Sections~2.4 -- 2.4.1 (sensitivity analysis under the sharp framework) of \fullcite{leavittmiratrix2026} \end{verse}

\begin{verse}
  Chapter 9 of \fullcite{rosenbaum2017}
\end{verse}

\begin{verse} \fullcite{rosenbaum2018} \end{verse}

\paragraph*{Recommended}

\begin{verse} Chapter 4 of \fullcite{rosenbaum2002a} \end{verse}

\begin{verse} \fullcite{rosenbaumkrieger1990} \end{verse}

\begin{verse} \fullcite{gastwirthetal2000} \end{verse}

\begin{verse} \fullcite{hansenrosenbaumsmall2014} \end{verse}

\begin{verse} \fullcite{hsusmall2013} \end{verse}

\subsection{Sensitivity analysis II: The weak framework}

\paragraph*{Required}

\begin{verse} Section~2.4.2 (sensitivity analysis under the weak framework) of \fullcite{leavittmiratrix2026} \end{verse}

\begin{verse} \fullcite{fogarty2023} \end{verse}

\paragraph*{Recommended}

\begin{verse} \fullcite{fogarty2020b} \end{verse}

\begin{verse} \fullcite{fogartyetal2017} \end{verse}

\section{Additional topics}

\subsection{Design sensitivity}

\begin{verse} \fullcite{rosenbaum2004} \end{verse}

\begin{verse} \fullcite{helleretal2009} \end{verse}

\begin{verse} \fullcite{hsuetal2013} \end{verse}

\begin{verse} \fullcite{smalletal2013} \end{verse}

\subsection{Special topics in matching: nonbipartite, multilevel, risk-set, cardinality, coarsened exact, template, exterior, generalized full, and fine balance}\label{sec:nbp}

\begin{verse} \fullcite{luetal2001} \end{verse}

\begin{verse} \fullcite{luetal2011} \end{verse}

\begin{verse} Pages 207 -- 221 of \fullcite{rosenbaum2010} \end{verse}

\begin{verse} \fullcite{silberetal2014} \end{verse}

\begin{verse} \fullcite{zubizarretakeele2017} \end{verse}

\begin{verse} \fullcite{pimenteletal2018} \end{verse}

\begin{verse} \fullcite{lietal2001} \end{verse}

\begin{verse} \fullcite{zubizarretaetal2014} \end{verse}

\begin{verse} \fullcite{iacusetal2012} \end{verse}

\begin{verse} \fullcite{rosenbaumsilber2013} \end{verse}

\begin{verse} \fullcite{savjehigginssekhon2021} \end{verse}

\begin{verse} \fullcite{cochran1965} \end{verse}

\begin{verse} \fullcite{danieletal2008} \end{verse}

\begin{verse} \fullcite{yuetal2021} \end{verse}

\begin{verse} \fullcite{rosenbaumetal2007} \end{verse}

\begin{verse} \fullcite{yangetal2012} \end{verse}

\begin{verse} \fullcite{yu2023} \end{verse}

\begin{verse} \fullcite{pimenteletal2015a} \end{verse}

\begin{verse} \fullcite{pimenteletal2015b} \end{verse}

\subsection{Extensions to covariate balance testing}

\begin{verse} \fullcite{gagnon-bartschshem-tov2019} \end{verse}

\begin{verse} \fullcite{branson2021} \end{verse}

\begin{verse} \fullcite{hansensales2015} \end{verse}

\subsection{Residual imbalance on observed covariates}

\begin{verse} \fullcite{rosenbaum1988} \end{verse}

\begin{verse} \fullcite{pimentelhuang2024} \end{verse}

\begin{verse} \fullcite{chenetal2023} \end{verse}

\begin{verse} \fullcite{hengetal2025} \end{verse}

\begin{verse} \fullcite{zhuetal2025} \end{verse}

\subsection{Defining an interpretable study population}

\begin{verse} \fullcite{traskinsmall2011} \end{verse}

\begin{verse} \fullcite{fogartyetal2016} \end{verse}

\begin{verse} \fullcite{crumpetal2009} \end{verse}

\begin{verse} \fullcite{kingzeng2006} \end{verse}

\subsection{Rerandomization}

\begin{verse} \fullcite{morganrubin2012} \end{verse}

\begin{verse} \fullcite{morganrubin2015} \end{verse}

\begin{verse} \fullcite{lietal2018b} \end{verse}

\begin{verse} \fullcite{liding2020} \end{verse}

\begin{verse} \fullcite{bransonetal2023} \end{verse}

\begin{verse} \fullcite{wangetal2021} \end{verse}

\begin{verse} \fullcite{johanssonetal2021} \end{verse}

\begin{verse} \fullcite{luetal2023} \end{verse}

\begin{verse} \fullcite{cytrynbaum2024} \end{verse}

\begin{verse} \fullcite{ribeirobranson2025} \end{verse}

\subsection{Randomization versus optimal design}

\begin{verse} \fullcite{kiefer1959} \end{verse}

\begin{verse} \fullcite{fedorov1972} \end{verse}

\begin{verse} \fullcite{harville1975} \end{verse}

\begin{verse} \fullcite{basu1980} \end{verse}

\begin{verse} \fullcite{wu1981} \end{verse}

\begin{verse} \fullcite{senn1994a} \end{verse}

\begin{verse} \fullcite{senn2013} \end{verse}

\begin{verse} \fullcite{bertsimasetal2015} \end{verse}

\begin{verse} \fullcite{kasy2016} \end{verse}

\begin{verse} \fullcite{banerjeeetal2017} \end{verse}

\begin{verse} \fullcite{kallus2018} \end{verse}

\begin{verse} \fullcite{banerjeeetal2020} \end{verse}

\begin{verse} \fullcite{kallus2021} \end{verse}

\begin{verse} \fullcite{kapelneretal2021} \end{verse}

\begin{verse} \fullcite{kapelneretal2022} \end{verse}

\begin{verse} \fullcite{bai2023} \end{verse}

\begin{verse} \fullcite{harshawetal2024} \end{verse}

\begin{verse} \fullcite{martinezteira2024} \end{verse}

\subsection{Quasi-experimental devices}

\begin{verse} \fullcite{campbell1957} \end{verse}

\begin{verse} \fullcite{campbellstanley1963} \end{verse}

\begin{verse} \fullcite{shadishetal2002} \end{verse}

\begin{verse} Section~IV of \fullcite{rosenbaum2025} \end{verse}

\subsubsection{Difference-in-Differences}\label{sec:did}

\begin{verse} \fullcite{hammiratrix2024} \end{verse}

\begin{verse} Pages 162 -- 167 of \fullcite{rosenbaum2017} \end{verse}

\begin{verse} Section 4.1 of \fullcite{gerbergreen2012} \end{verse}

\begin{verse} Chapter 5 of \fullcite{angristpischke2008} \end{verse}

\begin{verse} \fullcite{lechner2011} \end{verse}

\begin{verse} \fullcite{manskipepper2018} \end{verse}

\begin{verse} \fullcite{dingli2019} \end{verse}

\begin{verse} \fullcite{imaikim2021} \end{verse}

\begin{verse} \fullcite{callawaysantanna2021} \end{verse}

\begin{verse} \fullcite{rambachanroth2023} \end{verse}

\begin{verse} \fullcite{roth2022} \end{verse}

\begin{verse} \fullcite{rothsantanna2023} \end{verse}

\begin{verse} \fullcite{rothetal2023} \end{verse}

\begin{verse} \fullcite{bakeretal2026} \end{verse}

\begin{verse} \fullcite{mikhaeilharshaw2025} \end{verse}

\begin{verse} \fullcite{leavitthatfield2025} \end{verse}

\begin{verse} \fullcite{leavitt2026b} \end{verse}

\subsubsection{Regression discontinuity designs}\label{sec:rdd}

\begin{verse} \fullcite{caugheysekhon2011} \end{verse}

\begin{verse} \fullcite{cattaneoetal2020} \end{verse}

\begin{verse} \fullcite{saleshansen2020} \end{verse}

\begin{verse} \fullcite{keeletitiunikzubizarreta2015} \end{verse}

\begin{verse} \fullcite{sekhontitiunik2017} \end{verse}

\begin{verse} \fullcite{sekhontitiunik2016} \end{verse}

\begin{verse} \fullcite{hahnetal2001} \end{verse}

\begin{verse} \fullcite{imbenslemieux2008} \end{verse}

\begin{verse} \fullcite{lee2008} \end{verse}

\begin{verse} \fullcite{gelmanimbens2019} \end{verse}

\begin{verse} \fullcite{mccrary2008} \end{verse}

\begin{verse} Chapter 6 of \fullcite{angristpischke2008} \end{verse}

\subsection{Interference}

\begin{verse} \fullcite{rosenbaum2007a} \end{verse}

\begin{verse} \fullcite{bowersetal2013} \end{verse}

\begin{verse} \fullcite{aronowsamii2017} \end{verse}

\begin{verse} \fullcite{manski2013b} \end{verse}

\begin{verse} \fullcite{atheyetal2018} \end{verse}

\begin{verse} \fullcite{savjeetal2021} \end{verse}

\begin{verse} \fullcite{wangetal2025} \end{verse}

\begin{verse} \fullcite{kandirosetal2024} \end{verse}

\begin{verse} \fullcite{bowersetal2018} \end{verse}

\begin{verse} \fullcite{harshawetal2026} \end{verse}

\subsection{Factorial and complex experiments}

\begin{verse} \fullcite{dasguptaetal2015} \end{verse}

\begin{verse} \fullcite{hainmuelleretal2014} \end{verse}

\begin{verse} \fullcite{harshawetal2026} \end{verse}

\begin{verse} \fullcite{egamiimai2019} \end{verse}

\begin{verse} \fullcite{gerberetal2010} \end{verse}

\begin{verse} \fullcite{blackwellpashley2026} \end{verse}

\begin{verse} \fullcite{goplerudetal2025} \end{verse}

\subsection{Joint inference of sharp and weak causal hypotheses}

\begin{verse} \fullcite{chungromano2013} \end{verse}

\begin{verse} \fullcite{ding2017} \end{verse}

\begin{verse} \fullcite{wuding2021} \end{verse}

\begin{verse} \fullcite{cohenfogarty2022} \end{verse}

\subsection{Attributable effects, the ATT, and alternative effect models}

\begin{verse} \fullcite{rosenbaum2001a} \end{verse}

\begin{verse} \fullcite{rosenbaum2003} \end{verse}

\begin{verse} \fullcite{hansenbowers2009} \end{verse}

\begin{verse} \fullcite{sekhonshem-tov2021} \end{verse}

\begin{verse} \fullcite{keeleetal2017} \end{verse}

\begin{verse} \fullcite{rosenbaum2007} \end{verse}

\begin{verse} \fullcite{rosenbaum1999} \end{verse}

\begin{verse} Pages 46 -- 49 of \fullcite{rosenbaum2010} \end{verse}

\subsection{Sensitivity analysis: regression-based approaches and recent extensions}\label{sec:regsens}

\begin{verse} \fullcite{hosmanetal2010} \end{verse}

\begin{verse} \fullcite{cinellihazlett2020} \end{verse}

\begin{verse} \fullcite{imbens2003} \end{verse}

\begin{verse} \fullcite{oster2019} \end{verse}

\begin{verse} \fullcite{wuli2025} \end{verse}

\begin{verse} \fullcite{li2025} \end{verse}

\begin{verse} \fullcite{fogarty2025} \end{verse}

\begin{verse} \fullcite{bekermanetal2026} \end{verse}

\begin{verse} \fullcite{linetal2026} \end{verse}

\subsection{External validity}

\begin{verse} \fullcite{kernetal2016} \end{verse}

\begin{verse} \fullcite{miratrixetal2018a} \end{verse}

\begin{verse} \fullcite{bennettetal2020} \end{verse}

\begin{verse} \fullcite{silberetal2014} \end{verse}

\begin{verse} \fullcite{westreichetal2019} \end{verse}

\begin{verse} \fullcite{egamihartman2022} \end{verse}

\subsection{Weighting methods}

\begin{verse} \fullcite{chattopadhyayetal2020} \end{verse}

\begin{verse} \fullcite{imairatkovic2014} \end{verse}

\begin{verse} \fullcite{zubizarreta2015} \end{verse}

\begin{verse} \fullcite{wongchan2018} \end{verse}

\begin{verse} \fullcite{tan2020} \end{verse}

\begin{verse} \fullcite{wangzubizarreta2020} \end{verse}

\begin{verse} \fullcite{hainmueller2012} \end{verse}

\subsection{Synthetic control}

\begin{verse} \fullcite{abadiegardeazabal2003} \end{verse}

\begin{verse} \fullcite{abadieetal2010} \end{verse}

\begin{verse} \fullcite{abadieetal2012} \end{verse}

\begin{verse} \fullcite{abadie2021} \end{verse}

\begin{verse} \fullcite{arkhangelskyetal2021} \end{verse}

\begin{verse} \fullcite{ben-michaeletal2021} \end{verse}

\subsection{Bayesian causal inference}

\begin{verse} \fullcite{rubin1978a} \end{verse}

\begin{verse} Chapter 8 of \fullcite{imbensrubin2015} \end{verse}

\begin{verse} \fullcite{leavitt2023} \end{verse}

\begin{verse} \fullcite{leavitt2026a} \end{verse}

\begin{verse} \fullcite{imbensrubin1997} \end{verse}

\begin{verse} \fullcite{keelequinn2017} \end{verse}

\begin{verse} \fullcite{dingmiratrix2019} \end{verse}

\begin{verse} \fullcite{liuetal2025} \end{verse}

\clearpage

\printbibliography[title=References]

%\section{Additional Notes}
%\clearpage
%\bibliographystyle{plain}
%\nobibliography{bib_2023}
%\nobibliography{master,abbrev_long,causalinference,biomedicalapplications,misc}
%\nobibliography{../2013/BIB/master,../2013/BIB/abbrev_long,../2013/BIB/causalinference,../2013/BIB/biomedicalapplications,../2013/BIB/misc}
%"master" is Ryan Moore's file (as of summer 2012).
%"causalinference" is Ben's, and lives (as of summer 2013) in
% http://dept.stat.lsa.umich.edu/~bbh/texmf/bibtex/bib/

\end{document}
