\documentclass{article}
\usepackage{natbib}
\usepackage{soul}

\title{Causal Inference Assignment 1}
\input{../courseedition}
\author{Due noon Friday, June 19, 2026}
\usepackage{../icpsr-classwork}
\renewcommand{\bm}[1]{\boldsymbol{#1}} % bold-italic vectors, matching the slides


\begin{document}
\maketitle

\begin{enumerate}
%\item Can the simulation method approximate exact, permutation-based
%  calculation to as much accuracy as is practically meaningful?
%  State a limit on the absolute error of a p-value,
%  $|p_{\text{approx}} - p_{\text{exact}}|$, that seems reasonable to
%  you to tolerate; then use the formula presented in
%  \texttt{unit01-Rex.pdf} to determine how many simulation replicates
%  are need to obtain simulation $p$-values deviating from their exact
%  counterparts by no more than this amount.
%\item For the coffee experiment, does the chi-square procedure report
%  a similar p-value as the fisher test?  (In R,
%  it's \texttt{chisq.test()}. For documentation, enter \texttt{?chisq.test}.)
%\item For this data set, are the chi-square test and the fisher test
%  both admissible? (Hint: what are the sample size requirements of the
%  chi-square procedure? As review, read R's warnings and documentation.)
\item Refer to the \texttt{acorn} data set that accompanies this assignment.  Using the mean of turnout proportions in
treatment-group precincts, $n_{1}^{-1}\bm{Z}^{\top}\bm{y}$, as
test statistic, simulate its randomization distribution under the
sharp null hypothesis of no effect, reporting: \label{q:simmoments}
\begin{enumerate}
  \item your simulation $p$-value;
  \item your simulation approximation of $\mathrm{E}[
    n_{1}^{-1}\bm{Z}^{\top}\bm{y}]$, the null expected value of the test
    statistic;
  \item your simulation approximation of
    $\mathrm{Var}[n_{1}^{-1}\bm{Z}^{\top}\bm{y}]$, this test
    statistic's variance under the null.
  \end{enumerate}
 Here $n_1$ is the number of treated units, $n_0 = N - n_1$ the number of control units, $N$ the total number of units, $\bm{Z}$ the random assignment vector, and $\bm{y}$ the observed outcome vector.  In the \texttt{acorn03} data set, the column \texttt{z} records the observed treatment assignment $\bm{z}$ (the realized GOTV assignment indicator), of which $\bm{Z}$ is the random counterpart that you simulate over; the column \texttt{vote03} records the observed outcome $\bm{y}$ (the 2003 turnout proportion in each precinct).
\item Calculate $\mathrm{E} [ n_{1}^{-1}\bm{Z}^{\top}\bm{y} ]$, the expected value of the
  treatment-group mean under the sharp null hypothesis of no effect, from first principles --- i.e., without simulations ---  using the data in $\bm{y}$.  \label{q:exactEV}
\item In this setup, $\mathrm{Var}\, [\bar{y}_{1}]  = \frac{1}{n_{1}}
  \frac{n_0}{N} \frac{\sum_{i=1}^{N} (y_{i} - \bar y)^{2}}{N-1} $,
  where $\bar{y}_{1} = n_{1}^{-1}\bm{Z}^{\top}\bm{y}$ is the mean of
  $y$s among the group assigned to treatment ($\{i: Z_{i} =1\}$) while
  $\bar{y} = N^{-1}\bm{1}^{\top}\bm{y}$ is the mean of $y$ over
  the full study population.  If you've seen formulas for the sampling
  variance of the mean in earlier stats
  courses, you probably saw
  $$
  \widehat{\mathrm{Var}} \left[\bar{y}_{1}\right] = \frac{s^{2}_{y}}{n_{1}} = \frac{1}{n_{1}}
  \frac{\sum_{i=1}^{n_{1}} (y_{i} - \bar{y}_{1})^{2}}{n_{1}-1},
\, \mathrm{and/or}\,
  \mathrm{Var} \left[\bar{y}_{1}\right] = \frac{\sigma_{y}^{2}}{n_{1}} = \frac{1}{n_{1}}
  \frac{\sum_{i=1}^{N} (y_{i} - \bar{y})^{2}}{N}.
  $$
Which gives a larger value, our new $\mathrm{Var}\,
[\bar{y}_{1}]$-formula or the  $\mathrm{Var} \left[\bar{y}_{1}\right]$
formula immediately above?  Explain why it makes
sense that the formula we've given would differ in the direction it does
from this other formula. %, in terms of the difference between this
%sampling situation and the sampling model that more commonly
%accompanies the Central Limit Theorem.
\item Calculate $\mathrm{Var}\, [n_{1}^{-1}\bm{Z}^{\top}\bm{y}]$ using the appropriate
  formula.   Determine the error of the simulation-based
  approximation to this quantity that you reported in question~\ref{q:simmoments},
  expressing it as a percentage of $\mathrm{Var}\,[ n_{1}^{-1}\bm{Z}^{\top}\bm{y}]$. \label{q:exactvar}
\item Determine the Normal-theory approximation to
  $\mathrm{Pr}(n_{1}^{-1}\bm{Z}^{\top}\bm{y} \geq
  n_{1}^{-1}\bm{z}^{\top}\bm{y} ) $.  (Hints: Use the variance and
  expected value calculated in~\ref{q:exactvar} and \ref{q:exactEV} to transform your
  observed treatment-group mean into a corresponding ``$z$-score.''
  To determine Normal quantiles corresponding to z-scores in R, use
  \texttt{pnorm()}; type \texttt{?pnorm} for help. Here $\bm{z}^{\top}$ refers to
  the observed ACORN treatment assignment.)
\item %\input{gg15i}
A researcher plans to ask six subjects to donate time to an adult
literacy program. Each subject will be asked to donate either 30
($Z=0$) or 60 ($Z=1$)
minutes. The researcher is considering three methods for randomizing
the treatment. Method I is to make independent decisions for each
subject, tossing a coin each time. Method C is to
write ``30'' and ``60'' on three playing cards each, and then shuffle
the six cards. Method P tosses one coin for each of the 3 pairs
$(1,2)$, $(3,4)$, $(5,6)$, asking for 30 (60) minutes from exactly one
member of each pair.
\begin{itemize}
\item[a] Discuss strengths \& weaknesses of each method.
\item[b] How would your answers to (a) change if $N: 6 \mapsto 600$?
\item[c] Determine $\EE[  Z_{1} ]$  under each method.
\item[d] Determine $\EE\big[  Z_{1} + Z_{2} + \cdots + Z_{6} \big]$ under each method.
\item[e] Calculate $\EE[\bm{Z}^{\top}\bm{1}]$ under each of the three methods.
\item[f] For which of the methods does $\EE
  \big[\big(\bm{Z}^{\top}\bm{1} -\EE [\bm{Z}^{\top}\bm{1}]\big)^{2}\big] =
  0$?\footnotemark[1]
\item[g] For two of the three methods, the linearity property of expected value%
  \footnotemark[2]
  entails that
  $\EE \big[\frac{\bm{Z}^{\top}\bm{x}}{\bm{Z}^{\top}\bm{1}}\big] = (x_{1}
  + x_{2}+ \ldots + x_{6})/6$.  Which two are
  these, and why doesn't the argument work for the third?
\end{itemize}
\end{enumerate}
\footnotetext[1]{I.e., for which does
    $\mathrm{Var}[\bm{Z}^{\top}\bm{1}] = 0$?  (In general,
    $\mathrm{Var}[V] = \EE \big[(V - \EE[V])^{2} \big] $.)}
\footnotetext[2]{For constants $\alpha$ and $\beta$, $\EE(\alpha + \beta Y)
    = \alpha + \beta \EE(Y)$. Also $\EE(\alpha X + \beta Y)
    = \alpha\EE(X) + \beta \EE(Y)$.}
\end{document}
