\documentclass{article}
\usepackage{natbib}

\title{Causal Inference Assignment 3}
\input{../courseedition}
\author{Due Tuesday 8/15, 3pm}
\usepackage{../icpsr-classwork}


\begin{document}
\maketitle

We challenge you to find an existing study (your own or one from someone else, published or not) and to replicate one of the key analyses in that study using tools learned in this class. For example, you might find a study using a linear model for adjustment (i.e. "controlling for") and you might redo that analysis using matching. And/or you might find a study where you could add or apply permutation-based testing (perhaps you are concerned about the false positive error rate of a study with a weak instrument, a small number of clusters, a skewed outcome, or a small sample size). If you would like to do something other than matching and/or permutation-based testing, let us know.

Data and/or study ideas:
\begin{itemize}
 \item You could use your own data if you are replicating one of your own previous studies.
 \item You could search the \href{http://dataverse.org/}{Dataverse} or the \href{http://www.icpsr.umich.edu/icpsrweb/deposit/pra/index.jsp}{ICPSR Replication Archive} or the \href{http://www.3ieimpact.org/evaluation/impact-evaluation-replication-programme/}{3IE Replication program}
 \item Many journals now maintain or point to replication archives.
 \item Many scholars now maintain replication archives.
 \item You could work on the Medellin data.
\end{itemize}
\end{document}
