---
title: Strategies for making matched designs
date: '`r format(Sys.Date(), "%B %d, %Y")`'
author: ICPSR 2019 Session 2
bibliography:
 - refs.bib
 - BIB/master.bib
 - BIB/misc.bib
fontsize: 10pt
geometry: margin=1in
graphics: yes
biblio-style: authoryear-comp
output:
  beamer_presentation:
    slide_level: 2
    keep_tex: true
    latex_engine: xelatex
    citation_package: biblatex
    template: icpsr.beamer
    includes:
        in_header:
           - defs-all.sty
---


<!-- Make this document using library(rmarkdown); render("day12.Rmd") -->


```{r include=FALSE, cache=FALSE}
# Some customization.  You can alter or delete as desired (if you know what you are doing).
# knitr settings to control how R chunks work.

require(knitr)

## This plus size="\\scriptsize" from https://stackoverflow.com/questions/26372138/beamer-presentation-rstudio-change-font-size-for-chunk

knitr::knit_hooks$set(mysize = function(before, options, envir) {
  if (before)
    return(options$size)
})

knit_hooks$set(plotdefault = function(before, options, envir) {
    if (before) par(mar = c(3, 3, .1, .1),oma=rep(0,4),mgp=c(1.5,.5,0))
})

opts_chunk$set(
  tidy=FALSE,     # display code as typed
  echo=TRUE,
  results='markup',
  strip.white=TRUE,
  fig.path='figs/fig',
  cache=FALSE,
  highlight=TRUE,
  width.cutoff=132,
  size='\\scriptsize',
  out.width='.8\\textwidth',
  fig.retina=FALSE,
  message=FALSE,
  comment=NA,
  mysize=TRUE,
  plotdefault=TRUE)

if(!file.exists('figs')) dir.create('figs')

options(digits=4,
	scipen=8,
	width=132
	)
```

```{r eval=FALSE, include=FALSE, echo=FALSE}
## Run this only once and then not again until we want a new version from github
library('devtools')
library('withr')
with_libpaths('./lib', install_github("markmfredrickson/RItools"), 'pre')
```


```{r eval=FALSE, echo=FALSE,include=FALSE}
## Having downloaded optmatch from Box OR http://jakebowers.org/ICPSR for mac (tgz) or windows (zip)
## For Mac
## with_libpaths('./lib',install.packages('optmatch_0.9-8.9003.tgz', repos=NULL),'pre')
## For Windows
## with_libpaths('./lib',install.packages('optmatch_0.9-8.9003.zip', repos=NULL),'pre')
```

```{r eval=FALSE, echo=FALSE}
## Or if you have all of the required libraries (like fortran and c++) for compilation use
with_libpaths('./lib', install_github("markmfredrickson/optmatch"), 'pre')
```

```{r echo=FALSE}
library(dplyr)
library(chemometrics)
library(RItools,lib.loc="./lib")
library(optmatch,lib.loc="./lib")
```

## Today

\begin{enumerate}
  \item Agenda: Exact matching; Calipers; Combining matrices; Discussion of with and without-replacement matching and optimal vs greedy matching; Maybe use those tools to improve your best matched design.
\item Reading for this week: DOS 8--9, 13 and \cite[\S~9.5]{gelman2006dau}, and \cite{ho:etal:07}
\item Questions arising from the reading or assignments or life?
\item Tomorrow: Estimation and Testing after matching; Information and blocked designs (some blockings have more information about treatment effects than others); Balance tests and the Sequential Intersection Union Principle
\end{enumerate}


```{r loaddat, echo=FALSE}
load(url("http://jakebowers.org/Data/meddat.rda"))
meddat$id <- row.names(meddat)
meddat<- mutate(meddat, HomRate03=(HomCount2003/Pop2003)*1000,
                HomRate08=(HomCount2008/Pop2008)*1000)
## mutate strips off row names
row.names(meddat) <- meddat$id
options(show.signif.stars=FALSE)
```

```{r echo=FALSE}
library(dplyr)
library(chemometrics)
library(RItools,lib.loc="./lib")
library(optmatch,lib.loc="./lib")
```

# Your reports

## Your reports?

What did you do? What kind of overall balance did you achieve?

# The Propensity Scores

## The propensity score

Given covariates $\mathbf{X} (=(X_1, \ldots, X_k))$, and a
treatment variable $Z$, $Z(u) \in \{0, 1\}$,  $\PP (Z \vert \mathbf{X})$ is known as the (true) \alert<1>{propensity score} (PS).  
$$ \phi( \mathbf{X} ) \equiv \log\left( \PP (Z=1 \vert \mathbf{X})/\PP (Z=0 \vert \mathbf{X}) \right)$$
is also known as the PS.  In practice, one works
with an estimated PS, $\hat{\PP} (Z \vert \mathbf{X})$ or
$\hat{\phi}(\mathbf{X})$.

Theoretically, propensity-score strata or matched sets both
\begin{enumerate}
\item reduce extrapolation; and
\item balance each of $X_1, \ldots, X_k$.
\end{enumerate}
They do this by making the comparison more "experiment-like", at least in terms of $X_1, \ldots, X_k$.

Theory also tells us that in the absence of hidden bias, such a stratification
\begin{enumerate}
\addtocounter{enumi}{2}
\item supports unbiased estimation of treatment effects.
\end{enumerate}


## Propensity scoring in practice
\begin{columns}
\column{.65\linewidth}
\begin{enumerate}
\item Fitted propensity scores help identify extrapolation.
\item In practice, stratification on $\hat{\phi}(\mathbf{X})$
helps balance each of $X_1, \ldots, X_k$.
% \item Adjustment based on the
% propensity score is arguably more transparent and more stable
% that adjustment based on multiple regression.
\item There are \emph{lots of cases} in which adjustment with the propensity
score fails to generate estimates that square with those of
randomized studies. 
\end{enumerate}
\column{.35\linewidth} \igrphx{meddatpsplot}
\end{columns}

\begin{enumerate}
\item There are various reasons for this, starting with: lots of observational studies that
don't measure quite enough $x$es.
\item (Propensity scores address bias on measured variables, not
unmeasured ones.)  \textit{hidden bias}.
\end{enumerate}


## The separation problem in logistic regression and the benefits of shrinkage/penalization

```{r echo=FALSE}
thecovs <- unique(c(names(meddat)[c(5:7,9:24)],"HomRate03"))
balfmla<-reformulate(thecovs,response="nhTrt")
library(splines)
psfmla <- update(balfmla,.~.+ns(HomRate03,2)+ns(nhPopD,2)+ns(nhHS,2))
glm0 <- glm(balfmla,data=meddat,family=binomial(link="logit"))
glm1 <- glm(psfmla,data=meddat,family=binomial(link="logit"))

# install.packages("arm")
library(arm)

bayesglm0 <- bayesglm(balfmla,data=meddat,family=binomial(link="logit"))
bayesglm1 <- bayesglm(psfmla,data=meddat,family=binomial(link="logit"))

psg1 <- predict(glm1,type="response")
psg0 <- predict(glm0,type="response")
psb1 <- predict(bayesglm1,type="response")
psb0 <- predict(bayesglm0,type="response")
```

Logistic regression is excellent at discriminating between groups \ldots often too excellent for us \autocite{gelman2008weakly}. First evidence of this is big and/or missing coefficients in the propensity score model. See the coefficients below:

```{r echo=FALSE}
rbind(glm0=coef(glm0)[1:5],
glm1=coef(glm1)[1:5],
bayesglm0=coef(bayesglm0)[1:5],
bayesglm1=coef(bayesglm1)[1:5]
)
```

## The separation problem in logistic regression

```{r out.width=".8\\textwidth", echo=FALSE}
par(mfrow=c(2,2),mar=c(3,3,2,.1))
boxplot(psg0~meddat$nhTrt,main=paste("Logit",length(coef(glm0))," parms",sep=" "))
stripchart(psg0~meddat$nhTrt,vertical=TRUE,add=TRUE)
boxplot(psg1~meddat$nhTrt,main=paste("Logit",length(coef(glm1))," parms",sep=" "))
stripchart(psg1~meddat$nhTrt,vertical=TRUE,add=TRUE)
boxplot(psb0~meddat$nhTrt,main=paste("Shrinkage Logit",length(coef(bayesglm0))," parms",sep=" "))
stripchart(psb0~meddat$nhTrt,vertical=TRUE,add=TRUE)
boxplot(psb1~meddat$nhTrt,main=paste("Shrinkage Logit",length(coef(bayesglm1))," parms",sep=" "))
stripchart(psb1~meddat$nhTrt,vertical=TRUE,add=TRUE)
```

# Matching Tricks of the Trade

## Calipers

The optmatch package allows calipers (which disallow certain pairs from being matched).^[You can implement penalties by hand.] Here, for example, we disallow comparisons which differ by more than 3 standard deviations on the propensity score.

```{r}
psdist <- match_on(bayesglm1,data=meddat)
psdist[1:3,1:3]
quantile(as.vector(psdist),seq(0,1,.1))
psdistCal <- psdist + caliper(psdist,3)
as.matrix(psdistCal)[1:3,1:3]
```
## Calipers

The optmatch package allows calipers (which disallow certain pairs from being matched).^[You can implement penalties by hand.] Here, for example, we disallow comparisons which differ by more than 3 standard deviations on the propensity score.

```{r}
fmCal1 <- fullmatch(psdist+caliper(psdist,5),data=meddat,tol=.00001)
summary(fmCal1,min.controls=0,max.controls=Inf)
pmCal1 <- pairmatch(psdist+caliper(psdist,5),data=meddat, remove.unmatchables=TRUE)
```

## Calipers

Another example: We may want to primarily match on mahalanobis distance but disallow any pairs with extreme propensity distance and/or extreme differences in baseline homicide rates.
```{r}
mhdist <- match_on(balfmla,data=meddat,method="rank_mahalanobis")
tmp <- meddat$HomRate03
names(tmp) <- rownames(meddat)
absdist <- match_on(tmp, z = meddat$nhTrt,data=meddat)
absdist[1:3,1:3]
quantile(as.vector(absdist),seq(0,1,.1))
distCal <- mhdist + caliper(psdist,3) + caliper(absdist,2)
as.matrix(distCal)[1:3,1:3]
```

## Calipers

```{r}
fmCal2 <- fullmatch(distCal,data=meddat,tol=.00001)
summary(fmCal2,min.controls=0,max.controls=Inf)
```


## Exact Matching

We often have covariates that are categorical/nominal and on which we really care about strong balance. One approach to solve this problem is match **exactly** on one or more of such covariates. If `fullmatch` or `match_on` is going slow, this is also an approach to speed things up.

```{r echo=FALSE}
meddat$classLowHi <- ifelse(meddat$nhClass %in% c(2,3),"hi","lo")
```

```{r}
dist2 <- mhdist + exactMatch(nhTrt~classLowHi,data=meddat)
## or mhdist <- match_on(balfmla,within=exactMatch(nhTrt~classLowHi,data=meddat),data=meddat,method="rank_mahalanobis")
## or fmEx1 <- fullmatch(update(balfmla,.~.+strata(classLowHi)),data=meddat,method="rank_mahalanobis")
fmEx1 <- fullmatch(dist2,data=meddat,tol=.00001)
summary(fmEx1,min.controls=0,max.controls=Inf)
print(fmEx1,grouped=T)
```
## Exact Matching

\scriptsize
```{r}
ftable(Class=meddat$classLowHi,Trt=meddat$nhTrt,fmEx1,col.vars=c("Class","Trt"))
```

# Matching modes

##  Optimal (communitarian) vs greedy (individualistic) matching

\begin{center}
    \begin{tabular}{l|cccc}
      & \multicolumn{4}{c}{Illustrator} \\
Writer & Mo& John & Terry  & Pat \\ \hline
Ben    & 0 & 1 & 1 & 10 \\
Jake   & 10& 0 & 10 & 10 \\
Tom    &  1& 1 & $\infty$ & $\infty$ \\ \hline
    \end{tabular}
  \end{center}

```{r}
bookmat <- matrix(c(0,1,1,10,10,0,10,10,1,1,Inf,Inf),nrow=3,byrow=TRUE)
dimnames(bookmat) <- list(c("Ben","Jake","Tom"),c("Mo","John","Terry","Pat"))
## fullmatch(bookmat)
```
\note{
*Greedy:* Ben-Mo (0) , Jake-John (0), Tom-?
*Optimal:* Ben-Terry (1), Jake-John (0), Tom-Mo (1)
}

## Matching with and without replacement
*Example*: Costs of nuclear plants. When expanding, is it cheaper to build on the site of an existing plant?

\begin{columns}
\column{.5\textwidth}
\begin{center}
  \igrphx[width=.5\textwidth]{coxSnell}
\end{center}

\column{.5\textwidth}
  \igrphx[width=\textwidth]{Nuclear_plant_at_Grafenrheinfeld}
\end{columns}

\note{This is a \textit{prospective} design.  Explain.

(Presentation version has Nuclear plant image on 2nd overlay)
}


## New vs refurbished nuclear plants
\note{
  \begin{itemize}
  \item This is what many understand by matching.  [Greedy pair match follows.]
  \item Sometimes matching means  ``with-replacement'' matching. We'll permit that too, but in a structured way, and keeping track of how many replacements. 
  \item Why keep track of the replacements?  --B/c they affect standard errors of the matched comparison.  (``Effective sample size'' to be introduced a few slides later.)
  \end{itemize}
}
\begin{minipage}[t]{2in}
\begin{center}
Existing site\\
{\small
% latex table generated in R 3.0.1 by xtable 1.7-1 package
% Wed Aug 14 15:26:53 2013
\begin{tabular}{lr}
  \hline
 & capacity \\ 
  \hline
A & {660} {\mlpnode{NA}} \\ 
  B & {660} {\mlpnode{NB}} \\ 
  C & {420} {\mlpnode{NC}} \\ 
  D & {130} {\mlpnode{ND}} \\ 
  E & {650} {\mlpnode{NE}} \\ 
  F & {430} {\mlpnode{NF}} \\ 
  G & {420} {\mlpnode{NG}} \\ 
   \hline
\end{tabular}}
\end{center}
\bigskip
\bigskip
{\footnotesize  "capacity" is net capacity of the power plant, in MWe above
400.

\only<9->{\alert{This is matching "without replacement."}}
\only<10->{\alert{Contrast with matching "with replacement."}}
}
\end{minipage}
\begin{minipage}[t]{2in}
\begin{center}
New site\\
{\scriptsize
% latex table generated in R 3.0.1 by xtable 1.7-1 package
% Wed Aug 14 15:26:53 2013
\begin{tabular}{lr}
  \hline
 & capacity \\ 
  \hline
{\mlpnode{NH}\mbox{}} {H} & 290 \\ 
  {\mlpnode{NI}\mbox{}} {I} & 660 \\ 
  {\mlpnode{NJ}\mbox{}} {J} & 660 \\ 
  {\mlpnode{NK}\mbox{}} {K} & 110 \\ 
  {\mlpnode{NL}\mbox{}} {L} & 420 \\ 
  {\mlpnode{NM}\mbox{}} {M} &  60 \\ 
  {\mlpnode{NN}\mbox{}} {N} & 390 \\ 
  {\mlpnode{NO}\mbox{}} {O} & 160 \\ 
  {\mlpnode{NP}\mbox{}} {P} & 390 \\ 
  {\mlpnode{NQ}\mbox{}} {Q} & 130 \\ 
  {\mlpnode{NR}\mbox{}} {R} & 650 \\ 
  {\mlpnode{NS}\mbox{}} {S} & 450 \\ 
  {\mlpnode{NT}\mbox{}} {T} & 380 \\ 
  {\mlpnode{NU}\mbox{}} {U} & 440 \\ 
  {\mlpnode{NV}\mbox{}} {V} & 690 \\ 
  {\mlpnode{NW}\mbox{}} {W} & 510 \\ 
  {\mlpnode{NX}\mbox{}} {X} & 390 \\ 
  {\mlpnode{NY}\mbox{}} {Y} & 140 \\ 
  {\mlpnode{NZ}\mbox{}} {Z} & 730 \\ 
   \hline
\end{tabular}}
\end{center}
\end{minipage}


\begin{tikzpicture}[overlay]
  \path[draw,gray]<2-> (NA) edge (NI);
 \path[draw,gray]<3-> (NB) edge (NJ);
 \path[draw,gray]<4-> (NC) edge (NL);
 \path[draw,gray]<5-> (ND) edge (NQ);
 \path[draw,gray]<6-> (NE) edge (NR);
 \path[draw,gray]<7-> (NF) edge (NU);
 \path[draw,gray]<8-> (NG) edge (NN);
 \end{tikzpicture}\only<10>{
\begin{tikzpicture}[overlay]
  \path[draw,red] (NA) edge [out=0, in=-180] (NI);
 \path[draw,red] (NB) edge [out=0, in=-180] (NJ);
 \path[draw,red] (NC) edge [out=0, in=-180] (NL);
 \path[draw,red] (NF) edge [out=0, in=-180] (NL);
 \path[draw,red] (NG) edge [out=0, in=-180] (NL);
 \path[draw,red] (ND) edge [out=0, in=-180] (NQ);
 \path[draw,red] (NE) edge [out=0, in=-180] (NR);
 \end{tikzpicture}}



## Role of the matching algorithm: greedy matching

\begin{minipage}[t]{2in}
\begin{center}
Existing site\\
{\small
% latex table generated in R 3.0.1 by xtable 1.7-1 package
% Wed Aug 14 15:26:57 2013
\begin{tabular}{lrr}
  \hline
 & z.date & z.cap \\ 
  \hline
A & -1.6 & {1.2} {\mlpnode{NA}} \\ 
  B & -0.9 & {1.2} {\mlpnode{NB}} \\ 
  C & -0.4 & {0} {\mlpnode{NC}} \\ 
  D & -0.4 & {-1.4} {\mlpnode{ND}} \\ 
  E & 0.1 & {1.1} {\mlpnode{NE}} \\ 
  F & 2.2 & {0} {\mlpnode{NF}} \\ 
  G & 1.3 & {0} {\mlpnode{NG}} \\ 
   \hline
\end{tabular}}
\end{center}
\bigskip
\bigskip
\bigskip
{\footnotesize Here, ``date'' is \emph{rank of} date of
construction, in years after 1965, and  ``capacity'' is
\emph{rank of} net capacity of the power plant, in MWe
above 400.}
\end{minipage}
\begin{minipage}[t]{2in}
\begin{center}
New site\\
{\scriptsize
% latex table generated in R 3.0.1 by xtable 1.7-1 package
% Wed Aug 14 15:26:57 2013
\begin{tabular}{lrr}
  \hline
 & z.date & z.cap \\ 
  \hline
{\mlpnode{NH}\mbox{}} {H} & -0.3 & -0.7 \\ 
  {\mlpnode{NI}\mbox{}} {I} & -1.6 & 1.2 \\ 
  {\mlpnode{NJ}\mbox{}} {J} & -0.9 & 1.2 \\ 
  {\mlpnode{NK}\mbox{}} {K} & -0.9 & -1.5 \\ 
  {\mlpnode{NL}\mbox{}} {L} & -0.7 & -0.0 \\ 
  {\mlpnode{NM}\mbox{}} {M} & -0.4 & -1.8 \\ 
  {\mlpnode{NN}\mbox{}} {N} & -0.5 & -0.2 \\ 
  {\mlpnode{NO}\mbox{}} {O} & -0.3 & -1.3 \\ 
  {\mlpnode{NP}\mbox{}} {P} & -0.1 & -0.2 \\ 
  {\mlpnode{NQ}\mbox{}} {Q} & -0.4 & -1.4 \\ 
  {\mlpnode{NR}\mbox{}} {R} & 0.1 & 1.1 \\ 
  {\mlpnode{NS}\mbox{}} {S} & 0.1 & 0.1 \\ 
  {\mlpnode{NT}\mbox{}} {T} & -0.4 & -0.2 \\ 
  {\mlpnode{NU}\mbox{}} {U} & 0.7 & 0.1 \\ 
  {\mlpnode{NV}\mbox{}} {V} & 0.4 & 1.3 \\ 
  {\mlpnode{NW}\mbox{}} {W} & -0.1 & 0.4 \\ 
  {\mlpnode{NX}\mbox{}} {X} & 0.9 & -0.2 \\ 
  {\mlpnode{NY}\mbox{}} {Y} & 1.7 & -1.4 \\ 
  {\mlpnode{NZ}\mbox{}} {Z} & 2.3 & 1.5 \\ 
   \hline
\end{tabular}}
\end{center}
\end{minipage}

\begin{tikzpicture}[overlay]
  \path[draw,red]<2-> (NA) edge (NI);
 \path[draw,red]<3-> (NA) edge (NJ);
 \path[draw,red]<4-> (NB) edge (NR);
 \path[draw,red]<5-> (NB) edge (NW);
 \path[draw,red]<6-> (NC) edge (NN);
 \path[draw,red]<7-> (NC) edge (NT);
 \path[draw,red]<8-> (ND) edge (NO);
 \path[draw,red]<9-> (ND) edge (NQ);
 \path[draw,red]<10-> (NE) edge (NS);
 \path[draw,red]<11-> (NE) edge (NV);
 \path[draw,red]<12-> (NF) edge (NU);
 \path[draw,red]<13-> (NF) edge (NX);
 \path[draw,red]<14-> (NG) edge (NM);
 \path[draw,red]<15-> (NG) edge (NZ);
 \end{tikzpicture}

## Optimal Matching

\begin{minipage}[t]{2in}
\begin{center}
Existing site\\
{\small
% latex table generated in R 3.0.1 by xtable 1.7-1 package
% Wed Aug 14 15:26:57 2013
\begin{tabular}{lrr}
  \hline
 & z.date & z.cap \\ 
  \hline
A & -1.6 & {1.2} {\mlpnode{NA}} \\ 
  B & -0.9 & {1.2} {\mlpnode{NB}} \\ 
  C & -0.4 & {0} {\mlpnode{NC}} \\ 
  D & -0.4 & {-1.4} {\mlpnode{ND}} \\ 
  E & 0.1 & {1.1} {\mlpnode{NE}} \\ 
  F & 2.2 & {0} {\mlpnode{NF}} \\ 
  G & 1.3 & {0} {\mlpnode{NG}} \\ 
   \hline
\end{tabular}}
\end{center}
\bigskip

\textcolor{blue}{Optimal vs. Greedy matching}
\bigskip

{\footnotesize By evaluating potential matches all together rather than
  sequentially, optimal matching (\textcolor{blue}{blue lines}) reduces
  the mean matched distance from 1.17 to
 0.56.}

\end{minipage}
\begin{minipage}[t]{2in}
\begin{center}
New site\\
{\scriptsize
% latex table generated in R 3.0.1 by xtable 1.7-1 package
% Wed Aug 14 15:26:57 2013
\begin{tabular}{lrr}
  \hline
 & z.date & z.cap \\ 
  \hline
{\mlpnode{NH}\mbox{}} {H} & -0.3 & -0.7 \\ 
  {\mlpnode{NI}\mbox{}} {I} & -1.6 & 1.2 \\ 
  {\mlpnode{NJ}\mbox{}} {J} & -0.9 & 1.2 \\ 
  {\mlpnode{NK}\mbox{}} {K} & -0.9 & -1.5 \\ 
  {\mlpnode{NL}\mbox{}} {L} & -0.7 & -0.0 \\ 
  {\mlpnode{NM}\mbox{}} {M} & -0.4 & -1.8 \\ 
  {\mlpnode{NN}\mbox{}} {N} & -0.5 & -0.2 \\ 
  {\mlpnode{NO}\mbox{}} {O} & -0.3 & -1.3 \\ 
  {\mlpnode{NP}\mbox{}} {P} & -0.1 & -0.2 \\ 
  {\mlpnode{NQ}\mbox{}} {Q} & -0.4 & -1.4 \\ 
  {\mlpnode{NR}\mbox{}} {R} & 0.1 & 1.1 \\ 
  {\mlpnode{NS}\mbox{}} {S} & 0.1 & 0.1 \\ 
  {\mlpnode{NT}\mbox{}} {T} & -0.4 & -0.2 \\ 
  {\mlpnode{NU}\mbox{}} {U} & 0.7 & 0.1 \\ 
  {\mlpnode{NV}\mbox{}} {V} & 0.4 & 1.3 \\ 
  {\mlpnode{NW}\mbox{}} {W} & -0.1 & 0.4 \\ 
  {\mlpnode{NX}\mbox{}} {X} & 0.9 & -0.2 \\ 
  {\mlpnode{NY}\mbox{}} {Y} & 1.7 & -1.4 \\ 
  {\mlpnode{NZ}\mbox{}} {Z} & 2.3 & 1.5 \\ 
   \hline
\end{tabular}}
\end{center}
\end{minipage}

\begin{tikzpicture}[overlay]
  \path[draw,red] (NA) edge (NI);
 \path[draw,red] (NA) edge (NJ);
 \path[draw,red] (NB) edge (NR);
 \path[draw,red] (NB) edge (NW);
 \path[draw,red] (NC) edge (NN);
 \path[draw,red] (NC) edge (NT);
 \path[draw,red] (ND) edge (NO);
 \path[draw,red] (ND) edge (NQ);
 \path[draw,red] (NE) edge (NS);
 \path[draw,red] (NE) edge (NV);
 \path[draw,red] (NF) edge (NU);
 \path[draw,red] (NF) edge (NX);
 \path[draw,red] (NG) edge (NM);
 \path[draw,red] (NG) edge (NZ);
 \end{tikzpicture}\begin{tikzpicture}[overlay]
  \path[draw,blue] (NA) edge [out=0, in=-180] (NI);
 \path[draw,blue] (NA) edge [out=0, in=-180] (NJ);
 \path[draw,blue] (NB) edge [out=0, in=-180] (NL);
 \path[draw,blue] (NB) edge [out=0, in=-180] (NW);
 \path[draw,blue] (NC) edge [out=0, in=-180] (NN);
 \path[draw,blue] (NC) edge [out=0, in=-180] (NT);
 \path[draw,blue] (ND) edge [out=0, in=-180] (NO);
 \path[draw,blue] (ND) edge [out=0, in=-180] (NQ);
 \path[draw,blue] (NE) edge [out=0, in=-180] (NR);
 \path[draw,blue] (NE) edge [out=0, in=-180] (NV);
 \path[draw,blue] (NF) edge [out=0, in=-180] (NY);
 \path[draw,blue] (NF) edge [out=0, in=-180] (NZ);
 \path[draw,blue] (NG) edge [out=0, in=-180] (NU);
 \path[draw,blue] (NG) edge [out=0, in=-180] (NX);
 \end{tikzpicture}

## Estimation and Testing after Matching.

## Summary:

Next:
 - Estimation and Testing after Matching (and then Sensitivity Analysis)
 - When to stop matching? (When is balance good enough?)
 - How to think about information and effective sample size.

Summary:
 - Greedy matching can either leave out potentially matchable units and/or produce worse overall matched set configurations \autocite{hans:04,rose:char:1991}.


## Anything Else?

## References
