---
title: |
 | Statistical Adjustment in Observational Studies,
 | Assessment of Adjustment,
 | Matched Stratification for One and Multiple Variables.
date: '`r format(Sys.Date(), "%B %d, %Y")`'
author: |
  | ICPSR 2023 Session 1
  | Jake Bowers \& Tom Leavitt
bibliography:
 - 'BIB/MasterBibliography.bib'
fontsize: 10pt
geometry: margin=1in
graphics: yes
biblio-style: authoryear-comp
biblatexoptions:
  - natbib=true
urlcolor: orange
linkcolor: orange
output:
  beamer_presentation:
    slide_level: 2
    keep_tex: true
    latex_engine: xelatex
    citation_package: biblatex
    template: styles/icpsr-beamer-template
    incremental: true
    includes:
        in_header:
           - defs-all.sty
    md_extensions: +raw_attribute-tex_math_single_backslash+autolink_bare_uris+ascii_identifiers+tex_math_dollars
    pandoc_args: [ "--csl", "chicago-author-date.csl" ]
---


<!-- To show notes  -->
<!-- https://stackoverflow.com/questions/44906264/add-speaker-notes-to-beamer-presentations-using-rmarkdown -->

```{r setup1_env, echo=FALSE, include=FALSE}
library(here)
source(here::here("rmd_setup.R"))
```

```{r setup2_loadlibs, echo=FALSE, include=FALSE}
## Load all of the libraries that we will use when we compile this file
## We are using the renv system. So these will all be loaded from a local library directory
library(dplyr)
library(ggplot2)
library(coin)
library(RItools)
library(optmatch)
```


## Today

  1. Agenda:  Continue to think about an adjustment strategy to enhance
     counterfactual causal interpretations in non-randomized studies:
      - Workflow:
         a. choose basis or covariates that need adjusting $\rightarrow$
         b. create a stratification of the data / a stratified research design (ex. `fullmatch`) $\rightarrow$
         c. choose one or more approaches to evaluate the stratification (ex. `balanceTest`) $\rightarrow$
         d. (probably) improve the stratified design / iterate 
      - How to do this with one variable?
      - How to do this with  more than one variable?
      - How to create strata to clarify comparison along continuous or multi-valued
        "treatment" variables ("non-bipartite" matching) Feel free to read
        ahead.
 2. Questions arising from the reading or assignments or life?


```{r echo=FALSE, cache=TRUE}
load(url("http://jakebowers.org/Data/meddat.rda"))
meddat <- mutate(meddat,
  HomRate03 = (HomCount2003 / Pop2003) * 1000,
  HomRate08 = (HomCount2008 / Pop2008) * 1000
)
row.names(meddat) <- meddat$nh
```

## The optmatch workflow: The distance matrix

To minimize overall differences `optmatch` requires a matrix
of those differences (in general terms, a matrix of distances between the
treated and control units)

```{r optm1, echo=TRUE}
tmp <- meddat$nhAboveHS
names(tmp) <- rownames(meddat)
absdist <- match_on(tmp, z = meddat$nhTrt, data = meddat)
absdist[1:3, 1:3]
meddat %>%
  group_by(nhTrt) %>%
  summarize(nhAboveHS[1])
abs(meddat$nhAboveHS[meddat$nhTrt == 1][1] - meddat$nhAboveHS[meddat$nhTrt == 0][1])
```

## Created a Stratified Research Design

`fullmatch` and `pairmatch` use a distance matrix to produce strata.

```{r fm1, echo=TRUE}
fm1 <- fullmatch(absdist, data = meddat)
summary(fm1, min.controls = 0, max.controls = Inf)
table(meddat$nhTrt, fm1)
pm1 <- pairmatch(absdist, data = meddat)
summary(pm1, min.controls = 0, max.controls = Inf)
table(meddat$nhTrt, pm1, exclude = c())
fm2 <- fullmatch(absdist + caliper(absdist,.06), data = meddat, min.controls = .5)
stratumStructure(fm2)

summary(unlist(matched.distances(fm1,absdist)))

summary(unlist(matched.distances(fm2,absdist)))
```


## Evaluate the design: Within set differences

```{r echo=FALSE}
meddat$fm2 <- fm2
meddat$fm1 <- fm1
meddat$pm1 <- pm1
```

Differences within sets versus raw differences.

```{r echo=FALSE, out.width=".9\\textwidth"}
library(gridExtra)
bpfm1 <- ggplot(meddat, aes(x = fm1, y = nhAboveHS)) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", shape = 20, size = 3, color = "red", fill = "red")
meddat$nostrata <- rep(1, 45)
bporig <- ggplot(meddat, aes(x = nostrata, y = nhAboveHS)) +
  geom_boxplot() +
  stat_summary(
    fun = mean, geom = "point",
    shape = 20, size = 3, color = "red", fill = "red"
  )

grid.arrange(bpfm1, bporig, ncol = 2, layout_matrix = matrix(c(1, 1, 1, 1, 2), nrow = 1))
```

## Evaluate the design: Within-set differences
Differences within sets versus raw differences.

```{r echo=FALSE, out.width=".9\\textwidth"}
bpfm2 <- ggplot(meddat, aes(x = fm2, y = nhAboveHS)) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", shape = 20, size = 3, color = "red", fill = "red")

grid.arrange(bpfm2, bporig, ncol = 2, layout_matrix = matrix(c(1, 1, 1, 1, 2), nrow = 1))
```



## Evaluate the design: Within-set differences
Differences within sets versus raw differences.

```{r echo=FALSE, out.width=".9\\textwidth"}
bppm1 <- ggplot(meddat, aes(x = pm1, y = nhAboveHS)) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", shape = 20, size = 3, color = "red", fill = "red")

grid.arrange(bppm1, bporig, ncol = 2, layout_matrix = matrix(c(1, 1, 1, 1, 2), nrow = 1))
```

## Evaluate the design: Inspect within-set differences

Use what we know about Medellin neighborhoods, about Metrocable, to assess the research design:

```{r sdiffs, echo=FALSE}
rawmndiffs <- with(meddat, mean(nhAboveHS[nhTrt == 1]) - mean(nhAboveHS[nhTrt == 0]))
setdiffsfm1 <- meddat %>%
  group_by(fm1) %>%
  summarize(
    mneddiffs =
      mean(nhAboveHS[nhTrt == 1]) -
        mean(nhAboveHS[nhTrt == 0]),
    mnAboveHS = mean(nhAboveHS),
    minAboveHS = min(nhAboveHS),
    maxAboveHS = max(nhAboveHS)
  )

setdiffsfm1 %>% arrange(desc(abs(mneddiffs)))
# summary(setdiffs$mneddiffs)
# quantile(setdiffs$mneddiffs, seq(0,1,.1))
```



## Evaluate the design: Inspect within set differences

Use what we know about Medellin neighborhoods, about Metrocable, to assess the research design:

```{r echo=FALSE, warnings=FALSE}
setdiffspm1 <- meddat %>%
  group_by(pm1) %>%
  summarize(
    mneddiffs =
      mean(nhAboveHS[nhTrt == 1]) -
        mean(nhAboveHS[nhTrt == 0]),
    mnAboveHS = mean(nhAboveHS),
    minAboveHS = min(nhAboveHS),
    maxAboveHS = max(nhAboveHS)
  )

setdiffspm1 %>% arrange(desc(abs(mneddiffs)))
```


## Evaluate the design: Compare to a randomized experiment. {.allowframebreaks}

The observed within-set differences differ from those that would be expected
from a randomized experiment. Here, we compare 3 designs that keep all
observations, just organized into sets differently.

```{r xbhs2, echo=TRUE}
xbfm1 <- balanceTest(nhTrt ~ nhAboveHS + strata(fm1) + strata(fm2) + strata(pm1),
  data = meddat
)
xbfm1$results[,,]
xbfm1$overall
```


## What is balanceTest doing?

```{r xbagain, echo=TRUE}
setmeanDiffs_fm1 <- meddat %>%
  group_by(fm1) %>%
  summarise(
    diffAboveHS = mean(nhAboveHS[nhTrt == 1]) - mean(nhAboveHS[nhTrt == 0]),
    nb = n(),
    nTb = sum(nhTrt),
    nCb = sum(1 - nhTrt),
    hwt = (2 * (nCb * nTb) / (nTb + nCb))
  )
setmeanDiffs_fm1 %>% arrange(desc(abs(diffAboveHS)))
```

## What is balanceTest doing with multiple sets/blocks?

The test statistic is a weighted average of the set-specific differences (same
approach as we would use to test the null in a block-randomized experiment)

```{r wtmns, echo=TRUE}
## The descriptive mean difference using block-size weights
with(setmeanDiffs_fm1, sum(diffAboveHS*nTb/sum(nTb)))
## The mean diff used as the observed value in the testing
with(setmeanDiffs_fm1, sum(diffAboveHS * hwt / sum(hwt)))
## Compare to balanceTest output
xbfm1$results[, , "fm1"]
```

## What do we mean by "compare to a randomized experiment"?

Recall that $p$-values require distributions, and distributions require processes that can be repeated (even in a Bayesian formulation). What are we repeating in an experiment? (The assignment mechanism). 

```{r d1v2, echo=TRUE}
## This is another version that might be more clear in regards what is going on.
dstatv3 <- function(zz, mm, ss) {
  dat <- data.frame(mm = mm, z = zz, s = ss)
  datb <- dat %>%
    group_by(s) %>%
    summarize(
      mndiff = mean(mm[z == 1]) - mean(mm[z == 0]),
      nb = n(),
      pib = mean(z),
      nbwt = nb / nrow(dat),
      hbwt0 = pib * (1 - pib) * nbwt
    )
  datb$hbwt <- datb$hbwt0 / sum(datb$hbwt0)
  adjmn <- with(datb, sum(mndiff * hbwt))
  return(adjmn)
}
```

## What do we mean by "compare to a randomized experiment"?

Recall that $p$-values require distributions, and distributions require processes that can be repeated (even in a Bayesian formulation). What are we repeating in an experiment? (The assignment mechanism). 


```{r xbdist, messages=FALSE, warning=FALSE}
newexp <- function(zz, ss) {
  ## newz <- unlist(lapply(split(zz,ss),sample),ss)
  newzdat <- data.frame(zz = zz, ss = ss) %>%
    group_by(ss) %>%
    mutate(newz = sample(zz))
  return(newzdat$newz)
}

set.seed(123455)
nulldist <- replicate(1000, with(meddat, dstatv3(zz = newexp(zz = nhTrt, ss = fm1), mm = nhAboveHS, ss = fm1)))
obsdstat <- with(meddat, dstatv3(zz = nhTrt, mm = nhAboveHS, ss = fm1))
perm_p_value <- 2 * min(mean(nulldist >= obsdstat), mean(nulldist <= obsdstat))
perm_p_value
xbfm1$results[,,"fm1"]
```


## What do we mean by "compare to a randomized experiment"?

Recall that $p$-values require distributions, and distributions require processes that can be repeated (even in a Bayesian formulation). What are we repeating in an experiment? (The assignment mechanism). 

```{r plotxbnulldist, out.width=".7\\textwidth"}
plot(density(nulldist))
curve(dnorm(x,sd=sd(nulldist)),from=min(nulldist),to=max(nulldist),col="blue",add=TRUE)
rug(nulldist)
abline(v = obsdstat)
```

#  Matching on Many Covariates: Using Mahalnobis Distance

## Dimension reduction using the Mahalanobis Distance

The general idea: dimension reduction. When we convert many columns into one column we reduce the dimensions of the dataset (to one column). If using distance matrices and optimal matching allows us to side-step the problem of choosing cut-points for stratification, we can use the idea of **multivariate distance** to produce distance matrices to minimize **multivariate distances**.


```{r out.width=".7\\textwidth"}
X <- meddat[, c("nhAboveHS", "nhPopD")]
plot(meddat$nhAboveHS, meddat$nhPopD, xlim = c(-.3, .6), ylim = c(50, 700))
```

## Dimension reduction using the Mahalanobis Distance

First, let's look at Euclidean distance: $\sqrt{ (x_1 - x_2)^2 + (y_1 - y_2)^2 }$

```{r echo=FALSE, out.width=".8\\textwidth"}
par(mgp = c(1.25, .5, 0), oma = rep(0, 4), mar = c(3, 3, 0, 0))
plot(meddat$nhAboveHS, meddat$nhPopD, xlim = c(-.3, .6), ylim = c(50, 700))
points(mean(X[, 1]), mean(X[, 2]), pch = 19, cex = 1)
arrows(mean(X[, 1]), mean(X[, 2]), X["407", 1], X["407", 2])
text(.4, 200, label = round(dist(rbind(colMeans(X), X["407", ])), 2))
text(.3,100, label="Distance from the mean of both dimensions")
text(.3,75,bquote("(" * bar(x) * " = " * .(mean(X[,1])) * ", " *  bar(y) * " = " * .(mean(X[,2])) * ")"))
```

## Dimension reduction using the Mahalanobis Distance

First, let's look at Euclidean distance: $\sqrt{ (x_1 - x_2)^2 + (y_1 - y_2)^2 }$

```{r echo=FALSE, out.width=".5\\textwidth"}
par(mgp = c(1.25, .5, 0), oma = rep(0, 4), mar = c(3, 3, 0, 0))
plot(meddat$nhAboveHS, meddat$nhPopD, xlim = c(-.3, .6), ylim = c(50, 700))
points(mean(X[, 1]), mean(X[, 2]), pch = 19, cex = 1)
arrows(mean(X[, 1]), mean(X[, 2]), X["407", 1], X["407", 2])
text(.4, 200, label = round(dist(rbind(colMeans(X), X["407", ])), 2))
```

Distance between point in middle of the plot and unit "407".

```{r}
tmp <- rbind(colMeans(X), X["407", ])
tmp
sqrt((tmp[1, 1] - tmp[2, 1])^2 + (tmp[1, 2] - tmp[2, 2])^2)
```

Problem: overweights variables with bigger scales (Population Density dominates).

## Dimension reduction using the Mahalanobis Distance

Now the Euclidean distance (on a standardized scale) so neither variable is overly dominant.

```{r echo=TRUE}
Xsd <- scale(X)
apply(Xsd, 2, sd) ## should be 1
round(apply(Xsd, 2, mean),8)
```
```{r echo=FALSE,out.width=".6\\textwidth"}
plot(Xsd[, 1], Xsd[, 2], xlab = "nhAboveHS/sd", ylab = "nhPopD/sd")
points(mean(Xsd[, 1]), mean(Xsd[, 2]), pch = 19, cex = 1)
arrows(mean(Xsd[, 1]), mean(Xsd[, 2]), Xsd["407", 1], Xsd["407", 2])
text(2, -1.2, label = round(dist(rbind(colMeans(Xsd), Xsd["407", ])), 2))
```


## Dimension reduction using the Mahalanobis Distance

The mahalanobis distance avoids the scale problem in the euclidean distance.^[For more [see here](https://stats.stackexchange.com/questions/62092/bottom-to-top-explanation-of-the-mahalanobis-distance)] Here each circle are points of the same MH distance.

```{r mhfig, echo=FALSE,out.width=".6\\textwidth"}
library(chemometrics) ## for drawMahal
library(mvtnorm)

par(mgp = c(1.5, .5, 0), oma = rep(0, 4), mar = c(3, 3, 0, 0))
mh <- mahalanobis(X, center = colMeans(X), cov = cov(X))
drawMahal(X,
  center = colMeans(X), covariance = cov(X),
  quantile = c(0.975, 0.75, 0.5, 0.25)
)
abline(v = mean(meddat$nhAboveHS), h = mean(meddat$nhPopD))
pts <- c("401", "407", "411", "202")
arrows(rep(mean(X[, 1]), 4), rep(mean(X[, 2]), 4), X[pts, 1], X[pts, 2])
text(X[pts, 1], X[pts, 2], labels = round(mh[pts], 2), pos = 1)
```

```{r}
Xsd <- scale(X)
tmp <- rbind(c(0, 0), Xsd["407", ])
mahalanobis(tmp, center = c(0, 0), cov = cov(Xsd))
edist <- sqrt((tmp[1, 1] - tmp[2, 1])^2 + (tmp[1, 2] - tmp[2, 2])^2)
edist
```

## Dimension reduction using the Mahalanobis Distance

The standardized data

```{r}
plot(Xsd[, 1], Xsd[, 2], xlab = "nhAboveHS/sd", ylab = "nhPopD/sd")
```

## Dimension reduction using the Mahalanobis Distance

With mahalanobis distance contours calculated on standardized data, but then
applied to the raw data.

```{r}
drawMahal(X, center = colMeans(X), covariance = cov(X), quantile = c(.1, .2))
```

```{r}
covX <- cov(X)
newcovX <- covX
newcovX[1, 2] <- 0
newcovX[2, 1] <- 0
```

## Dimension reduction using the Mahalanobis Distance

Using a new covariance matrix (assuming no covariance between the variables).

```{r}
drawMahal(X, center = colMeans(X), covariance = newcovX, quantile = c(.1, .2))
```

## Matching on the Mahalanobis Distance

Here using the rank based Mahalanobis distance following DOS Chap. 8 (but comparing to the ordinary version).

```{r echo=TRUE}
mhdist <- match_on(nhTrt ~ nhPopD + nhAboveHS, data = meddat, method = "rank_mahalanobis")
mhdist[1:3, 1:3]
mhdist2 <- match_on(nhTrt ~ nhPopD + nhAboveHS, data = meddat)
mhdist2[1:3, 1:3]
mhdist2[, "407"]
```

## Matching on the Mahalanobis Distance

Here using the rank based Mahalanobis distance following DOS Chap. 8 (but comparing to the ordinary version).

```{r out.width=".5\\textwidth"}
par(mgp = c(1.5, .5, 0), oma = rep(0, 4), mar = c(3, 3, 0, 0))
drawMahal(X,
  center = colMeans(X), covariance = cov(X),
  quantile = c(0.975, 0.75, 0.5, 0.25)
)
abline(v = mean(meddat$nhAboveHS), h = mean(meddat$nhPopD))
cpts <- c("401", "407", "411")
tpts <- c("101", "102", "202")
arrows(X[tpts, 1], X[tpts, 2], rep(X["407", 1]), rep(X["407", 2]))
text(X[tpts, 1], X[tpts, 2], labels = round(mhdist2[tpts, "407"], 2), pos = 1)
```

```{r echo=TRUE}
mhdist2[tpts, "407"]
```

## More on the Mahalanobis distance

To Review: The Mahalanobis distance \citep{mahalanobis1930test}, avoids the scale and correlation problem in the euclidean distance.^[For more see <https://stats.stackexchange.com/questions/62092/bottom-to-top-explanation-of-the-mahalanobis-distance>] $dist_M =  \sqrt{ (\mathbf{x} - \mathbf{\bar{x}})^T \mathbf{M}^{-1} (\mathbf{y} - \mathbf{\bar{y}}) }$ where $\mathbf{M}=\begin{bmatrix} \var(x) & \cov(x,y)  \\ \cov(x,y) & \var(y) \end{bmatrix}$


Here, using simulated data: The contour lines show points with the same
Mahalanobis distance and the numbers are Euclidean distance.

```{r}
set.seed(12345)
newX <- rmvnorm(n=45,mean=colMeans(X),sigma=matrix(c(.03,-7,-7,14375),2,2))
row.names(newX) <- row.names(X)
#plot(newX)
cor(newX)
```

```{r echo=FALSE, out.width=".4\\textwidth"}
mhnew <- mahalanobis(newX,center=colMeans(newX),cov=cov(newX))
drawMahal(newX,center=colMeans(newX),covariance=cov(newX),
          quantile = c(0.975, 0.75, 0.5, 0.25))
abline(v=mean(newX[,1]),h=mean(newX[,2]),col="gray")
points(mean(newX[,1]),mean(newX[,2]),pch=19,cex=1)
newpts <- c(3,16,17,20,22,30,42)
row.names(newX[newpts,])
arrows(mean(newX[,1]),mean(newX[,2]),newX[newpts,1],newX[newpts,2],length=.1)
edist <- as.matrix(dist(rbind(centers=colMeans(newX),newX[newpts,])))
text(newX[newpts,1]-.02,newX[newpts,2]-10,
     label=round(edist[-1,"centers"],2),font=2)
```

## Dimension reduction using the Mahalanobis distance

The contour lines show points with the same
Mahalanobis distance, the numbers are Euclidean distance. Notice that the point with Euclidean distance of 161 is farther from the center than 250 in Mahalanobis terms.

```{r echo=FALSE, results="hide", out.width=".75\\textwidth"}
mhnew <- mahalanobis(newX,center=colMeans(newX),cov=cov(newX))
drawMahal(newX,center=colMeans(newX),covariance=cov(newX),
          quantile = c(0.975, 0.75, 0.5, 0.25))
abline(v=mean(newX[,1]),h=mean(newX[,2]),col="gray")
points(mean(newX[,1]),mean(newX[,2]),pch=19,cex=1)
newpts <- c(3,16,17,20,22,30,42)
row.names(newX[newpts,])
arrows(mean(newX[,1]),mean(newX[,2]),newX[newpts,1],newX[newpts,2],length=.1)
edist <- as.matrix(dist(rbind(centers=colMeans(newX),newX[newpts,])))
text(newX[newpts,1]-.02,newX[newpts,2]-10,
     label=round(edist[-1,"centers"],2),font=2)
```


## Dimension reduction using the Mahalanobis distance

The contour lines show points with the same
Mahalanobis distance and the numbers are Euclidean distance (on the
standardized variables). (notice that 1.63 is farther from the center in
Mahalanobis terms than 2.11, but 2.11 is farther in Eucliean terms.)


```{r echo=FALSE}
newXsd <- scale(newX)
drawMahal(newXsd,center=colMeans(newXsd),covariance=cov(newXsd),
          quantile = c(0.975, 0.75, 0.5, 0.25))
abline(v=mean(newXsd[,1]),h=mean(newXsd[,2]),col="gray")
points(mean(newXsd[,1]),mean(newXsd[,2]),pch=19,cex=1)
arrows(mean(newXsd[,1]),mean(newXsd[,2]),newXsd[newpts,1],newXsd[newpts,2],length=.1)
edistSd <- as.matrix(dist(rbind(centers=colMeans(newXsd),newXsd[newpts,])))
text(newXsd[newpts,1]-.1,newXsd[newpts,2]-.1,
     label=round(edistSd[-1,"centers"],2),font=2)
```


## Matching on the Mahalanobis Distance

```{r echo=TRUE}
mhdist <- match_on(nhTrt ~ nhPopD + nhAboveHS, data = meddat, method = "rank_mahalanobis")

fmMh <- fullmatch(mhdist, data = meddat)
summary(fmMh, min.controls = 0, max.controls = Inf)
summary(unlist(matched.distances(fmMh,mhdist)))

quantile(as.vector(mhdist),seq(0,1,.1))
fmMh1 <- fullmatch(mhdist + caliper(mhdist,1) , data = meddat) #, min.controls = 1)
summary(fmMh1, min.controls = 0, max.controls = Inf)
summary(unlist(matched.distances(fmMh1,mhdist)))

```

```{r echo=TRUE}

xbMh <- balanceTest(nhTrt ~ nhAboveHS + nhPopD+strata(fmMh)+strata(fmMh1),  data = meddat)
xbMh$results[,,]
xbMh$overall
```


#  Matching on Many Covariates: Using Propensity Scores

## Matching on the propensity score

**Make the score**^[Note that we will be using `brglm` or `bayesglm` in the
future because of logit separation problems when the number of covariates
increases.]

```{r echo=TRUE}
theglm <- glm(nhTrt ~ nhPopD + nhAboveHS + HomRate03, data = meddat, family = binomial(link = "logit"))
thepscore <- theglm$linear.predictor
thepscore01 <- predict(theglm, type = "response")
```

We tend to match on the linear predictor rather than the version required to
range only between 0 and 1.

```{r echo=FALSE, out.width=".7\\textwidth"}
par(mfrow = c(1, 2), oma = rep(0, 4), mar = c(3, 3, 2, 0), mgp = c(1.5, .5, 0))
boxplot(split(thepscore, meddat$nhTrt), main = "Linear Predictor (XB)")
stripchart(split(thepscore, meddat$nhTrt), add = TRUE, vertical = TRUE)

boxplot(split(thepscore01, meddat$nhTrt), main = "Inverse Link Function (g^-1(XB)")
stripchart(split(thepscore01, meddat$nhTrt), add = TRUE, vertical = TRUE)
```

## Matching on the propensity score

```{r echo=TRUE}
psdist <- match_on(theglm, data = meddat)
psdist[1:4, 1:4]
fmPs <- fullmatch(psdist, data = meddat)
summary(fmPs, min.controls = 0, max.controls = Inf)
```

## Can you do better?

**Challenge:** Improve the matched design by adding covariates or functions of
covariates using either or both of the propensity score or mahalanobis distance
(rank- or not-rank based). So far we have:

```{r echo=TRUE}
thecovs <- unique(c(names(meddat)[c(5:7, 9:24)], "HomRate03"))
balfmla0 <- reformulate(thecovs, response = "nhTrt")
balfmla <- update(balfmla0,.~.+strata(fmMh)+strata(fmPs))
xb5 <- balanceTest(balfmla, data = meddat)
xb5$overall
```

## Can you do better?

Challenge: Improve the matched design. So far we have:

```{r}
plot(xb5)
```

# A pre-vs-post design?

## Why not use a pre-vs-post design for these data?  {.allowframebreaks}

We have homocide rates in 2003, indicator of Metrocable station completion in
2004, and homocide rates in 2008. Why not use a difference-in-differences design
here? Or a lagged dependent variable approach (Ding and Li 2019, Blackwell and
Glynn 2018).

```{r echo=TRUE}
library(estimatr)
library(tidyverse)

## Reshape the data to "long" form
meddat_long <- meddat %>% dplyr::select(one_of("nh","nhTrt","HomRate08","HomRate03")) %>% pivot_longer(cols=c("HomRate08","HomRate03"),names_to="year",names_prefix="HomRate0",values_to="HomRate")
head(meddat_long)
meddat_long$year <- as.numeric(meddat_long$year) + 2000
head(meddat_long)

## Calc by hand
the_means <- meddat_long %>% group_by(nhTrt,year) %>% summarize(bary=mean(HomRate)) %>% ungroup()
the_means
the_diffs <- the_means %>% summarize(trt_diff=bary[nhTrt==1&year==2008] - bary[nhTrt==1&year==2003],
ctrl_diff=bary[nhTrt==0&year==2008] - bary[nhTrt==0&year==2003])
the_diffs
did0 <- with(the_diffs, trt_diff - ctrl_diff)
did0

## Calc with OLS
did2 <- lm_robust(HomRate~nhTrt*I(year==2008),meddat_long,cluster=nh)
did2

## Using the wide format data
the_diffs2 <- meddat %>% summarize(trt_diff=mean(HomRate08[nhTrt==1]) - mean(HomRate03[nhTrt==1]),
ctrl_diff = mean(HomRate08[nhTrt==0]) - mean(HomRate03[nhTrt==0]))
the_diffs2 
did0_wide <- with(the_diffs2, trt_diff - ctrl_diff)
did0_wide

did3 <- lm_robust(I(HomRate08-HomRate03)~nhTrt,data=meddat)
did3

## Following Ding and Li 2019 to "bracket" the ATT using the DID and Lag DV
lag_dv <- lm_robust(HomRate08~nhTrt+HomRate03,data=meddat)
lag_dv
```

## Another pre-post-design

```{r}
## Make std diffs on baseline
baseline_dist <- match_on(nhTrt~HomRate03,data=meddat)

quantile(baseline_dist,seq(0,1,.1))
pm_hr03 <- pairmatch(baseline_dist + caliper(baseline_dist,.4),data=meddat,remove.unmatchables = TRUE)
stratumStructure(pm_hr03)
summary(unlist(matched.distances(pm_hr03,baseline_dist)))
meddat$pm_hr03 <- pm_hr03
baltest_hr03 <- balanceTest(nhTrt~HomRate03+strata(pm_hr03),data=meddat,subset=!is.na(pm_hr03))
baltest_hr03$overall

grpdat <- meddat %>% filter(!is.na(pm_hr03)) %>% 
  group_by(pm_hr03) %>% 
  summarize(diffHR03=mean(HomRate03[nhTrt==1]) - mean(HomRate03[nhTrt==0])) %>% 
  arrange(diffHR03)

## Within set very homogeneous on baseline homocide rates:
est1<- lm_robust(HomRate08~nhTrt,fixed_effects = ~pm_hr03,data=meddat,subset=!is.na(pm_hr03))
est1
## Look only at the difference
est2<- lm_robust(I(HomRate08-HomRate03)~nhTrt,fixed_effects = ~pm_hr03,data=meddat,subset=!is.na(pm_hr03))
est2

```

## Questions to answer with a pre-vs-post design for these data

Some questions to answer:

 - What justifies claims about unbiasedness or at least consistency of the
estimator of the Average effect of treatment on the treated  (the ATT or Effect
of Treatment on the Treated (ETT))?

 - What justifies use of a Normal (or t-distribution) as a description of the
hypothesize of no average effects? (It is **not** an approximation to an
as-if-randomized distribution. Is this a random sample from a population? Do we
believe that homocide rates arise from an IID Normal DGP?)

 - What if we wanted to test another hypothesis or use a different test statistic?
Say, the sharp null of no effects using a rank based test?

## Summary:

What do you think?

 - Statistical adjustment with linear regression models is hard to justify.
 - Stratification via matching is easier to justify and assess (and describe).
 - Matching solves the problem of making comparisons that are transparent
   (Question: "Did you adjust enough for X?" Ans: "Here is some evidence about
   how well I did.")
 - You can adjust for one variable or more than one (if more than one, you
   need to choose one or more methods for reducing many columns to one
   column).
 - The workflow involves the creation of a distance matrix, asking an
   algorithm to find the best configuration of sets that minimize the
   distances within set, and checking balance. (Eventually, it will also be
   concerned about the effective sample size.)
  - Next: We will get more into the
   differences between full matching, optimal matching, greedy matching,
   matching with and without replacement, etc.. (Also: handling
   missing data, calipers and other methods of improving design).
  - Next: How to estimate causal effects and test causal hypotheses with a
    matched design (preview: treat the matched design as a stratified randomized
    experiment aka an experiment with randomization within blocks.)



## References

