---
title: "Day 1 Exercise Sketch: Two Versions of Randomization Inference, Fisher Style."
author: Jake Bowers
---

The version using the central limit theorem for the randomization distribution

```{r}
n0 <- 201229
n1 <- 200745
n <- n0 + n1
obs_test_stat <- 0/n1

expected_test_stat_h0 <- 4/n
these_test_stat_h0 <- 5e-6

## The p-value
pnorm( (obs_test_stat-expected_test_stat_h0)/these_test_stat_h0 )
```

Another way:

```{r}

dat <- data.frame(Z=c(rep(0,n0),rep(1,n1)),Y=c(rep(1,4),rep(0, (n1+n0)-4 )))
with(dat,table(Z,Y))

test_stat <- function(z,y){
    sum(z*y)/sum(z)
}

## Calculate the observed summary of the treatment and outcome relationship
obs_test_stat2 <- with(dat,test_stat(z=Z,y=Y))
## Test the code: the two observed numbers should be the same
stopifnot(all.equal(obs_test_stat2,obs_test_stat))

```

Now, shuffle treatment assignment and recalculate the test statistic representing the null hypothesis of no effect (shuffling breaks the relationship between treatment assignment and outcome, thus representing "no effects")

```{r}

## This function shuffles treatment assignment
new_assignment <- function(z){
    sample(z)
}

## Does new_assignment() do the right thing?

set.seed(12345)
nsims <- 1000
randomization_distribution <- replicate(nsims,test_stat(z=new_assignment(dat$Z),y=dat$Y))


summary(randomization_distribution)
table(randomization_distribution)
table(randomization_distribution)/nsims

expected_test_stat_h0_2 <- mean(randomization_distribution)

expected_test_stat_h0_2 - expected_test_stat_h0

p_value_2 <- mean(randomization_distribution <= obs_test_stat2)

p_value_2

```

Compare the asymptotic and the permutation-based randomization reference distributions:


```{r}
the_rand_dist_tab <- table(randomization_distribution)/nsims
the_dist_vals <- as.numeric(names(the_rand_dist_tab))

set.seed(12345)
the_asymp_rand_dist <- rnorm(10000,mean=expected_test_stat_h0,sd=these_test_stat_h0)
dens_the_asymp_rand_dist <- density(the_asymp_rand_dist)

hist(randomization_distribution,probability = TRUE,xlim=range(the_asymp_rand_dist))
points(x=obs_test_stat2,y=-.01,pch=17,cex=2)
curve(dnorm(x,mean=expected_test_stat_h0,sd=these_test_stat_h0),from=min(the_asymp_rand_dist),to=max(the_asymp_rand_dist),add=TRUE)
lines(dens_the_asymp_rand_dist$x,dens_the_asymp_rand_dist$y)

```


# From Tom's Slides


```{r }


y_c <- c(10, 15, 20, 20, 10, 15, 15)
y_t <- c(15, 15, 30, 15, 20, 15, 30)

true_ate <- mean(y_t - y_c)

z1 <- c(1,1,0,0,0,0,0)

## Notice these two are the same
Y_obs_1 <- z1*y_t + (1-z1)*y_c
Y_obs_1a <- ifelse(z1==1,y_t,y_c)

est1 <- mean(Y_obs_1[z1==1]) - mean(Y_obs_1[z1==0])

true_ate
est1



```


